### 2023-06-24 scripts/Introduction.Rmd
intro moved to independent location

# Introduction

Arid and semi-arid lands constitute the plurality of earths terrestrial surface. 
These areas are used extensively for raising agricultural animals, such as cattle and sheep, without wholesale conversion of the land to agronomic settings, i.e. pastoral or rangelands. 
Accordingly arid lands represent an enormous proportion of Earth's natural areas.
However agricultural related use of these land have often lead to degradation of them
Historic actions and phenomena such as fire suppression, shrub encroachment, and overgrazing, in combination with climate change and drought are resulting in wildfires and their subsequent burns scars which are often dominated by just a handful of invasive species.
The prevalence of these invasive species decrease the ability of the land to serve multiple ecosystem services, such as to foster both wildlife and pastoral goals, and in extreme cases may preclude uses of the land by both wildlife and humans.

In order to maintain the ecosystem services of burned areas restoration plantings have long been used. 
More recently a focus on utilizing native species, and now locally adapted genotypes, has been adopted as these species and populations have been shown to provide ecosystem services to a variety of wildlife and stakeholders, and they are often more capable of establishing at sites. 
However, the acquisition of native plant materials in the volumes required for restoration, especially in fire prone years, is not yet feasible. 
In an attempt to provide for the demand of locally adapted native plant materials partnerships with farmers who are planting wild harvested seeds, and harvesting their seed for sale to restoration performing agencies has increased. However, the implementation of these goals still face significant challenges.

We anticipate that three particular challenges which are currently being faced in developing and delivering a native seed supply in the Great Basin Ecoregion are ubiquitous to all supply chains. 
1) Locating populations of adequate size that they can have enough viable seed collected from that that it is economically viable for farmers to grow out from the seed lots while not adversely affecting the demographic vital rates of the population. 
2) Minimizing the adverse affects of artificial selection of these seed lines while being increased in agricultural settings. 
3) Ensuring that farmed seeds are applied to the areas in burn scars which they have the highest probability of success. 
We posit that Species Distribution Modelling (SDMs) (or Environmental Niche Modelling (ENM)) may be used as an heuristic to increase the probability of success at each of these stages, and the development of these models are markedly cheaper than purely empirical approaches.

Locating populations of native species which are large enough to support a seed collection which is large enough to establish an agricultural lot is difficult. 
The areas over which native seed may feasibly be collected in semi-arid areas, which often times have higher proportion of natural areas relative to other settings, are enormous and have generally had their floras coarsely mapped. 
Restoration approaches for sourcing locally adapted seed such as Seed Transfer Zones (STZ's), predictive provincing, or Omernik Level 4 ecoregions, increase the relevancy of identifying many populations of suitable for collections across ranges. 
Locating such populations is difficult given the lengths of growing seasons, human resources, and the length of time required to collect enough viable seed.

Agricultural cultivation presents multiple steps which alter the frequency of gene forms (alleles) within the seed lots via unintentional artificial selection.
It is theorized that reducing these effects will result in seed lots with a proportion of more individuals with gene forms more typical of the wild population (F~0~), than the F1, and hence have more individuals likely to establish in a restoration. 
A method to reduce the demographic bottlenecks associated with recruitment of an individual, and to reduce over-representation of seed from species which are more fit in cultivated settings than the wild may be to reduce the environmental dissimilarity between the wild and cultivated settings. 
These models may also help farmers prioritize which species to grow to have the most successful crops, and to help recruit new farmers to dedicate portions of there fields to developing native plant materials.

Planning a restoration following a wildfire is a difficult process, due to limitations of both fiscal budget, time, human resources, and knowledge of the heterogeneity of the affected areas. 
Species Distribution Models offer a line of evidence which practitioners may use to identify species which are most suitable for the burned areas in a simple readily query friendly format. 
The utilization of them

### 2023-06-24 scripts/Introduction.Rmd
working on cost surface outline

@@
-# Introduction
+---
+title: "Introduction"
+date: '`r format(Sys.Date(), "%B %d, %Y")`'
+output: html_document
+---
 
-Arid and semi-arid lands constitute the plurality of earths terrestrial surface. 
-These areas are used extensively for raising agricultural animals, such as cattle and sheep, without wholesale conversion of the land to agronomic settings, i.e. pastoral or rangelands. 
-Accordingly arid lands represent an enormous proportion of Earth's natural areas.
-However agricultural related use of these land have often lead to degradation of them
-Historic actions and phenomena such as fire suppression, shrub encroachment, and overgrazing, in combination with climate change and drought are resulting in wildfires and their subsequent burns scars which are often dominated by just a handful of invasive species.
-The prevalence of these invasive species decrease the ability of the land to serve multiple ecosystem services, such as to foster both wildlife and pastoral goals, and in extreme cases may preclude uses of the land by both wildlife and humans.
+Drylands constitute roughly ~45% of earths surface (@pruavualie2016drylands). 
+These water-limited areas are commonly used as range or pasture lands, hence much of the land is maintained in natural or semi-natural conditions (@reynolds2007globa, @asner2004grazingl). 
+Accordingly drylands lands represent an enormous proportion of Earth's remaining natural areas, while fostering roughly 2.5 billion persons, however many of these areas are in ecological states which do not maximize their ecosystem services to either human or wildlife (@reynolds2007global, @bestelmeyer2015desertification).
+Historic actions and phenomena such as fire suppression, shrub encroachment, and overgrazing, in combination with climate change and drought are resulting in severe wildfires and their subsequent burns scars which may lead to a regime shifts wherein invasive species dominate burned sites for the foreseeable future (@bestelmeyer2015desertification).
+The prevalence of these invasive species decrease the ability of the land to serve multiple ecosystem services, such as to foster both wildlife and pastoral goals, and in extreme cases may preclude uses of the land by both wildlife and humans, and may have require high fiscal costs to return to a more desirable state.
 
@@
 The areas over which native seed may feasibly be collected in semi-arid areas, which often times have higher proportion of natural areas relative to other settings, are enormous and have generally had their floras coarsely mapped.
-Restoration approaches for sourcing locally adapted seed such as Seed Transfer Zones (STZ's), predictive provincing, or Omernik Level 4 ecoregions, increase the relevancy of identifying many populations of suitable for collections across ranges.
+Restoration approaches for sourcing locally adapted seed such as Seed Transfer Zones (STZ's), predictive provincing, or Omernik Level 4 ecoregions, increase the relevancy of identifying many populations suitable for collections across a species range.
 Locating such populations is difficult given the lengths of growing seasons, human resources, and the length of time required to collect enough viable seed.
 
-Agricultural cultivation presents multiple steps which alter the frequency of gene forms (alleles) within the seed lots via unintentional artificial selection.
-It is theorized that reducing these effects will result in seed lots with a proportion of more individuals with gene forms more typical of the wild population (F~0~), than the F1, and hence have more individuals likely to establish in a restoration.
+Agricultural cultivation presents multiple steps which may alter the frequency of gene forms (alleles) within the seed lots via unintentional artificial selection.
+It is theorized that reducing these effects will result in seed lots with a proportion of more individuals with gene forms more typical of the planted wild population, than the offspring derived from it, and hence have more individuals likely to establish in a restoration.
 A method to reduce the demographic bottlenecks associated with recruitment of an individual, and to reduce over-representation of seed from species which are more fit in cultivated settings than the wild may be to reduce the environmental dissimilarity between the wild and cultivated settings.
@@
 Species Distribution Models offer a line of evidence which practitioners may use to identify species which are most suitable for the burned areas in a simple readily query friendly format. 
-The utilization of them
+The utilization of them ...
+
+Herein we develop simple Species Distribution Models, using a variety of predictors which are global in coverage - or for which global analogs exist, and species occurrences from publicly available and rangeland monitoring datasets, to assist in an ongoing effort to develop native plant materials in the Great Basin Ecoregion (U.S.A.).
+Two sets of models are generated for each species, one set which reflects the natural, in particular biophysical the settings of the species, and another set which more closely reflects the climatic envelope and soil properties of the species.
+The former are to assist in challenges of finding populations and restoration efforts while the latter is to support seed increase endeavors.
+In attempts to reduce the discrepancies between the fundamental and realized niche, to assist field workers in finding populations, known occurrences are combined with cost-surfaces and predictions of suitability models to provide a consensus surface and emulate the restrictions of dispersal to species distributions.
+Finally, cell statistics are used to create aggregate scores for patches of suitable habitat which have been shown to correlate with abundance.
+
+Distance = aggregate
+
+% Species Cover on plot = sum(aggregate suitable cells)
+% Species Cover on plot = sqrt(sum(aggregate suitable cells))
+% Species Cover on plot = log(sum(aggregate suitable cells))

### 2023-06-24 scripts/patches.R
deriving patches

@@ -1,0 +1,17 @@
+library(terra)
+library(tidyverse)
+r <- rast(nrows=18, ncols=36, xmin=0)
+r[1:2, 5:8] <- 1.1
+r[5:8, 2:6] <- 1.2
+r[7:12, 22:36] <- 1.3
+r[15:16, 18:29] <- 1.4
+p <- patches(r)
+
+plot(p)
+
+tibble(
+  Patch = values(p)[!is.na(values(p))],
+  Suitability = values(r)[!is.na(values(p))]
+) %>% 
+  sample_n(15)
+

### 2023-10-14 scripts/download_data.Rmd
new climate source and download protocal

---
title: "Download data"
author: "steppe"
date: "2023-10-14"
output: html_document
---

```{r}
library(tidyverse)
```

The main bioclim variables can be found at the [link here](https://envicloud.wsl.ch/#/?bucket=https%3A%2F%2Fos.zhdk.cloud.switch.ch%2Fenvicloud%2F&prefix=chelsa%2Fchelsa_V2%2FGLOBAL%2Fclimatologies%2F1981-2010%2Fbio%2F) an S3 bucket for CHELS*A* (not the neighborhood!).
All files can be downloaded as a textfile for further use with wget. 
We will subset these files to only download the layers of interest.

```{r Subset Chelsea bioclim layers}

lyrs2grep <- c('bio1', 'bio4', 'bio5', 'bio6', 'bio10', 'bio11', 'bio12',
               'bio18', 'bio19', # bioclim variables
               'gdd5', 'ngd5', 'gdd10', 'ngd10', # growing degree days
               'gdgfgd5', 'gdgfgd10', # first growing degree day above 5/10 *c
               'tcc_mean', 'vpd_mean', # mean cloud cover, vapour press deficit
               'hurs_mean', # mean monthly near surface humidity
               'scd', # number of days with snow cover
               'swe', # snow water equivalent. 
               'fcf' # number of events where temps go from to/below 0*c
               )
lyrs2grep <- paste0('_', lyrs2grep, '_', collapse = "|")

bioclim_paths <- read.delim(
  '../data/raw/envidatS3paths-mainbioclim.txt', header = F) %>% 
  rename(Path = V1) %>% 
  filter(str_detect(Path, lyrs2grep))

write.table(bioclim_paths, 
            file = '../data/raw/ClimateVariables.txt', 
            row.names = F ,col.names = F, quote = F)

rm(lyrs2grep, bioclim_paths)
```

```{sh download Chelsa climate variables, eval = F}
URL_LIST=$(<../data/raw/ClimateVariables.txt)
## echo $URL_LIST | xargs -n 1 -P 8 wget -q -P /path/to/folder/
```

```{r download soil grids data eval = F}
library(terra)
library(sf)
library(gdalUtilities)

## download soil grids data

igh = '+proj=igh +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs' # proj string for Homolosine projection

domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
crs(domain) <- "EPSG:4326" # define the projection
domain <- as.polygons(domain) |>  # convert to vector data
  st_as_sf() |> # convert to simple feature
  st_transform(igh) %>% 
  st_bbox()

soilgrids_download <- function(list_vrt, # download url
                               list_lfile, # destination path
                               shape_igh, # AOI shape in IGH proj
                               destproj) { # desired projection

  terra::rast(paste0(sg_url, list_vrt)) %>% # read raster
    terra::crop(ext(vect(shape_igh))) %>% # crop to bounding box
    terra::project(destproj) %>% # reproject
    terra::writeRaster(list_lfile,
                       overwrite = T
    ) # @ Tsyplenkov, Anatoly
}

######### Create directory to hold these raw data ###########

if (!dir.exists("soilgrid")) {
  dir.create("soilgrid")
}

# Create paths
lfile <- paste0(
  "soilgrid/",
  props, "_",
  rep(layers, 4),
  ".tif"
)

sg_url = "/vsicurl?max_retry=3&retry_delay=1&list_dir=no&url=https://files.isric.org/soilgrids/latest/data/"

### soil physical-texture properties ###

var <- c('silt', 'clay', 'sand') # these three have all three vars downloaded
depth <- c('0-5', '5-15', '15-30')

vrts <- paste0(var, '/', var, '_', rep(depth, each = length(var)), 'cm_mean.vrt')

gdal_translate(
  paste0(sg_url, vrts),
               "./crop_roi_igh_r.vrt",
               of = "VRT", tr = c(250, 250),
               projwin = domain,
               projwin_srs = igh)

gdalwarp(source = "./crop_roi_igh_r.vrt",
         destination = "./crop_roi_ll_r.vrt", 
         s_srs = igh, 
         t_srs = "EPSG:5070", 
         of = "VRT",
         overwrite = TRUE)

gdal_translate("./crop_roi_ll_r.vrt",  
               "./crop_roi_ll_r.tif", 
               co=c("TILED=YES", "COMPRESS=DEFLATE", "PREDICTOR=2", "BIGTIFF=YES"))

#### hydroxide content & coarse fragments ####
var <- c('phh2o', 'cfvo')  
depth <- c('0-5', '30-60')

#### soil organic carbon ####
var <- 'soc'  # these only have these depths downloaded
depth <- c('0-5')

#### salinity ####

#### Soil USDA Class ####

```

### 2023-12-04 scripts/Regress_AIM_cover_Patch.Rmd
adding empty scripts for steps

---
title: "Regress Percent Cover of Species in AIM plots against Mean Percent Patch Suitability"
author: "steppe"
date: "2023-12-04"
output: html_document
---

Species Distribution Models have been shown to be semi-quantitative, i.e. be able to predict species abundance from suitability scores,  under various scenarios.
Here we determine whether a positive relationship exists between the mean predicted habitat suitability of a patch, or a sector of large patches, and the percent aerial cover of the taxon on an AIM plot.

### 2023-12-04 scripts/download_dependent_variables.Rmd
adding empty scripts for steps

---
title: "Download data"
author: "steppe"
date: "2023-10-14"
output: html_document
---

```{r load packages}
library(tidyverse)
library(BIEN)
library(ridigbio)
library(rgbif)
source('functions.R')
```

```{r import and clean target species}
targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>% 
  filter(species != 'tridentata') # purshia, artemisia, and larrea !

target_taxa <- unite(targets, 'taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% 
  pull(taxon) %>% 
  unique() # 293 species.

## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources. 
```

```{r Perform gbif queries, eval = F}
user <- "steppe" # your gbif.org username 
pwd <- "blmquery" # your gbif.org password # this one is now old and serves as an example
email <- "reedbenkendorf27@gmail.com" # your email

gbif_taxon_keys <- target_taxa %>% 
  name_backbone_checklist()  %>% 
  filter(!matchType == "NONE") %>% 
  pull(usageKey)

dl_key <- occ_download(
  pred_in("taxonKey", gbif_taxon_keys),
  user=user, pwd=pwd, email=email,
  pred("country", "US"),
  pred("hasCoordinate", TRUE),
  pred("hasGeospatialIssue", FALSE),
  pred_gte("year", 1900)
)

occ_download_wait(dl_key)
occ_download_get( dl_key )

rm(user, pwd, email, gbif_taxon_keys, dl_key)
```

```{r Perform bien queries, eval = F}

bien_recs <- lapply(target_taxa, BIEN_occurrence_species, collection.info = TRUE,
                    observation.type = TRUE, new.world = TRUE)

```

```{r Perform idigbio queries, eval = F}

ob <- lapply(target_taxa, idig_records)
t <- bind_rows(ob)

```

```{r Import and Query AIM data, eval = F}

file.path(p, list.files(p, pattern = 'gdb$', recursive = T))
# for some strange reason this not exactly working! a weird /gdb is being tagged
# onto the end of it... we can manually specify the data beneath
st_layers(dsn = '../data/2022_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb')

AIM_points <- st_read(dsn = '../data/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb', 
        layer = 'TerrADat', quiet = T) %>%
  filter(State %in% c('NV', 'OR', 'CA', 'ID', 'UT')) %>% 
  select(PrimaryKey, geometry = Shape) %>% 
  st_transform(5070)

AIM_points <- AIM_points[st_covered_by(AIM_points, field_off) %>% lengths > 0,]

spp_richness <- st_read(dsn = '../data/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb', 
        layer = 'tblSpecRichDetail', quiet = T) %>% 
  select(SpeciesList, PrimaryKey) %>% 
  right_join(AIM_points, by = 'PrimaryKey')

spp_found <- inner_join(target_taxa_lkp, spp_richness, 
                        by = c('USDA.Code' = 'SpeciesList')) %>% 
  select(species, PrimaryKey, geometry) %>% 
  st_as_sf()

spp_occurrences <- bind_rows(spp_found, pub_presence) %>% 
  arrange(species) %>% 
  relocate(c('year', 'date'), .before = geometry) %>% 
  mutate(species = str_trim(species)) %>% 
  filter(!species %in% c('Artemisia tridentata',
                         'Elymus elymoides',
                         'Poa secunda'))

```

### 2023-12-04 scripts/functions.R

@@ -1,0 +1,17 @@
+idig_records <- function(query){
+  
+  result <- ridigbio::idig_search_records(rq = list(scientificname = query)) |> 
+    dplyr::select(scientificname, datecollected, collector, geopoint.lon, geopoint.lat) |> 
+    dplyr::drop_na(geopoint.lon, geopoint.lat) |> 
+    dplyr::mutate(datecollected = as.Date(datecollected)) |> 
+    dplyr::filter(datecollected > '1950-01-01')
+  
+  if(nrow(result) > 0) {result$status = 'found'} else {
+    result <- data.frame(
+      'scientificname' = query, 
+      'datecollected' = NA, 'collector' = NA, 'geopoint.lon' = NA, 'geopoint.lat' = NA, 
+      'status' = 'not found'
+    )
+  }
+  return(result)
+}

### 2023-12-05 scripts/download_dependent_variables.Rmd
all queries ran

@@
 library(rgbif)
+library(sf)
 source('functions.R')
@@
 targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>% 
-  filter(species != 'tridentata') # purshia, artemisia, and larrea !
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) # purshia, artemisia, and larrea !
 
@@
 ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+
+## we will add on USDA codes so we can interface with AIM data.
+
 ```
@@
 
-gbif_taxon_keys <- target_taxa %>% 
+gbif_taxon_keys <- target_taxa[294:364] %>% 
   name_backbone_checklist()  %>% 
@@
 
+BIEN is pretty finicky these days and is prone to crashing.
+There is a paper by Maitner with Enquist, Wilson, Merow etc.
+about the need for open source devs to be rewarded academically, I take that has a sign of Brian stepping away from it.
+We will use a for loop to download from BIEN.
 ```{r Perform bien queries, eval = F}
 
-bien_recs <- lapply(target_taxa, BIEN_occurrence_species, collection.info = TRUE,
-                    observation.type = TRUE, new.world = TRUE)
+for(i in seq_along(target_taxa)){
+  print(target_taxa[i])
+  dl <- BIEN_occurrence_species(target_taxa[i], collection.info = TRUE, observation.type = TRUE, new.world = TRUE)
+  write.csv(dl, file.path('../data/raw/occurrence/bien', paste0(gsub(' ', '_', target_taxa[i]), '.csv')) )#, row.names = FALSE)
+}
+```
 
+```{r stitch together BIEN queries}
+
+p <- '../data/raw/occurrence/bien'
+files <- file.path(p, list.files(p))
+
+lapply(Filter(function(x) R.utils::countLines(x)==1, files), unlink) # delete empty CSV's in directory
+files <- file.path(p, list.files(p))
+
+bien_data <-  do.call(rbind, lapply(files, read.csv)) %>% 
+  filter(datasource != 'GBIF' & date_collected > '1950-01-01', ! scrubbed_species_binomial %in% c('Acer negundo', 'Prosopis glandulosa')) %>% 
+  drop_na(date_collected, latitude) %>% 
+  select(scrubbed_species_binomial:datasource, observation_type)
+
+write.csv(bien_data, '../data/raw/occurrence/BIEN_records.csv', row.names = FALSE)
 ```
@@
 ob <- lapply(target_taxa, idig_records)
-t <- bind_rows(ob)
+idig <- bind_rows(ob) %>% 
+  filter(datecollected > '1950-01-01')
+
+write.csv(idig, '../data/raw/occurrence/IDIG.csv')
 
@@
 
-file.path(p, list.files(p, pattern = 'gdb$', recursive = T))
+p1 <- '../data/raw/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb'
+
 # for some strange reason this not exactly working! a weird /gdb is being tagged
 # onto the end of it... we can manually specify the data beneath
-st_layers(dsn = '../data/2022_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb')
+st_layers(dsn = p1)
 
-AIM_points <- st_read(dsn = '../data/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb', 
+AIM_points <- st_read(dsn = p1, 
         layer = 'TerrADat', quiet = T) %>%
-  filter(State %in% c('NV', 'OR', 'CA', 'ID', 'UT')) %>% 
+#  filter(State %in% c('NV', 'OR', 'CA', 'ID', 'UT')) %>% 
   select(PrimaryKey, geometry = Shape) %>% 
@@
 
-AIM_points <- AIM_points[st_covered_by(AIM_points, field_off) %>% lengths > 0,]
+# AIM_points <- AIM_points[st_covered_by(AIM_points, field_off) %>% lengths > 0,]
 
-spp_richness <- st_read(dsn = '../data/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb', 
+spp_richness <- st_read(dsn = p1, 
         layer = 'tblSpecRichDetail', quiet = T) %>% 

### 2023-12-05 scripts/functions.R

@@ -2,7 +2,7 @@ idig_records <- function(query){
   
   result <- ridigbio::idig_search_records(rq = list(scientificname = query)) |> 
     dplyr::select(scientificname, datecollected, collector, geopoint.lon, geopoint.lat) |> 
-    dplyr::drop_na(geopoint.lon, geopoint.lat) |> 
+    tidyr::drop_na(geopoint.lon, geopoint.lat) |> 
     dplyr::mutate(datecollected = as.Date(datecollected)) |> 
     dplyr::filter(datecollected > '1950-01-01')
   

### 2023-12-05 scripts/download_dependent_variables.Rmd
adding usda codes manual review

@@
 targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>% 
-  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) # purshia, artemisia, and larrea !
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') # artemisia, and larrea !
 
-target_taxa <- unite(targets, 'taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% 
+target_taxa <- targets %>% 
   pull(taxon) %>% 
@@
 ## we will add on USDA codes so we can interface with AIM data. 
+symbols <- read.csv('../data/raw/USDA_PLANTS.csv') %>% 
+  right_join(., targets, by = c('' = 'taxon'))
 

### 2023-12-06 scripts/download_dependent_variables.Rmd
AIM import for presence

@@
 
+targets %>% 
+  group_by(symbol, taxon) %>% 
+  count()
+
 target_taxa <- targets %>% 
@@
 
+target_symbols <- targets %>% 
+  pull(symbol) %>% 
+  unique()
+
 ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
-
-## we will add on USDA codes so we can interface with AIM data. 
-symbols <- read.csv('../data/raw/USDA_PLANTS.csv') %>% 
-  right_join(., targets, by = c('' = 'taxon'))
 
@@
 
-# for some strange reason this not exactly working! a weird /gdb is being tagged
-# onto the end of it... we can manually specify the data beneath
 st_layers(dsn = p1)
@@
         layer = 'TerrADat', quiet = T) %>%
-#  filter(State %in% c('NV', 'OR', 'CA', 'ID', 'UT')) %>% 
-  select(PrimaryKey, geometry = Shape) %>% 
+  select(PrimaryKey, DateVisited, geometry = Shape) %>% 
   st_transform(5070)
-
-# AIM_points <- AIM_points[st_covered_by(AIM_points, field_off) %>% lengths > 0,]
 
@@
   select(SpeciesList, PrimaryKey) %>% 
-  right_join(AIM_points, by = 'PrimaryKey')
+  filter(SpeciesList %in% target_symbols) %>% 
+  right_join(AIM_points, by = 'PrimaryKey') %>% 
+  st_as_sf()  %>% 
+  mutate(YearSampled = gsub('-.*$', '', DateVisited), .before = geometry) %>% 
+  select(-DateVisited) %>% 
+  arrange(SpeciesList) %>% 
+  right_join(., select(targets, taxon, symbol) %>% 
+               distinct(), by = c('SpeciesList' = 'symbol' ))
 
-spp_found <- inner_join(target_taxa_lkp, spp_richness, 
-                        by = c('USDA.Code' = 'SpeciesList')) %>% 
-  select(species, PrimaryKey, geometry) %>% 
-  st_as_sf()
+st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records.shp')
 
-spp_occurrences <- bind_rows(spp_found, pub_presence) %>% 
-  arrange(species) %>% 
-  relocate(c('year', 'date'), .before = geometry) %>% 
-  mutate(species = str_trim(species)) %>% 
-  filter(!species %in% c('Artemisia tridentata',
-                         'Elymus elymoides',
-                         'Poa secunda'))
-
+rm(p1, AIM_points)
 ```

### 2023-12-06 scripts/generate_absences.Rmd
fleshing fn to draw absences

---
title: "Generate Real and Pseudo-absences"
author: "steppe"
date: "2023-12-06"
output: html_document
---

```{r load packages}
library(tidyverse)
library(sf)
source('functions.R')
```

We use absences while creating species distribution models. 
Both true absences, acquired from vegetation monitoring plots, and pseudo absences generated via geographic and environmental space are used.

```{r Create simplified BLM surface management}

```

```{r}
library(geosphere)

st_area(x, ...)

```

### 2023-12-06 scripts/absence_drawer.R

@@ -1,0 +1,48 @@
+absence_drawer <- function(x, bg_abs){
+  
+  
+  # create the observed hull, the buffered hull for random sample selection, 
+  # and a hull to form the basis for a bounding box to generate regular absences in
+  
+  observed_hull <- sf::st_union(x) |>
+    sf::st_convex_hull()
+  
+  obs_area <- sf::st_area(observed_hull)
+  buffer_hull <- st_buffer(observed_hull, dist = (obs_area*0.25)/2  ) 
+  buf_area <- sf::st_area(buffer_hull)
+  box_hull <- sf::st_buffer(buffer_hull, dist = (buf_area*0.1)/2)
+  
+  bbox <- sf::st_bbox(box_hull) |>
+    sf::st_as_sfc() 
+  
+  box_hull <- sf::st_difference(buffer_hull, box_hull)
+  
+  # perform the outline background absence sample
+  reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular')
+  
+  # identify AIM plots and generate absences. 
+  
+  blm_surf_area <- sf::st_area(blm_surf)
+  prcnt_blm <- (blm_surf_area / buf_area) * 100
+
+  aim_recs <- (1 - bg_abs) * round(prcnt_blm, 0) * 0.1
+  
+  # generate AIM plot absences
+  
+  
+#  st_intersection(aim_plots, blm_admin_surf)
+  
+  
+  # generate Random absences
+  
+#  rand_recs <- ((1 - bg_abs) * (1 - prcnt_blm)) * 0.1
+#  buffered_presences <- sf::st_buffer(x, dist = 10000) |>
+#    sf::st_union()
+  
+#  non_blm_area <- sf::st_difference(blm_admin_surf, buffer_hull) 
+#  non_blm_area <- sf::st_difference(non_blm_area, buffered_presences)
+  
+#  random_samples <- sf::st_sample(non_blm_area, prcnt_blm)
+  
+}
+

### 2023-12-06 scripts/generate_absences.Rmd
flesh fn

@@
 
+```{r Create simplified BLM surface management, eval = F}
+p1 <- '/media/steppe/hdd/geospatial/SMA_WM.gdb'
+
+st_layers(p1)
+ag <- st_read(p1, 'SurfaceMgtAgy_BLM') %>% 
+  filter(ADMIN_ST != 'AK') %>% 
+  rename(geometry = SHAPE) %>% 
+  select(geometry) %>% 
+  st_make_valid() %>% 
+  st_union()
+
+ag <- st_simplify(ag, dTolerance = 1000)
+plot(ag)
+
+pout <- '/media/steppe/hdd/SDM_restorations/data/raw/coarse_blm/'
+st_write(ag, dsn = paste0(pout, 'coarse_surface.shp'))
+```
+
 We use absences while creating species distribution models. 
@@
 
-```{r Create simplified BLM surface management}
+```{r load aim data for absences}
 
+AIM_points <- st_read(dsn = p, layer = 'TerrADat', quiet = T) %>%
+  select(PrimaryKey, DateVisited, geometry = Shape) %>% 
+  st_transform(5070)
+
+spp_richness <- st_read(dsn = p1, 
+        layer = 'tblSpecRichDetail', quiet = T) %>% 
+  select(SpeciesList, PrimaryKey) %>% 
+  filter(SpeciesList %in% target_symbols) %>% 
+  right_join(AIM_points, by = 'PrimaryKey') %>% 
+  st_as_sf()
+
+pout <- '/media/steppe/hdd/SDM_restorations/data/raw/coarse_blm/'
+blm_surf <- st_read(paste0(pout, 'coarse_surface.shp'))
+blm_surf <- st_simplify(blm_surf, dTolerance = 500) # just make it a bit smaller. 
 ```
@@
 ```{r}
-library(geosphere)
 
-st_area(x, ...)
+aim <- st_read('../data/raw/occurrence/AIM/AIM_records.shp') %>% 
+  filter(SpcsLst == 'ACHY')
+
+ob_hull <- absence_drawer(aim, bg_abs = 0.15)
+buf_hull <- absence_drawer(aim, bg_abs = 0.15)
+box_hull <- absence_drawer(aim, bg_abs = 0.15)
+bbox <- absence_drawer(aim, bg_abs = 0.15)
+
+ggplot() +
+ # geom_sf(data = ob_hull, fill = NA) + 
+#  geom_sf(data = buf_hull, fill = NA) + 
+ # geom_sf(data = box_hull, fill = NA) +
+  geom_sf(data = bbox, fill = 'red')
 

### 2023-12-06 scripts/absence_drawer.R

@@ -5,27 +5,39 @@ absence_drawer <- function(x, bg_abs){
   # and a hull to form the basis for a bounding box to generate regular absences in
   
   observed_hull <- sf::st_union(x) |>
-    sf::st_convex_hull()
+    sf::st_convex_hull() |>
+    sf::st_sfc() 
   
-  obs_area <- sf::st_area(observed_hull)
-  buffer_hull <- st_buffer(observed_hull, dist = (obs_area*0.25)/2  ) 
-  buf_area <- sf::st_area(buffer_hull)
-  box_hull <- sf::st_buffer(buffer_hull, dist = (buf_area*0.1)/2)
+  obs_area <- as.numeric(st_area(observed_hull)) * 0.15 / 2
+  obs_area <- if(obs_area > 160934){obs_area = 160934} # cap area with 100 mi buffer. 
+  
+  buffer_hull <- st_buffer(observed_hull, obs_area) 
+  
+  buf_area <- as.numeric(sf::st_area(buffer_hull) *0.1) / 2
+  obs_area <- if(buf_area > 40233){buf_area = 40233} # cap area with 25 mi buffer.
+  
+  box_hull <- sf::st_buffer(buffer_hull, dist = obs_area)
   
   bbox <- sf::st_bbox(box_hull) |>
     sf::st_as_sfc() 
   
-  box_hull <- sf::st_difference(buffer_hull, box_hull)
+  box_hull <- sf::st_difference(bbox, buffer_hull)
   
   # perform the outline background absence sample
   reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular')
   
+  rm(box_hull, bbox, observed_hull)
+  
+
   # identify AIM plots and generate absences. 
   
-  blm_surf_area <- sf::st_area(blm_surf)
+  blm_surf_area <- as.numeric(sf::st_area(blm_surf))
   prcnt_blm <- (blm_surf_area / buf_area) * 100
-
   aim_recs <- (1 - bg_abs) * round(prcnt_blm, 0) * 0.1
+  
+  AIM_points <- AIM_points[lengths(st_intersects(AIM_points, buffer_hull)) > 0,]
+  AIM_points <- AIM_points |> 
+    filter(! PrimaryKey %in% x$PrimaryKey ) # these are presences
   
   # generate AIM plot absences
   

### 2023-12-06 scripts/generate_absences.Rmd
draft absence pseudo abs pt generator

@@
 
-```{r load aim data for absences}
+```{r load aim data for absences, eval = F}
 
+p  <- '../data/raw/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb'
 AIM_points <- st_read(dsn = p, layer = 'TerrADat', quiet = T) %>%
@@
 
-spp_richness <- st_read(dsn = p1, 
-        layer = 'tblSpecRichDetail', quiet = T) %>% 
-  select(SpeciesList, PrimaryKey) %>% 
-  filter(SpeciesList %in% target_symbols) %>% 
-  right_join(AIM_points, by = 'PrimaryKey') %>% 
-  st_as_sf()
-
 pout <- '/media/steppe/hdd/SDM_restorations/data/raw/coarse_blm/'
@@
 blm_surf <- st_simplify(blm_surf, dTolerance = 500) # just make it a bit smaller. 
+blm_surf <- st_make_valid(blm_surf)
 ```
@@
 
-ob_hull <- absence_drawer(aim, bg_abs = 0.15)
-buf_hull <- absence_drawer(aim, bg_abs = 0.15)
-box_hull <- absence_drawer(aim, bg_abs = 0.15)
 bbox <- absence_drawer(aim, bg_abs = 0.15)
 
-ggplot() +
- # geom_sf(data = ob_hull, fill = NA) + 
-#  geom_sf(data = buf_hull, fill = NA) + 
- # geom_sf(data = box_hull, fill = NA) +
-  geom_sf(data = bbox, fill = 'red')
-
 ```

### 2023-12-06 scripts/absence_drawer.R

@@ -24,37 +24,54 @@ absence_drawer <- function(x, bg_abs){
   box_hull <- sf::st_difference(bbox, buffer_hull)
   
   # perform the outline background absence sample
-  reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular')
+  reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular') |>
+    sf::st_as_sf() |>
+    dplyr::mutate(Occurrence = 0, PtType = 'Background Absence') |>
+    dplyr::rename(geometry = x)
   
-  rm(box_hull, bbox, observed_hull)
+  rm(bbox, observed_hull)
   
-
   # identify AIM plots and generate absences. 
   
   blm_surf_area <- as.numeric(sf::st_area(blm_surf))
-  prcnt_blm <- (blm_surf_area / buf_area) * 100
-  aim_recs <- (1 - bg_abs) * round(prcnt_blm, 0) * 0.1
+  prcnt_blm <- (blm_surf_area / as.numeric(sf::st_area(buffer_hull))) 
+  aim_recs <- nrow(x) * (1 - bg_abs) * prcnt_blm
   
   AIM_points <- AIM_points[lengths(st_intersects(AIM_points, buffer_hull)) > 0,]
   AIM_points <- AIM_points |> 
     filter(! PrimaryKey %in% x$PrimaryKey ) # these are presences
   
-  # generate AIM plot absences
+  if(aim_recs >= nrow(AIM_points)){aim_recs <- nrow(AIM_points)} # in case species v. abundant. 
   
-  
-#  st_intersection(aim_plots, blm_admin_surf)
-  
-  
+  AIM_absence <- sample_n(AIM_points, size = round(aim_recs * 1.1, 0), replace = FALSE) |> 
+    dplyr::mutate(Occurrence = 0, PtType = 'AIM Absence') |>
+    dplyr::select(Occurrence, PtType, geometry) |>
+    sf::st_transform(sf::st_crs(x))
+
   # generate Random absences
   
-#  rand_recs <- ((1 - bg_abs) * (1 - prcnt_blm)) * 0.1
-#  buffered_presences <- sf::st_buffer(x, dist = 10000) |>
-#    sf::st_union()
+  rand_recs <- nrow(x) * ((1 - bg_abs) * (1 - prcnt_blm))
+  buffered_presences <- sf::st_buffer(x, dist = 10000) |>
+    sf::st_union() |>
[…]

### 2023-12-06 scripts/download_dependent_variables.Rmd
working on thinning to one record per cell

@@
 
-```{r stitch together BIEN queries}
+```{r stitch together BIEN queries, eval = F}
 
@@
   st_as_sf()  %>% 
-  mutate(YearSampled = gsub('-.*$', '', DateVisited), .before = geometry) %>% 
-  select(-DateVisited) %>% 
+  mutate(DateVisited = as.Date(DateVisited), .before = geometry) %>% 
   arrange(SpeciesList) %>% 
@@
 
-st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records.shp')
+st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records.shp', append = F)
 
@@
 ```
+
+```{r}
+rm(idig_records, target_symbols, target_taxa)
+```
+
+```{r Create consensus presence records}
+
+base_p <- '../data/raw/occurrence'
+
+bien <- read.csv(file.path(base_p, 'BIEN_records.csv')) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) %>% 
+  mutate(date = as.Date(date_collected)) %>% 
+  select(taxon = scrubbed_species_binomial, date)
+
+idig <- read.csv(file.path(base_p, 'IDIG.csv')) %>% 
+  st_as_sf(coords = c('geopoint.lon', 'geopoint.lat'), crs = 4326) %>% 
+  mutate(date = as.Date(datecollected)) %>% 
+  select(taxon = scientificname)
+
+aim <- st_read(file.path(base_p, 'AIM', 'AIM_records.shp')) %>% 
+  mutate(date = as.Date(DatVstd)) %>% 
+  select(taxon, date) %>% 
+  st_transform(4326)
+
+gbif1 <- read.delim(file.path(base_p, 'gbif1', 'occurrence.txt'), na.strings = "") # %>% 
+  drop_na(decimalLongitude, decimalLatitude) %>% 
+  filter(str_detect(decimalLatitude, '[A-Z]|[a-z]', negate = T)) %>% 
+  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'), crs = 4326) %>% 
+  mutate(date = as.Date(eventDate), .before = geometry, 
+         coordinateUncertaintyInMeters = as.numeric(coordinateUncertaintyInMeters)) %>% 
+  filter(coordinateUncertaintyInMeters <= 250) %>% 
+  select(taxon = species, date)
+
+gbif2 <- read.delim(file.path(base_p, 'gbif2', 'occurrence.txt'), na.strings = '')%>% 
+  drop_na(decimalLongitude, decimalLatitude) %>% 
+  filter(str_detect(decimalLatitude, '[A-Z]|[a-z]', negate = T)) %>% 
+  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'), crs = 4326) %>% 
+  mutate(date = as.Date(eventDate), .before = geometry, 
+         coordinateUncertaintyInMeters = as.numeric(coordinateUncertaintyInMeters)) %>% 
+  filter(coordinateUncertaintyInMeters <= 250) %>% 
+  select(taxon = species, date)
+
+records <- bind_rows(aim, bien, gbif1, gbif2, idig) %>% 
+  mutate(taxon = str_to_sentence(taxon)) %>% 
+  arrange(taxon) # recs need be > 90 m from each other
+
+records <- st_transform(records, 5070)
+splicies <- split(records, records$taxon)[1:2]
+
+splicies <- lapply(splicies, st_buffer, dist = 90)
+st_buffer(x, dist = 90)
+
+ONEperCELL <- function(x){
+
+  dists <- sf::st_distance(x)
+}
+
+test <- st_intersects(splicies[[1]])
+
+test1 <- test[ lengths(test) > 1 ]
+untest <- unique(test1)
+
+sum( lengths(test)  > 1 ) # 637 records
+sum( lengths(test) == 2 ) # 285 records dupe in paired, 142 records...
+sum( lengths(test)  > 3 )
+
+test <- test[sum(lengths(test)  > 1 )]
+
+out <- unique(test)
+
+# identify each
+
+# find records < 90m distance
+
+```

### 2023-12-06 scripts/absence_drawer.R

@@ -68,7 +68,8 @@ absence_drawer <- function(x, bg_abs){
     sf::st_transform(sf::st_crs(x)) |>
     dplyr::rename(geometry = x)
   
-  absences <- bind_rows(reg_pts, AIM_absence, random_samples)
+  absences <- dplyr::bind_rows(reg_pts, AIM_absence, random_samples) |>
+    dplyr::mutate(Taxon = unique(x$taxon), .before = 'geometry')
   
   return(absences)
   

### 2023-12-07 scripts/download_dependent_variables.Rmd
fn to select newest record per cell

@@
   mutate(date = as.Date(datecollected)) %>% 
-  select(taxon = scientificname)
+  select(taxon = scientificname, date)
 
-aim <- st_read(file.path(base_p, 'AIM', 'AIM_records.shp')) %>% 
+aim <- st_read(file.path(base_p, 'AIM', 'AIM_records.shp'), quiet = TRUE) %>% 
   mutate(date = as.Date(DatVstd)) %>% 
@@
 
-gbif1 <- read.delim(file.path(base_p, 'gbif1', 'occurrence.txt'), na.strings = "") # %>% 
+gbif1 <- read.delim(file.path(base_p, 'gbif1', 'occurrence.txt'), na.strings = "") %>% 
   drop_na(decimalLongitude, decimalLatitude) %>% 
@@
 
-records <- st_transform(records, 5070)
-splicies <- split(records, records$taxon)[1:2]
-
-splicies <- lapply(splicies, st_buffer, dist = 90)
-st_buffer(x, dist = 90)
+splicies <- split(records, records$taxon)[1:10]
 
@@
 
-  dists <- sf::st_distance(x)
+  records <- sf::st_transform(x, 5070) |>
+    sf::st_buffer( dist = 90)
+
+  intersects <- sf::st_intersects(records)
+  intersect_sets <- unique(intersects) # identify intersected points
+
+  records_list <- vector(mode = 'list', length = length(intersect_sets))
+  for (i in seq_along(intersect_sets)){
+    records_list[[i]] <- records[unlist((intersect_sets[i])),]
   }
 
-test <- st_intersects(splicies[[1]])
+  records_list <- bind_rows(
+    lapply(records_list, function(x) x[which.max(x$date),])
+  )
 
-test1 <- test[ lengths(test) > 1 ]
-untest <- unique(test1)
+  return(records_list)
 
-sum( lengths(test)  > 1 ) # 637 records
-sum( lengths(test) == 2 ) # 285 records dupe in paired, 142 records...
-sum( lengths(test)  > 3 )
+}
 
-test <- test[sum(lengths(test)  > 1 )]
-
-out <- unique(test)
-
-# identify each
-
-# find records < 90m distance
+out <- lapply(splicies, ONEperCELL)
 
 ```
+
+```{r}
+multiplesPERcell <- function(x){
+
+  records <- sf::st_transform(x, 5070) |>
+    sf::st_buffer( dist = 90)
+
+  intersects <- sf::st_intersects(records)
+  intersect_sets <- unique(intersects) # identify intersected points
+
+  records_list <- vector(mode = 'list', length = length(intersect_sets))
+  for (i in seq_along(intersect_sets)){
+    records_list[[i]] <- records[unlist((intersect_sets[i])),]
+  }
+
+  max <- bind_rows(
+    lapply(records_list, function(x) x[which.max(x$date),])
+  )
+
+}
+```

### 2023-12-07 scripts/download_dependent_variables.Rmd
generating thinned records

@@
   filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
-  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') # artemisia, and larrea !
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
 
@@
 records <- bind_rows(aim, bien, gbif1, gbif2, idig) %>% 
-  mutate(taxon = str_to_sentence(taxon)) %>% 
+  mutate(taxon = str_to_sentence(taxon), 
+         taxon = str_squish(taxon)) %>% 
+  filter(taxon %in% unique(targets$taxon)) %>% 
   arrange(taxon) # recs need be > 90 m from each other
 
-splicies <- split(records, records$taxon)[1:10]
+rm(aim, bien, gbif1, gbif2, idig)
 
-ONEperCELL <- function(x){
+splicies <- split(records, records$taxon)
 
-  records <- sf::st_transform(x, 5070) |>
-    sf::st_buffer( dist = 90)
+thinned <- lapply(splicies, ONEperCELL)
+thinned <- bind_rows(thinned)
+write.csv(thinned, '../data/raw/occurrence/thinned_occurrences.csv', row.names = F)
+repeated_occurrences <- lapply(splicies, multiplesPERcell)
+repeated_occurrences <- purrr::discard(repeated_occurrences, ~nrow(.) < 2)
+repeated_occurrences <- bind_rows(repeated_occurrences)
+write.csv(repeated_occurrences, '../data/raw/occurrence/repeated_occurrences.csv', row.names = F)
 
-  intersects <- sf::st_intersects(records)
-  intersect_sets <- unique(intersects) # identify intersected points
-
-  records_list <- vector(mode = 'list', length = length(intersect_sets))
-  for (i in seq_along(intersect_sets)){
-    records_list[[i]] <- records[unlist((intersect_sets[i])),]
-  }
-
-  records_list <- bind_rows(
-    lapply(records_list, function(x) x[which.max(x$date),])
-  )
-
-  return(records_list)
-
-}
-
-out <- lapply(splicies, ONEperCELL)
-
+rm(ONEperCELL, multiplesPERcell, records, splicies, repeated_occurrences, base_p)
 ```
-
-```{r}
-multiplesPERcell <- function(x){
-
-  records <- sf::st_transform(x, 5070) |>
-    sf::st_buffer( dist = 90)
-
-  intersects <- sf::st_intersects(records)
-  intersect_sets <- unique(intersects) # identify intersected points
-
-  records_list <- vector(mode = 'list', length = length(intersect_sets))
-  for (i in seq_along(intersect_sets)){
-    records_list[[i]] <- records[unlist((intersect_sets[i])),]
-  }
-
-  max <- bind_rows(
-    lapply(records_list, function(x) x[which.max(x$date),])
-  )
-
-}
-```

### 2023-12-07 scripts/functions.R

@@ -15,3 +15,54 @@ idig_records <- function(query){
   }
   return(result)
 }
+
+ONEperCELL <- function(x){
+  
+  records <- sf::st_transform(x, 5070) |>
+    sf::st_buffer( dist = 90)
+  
+  intersects <- sf::st_intersects(records)
+  intersect_sets <- unique(intersects) # identify intersected points
+  
+  records_list <- vector(mode = 'list', length = length(intersect_sets))
+  for (i in seq_along(intersect_sets)){
+    records_list[[i]] <- records[unlist((intersect_sets[i])),]
+  }
+  
+  records_list <- bind_rows(
+    lapply(records_list, function(x) x[which.max(x$date),])
+  )
+  
+  return(records_list)
+  
+}
+
+multiplesPERcell <- function(x){
+  
+  records <- sf::st_transform(x, 5070) |>
+    sf::st_buffer( dist = 90)
+  
+  intersects <- sf::st_intersects(records)
+  intersect_sets <- unique(intersects[ lengths(intersects) > 1])
+  
+  records_list <- vector(mode = 'list', length = length(intersect_sets))
+  for (i in seq_along(intersect_sets)){
+    records_list[[i]] <- records[unlist((intersect_sets[i])),]
+  }
+  
+  min <- lapply(records_list, function(x) x[which.min(x$date), ])
+  max <- lapply(records_list, function(x) x[which.max(x$date), ])
+  dates_min <- dplyr::bind_rows(min); dates_max <- dplyr::bind_rows(max)
+  subsets <- dates_min$date != dates_max$date
+  
+  min <- min[subsets] ; max <- max[subsets]
+  same_sites <- dplyr::bind_rows(min, max) 
+  if(nrow(same_sites) > 1){
+  	same_sites <- dplyr::mutate(same_sites, SitePair = rep(1:length(min), times = 2), .before = geometry) |>
+      dplyr::arrange(SitePair)} else {
+    same_sites <- tibble(taxon = NA, date = NA, SitePair = NA, geometry = NA)}
+  
+  return(same_sites)
+  
+}
+

### 2023-12-07 scripts/download_dependent_variables.Rmd
generation of raw absences

@@
   mutate(date = as.Date(DatVstd)) %>% 
-  select(taxon, date) %>% 
+  select(taxon, date, PrmryKy) %>% 
   st_transform(4326)
@@
 thinned <- bind_rows(thinned)
-write.csv(thinned, '../data/raw/occurrence/thinned_occurrences.csv', row.names = F)
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences.shp', append = F)
 repeated_occurrences <- lapply(splicies, multiplesPERcell)
@@
 repeated_occurrences <- bind_rows(repeated_occurrences)
-write.csv(repeated_occurrences, '../data/raw/occurrence/repeated_occurrences.csv', row.names = F)
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences.shp', append = F)
 
@@
 ```
+
+```{r remove points outside analytical area}
+library(terra)
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+
+domain <- as.polygons(domain) |>  # convert to vector data
+  st_as_sf() |> # convert to simple feature
+  st_transform(5070)
+
+thinned <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE)
+thinned <- thinned[lengths(st_disjoint(thinned, domain)) ==  0, ]
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences.shp', append = FALSE)
+
+repeated_occurrences <- st_read('../data/raw/occurrence/repeated/repeated_occurrences.shp', quiet = TRUE)
+repeated_occurrences <- thinned[lengths(st_disjoint(repeated_occurrences, domain)) ==  0, ]
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences.shp', append = FALSE)
+```

### 2023-12-07 scripts/generate_absences.Rmd
generation of raw absences

@@
 pout <- '/media/steppe/hdd/SDM_restorations/data/raw/coarse_blm/'
-blm_surf <- st_read(paste0(pout, 'coarse_surface.shp'))
+blm_surf <- st_read(paste0(pout, 'coarse_surface.shp'), quiet = TRUE)
 blm_surf <- st_simplify(blm_surf, dTolerance = 500) # just make it a bit smaller. 
@@
 
-```{r}
+```{r generate raw absences, eval = F}
 
-aim <- st_read('../data/raw/occurrence/AIM/AIM_records.shp') %>% 
-  filter(SpcsLst == 'ACHY')
+records <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE)
+records <- records %>% 
+  group_by(taxon) %>% 
+  filter(n() >= 15)
+records <- split(records, f = records$taxon)
 
-bbox <- absence_drawer(aim, bg_abs = 0.15)
-
+out <- lapply(records, absence_drawer, bg_abs = 0.15)
+out <- bind_rows(out)
+st_write(out, '../data/raw/occurrence/absences.shp')
 ```

### 2023-12-07 scripts/absence_drawer.R

@@ -1,78 +1,0 @@
-absence_drawer <- function(x, bg_abs){
-  
-  
-  # create the observed hull, the buffered hull for random sample selection, 
-  # and a hull to form the basis for a bounding box to generate regular absences in
-  
-  observed_hull <- sf::st_union(x) |>
-    sf::st_convex_hull() |>
-    sf::st_sfc() 
-  
-  obs_area <- as.numeric(st_area(observed_hull)) * 0.15 / 2
-  obs_area <- if(obs_area > 160934){obs_area = 160934} # cap area with 100 mi buffer. 
-  
-  buffer_hull <- st_buffer(observed_hull, obs_area) 
-  
-  buf_area <- as.numeric(sf::st_area(buffer_hull) *0.1) / 2
-  obs_area <- if(buf_area > 40233){buf_area = 40233} # cap area with 25 mi buffer.
-  
-  box_hull <- sf::st_buffer(buffer_hull, dist = obs_area)
-  
-  bbox <- sf::st_bbox(box_hull) |>
-    sf::st_as_sfc() 
-  
-  box_hull <- sf::st_difference(bbox, buffer_hull)
-  
-  # perform the outline background absence sample
-  reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular') |>
-    sf::st_as_sf() |>
-    dplyr::mutate(Occurrence = 0, PtType = 'Background Absence') |>
-    dplyr::rename(geometry = x)
-  
-  rm(bbox, observed_hull)
-  
-  # identify AIM plots and generate absences. 
-  
-  blm_surf_area <- as.numeric(sf::st_area(blm_surf))
-  prcnt_blm <- (blm_surf_area / as.numeric(sf::st_area(buffer_hull))) 
-  aim_recs <- nrow(x) * (1 - bg_abs) * prcnt_blm
-  
-  AIM_points <- AIM_points[lengths(st_intersects(AIM_points, buffer_hull)) > 0,]
-  AIM_points <- AIM_points |> 
-    filter(! PrimaryKey %in% x$PrimaryKey ) # these are presences
-  
-  if(aim_recs >= nrow(AIM_points)){aim_recs <- nrow(AIM_points)} # in case species v. abundant. 
-  
-  AIM_absence <- sample_n(AIM_points, size = round(aim_recs * 1.1, 0), replace = FALSE) |> 
[…]

### 2023-12-07 scripts/functions.R

@@ -31,7 +31,8 @@ ONEperCELL <- function(x){
   
   records_list <- bind_rows(
     lapply(records_list, function(x) x[which.max(x$date),])
-  )
+  ) |>
+    sf::st_centroid()
   
   return(records_list)
   
@@ -59,10 +60,89 @@ multiplesPERcell <- function(x){
   same_sites <- dplyr::bind_rows(min, max) 
   if(nrow(same_sites) > 1){
   	same_sites <- dplyr::mutate(same_sites, SitePair = rep(1:length(min), times = 2), .before = geometry) |>
-      dplyr::arrange(SitePair)} else {
+      dplyr::arrange(SitePair)|>
+  	  sf::st_centroid() } else {
     same_sites <- tibble(taxon = NA, date = NA, SitePair = NA, geometry = NA)}
   
   return(same_sites)
   
 }
 
+absence_drawer <- function(x, bg_abs){
+  
+  
+  # create the observed hull, the buffered hull for random sample selection, 
+  # and a hull to form the basis for a bounding box to generate regular absences in
+  
+  observed_hull <- sf::st_union(x) |>
+    sf::st_convex_hull() |>
+    sf::st_sfc() 
+  
+  obs_area <- as.numeric(st_area(observed_hull)) * 0.15 / 2
+  obs_area <- if(obs_area > 160934){obs_area = 160934} # cap area with 100 mi buffer. 
+  
+  buffer_hull <- st_buffer(observed_hull, obs_area) 
+  
+  buf_area <- as.numeric(sf::st_area(buffer_hull) *0.1) / 2
+  obs_area <- if(buf_area > 40233){buf_area = 40233} # cap area with 25 mi buffer.
+  
+  box_hull <- sf::st_buffer(buffer_hull, dist = obs_area)
+  
+  bbox <- sf::st_bbox(box_hull) |>
+    sf::st_as_sfc() 
+  
+  box_hull <- sf::st_difference(bbox, buffer_hull)
+  
+  # perform the outline background absence sample
+  reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular') |>
+    sf::st_as_sf() |>
+    dplyr::mutate(Occurrence = 0, PtType = 'Background Absence') |>
+    dplyr::rename(geometry = x)
+  
+  rm(bbox, observed_hull)
+  
[…]

### 2023-12-08 scripts/functions.R
outlier detection methods

@@ -146,3 +146,14 @@ absence_drawer <- function(x, bg_abs){
   return(absences)
   
 }
+
+
+
+
+### calculate distances between records
+
+recordDist <- function(x){
+  
+  st_distance(pts, circles[nearest], by_element = TRUE)
+  
+}

### 2023-12-08 scripts/functions.R
manual review of outlier detection pilot

@@ -150,10 +150,56 @@ absence_drawer <- function(x, bg_abs){
 
 
 
-### calculate distances between records
 
-recordDist <- function(x){
+quantile_flagger <- function(x, quant, flag_threshold){
   
-  st_distance(pts, circles[nearest], by_element = TRUE)
+  if(missing(quant)){quant <- 0.025}
+  if(missing(flag_threshold)){flag_threshold <- 0.2}
+  
+  # perform calculations# 
+  quant_horse <- function(x, quant){
+    lwr <- quantile(x, quant, na.rm = TRUE)
+    upr <- quantile(x, 1 - quant, na.rm = TRUE)
+    
+    lwr_flag <- which(x <= lwr)
+    upr_flag <- which(x >= upr)
+    flags <- c(lwr_flag, upr_flag)
+    
+    return(list(flags))
+  }
+  x_ng <- sf::st_drop_geometry(x)
+  flagged_vals <- apply(x_ng[, grepl(pattern = 'bio*', colnames(x_ng))], MARGIN = 2, 
+                        FUN = quant_horse, quant)
+  
+  nvars <- ncol(x[, grepl(pattern = 'bio*', colnames(x))])
+  colsReq2flag <- round(flag_threshold * nvars, 0)
+  if(colsReq2flag < 2){colsReq2flag <- 2}
+  
+  counts <- table(unlist(flagged_vals)) >= colsReq2flag
+  flagged <- as.numeric(names(counts[counts == TRUE]))
+  
+  x['EnvFlag'] <- FALSE
+  x[flagged,'EnvFlag'] <- TRUE
+  
+  # now flag geographically farflung records
+  
+  dist <- as.numeric(
+    sf::st_distance(x, x[sf::st_nearest_feature(x),], by_element = TRUE)
+  )
+  
+  ub_geo <- quantile(dist, (1 - quant) ) 
+  x_out <- dplyr::mutate(x, distance = dist, 
+                         DistFlag = dplyr::if_else(dist > ub_geo, T, F),
+                         OutlierFlag = dplyr::case_when(
+                           DistFlag == T & EnvFlag == T ~ 'Dist-Env',
+                           DistFlag == T & EnvFlag == F ~ 'Dist',
+                           DistFlag == F & EnvFlag == T ~ 'Env',
+                           .default = NA,
+                         ),
+                         .before = geometry) |>
+    dplyr::select(-EnvFlag, DistFlag)
+  
+  return(x_out)
   
 }
+

### 2023-12-12 scripts/functions.R
more

@@ -197,7 +197,7 @@ quantile_flagger <- function(x, quant, flag_threshold){
                            .default = NA,
                          ),
                          .before = geometry) |>
-    dplyr::select(-EnvFlag, DistFlag)
+    dplyr::select(-EnvFlag, -DistFlag)
   
   return(x_out)
   

### 2023-12-14 scripts/download_dependent_variables.Rmd
downloaded misspelled taxa

@@
 targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
   filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
@@
 
-## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+## these species had initial errors, fix on Dec 14th, 
+target_taxa <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
 
@@
 
-gbif_taxon_keys <- target_taxa[294:364] %>% 
+gbif_taxon_keys <- target_taxa %>% 
   name_backbone_checklist()  %>% 
@@
 occ_download_wait(dl_key)
-occ_download_get( dl_key )
+occ_download_get(dl_key)
 
@@
 }
+
+rm(i)
 ```
@@
 
+```{r stitch together BIEN queries 2, eval = F}
+
+p <- '../data/raw/occurrence/bien'
+files <- file.path(p, list.files(p))
+# files <- files[ grep("Adenostoma_fasciculatum.csv|Brodiaea_elegans.csv|Krameria_bicolor.csv|Lotus_humistratus.csv", files) ]
+
+bien_data <-  do.call(rbind, lapply(files, read.csv)) %>% 
+  filter(datasource != 'GBIF' & date_collected > '1950-01-01', ! scrubbed_species_binomial %in% c('Acer negundo', 'Prosopis glandulosa')) %>%
+  drop_na(date_collected, latitude) %>% 
+  select(scrubbed_species_binomial:datasource, observation_type)
+
+write.csv(bien_data, '../data/raw/occurrence/BIEN_records2.csv', row.names = FALSE)
+```
+
 ```{r Perform idigbio queries, eval = F}
@@
 
-write.csv(idig, '../data/raw/occurrence/IDIG.csv')
-
+write.csv(idig, '../data/raw/occurrence/IDIG2.csv')
 ```
@@
 
+target_symbols <- c('ADFA', 'ADFAF', 'ADFAO', 'BREL', 'BRELE', 'KRGR', 'LOHU2')
+targets <- filter(targets, symbol %in% target_symbols) %>% 
+  distinct()
+
 spp_richness <- st_read(dsn = p1, 
@@
   select(SpeciesList, PrimaryKey) %>% 
+  # perform this a second time with new targets #
   filter(SpeciesList %in% target_symbols) %>% 
@@
 
-st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records.shp', append = F)
+st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records2.shp', append = F)
 
@@
 
-```{r Create consensus presence records}
+```{r Create consensus presence records 1}
 
@@
 
-```{r remove points outside analytical area}
+```{r Create consensus presence records}
+
+base_p <- '../data/raw/occurrence'
+
+bien <- read.csv(file.path(base_p, 'BIEN_records2.csv')) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) %>% 
+  mutate(date = as.Date(date_collected)) %>% 
+  select(taxon = scrubbed_species_binomial, date)
+
+idig <- read.csv(file.path(base_p, 'IDIG2.csv')) %>% 
+  st_as_sf(coords = c('geopoint.lon', 'geopoint.lat'), crs = 4326) %>% 
+  mutate(date = as.Date(datecollected)) %>% 
+  select(taxon = scientificname, date)
+
+aim <- st_read(file.path(base_p, 'AIM', 'AIM_records2.shp'), quiet = TRUE) %>% 
+  mutate(date = as.Date(DatVstd)) %>% 
+  select(taxon, date, PrmryKy) %>% 
+  st_transform(4326)
+
+gbif3 <- read.delim(file.path(base_p, 'gbif3', 'occurrence.txt'), na.strings = "") %>% 
+  drop_na(decimalLongitude, decimalLatitude) %>% 
+  filter(str_detect(decimalLatitude, '[A-Z]|[a-z]', negate = T)) %>% 
+  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'), crs = 4326) %>% 
+  mutate(date = as.Date(eventDate), .before = geometry, 
+         coordinateUncertaintyInMeters = as.numeric(coordinateUncertaintyInMeters)) %>% 
+  filter(coordinateUncertaintyInMeters <= 250) %>% 
+  select(taxon = species, date)
+
+records <- bind_rows(aim, bien, gbif3,  idig) %>% 
+  mutate(taxon = str_to_sentence(taxon), 
+         taxon = str_squish(taxon), 
+         taxon = str_replace(taxon, 'Adenostoma fasiculatum', 'Adenostoma fasciculatum'), 
+         taxon = str_replace(taxon, 'Brodiaeae elegans', 'Brodiaea elegans'), 
+         taxon = str_replace(taxon, 'Lotus humistratus|Acmispon humistratus', 'Acmispon brachycarpus')
+         ) %>% 
+  arrange(taxon) # recs need be > 90 m from each other
+
+rm(aim, bien, gbif3,  idig)
+
+splicies <- split(records, records$taxon)
+thinned <- lapply(splicies, ONEperCELL)
+thinned <- bind_rows(thinned)
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences2.shp', append = F)
+repeated_occurrences <- lapply(splicies, multiplesPERcell)
+repeated_occurrences <- purrr::discard(repeated_occurrences, ~nrow(.) < 2)
+repeated_occurrences <- bind_rows(repeated_occurrences)
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences2.shp', append = F)
+
+rm(ONEperCELL, multiplesPERcell, records, splicies, repeated_occurrences, base_p)
+```
+
+```{r remove points outside analytical area, eval = F}
 library(terra)
@@
 ```
+
+```{r remove points outside analytical area2, eval = F}
+library(terra)
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+
+domain <- as.polygons(domain) |>  # convert to vector data
+  st_as_sf() |> # convert to simple feature
+  st_transform(5070)
+
+thinned <- st_read('../data/raw/occurrence/thinned/thinned_occurrences2.shp', quiet = TRUE)
+thinned <- thinned[lengths(st_disjoint(thinned, domain)) ==  0, ]
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences2.shp', append = FALSE)
+
+repeated_occurrences <- st_read('../data/raw/occurrence/repeated/repeated_occurrences2.shp', quiet = TRUE)
+repeated_occurrences <- thinned[lengths(st_disjoint(repeated_occurrences, domain)) ==  0, ]
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences2.shp', append = FALSE)
+```

### 2023-12-16 scripts/functions.R
reading in data

@@ -203,3 +203,21 @@ quantile_flagger <- function(x, quant, flag_threshold){
   
 }
 
+
+colon_balloon <- function(x){
+  
+  if(stringr::str_detect(x, ':')){
+    
+    suppressWarnings({
+      ranges <- as.vector(str_extract_all(x, '[:digit:]{1,}:[:digit:]{1,}'))
+      lower <- stringr::str_extract_all(ranges, '[:digit:]{1,}(?=:)') # lower end of the sequence
+      upper <- stringr::str_extract_all(ranges, '(?<=:)[:digit:]{1,}') # upper end of the sequence
+      lower <- unlist(lapply(lower, as.numeric)) ; upper <- unlist(lapply(upper, as.numeric))
+      numeric_ranges <- as.numeric(unlist( mapply(seq, from = lower, to = upper)))
+    })
+    
+    return(numeric_ranges)
+  } else {return(NA)}
+  
+}
+

### 2023-12-17 scripts/generate_absences.Rmd
preparing lda fn for PA removals

@@
 st_write(out, '../data/raw/occurrence/absences.shp')
+
+ab <- st_read('../data/raw/occurrence/absences.shp')
 ```
+
+```{r remove pseudo absences in presence niche space}
+
+library(klaR)
+library(psych)
+library(MASS)
+library(ggord)
+
+# to determine whether the pseudo-random absences are in areas which are possibly suitable
+# we will use linear discrimination analysis (LDA). LDA uses a label, e.g. presence or absence
+# as a response which can be predicted by independent variables which are linear combinations 
+# generated by reducing a higher dimensional data set to two axis. LDA will be trained on a data set 
+# of AIM Plots with true absences, geographic absences, and a subset of the Psuedo-absences. 
+# Only up to 750:750 presences:absences per species will be used to train a model. The maximum
+# imbalance between presences to absences will be 750:250, or 3:1.
+
+lda_PA_dropper <- function(x, path){
+
+  pres <- x[x$Occurrence == 'Presence', ]
+  real_abs <- x[x$PtType == 'AIM Absence', ]
+  geo_abs <- x[x$PtType == 'Background Absence', ]
+  model_abs <- dplyr::bind_rows(real_abs, geo_abs)
+
+  if(nrow(pres) > 750){pres <- sample(1:nrow(pres), size = 750, replace = FALSE)}
+  if(nrow(model_abs) > 750){model_abs <- sample(1:nrow(model_abs), size = 750, replace = FALSE)}
+
+  training <- dplyr::bind_rows(pres, model_abs)
+  testing <- filter(x, ! ID %in% training$ID)
+
+  LDA_model <- lda(Species ~ ., training) # generate model
+
+  # Here the accuracy of the trained modeled is evaluated
+  LDA_train <- predict(LDA_model, training)$class
+  tab <- table(Predicted = LDA_train, Actual = training$Species)
+  ACC_train <- sum(diag(tab))/sum(tab) # accuracy
+
+  # Here the model is applied to unclassified data
+  LDA_test <- predict(LDA_model, testing)$class
+  testing$Predicted <- predict(LDA_model, testing)$class
+  tab1 <- table(Predicted = LDA_test, Actual = testing$Species)
+  ACC_test <- sum(diag(tab1))/sum(tab1) # accuracy.
+
+  n_pres <- nrow(x[x$Occurrence == 'Presence',])
+  n_abs <- nrow(x[x$Occurrence != 'Presence',])
+  conflicted <- filter(testing, Predicted != Occurrence) 
+  max_rcs_2_remove <- n_pres - n_ab
+
+  if(nrow(conflicted) == 0){ # identify the records we want to remove from the test data. 
+    removals <- NA
+  } else if (nrow(conflicted) > max_rcs_2_remove) {
+    removals <- conflicted[sample(1:nrow(conflicted), size = max_rcs_2_remove, replace = FALSE), 'ID']
+  } else {
+    removals <- conflicted$ID
+  }
+
+  if(!is.na(removals)){data_out <- dplyr::filter(x, ! ID %in% removals)}
+
+  metrics <- data.frame(
+    Presences = nrow(pres), 
+    Absences = nrow(model_abs),
+    ACC_train = ACC_train, 
+    ACC_test = ACC_test,
+    Flagged = nrow(conflicted)
+    )
+
+  species <- x$Taxon[1]
+  message(length(removals), ' records dropped from ', species)
+  species <- gsub(' ', '_', species)
+  write.csv(metrics, file.path(path, paste0(species, '.csv')), row.names = FALSE)
+
+  return(data_out)
+}
+
+```

### 2023-12-18 scripts/generate_absences.Rmd
remove ocean from absence_drawer, keep distance same and increase absence density

@@
 library(sf)
+library(terra)
 source('functions.R')
@@
 
+# here we subset a terrestrial surface to our area of analysis - we don't want all of
+# our random points to fall in the ocean
+
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+domain <- as.polygons(domain) |>  # convert to vector data
+  st_as_sf() |> # convert to simple feature
+  st_bbox()
+
+land <- rnaturalearth::ne_download(type= 'land', scale = 10, category= 'physical') |>
+  sf::st_as_sf()
+land <- st_crop(land, domain) |>
+  st_transform(5070)
+
+rm(domain)
+
 records <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE)
@@
 
-out <- lapply(records, absence_drawer, bg_abs = 0.15)
+out <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
 out <- bind_rows(out)
@@
 ab <- st_read('../data/raw/occurrence/absences.shp')
+
+table(records$taxon)
 ```
@@
 
-  LDA_model <- lda(Species ~ ., training) # generate model
+  LDA_model <- MASS::lda(Species ~ ., training) # generate model
 

### 2023-12-18 scripts/functions.R

@@ -68,22 +68,22 @@ multiplesPERcell <- function(x){
   
 }
 
-absence_drawer <- function(x, bg_abs){
+absence_drawer <- function(x, bg_abs, terrestrial){
   
   
   # create the observed hull, the buffered hull for random sample selection, 
   # and a hull to form the basis for a bounding box to generate regular absences in
   
-  observed_hull <- sf::st_union(x) |>
+  observed_hull <- sf::st_union(x) |> # this is the true hull around the observations
     sf::st_convex_hull() |>
     sf::st_sfc() 
   
-  obs_area <- as.numeric(st_area(observed_hull)) * 0.15 / 2
+  obs_area <- as.numeric(st_area(observed_hull)) * 0.15 / 2 # this hull should go at most 100 miles around the inner hull
   obs_area <- if(obs_area > 160934){obs_area = 160934} # cap area with 100 mi buffer. 
   
   buffer_hull <- st_buffer(observed_hull, obs_area) 
   
-  buf_area <- as.numeric(sf::st_area(buffer_hull) *0.1) / 2
+  buf_area <- as.numeric(sf::st_area(buffer_hull) *0.1) / 2 # this final hull will be used to create a box around the buffer hull
   obs_area <- if(buf_area > 40233){buf_area = 40233} # cap area with 25 mi buffer.
   
   box_hull <- sf::st_buffer(buffer_hull, dist = obs_area)
@@ -92,6 +92,11 @@ absence_drawer <- function(x, bg_abs){
     sf::st_as_sfc() 
   
   box_hull <- sf::st_difference(bbox, buffer_hull)
+  
+  # remove large oceans, or lakes, or what have you. 
+  observed_hull <- st_intersection(observed_hull, terrestrial)
+  buffer_hull <- st_intersection(buffer_hull, terrestrial)
+  box_hull <- st_intersection(box_hull, terrestrial)
   
   # perform the outline background absence sample
   reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular') |>

### 2023-12-18 scripts/generate_absences.Rmd
problems with random sampler, aim sample size resolved

@@
 blm_surf <- st_make_valid(blm_surf)
+
+rm(p, pout)
 ```
 
+After we have removed spurious occurrence records (see chunk: 'subset occurrence records for final clean batch', in clean_records), we can generate absences for each species. 
 ```{r generate raw absences, eval = F}
@@
 
-out <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
-out <- bind_rows(out)
-st_write(out, '../data/raw/occurrence/absences.shp')
+out1 <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
 
-ab <- st_read('../data/raw/occurrence/absences.shp')
+out <- bind_rows(out1) %>% 
+  rename(taxon = Taxon) %>% 
+  mutate(date = as.Date('2023-12-18'), .before = geometry)
 
-table(records$taxon)
+st_write(out, '../data/raw/occurrence/absences/absences.shp', append = FALSE)
+
+rm(out1, blm_surf, AIM_points, records)
+```
+
+```{r extract environmental predictors to records}
+
+absences <- st_read('../data/raw/occurrence/absences/absences.shp', quiet = TRUE)
+presences <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE) %>% 
+  mutate(Occurrence = 1, 
+         PtType = if_else(is.na(PrmryKy), 'AIM Presence', 'Opportunistic Presence'), .before = geometry)
+
+abvi_p <- filter(presences, taxon == 'Abronia villosa')
+abvi_a <- filter(absences, taxon == 'Abronia villosa')
+
+abvi <- bind_rows(abvi_p, abvi_a) %>% 
+  st_as_sf() %>% 
+  mutate(ID = 1:n())
+
+table(abvi$Occurrence)
+
+p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'
+paths <- file.path(p, list.files(p, recursive = F))
+files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
+names(files) <- basename(paths)
+bio <- lapply(files, vrt)
+bio <- rast(bio)
+
+vif_vals <- spatSample(bio, size = 5000)
+collinear_probs <- vifstep(vif_vals, th = 10)
+exclude()
+
+abvi_rv <- extract(bio, abvi)
+abvi_rv <- select(abvi_rv, -gdgfgd10, -gdgfgd5, -swe) %>% 
+  drop_na()
+abvi_rv_scaled <- data.frame(apply(abvi_rv[,2:17], MARGIN = 2, FUN = scales::rescale, to = c(0,1)))
+abvi_rv_scaled$ID <- abvi_rv$ID
+
+table(abvi_rv_scaled$Occurrence)
+
+abvi <- right_join(abvi, abvi_rv_scaled, by = 'ID')
+
+table(abvi$Occurrence)
+
+rm(abvi_a, abvi_p, p, paths, files, abvi_rv)
 ```
@@
 
-lda_PA_dropper <- function(x, path){
+lda_PA_dropper <- function(x, path, column_range){
 
-  pres <- x[x$Occurrence == 'Presence', ]
+  pres <- x[x$Occurrence == 1, ]
   real_abs <- x[x$PtType == 'AIM Absence', ]
@@
 
-  if(nrow(pres) > 750){pres <- sample(1:nrow(pres), size = 750, replace = FALSE)}
-  if(nrow(model_abs) > 750){model_abs <- sample(1:nrow(model_abs), size = 750, replace = FALSE)}
+  if(nrow(pres) > 750){pres <- pres[sample(1:nrow(pres), size = 750, replace = FALSE),]}
+  if(nrow(model_abs) > 750){model_abs <- model_abs[sample(1:nrow(model_abs), size = 750, replace = FALSE),]}
 
   training <- dplyr::bind_rows(pres, model_abs) 
+  training <- training[, c('Occurrence', colnames(abvIng)[7:22])]
   testing <- filter(x, ! ID %in% training$ID)
 
-  LDA_model <- MASS::lda(Species ~ ., training) # generate model
+  LDA_model <- MASS::lda(Occurrence ~ ., training) # generate model
 
@@
   LDA_train <- predict(LDA_model, training)$class
-  tab <- table(Predicted = LDA_train, Actual = training$Species)
+  tab <- table(Predicted = LDA_train, Actual = training$Occurrence)
   ACC_train <- sum(diag(tab))/sum(tab) # accuracy
@@
   testing$Predicted <- predict(LDA_model, testing)$class
-  tab1 <- table(Predicted = LDA_test, Actual = testing$Species)
+  tab1 <- table(Predicted = LDA_test, Actual = testing$Occurrence)
   ACC_test <- sum(diag(tab1))/sum(tab1) # accuracy.
 
-  n_pres <- nrow(x[x$Occurrence == 'Presence',])
-  n_abs <- nrow(x[x$Occurrence != 'Presence',])
+  n_pres <- nrow(x[x$Occurrence == 1,])
+  n_abs <- nrow(x[x$Occurrence == 0,])
   conflicted <- filter(testing, Predicted != Occurrence) 
-  max_rcs_2_remove <- n_pres - n_ab
+  max_rcs_2_remove <- n_abs - n_pres
 
@@
   metrics <- data.frame(
-    Presences = nrow(pres), 
-    Absences = nrow(model_abs),
+    Train_Presences = nrow(pres), 
+    Train_Absences = nrow(model_abs),
     ACC_train = ACC_train, 
     ACC_test = ACC_test,
-    Flagged = nrow(conflicted)
+    Flagged = nrow(conflicted), 
+    Removed = length(removals)
     )
 
-  species <- x$Taxon[1]
+  species <- x$taxon[1]
   message(length(removals), ' records dropped from ', species)
@@
 
+abvIng <- st_drop_geometry(abvi) %>% 
+  mutate(Occurrence = as.factor(Occurrence))
+
+TEST <- lda_PA_dropper(abvIng, path = '../results/lda/', column_range = 7:22)
+
+mapme <- left_join(select(abvi, ID, geometry), TEST)
+
+ggplot() +
+  geom_sf(data = mapme, aes(color = Occurrence))
+
+table(TEST$Occurrence)
+
 ```

### 2023-12-18 scripts/functions.R

@@ -87,16 +87,16 @@ absence_drawer <- function(x, bg_abs, terrestrial){
   obs_area <- if(buf_area > 40233){buf_area = 40233} # cap area with 25 mi buffer.
   
   box_hull <- sf::st_buffer(buffer_hull, dist = obs_area)
-  
   bbox <- sf::st_bbox(box_hull) |>
     sf::st_as_sfc() 
   
   box_hull <- sf::st_difference(bbox, buffer_hull)
   
   # remove large oceans, or lakes, or what have you. 
-  observed_hull <- st_intersection(observed_hull, terrestrial)
-  buffer_hull <- st_intersection(buffer_hull, terrestrial)
-  box_hull <- st_intersection(box_hull, terrestrial)
+  observed_hull <- sf::st_intersection(observed_hull, terrestrial) |> sf::st_union()
+  buffer_hull <- sf::st_intersection(buffer_hull, terrestrial) |> sf::st_union()
+  box_hull <- sf::st_intersection(box_hull, terrestrial) |> sf::st_union()
+  bbox <- sf::st_intersection(bbox, terrestrial) |> sf::st_union()
   
   # perform the outline background absence sample
   reg_pts <- sf::st_sample(box_hull, size = round(nrow(x) * bg_abs, 0), type = 'regular') |>
@@ -112,21 +112,20 @@ absence_drawer <- function(x, bg_abs, terrestrial){
     sf::st_transform(buffer_hull, sf::st_crs(blm_surf)), blm_surf)
   blm_surf_area <- as.numeric(sf::st_area(blm_surf))
   prcnt_blm <- (blm_surf_area / as.numeric(sf::st_area(buffer_hull))) 
-  aim_recs <- nrow(x) * (1 - bg_abs) * prcnt_blm
+  aim_recs <- round( nrow(x) * (1 - bg_abs) * prcnt_blm, 0)
   
   AIM_points <- AIM_points[lengths(st_intersects(AIM_points, buffer_hull)) > 0,]
   AIM_points <- AIM_points |> 
     filter(! PrimaryKey %in% x$PrmryKy ) # these are presences
   
-  if(aim_recs >= nrow(AIM_points)){aim_recs <- nrow(AIM_points)} # in case species v. abundant. 
+  if(aim_recs >= nrow(AIM_points)){aim_recs <- nrow(AIM_points)}
   
-  AIM_absence <- sample_n(AIM_points, size = round(aim_recs * 1.1, 0), replace = FALSE) |> 
[…]

### 2023-12-19 scripts/generate_absences.Rmd
notes on 0 variance error trip

--- 2023-12-18 scripts/generate_absences.Rmd
@@
 
-out <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
-out <- bind_rows(out)
-st_write(out, '../data/raw/occurrence/absences.shp')
+out1 <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
 
-ab <- st_read('../data/raw/occurrence/absences.shp')
+out <- bind_rows(out1) %>% 
+  rename(taxon = Taxon) %>% 
+  mutate(date = as.Date('2023-12-18'), .before = geometry)
 
-table(records$taxon)
+st_write(out, '../data/raw/occurrence/absences/absences.shp', append = FALSE)
+
+rm(out1, blm_surf, AIM_points, records)
+```
+
+```{r import raster stack}
+
+p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'
+paths <- file.path(p, list.files(p, recursive = F))
+files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
+names(files) <- basename(paths)
+bio <- lapply(files, vrt)
+bio <- rast(bio)
+bio <- subset(bio, c(1:11, 14:17, 19)) # these 'empty' need investigate
+
+rm(p, paths, files)
+```
+
+```{r subset rasterstack, eval = F}
+vif_vals <- spatSample(bio, size = 5000, values = TRUE) # this steps takes forever... 
+vif_vals <- vif_vals[ complete.cases(vif_vals), ] # 4148 records used
+collinear_probs <- usdm::vifstep(vif_vals, th = 10)
+bio <- usdm::exclude(bio, collinear_probs)
+
+write_rds(collinear_probs, '../results/bioclim_VIF_PA_drop.rda')
+
+rm(p, paths, files)
+```
+
+```{r import potential records}
+
+absences <- st_read('../data/raw/occurrence/absences/absences.shp', quiet = TRUE)
+presences <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE) %>% 
+  mutate(Occurrence = 1, 
+         PtType = if_else(is.na(PrmryKy), 'AIM Presence', 'Opportunistic Presence'), .before = geometry)
+
+recs <- bind_rows(absences, presences) %>% 
+  group_by(taxon) %>% 
+  mutate(ID = 1:n())
+
+collinear_probs <- read_rds( '../results/bioclim_VIF_PA_drop.rda')
+bio <- usdm::exclude(bio, collinear_probs)
+
+rm(absences, presences, collinear_probs)
 ```
@@
 
-library(klaR)
-library(psych)
-library(MASS)
-library(ggord)
+set.seed(27)
 
-# to determine whether the pseudo-random absences are in areas which are possibly suitable
-# we will use linear discrimination analysis (LDA). LDA uses a label, e.g. presence or absence
-# as a response which can be predicted by independent variables which are linear combinations 
-# generated by reducing a higher dimensional data set to two axis. LDA will be trained on a data set 
-# of AIM Plots with true absences, geographic absences, and a subset of the Psuedo-absences. 
-# Only up to 750:750 presences:absences per species will be used to train a model. The maximum
-# imbalance between presences to absences will be 750:250, or 3:1.
+cn <- names(bio)
+# splicies <- split(recs, f = recs$taxon)
+lapply(splicies, lda_PA_dropper, path = '../results/lda/', col_names = cn)
 
-lda_PA_dropper <- function(x, path){
-
-  pres <- x[x$Occurrence == 'Presence', ]
-  real_abs <- x[x$PtType == 'AIM Absence', ]
-  geo_abs <- x[x$PtType == 'Background Absence', ]
-  model_abs <- dplyr::bind_rows(real_abs, geo_abs)
-
-  if(nrow(pres) > 750){pres <- sample(1:nrow(pres), size = 750, replace = FALSE)}
-  if(nrow(model_abs) > 750){model_abs <- sample(1:nrow(model_abs), size = 750, replace = FALSE)}
-
-  training <- dplyr::bind_rows(pres, model_abs)
-  testing <- filter(x, ! ID %in% training$ID)
-
-  LDA_model <- MASS::lda(Species ~ ., training) # generate model
-
-  # Here the accuracy of the trained modeled is evaluated
-  LDA_train <- predict(LDA_model, training)$class
-  tab <- table(Predicted = LDA_train, Actual = training$Species)
-  ACC_train <- sum(diag(tab))/sum(tab) # accuracy
-
-  # Here the model is applied to unclassified data
-  LDA_test <- predict(LDA_model, testing)$class
-  testing$Predicted <- predict(LDA_model, testing)$class
-  tab1 <- table(Predicted = LDA_test, Actual = testing$Species)
-  ACC_test <- sum(diag(tab1))/sum(tab1) # accuracy.
-
-  n_pres <- nrow(x[x$Occurrence == 'Presence',])
-  n_abs <- nrow(x[x$Occurrence != 'Presence',])
-  conflicted <- filter(testing, Predicted != Occurrence) 
-  max_rcs_2_remove <- n_pres - n_ab
-
-  if(nrow(conflicted) == 0){ # identify the records we want to remove from the test data. 
-    removals <- NA
-  } else if (nrow(conflicted) > max_rcs_2_remove) {
-    removals <- conflicted[sample(1:nrow(conflicted), size = max_rcs_2_remove, replace = FALSE), 'ID']
-  } else {
-    removals <- conflicted$ID
-  }
-
-  if(!is.na(removals)){data_out <- dplyr::filter(x, ! ID %in% removals)}
-
-  metrics <- data.frame(
-    Presences = nrow(pres), 
-    Absences = nrow(model_abs),
-    ACC_train = ACC_train, 
-    ACC_test = ACC_test,
-    Flagged = nrow(conflicted)
-    )
-
-  species <- x$Taxon[1]
-  message(length(removals), ' records dropped from ', species)
-  species <- gsub(' ', '_', species)
-  write.csv(metrics, file.path(path, paste0(species, '.csv')), row.names = FALSE)
-
-  return(data_out)
-}
-
+which(names(splicies)[130] == 'Dudleya cymosa')
+# need to restart on record 130, 
+# which is Dudleya cymosa ssp. cymosa, and which should be subsumed into Dudleya cymosa!
+# the problem was that a variable displayed 0 variance because there were only 2 records!!!
 ```

### 2023-12-19 scripts/functions.R

@@ -225,3 +225,95 @@ colon_balloon <- function(x){
   
 }
 
+
+
+
+
+
+# to determine whether the pseudo-random absences are in areas which are possibly suitable
+# we will use linear discrimination analysis (LDA). LDA uses a label, e.g. presence or absence
+# as a response which can be predicted by independent variables which are linear combinations 
+# generated by reducing a higher dimensional data set to two axis. LDA will be trained on a data set 
+# of AIM Plots with true absences, geographic absences, and a subset of the Psuedo-absences. 
+# Only up to 750:750 presences:absences per species will be used to train a model. The maximum
+# imbalance between presences to absences will be 750:250, or 3:1. 
+
+lda_PA_dropper <- function(x, path, col_names){
+  
+  recs_w_vars <- terra::extract(bio, x)
+  
+  recs_w_vars_scaled <- data.frame(apply(recs_w_vars[,2:ncol(recs_w_vars)], MARGIN = 2, FUN = scales::rescale, to = c(0,1)))
+  recs_w_vars_scaled$ID <- x$ID
+  recs_w_vars_scaled<- tidyr::drop_na(recs_w_vars_scaled)
+  x1 <- dplyr::right_join(sf::st_drop_geometry(x), recs_w_vars_scaled, by = 'ID') 
+  
+  
+  pres <- x1[x1$Occurrence == 1, ]
+  real_abs <- x1[x1$PtType == 'AIM Absence', ]
+  geo_abs <- x1[x1$PtType == 'Background Absence', ]
+  model_abs <- dplyr::bind_rows(real_abs, geo_abs)
+  
+  if(nrow(pres) > 750){pres <- pres[sample(1:nrow(pres), size = 750, replace = FALSE),]}
+  if(nrow(model_abs) > 750){model_abs <- model_abs[sample(1:nrow(model_abs), size = 750, replace = FALSE),]}
+  
+  training <- dplyr::bind_rows(pres, model_abs) |>
+    sf::st_drop_geometry()
+  training <- training[, c('Occurrence', 'ID', col_names)]
+  
+  testing <- dplyr::filter(x1, ! ID %in% training$ID) |>
+    sf::st_drop_geometry()
+  
+  LDA_model <- MASS::lda(Occurrence ~ ., training) # generate model
+  
+  # Here the accuracy of the trained modeled is evaluated
[…]

### 2023-12-29 scripts/download_dependent_variables.Rmd
found number of cases where species detected on plot before and after fire

@@
 thinned <- bind_rows(thinned)
-st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences.shp', append = F)
+# st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences.shp', append = F)
 repeated_occurrences <- lapply(splicies, multiplesPERcell)
@@
 
-```{r Create consensus presence records}
+```{r Create consensus presence records 2, eval = F}
 

### 2023-12-29 scripts/generate_absences.Rmd
found number of cases where species detected on plot before and after fire

@@
 
+collinear_probs <- read_rds('../results/bioclim_VIF_PA_drop.rda')
+bio <- usdm::exclude(bio, collinear_probs)
+
 absences <- st_read('../data/raw/occurrence/absences/absences.shp', quiet = TRUE)
@@
 recs <- bind_rows(absences, presences) %>% 
+  mutate(taxon = str_replace(taxon, 'Dudleya cymosa ssp. cymosa', 'Dudleya cymosa')) %>% 
   group_by(taxon) %>% 
   mutate(ID = 1:n())
-
-collinear_probs <- read_rds( '../results/bioclim_VIF_PA_drop.rda')
-bio <- usdm::exclude(bio, collinear_probs)
 
@@
 
-```{r remove pseudo absences in presence niche space}
+```{r remove pseudo absences in presence niche space, eval = F}
 
 set.seed(27)
+cn <- names(bio)
+splicies <- split(recs, f = recs$taxon)
 
-cn <- names(bio)
-# splicies <- split(recs, f = recs$taxon)
 lapply(splicies, lda_PA_dropper, path = '../results/lda/', col_names = cn)
-
-which(names(splicies)[130] == 'Dudleya cymosa')
-# need to restart on record 130, 
-# which is Dudleya cymosa ssp. cymosa, and which should be subsumed into Dudleya cymosa!
-# the problem was that a variable displayed 0 variance because there were only 2 records!!!
 ```

### 2023-12-29 scripts/download_independent_variables.Rmd
adding empty scripts for steps
outlier detection methods
stuff

--- 2023-10-14 scripts/download_data.Rmd
@@
 library(tidyverse)
+library(terra)
 ```
@@
 The main bioclim variables can be found at the [link here](https://envicloud.wsl.ch/#/?bucket=https%3A%2F%2Fos.zhdk.cloud.switch.ch%2Fenvicloud%2F&prefix=chelsa%2Fchelsa_V2%2FGLOBAL%2Fclimatologies%2F1981-2010%2Fbio%2F) an S3 bucket for CHELS*A* (not the neighborhood!).
-All files can be downloaded as a textfile for further use with wget. 
+All files can be downloaded as a textile for further use with wget. 
 We will subset these files to only download the layers of interest.
@@
 
-#### Soil USDA Class ####
+```
 
+process dem
+```{r}
+p2dem <- '/media/steppe/hdd/Geospatial_data/merit-dem'
+files <- file.path(p2dem, list.files(p2dem, recursive = TRUE, pattern= 'flt'))
+files <- files[15:length(file)] # lower files south of extent
+
+dem <- vrt(files)
+dem <- crop(dem, project(domain, crs(dem)))
+dem <- project(dem, 'EPSG:5070')
+
+pout <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/dem/' 
+filename <- paste0(pout, 'dem_.tif')
+
+bio1 <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa/bio1'
+target_cells <- vrt(file.path(bio1, list.files(bio1)))
+dem <- resample(dem, target_cells)
+
+template <- rast(nrows = 25, ncols = 25) # create a big empty raster, you can go in through sf too. 
+ext(template) <- c( -125.5, -100, 27,  50) # set the extent
+crs(template) <- "EPSG:4326" 
+template <- project(template, 'EPSG:5070')
+
+makeTiles(dem, template, filename)
+
+rm(p2dem, files, dem, pout, filename, bio1, target_cells, template)
 ```

### 2023-12-29 scripts/download_independent_variables.Rmd
replacing N NAs with 0 in three rasters

@@
 ```
+
+these layers have NAs which we will treat as 0s
+```{r eval = F}
+
+p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'
+paths <- list.dirs(p, recursive = T)
+
+maskR <- vrt(file.path(p, 'bio1', list.files(file.path(p, 'bio1'))))
+
+gdgfgd10 <- vrt(file.path(p, 'gdgfgd10_old', list.files(file.path(p, 'gdgfgd10_old'))))
+gdgfgd5 <- vrt(file.path(p, 'gdgfgd5_old', list.files(file.path(p, 'gdgfgd5_old'))))
+swe <- vrt(file.path(p, 'swe_old', list.files(file.path(p, 'swe_old'))))
+
+gdgfgd10 <- subst(gdgfgd10, NA, 0)
+gdgfgd10 <- mask(gdgfgd10, maskR) # remove the ocean/set extent again
+filename <- paste0(p, '/gdgfgd10/', 'gdgfgd10_.tif')
+makeTiles(gdgfgd10, template, filename)
+
+gdgfgd5 <- subst(gdgfgd5, NA, 0)
+gdgfgd5 <- mask(gdgfgd5, maskR) # remove the ocean/set extent again
+filename <- paste0(p, '/gdgfgd5/', 'gdgfgd5_.tif')
+makeTiles(gdgfgd5, template, filename)
+
+swe <- subst(swe, NA, 0)
+swe <- mask(swe, maskR) # remove the ocean/set extent again
+filename <- paste0(p, '/swe/', 'swe_.tif')
+makeTiles(swe, template, filename)
+
+rm(gdgfgd10, gdgfgd5, swe, p, paths, maskR)
+```

### 2023-12-29 scripts/generate_RF_models.Rmd
replacing N NAs with 0 in three rasters

---
title: "Generate Random Forest Models"
author: "steppe"
date: "2023-10-14"
output: html_document
---

```{r}
library(tidyverse)
library(terra)
source('functions.R')
```

```{r read in predictors}

p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2'
paths <- list.dirs(p, recursive = T)
paths <- paths[2:length(paths)]
paths <- paths[! grepl('/chelsa$', paths) ]

# soils need to be uploaded seperately, e.g. by the depths
paths <- paths[! grepl('/soilgrids$', paths) ]

files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
names(files) <- basename(paths)

layers <- lapply(files, vrt)
layers <- rast(layers)
names(layers) <- basename(paths)
plot(layers[[1:3]])

# gdgfgd10", "gdgfgd5, "swe"
# both need NA replaced with 0, if terrestrial...

layers[[19]] <- subst(layers[[19]], NA, 0)
layers[[19]] <- mask(layers[[19]], layers[[1]]) # remove the ocean/set extent again

plot(layers[[22]])

```

```{r read in response}
p <- '../data/raw/occurrence/combined_records'
p2f <- file.path(p, list.files(p, pattern = '.shp$'))
train_dat <- lapply(p2f, sf::st_read, quiet = TRUE)
train_dat <- bind_rows(train_dat) %>% 
  st_transform(crs(layers))

abvi <- filter(train_dat, taxon == 'Abronia villosa')

abvi_test <- terra::extract(layers, abvi)
```

### 2023-12-30 scripts/download_independent_variables.Rmd
replacing NA with biological 0 in 3 rasters
processedsed earthenv landcover

@@
 
-rm(gdgfgd10, gdgfgd5, swe, p, paths, maskR)
+rm(gdgfgd10, gdgfgd5, swe, p, paths, maskR, filename, template)
 ```
+
+make tiles of shrub and herbaceous cover, and create total percent tree cover layers
+
+```{r}
+p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa/bio1'
+bio1 <- vrt(file.path(p, list.files(p))) # use this for resampling.
+
+p <- '/media/steppe/hdd/Geospatial_data/earthenv'
+f <- list.files(p)
+
+shrub <- vrt(file.path(p, f[5]))
+shrub <- crop(shrub, project(domain, crs(shrub)))
+shrub <- project(shrub, 'EPSG:5070')
+shrub <- resample(shrub, bio1)
+
+pout <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/veg_cover'
+filename <- paste0(pout, '/shrub/', 'shrub_.tif')
+makeTiles(shrub, template, filename)
+rm(shrub)
+
+herbs <- vrt(file.path(p, f[6]))
+herbs <- crop(herbs, project(domain, crs(herbs)))
+herbs <- project(herbs, 'EPSG:5070')
+herbs <- resample(herbs, bio1)
+
+filename <- paste0(pout, '/herb/', 'herb_.tif')
+makeTiles(herbs, template, filename)
+rm(herbs)
+
+layers <- lapply(file.path(p, f[1:4]), vrt)
+layers <- rast(layers)
+
+layers <- crop(layers, project(domain, crs(layers)))
+layers <- project(layers, 'EPSG:5070')
+layers <- resample(layers, bio1)
+
+tree <- sum(layers)
+filename <- paste0(pout, '/tree/', 'tree_.tif')
+makeTiles(tree, template, filename)
+rm(tree, layers)
+
+rm(p, f, pout, filename)
+```

### 2023-12-30 scripts/download_independent_variables.Rmd
reran soil grids

@@
 
-```{r download soil grids data eval = F}
+```{r download soil grids data,  eval = F}
 library(terra)
@@
 vrts <- paste0(var, '/', var, '_', rep(depth, each = length(var)), 'cm_mean.vrt') 
+vrts
+
+lapply(paste0(sg_url, vrts)[1:2], gdal_translate,  dst_dataset = "./crop_roi_igh_r.vrt",
+               of = "VRT", tr = c(250, 250),
+               projwin = domain,
+               projwin_srs = igh)
 
 gdal_translate(
-  paste0(sg_url, vrts),
+  paste0(sg_url, vrts[1]),
                "./crop_roi_igh_r.vrt",
@@
 
-#### salinity ####
+```
 
+```{r create soil grids products, eval = F}
+
+p2grids <- '/media/steppe/hdd/Geospatial_data/soilgrids'
+prods <- list.files(p2grids)
+tifs <- list.files(p2grids, recursive = TRUE)
+tifs <- tifs[grepl('cm_mean', tifs)]
+
+# tifs is input.
+
+gridder <- function(x){
+
+  product <- terra::vrt(file.path(p2grids, x))
+  product <- terra::crop(product, terra::project(domain, terra::crs(product)))
+  product <- terra::project(product, 'EPSG:5070')
+  product <- resample(product, target_cells)
+
+  out_root <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/soilgrids'
+  paths <- gsub('cm_.*', '', x)
+  paths <- file.path(out_root, gsub('-', '_', paths))
+
+  filename <- file.path(paths, basename(x))
+  makeTiles(product, template, filename)
+
+}
+
+lapply(tifs, gridder)
+
+p1 <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/soilgrids/cfvo/cfvo_0_5'
+cfvo5 <- vrt(file.path(p1, list.files(p1)))
+plot(cfvo5)
 ```
@@
 makeTiles(herbs, template, filename)
-rm(herbs)
 

### 2023-12-30 scripts/generate_RF_models.Rmd
replacing NA with biological 0 in 3 rasters
reran soil grids

@@
 library(terra)
+library(sf)
 source('functions.R')
@@
 paths <- paths[2:length(paths)]
-paths <- paths[! grepl('/chelsa$', paths) ]
-
-# soils need to be uploaded seperately, e.g. by the depths
-paths <- paths[! grepl('/soilgrids$', paths) ]
+paths <- paths[! grepl('chelsa$|veg_cover$|soilgrids$|sand$|silt$|clay$|phh2o$|cfvo$', paths) ]
 
@@
 names(layers) <- basename(paths)
-plot(layers[[1:3]])
-
-# gdgfgd10", "gdgfgd5, "swe"
-# both need NA replaced with 0, if terrestrial...
-
-layers[[19]] <- subst(layers[[19]], NA, 0)
-layers[[19]] <- mask(layers[[19]], layers[[1]]) # remove the ocean/set extent again
-
-plot(layers[[22]])
-
 ```
@@
 abvi_test <- terra::extract(layers, abvi)
+
+# identify a single component of each pair for usage. 
 ```

### 2023-12-30 scripts/generate_RF_models.Rmd
funs for rf modelling

@@
 paths <- paths[2:length(paths)]
-paths <- paths[! grepl('chelsa$|veg_cover$|soilgrids$|sand$|silt$|clay$|phh2o$|cfvo$', paths) ]
+paths <- paths[! grepl('chelsa$|veg_cover$|soilgrids$|sand$|silt$|clay$|phh2o$|cfvo$|soc$', paths) ]
 
@@
 names(layers) <- basename(paths)
+
+rm(paths, files)
 ```
@@
 
-abvi <- filter(train_dat, taxon == 'Abronia villosa')
+splicies <- split(train_dat, f = train_dat$taxon)
 
-abvi_test <- terra::extract(layers, abvi)
+cores <- parallel::detectCores()
+lapply(splicies[3:335], modeller)
 
-# identify a single component of each pair for usage. 
 ```

### 2023-12-30 scripts/Boruta_fun.R

@@ -1,0 +1,35 @@
+cores <- parallel::detectCores() # run this outside fn to avoid repeat. 
+
+Boruta_var_selector <- function(x){
+  
+  BorutaRes <- Boruta::Boruta(Occurrence ~ ., data = x, num.threads = cores, doTrace = 1)
+  importance <- Boruta::attStats(BorutaRes)
+  rn <- rownames(importance)
+  important_vars <- Boruta::getSelectedAttributes(BorutaRes, withTentative = F)
+  
+  
+  ### Remove highly autocorrelated variables ###
+  co_pairs <- c('cfvo', 'phh2o', 'gdd', 'ngd', 'gdgfgd')
+  pair_removals <- unlist(lapply(co_pairs, function(x){ rn[min(grep(x, rn))]}))
+  
+  triplets <- c('sand', 'silt', 'clay')
+  triplet_removals <- unlist(lapply(triplets, function(x){ rn[grep(x, rn)[1:2]]}))
+
+  # reduce two the maximum of three soil textures
+  sand_imp <- mean(importance [ grep('sand', rn), 'meanImp'])
+  silt_imp <- mean(importance [ grep('silt', rn), 'meanImp'])
+  clay_imp <- mean(importance [ grep('clay', rn), 'meanImp'])
+  
+  texture2remove <- c('sand', 'silt', 'clay')[which.min(c(sand_imp, silt_imp, clay_imp))]
+  textures2remove <- rn[ grep(texture2remove, rn)]
+  
+  removals <- unique(c(pair_removals, triplet_removals, textures2remove))
+  
+  x_sub <- x[ , !names(x) %in% removals]# remove
+  x_sub <- x_sub[, names(x_sub) %in% c('Occurrence', important_vars)] # keep informative
+  
+  return(x_sub)
+}
+
+
+

### 2023-12-30 scripts/functions.R

@@ -317,3 +317,88 @@ lda_PA_dropper <- function(x, path, col_names){
 }
 
 
+
+Boruta_var_selector <- function(x){
+  
+  BorutaRes <- Boruta::Boruta(Occurrence ~ ., data = x, num.threads = cores, doTrace = 1)
+  importance <- Boruta::attStats(BorutaRes)
+  rn <- rownames(importance)
+  important_vars <- Boruta::getSelectedAttributes(BorutaRes, withTentative = F)
+  
+  
+  ### Remove highly autocorrelated variables ###
+  co_pairs <- c('cfvo', 'phh2o', 'gdd', 'ngd', 'gdgfgd')
+  pair_removals <- unlist(lapply(co_pairs, function(x){ rn[min(grep(x, rn))]}))
+  
+  triplets <- c('sand', 'silt', 'clay')
+  triplet_removals <- unlist(lapply(triplets, function(x){ rn[grep(x, rn)[1:2]]}))
+  
+  # reduce two the maximum of three soil textures
+  sand_imp <- mean(importance [ grep('sand', rn), 'meanImp'])
+  silt_imp <- mean(importance [ grep('silt', rn), 'meanImp'])
+  clay_imp <- mean(importance [ grep('clay', rn), 'meanImp'])
+  
+  texture2remove <- c('sand', 'silt', 'clay')[which.min(c(sand_imp, silt_imp, clay_imp))]
+  textures2remove <- rn[ grep(texture2remove, rn)]
+  
+  removals <- unique(c(pair_removals, triplet_removals, textures2remove))
+  
+  x_sub <- x[ , !names(x) %in% removals]
+  return(x_sub)
+}
+
+
+
+#' perform random forest modelling and save outputs to locations
+#' test_data output from Boruta_var_selector
+randomForests <- function(train_data, test_data, species){
+  
+  # identify the appropriate mtry
+  opt_mtry <- data.frame(
+    randomForest::tuneRF(train_data[,-1], train_data[,1], stepFactor=1.5))
+  opt_mtry <- opt_mtry[ which.min(opt_mtry$OOBError), 'mtry'] # here we subset the lowest OOB. 
+  
+  # model and predict
+  rf_model <- randomForest::randomForest(Occurrence ~ ., data = train_data, mtry = opt_mtry)
+  prediction <- predict(rf_model, test_data)
+  cmRestrat <- caret::confusionMatrix(prediction, test_data$Occurrence)
+  
+  # plot and save results
[…]

### 2023-12-31 scripts/generate_RF_models.Rmd
all models take

@@
 
-```{r read in response}
+```{r read in response data and model}
 p <- '../data/raw/occurrence/combined_records'
@@
 ```
+
+```{r summarize results from RF models}
+
+psum <- '../results/summary'
+f <- file.path(psum, list.files(psum))
+
+summary <- lapply(f, read.csv)
+names(summary) <- gsub('.csv', '',  basename(f))
+summary <- data.table::rbindlist(summary, idcol = 'species')
+
+filter(summary, Metric == 'Balanced Accuracy') %>% 
+  ggplot(aes(x = Value)) + 
+  geom_histogram() +
+  geom_density()
+
+filter(summary, Metric == 'Sensitivity') %>% 
+  ggplot(aes(x = Value)) + 
+  geom_histogram()
+```

### 2024-01-02 scripts/predict_models.Rmd
fn to predict results onto maps

---
title: "Predict Models"
author: "steppe"
date: "2024-01-02"
output: html_document
---

```{r}
library(tidyverse)
library(terra)
library(sf)
source('functions.R')
```

```{r read in predictors}

p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2'
paths <- list.dirs(p, recursive = T)
paths <- paths[2:length(paths)]
paths <- paths[! grepl('chelsa$|veg_cover$|soilgrids$|sand$|silt$|clay$|phh2o$|cfvo$|soc$', paths) ]

files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
names(files) <- basename(paths)

layers <- lapply(files, vrt)
layers <- rast(layers)
names(layers) <- basename(paths)

rm(paths, files)
```

```{r read in models}

pmodels <- '../results/rf_models'
f <- file.path(pmodels, list.files(pmodels))

occpath <- '../data/raw/occurrence/combined_records'
occ_files <- list.files(occpath, pattern = 'shp$')

models_occ <- data.frame(
  path2model = f, occurrence_data = occ_files
) %>% 
  mutate(
    species = gsub('-.*$', '', basename(path2model)),
    match = if_else(species == gsub('[.]shp$', '', occurrence_data), T, F )
  ) # all look good!

models_occ <- split(models_occ, f = models_occ$species)

predict_wrapper <- function(x){

  sdm <- readRDS(x$path2model)
  occ_path <- file.path(occpath, x$occurrence_data)

  # read in occurrences and subset prediction to 100 miles from border
  bound <- sf::st_read(occ_path, quiet = TRUE) |>
    dplyr::filter(Occurrence == 1) |>
    sf::st_union() |>
    sf::st_convex_hull() |>
    sf::st_buffer(160934) |>
    sf::st_bbox()

  pout <- '../results/suitability_maps'
  layer_subset <- terra::crop(layers, bound)
  terra::predict(
    type = 'prob', cores = 8,
    layer_subset, sdm, cpkgs="randomForest",
    filename = file.path(pout, paste0(x$species, '.tif')),
    wopt = c(names = 'predicted_suitability'))

}

lapply(models_occ[3], predict_wrapper)

r <- terra::rast('../results/suitability_maps/Acamptopappus_sphaerocephalus.tif')
plot(r)

```

```{r}

g <- readRDS('../results/rf_models/Abronia_villosa-2023-12-30_14:48:33.rds')

```

### 2024-01-02 scripts/functions.R

@@ -364,8 +364,9 @@ randomForests <- function(train_data, test_data, species){
   cmRestrat <- caret::confusionMatrix(prediction, test_data$Occurrence)
   
   # plot and save results
+  species <- gsub(' ', '_', species)
   saveRDS(rf_model,
-          file = paste0('../results/rf_models/', species, '-', Sys.time(), '.rds'))
+          file = paste0('../results/rf_models/', species, '-', gsub(' ', '_', Sys.time()), '.rds'))
   
   vip::vip(rf_model) +
     theme_bw() +
@@ -400,5 +401,29 @@ randomForests <- function(train_data, test_data, species){
     setNames(data.frame(as.numeric(result.roc$auc), 'AUC'), c('Value', 'Metric')), 
     summary, samples)
   write.csv(summary, paste0('../results/summary/', species, '.csv'), row.names = F)
+  
+}
+
+modeller <- function(x){
+  
+  species <- sf::st_drop_geometry(x) %>% 
+    dplyr::pull(taxon)
+  species <- species[1]
+  x1 <- terra::extract(layers, x)
+  
+  x_dat <- cbind(x$Occurrence, x1) %>% 
+    dplyr::mutate(Occurrence = as.factor(`x$Occurrence`)) %>% 
+    dplyr::select(-ID, -`x$Occurrence`) %>% 
+    dplyr::relocate(Occurrence, .before = 1) %>% 
+    dplyr::filter(if_all(2:ncol(.), ~ !is.na(.)))
+  
+  trainIndex <- caret::createDataPartition(x_dat$Occurrence, p = .8, 
+                                           list = FALSE, times = 1)
+  TRAIN <- x_dat[ trainIndex,]
+  TEST  <- x_dat[-trainIndex,]
+  
+  TRAIN_DATA <- Boruta_var_selector(TRAIN)
+  randomForests(train_data = TRAIN_DATA, test_data = TEST, species)
+  message(species, ' complete')
   
 }

### 2024-01-03 scripts/predict_models.Rmd
subset layers before crop

@@
 ```{r read in predictors}
+
+setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE=1024"))
 
@@
     sf::st_convex_hull() |>
-    sf::st_buffer(160934) |>
+    sf::st_buffer(80467) |>
     sf::st_bbox()
@@
   pout <- '../results/suitability_maps'
-  layer_subset <- terra::crop(layers, bound)
+  layers <- terra::subset(layers, term_grab(sdm))
+  layer_subset <- terra::crop(rast(layers, bound)
   terra::predict(
-    type = 'prob', cores = 8,
+    type = 'prob', cores = 16,
     layer_subset, sdm, cpkgs="randomForest",
@@
 
-lapply(models_occ[3], predict_wrapper)
+setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE"),  c(""))
 
-r <- terra::rast('../results/suitability_maps/Acamptopappus_sphaerocephalus.tif')
-plot(r)
-
+lapply(models_occ[c(5)], predict_wrapper)
 ```
@@
 ```{r}
-
-g <- readRDS('../results/rf_models/Abronia_villosa-2023-12-30_14:48:33.rds')
-
+achy <- rast('../results/suitability_maps/Achnatherum_hymenoides.tif')
+plot(achy[[2]])
 ```

### 2024-01-03 scripts/functions.R

@@ -427,3 +427,40 @@ modeller <- function(x){
   message(species, ' complete')
   
 }
+
+
+predict_wrapper <- function(x){
+  
+  sdm <- readRDS(x$path2model)
+  occ_path <- file.path(occpath, x$occurrence_data)
+  
+  # read in occurrences and subset prediction to 100 miles from border
+  bound <- sf::st_read(occ_path, quiet = TRUE) |>
+    dplyr::filter(Occurrence == 1) |>
+    sf::st_union() |>
+    sf::st_convex_hull() |>
+    sf::st_buffer(160934) |>
+    sf::st_bbox()
+  
+  pout <- '../results/suitability_maps'
+  layer_subset <- terra::crop(layers, bound)
+  terra::predict(
+    type = 'prob', cores = 16,
+    layer_subset, sdm, cpkgs="randomForest",
+    filename = file.path(pout, paste0(x$species, '.tif')),
+    wopt = c(names = 'predicted_suitability'))
+  
+}
+
+
+term_grab <- function(x){
+  
+  pieces <- unlist(
+    strsplit(
+      as.character(x[['terms']][[3]]),
+      " ")
+  )
+  
+  terms <- pieces[grepl('[A-z]', pieces)]
+  return(terms)
+}

### 2024-01-03 scripts/generate_RF_models_copy.Rmd
predict too slow, need to reduce feature using caret::rfe

--- 2023-12-31 scripts/generate_RF_models.Rmd
@@
 cores <- parallel::detectCores()
-lapply(splicies[3:335], modeller)
+lapply(splicies[1], modeller)
+
+```
+
+```{r}
+x <- filter(train_dat, taxon==  'Asclepias cryptoceras' )
+
+  species <- sf::st_drop_geometry(x) %>% 
+    dplyr::pull(taxon)
+  species <- species[1]
+  x1 <- terra::extract(layers, x)
+
+  x_dat <- cbind(x$Occurrence, x1) %>% 
+    dplyr::mutate(Occurrence = as.factor(`x$Occurrence`)) %>% 
+    dplyr::select(-ID, -`x$Occurrence`) %>% 
+    dplyr::relocate(Occurrence, .before = 1) %>% 
+    dplyr::filter(if_all(2:ncol(.), ~ !is.na(.)))
+
+  trainIndex <- caret::createDataPartition(x_dat$Occurrence, p = .8, 
+                                           list = FALSE, times = 1)
+  TRAIN <- x_dat[ trainIndex,]
+  TEST  <- x_dat[-trainIndex,]
+
+  library(caret)
+
+TRAIN <- Boruta_var_selector(TRAIN)
+
+ctrl <- rfeControl(functions = rfFuncs,
+                   method = "repeatedcv", repeats = 5)
+
+floor((ncol(TRAIN)-1)*0.1)*10 # upper subset
+subsets = c(30, 25, 20, 15, 10)
+
+rfProfile <- rfe(TRAIN[,2:ncol(TRAIN)], TRAIN$Occurrence,
+                 sizes = subsets,
+                 rfeControl = ctrl)
+
+# accept a percent change with less than -1.5% 
+results <- rfProfile[['results']] %>% 
+    mutate(pct_change = (Accuracy/lead(Accuracy) - 1) * 100)
+
+predictors(rfProfile)
+plot(rfProfile)
+
+v_no <- filter(results, pct_change > -1.5) %>% 
+  # identify all selections which performed well relative to maximal model 
+  slice_min(Variables, n = 1) %>%  # now identify the one with the fewest variables
+  pull(Variables)
+
+rfProfile[["variables"]] %>% 
+  filter(Variables == v_no) %>% 
+  group_by(var) %>% 
+  summarise(imp = mean(Overall)) %>% 
+  arrange(-imp) # tri, slope, cfvo_30_60, phh2o_0_5, ngd10
 

### 2024-01-03 scripts/predict_models.Rmd
predict too slow, need to reduce feature using caret::rfe

@@
   pout <- '../results/suitability_maps'
-  layers <- terra::subset(layers, term_grab(sdm))
-  layer_subset <- terra::crop(rast(layers, bound)
-  terra::predict(
-    type = 'prob', cores = 16,
-    layer_subset, sdm, cpkgs="randomForest",
-    filename = file.path(pout, paste0(x$species, '.tif')),
-    wopt = c(names = 'predicted_suitability'))
+  layer_subset <- terra::subset(layers, term_grab(sdm))
+  layer_subset <- terra::crop(layer_subset, bound)
+
+  return(layer_subset)
+  #terra::predict(
+  #  type = 'prob', cores = 16,
+  #  layer_subset, sdm, cpkgs="randomForest",
+  #  filename = file.path(pout, paste0(x$species, '.tif')),
+  #  wopt = c(names = 'predicted_suitability'))
 
@@
 
-setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE"),  c(""))
+ob <- lapply(models_occ[c(5)], predict_wrapper)
+ob1 <- ob[[1]]
+ob1
 
-lapply(models_occ[c(5)], predict_wrapper)
+predict_wrapper(models_occ[[5]])
 ```

### 2024-01-03 scripts/functions.R

@@ -349,6 +349,8 @@ Boruta_var_selector <- function(x){
 
 
 
+
+
 #' perform random forest modelling and save outputs to locations
 #' test_data output from Boruta_var_selector
 randomForests <- function(train_data, test_data, species){

### 2024-01-04 scripts/predict_models.Rmd
perform rfe to subset predictors to reasonable size for prediction onto a gridded surface incroporated into modeller

@@
   return(layer_subset)
-  #terra::predict(
-  #  type = 'prob', cores = 16,
-  #  layer_subset, sdm, cpkgs="randomForest",
-  #  filename = file.path(pout, paste0(x$species, '.tif')),
-  #  wopt = c(names = 'predicted_suitability'))
+  surface <- terra::subset(terra::predict(
+    type = 'prob', cores = 16,
+    layer_subset, sdm, cpkgs="randomForest",
+    wopt = c(names = 'predicted_suitability'))
+  , 2)
+  names(surface) <- 'Hyp.Suitability'
+  writeRaster(surface, filename = file.path(pout, paste0(x$species, '.tif')))
 

### 2024-01-04 scripts/functions.R

@@ -325,7 +325,6 @@ Boruta_var_selector <- function(x){
   rn <- rownames(importance)
   important_vars <- Boruta::getSelectedAttributes(BorutaRes, withTentative = F)
   
-  
   ### Remove highly autocorrelated variables ###
   co_pairs <- c('cfvo', 'phh2o', 'gdd', 'ngd', 'gdgfgd')
   pair_removals <- unlist(lapply(co_pairs, function(x){ rn[min(grep(x, rn))]}))
@@ -425,6 +424,7 @@ modeller <- function(x){
   TEST  <- x_dat[-trainIndex,]
   
   TRAIN_DATA <- Boruta_var_selector(TRAIN)
+  TRAIN_DATA <- rfe_var_selector(TRAIN_DATA)
   randomForests(train_data = TRAIN_DATA, test_data = TEST, species)
   message(species, ' complete')
   
@@ -466,3 +466,41 @@ term_grab <- function(x){
   terms <- pieces[grepl('[A-z]', pieces)]
   return(terms)
 }
+
+
+rfe_var_selector <- function(x){
+  
+  ctrl <- caret::rfeControl(functions = rfFuncs, method = "repeatedcv", number = 10,
+                            repeats = 5, rerank = TRUE, allowParallel = TRUE)
+  
+  subsets = seq(from = 10, to = (floor((ncol(x)-1)*0.1)*10), by = 5)
+  
+  cl <- parallel::makeCluster(cores, type='PSOCK')
+  doParallel::registerDoParallel(cl)
+  rfProfile <- caret::rfe(x[,2:ncol(x)], x$Occurrence,
+                          sizes = subsets, rfeControl = ctrl)
+  ParallelLogger::stopCluster(cl)
+  
+  # accept a percent change with less than -1.5% 
+  results <- rfProfile[['results']] %>% 
+    dplyr::mutate(pct_change = (Accuracy/lead(Accuracy) - 1) * 100)
+  
+  v_no <- if(any(results$pct_change > 0 & results$Variables < 30)){ # here if performance increases, select the
+    A <- results[results$pct_change  > 0,] # increased value with the fewest number of terms. Do not allow the 
+    v_no <- A[which.min(A$Variables), 'Variables'] # 30 vars!! that is essentially where we started.
+  } else { # if perfomance only moderately decreases across all models with fewer vars, select that with 
[…]

### 2024-01-04 scripts/predict_models.Rmd
more rfe subset logic

--- 2024-01-03 scripts/predict_models.Rmd
@@
   return(layer_subset)
-  #terra::predict(
-  #  type = 'prob', cores = 16,
-  #  layer_subset, sdm, cpkgs="randomForest",
-  #  filename = file.path(pout, paste0(x$species, '.tif')),
-  #  wopt = c(names = 'predicted_suitability'))
+  surface <- terra::subset(terra::predict(
+    layer_subset, sdm, cpkgs="randomForest",
+    type = 'prob', cores = 16,
+    wopt = c(names = 'predicted_suitability'))
+  , 2)
+  names(surface) <- 'Hyp.Suitability'
+  writeRaster(surface, filename = file.path(pout, paste0(x$species, '.tif')))
 

### 2024-01-04 scripts/functions.R

@@ -383,7 +383,7 @@ randomForests <- function(train_data, test_data, species){
     result.roc,
     print.thres="best",
     print.thres.best.method="closest.topleft")
-  dev.off()
+  dev.off(); graphics.off()
   
   summary <- setNames(
     rbind(stack(cmRestrat$byClass), stack(cmRestrat$overall)), 
@@ -400,6 +400,7 @@ randomForests <- function(train_data, test_data, species){
   )
   summary <- rbind(
     setNames(data.frame(as.numeric(result.roc$auc), 'AUC'), c('Value', 'Metric')), 
+    setNames(data.frame((ncol(train_data)-1), 'Independent_Variables'), c('Value', 'Metric')), 
     summary, samples)
   write.csv(summary, paste0('../results/summary/', species, '.csv'), row.names = F)
   
@@ -470,7 +471,7 @@ term_grab <- function(x){
 
 rfe_var_selector <- function(x){
   
-  ctrl <- caret::rfeControl(functions = rfFuncs, method = "repeatedcv", number = 10,
+  ctrl <- caret::rfeControl(functions = caret::rfFuncs, method = "repeatedcv", number = 10,
                             repeats = 5, rerank = TRUE, allowParallel = TRUE)
   
   subsets = seq(from = 10, to = (floor((ncol(x)-1)*0.1)*10), by = 5)
@@ -485,8 +486,8 @@ rfe_var_selector <- function(x){
   results <- rfProfile[['results']] %>% 
     dplyr::mutate(pct_change = (Accuracy/lead(Accuracy) - 1) * 100)
   
-  v_no <- if(any(results$pct_change > 0 & results$Variables < 30)){ # here if performance increases, select the
-    A <- results[results$pct_change  > 0,] # increased value with the fewest number of terms. Do not allow the 
+  v_no <- if(any(results$pct_change > 0 & results$Variables < 20)){ # here if performance increases, select the
+    A <- results[results$pct_change  > 0 & results$Variables < 20,] # increased value with the fewest number of terms. Do not allow the 
     v_no <- A[which.min(A$Variables), 'Variables'] # 30 vars!! that is essentially where we started.
   } else { # if perfomance only moderately decreases across all models with fewer vars, select that with 
     B <- results[results$pct_change > -1.5,] # the fewest terms, but with less than a 1.5% drop in performance
[…]

### 2024-01-06 scripts/predict_models.Rmd
pushed processed onto ssd, avoiding parallel to max RAM on single thread

--- 2024-01-03 scripts/predict_models.Rmd
@@
 library(sf)
+library(randomForest)
 source('functions.R')
+
+# unix::rlimit_as(1e32)
+# setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE"), c(""))
 ```
@@
 ```{r read in predictors}
-
-setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE=1024"))
-
-p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2'
+p <- '~/Documents/WesternPlantPredictors2'
 paths <- list.dirs(p, recursive = T)
@@
 f <- file.path(pmodels, list.files(pmodels))
+f <- data.frame(
+  path = f,  
+  species = gsub('-.*$', '', basename(f)), 
+  time = gsub('[A-Za-z]', '', basename(f))
+  ) |>
+  mutate(time = gsub('_-|[.]', '', time),
+         time = gsub('_', ' ', time), 
+         time = as.POSIXct(time)) |>
+  group_by(species) |>
+  slice_max(order_by = time, n = 1) |>
+  pull(path)
 
@@
 
-  pout <- '../results/suitability_maps'
+  pout <- '~/Documents/suitability_maps'
   layer_subset <- terra::subset(layers, term_grab(sdm))
@@
 
-  return(layer_subset)
-  #terra::predict(
-  #  type = 'prob', cores = 16,
-  #  layer_subset, sdm, cpkgs="randomForest",
-  #  filename = file.path(pout, paste0(x$species, '.tif')),
-  #  wopt = c(names = 'predicted_suitability'))
+  OneClass <- \(...) predict(...)[,2]
+  terra::predict(
+    layer_subset, sdm, cpkgs="randomForest",
+    type = 'prob', cores = 1, f = OneClass,
+    filename = file.path(pout, paste0(x$species, '-', gsub(' ', '_', Sys.time()), '.tif')),
+    wopt = c(names = 'predicted_suitability'))
 
@@
 
-ob <- lapply(models_occ[c(5)], predict_wrapper)
-ob1 <- ob[[1]]
-ob1
+lapply(models_occ[2], predict_wrapper)
 
-predict_wrapper(models_occ[[5]])
 ```
@@
 ```
+
+```{r}
+logo <- rast(system.file("ex/logo.tif", package="terra"))   
+names(logo) <- c("red", "green", "blue")
+p <- matrix(c(48, 48, 48, 53, 50, 46, 54, 70, 84, 85, 74, 84, 95, 85, 
+              66, 42, 26, 4, 19, 17, 7, 14, 26, 29, 39, 45, 51, 56, 46, 38, 31, 
+              22, 34, 60, 70, 73, 63, 46, 43, 28), ncol=2)
+
+a <- matrix(c(22, 33, 64, 85, 92, 94, 59, 27, 30, 64, 60, 33, 31, 9,
+              99, 67, 15, 5, 4, 30, 8, 37, 42, 27, 19, 69, 60, 73, 3, 5, 21,
+              37, 52, 70, 74, 9, 13, 4, 17, 47), ncol=2)
+
+xy <- rbind(cbind(1, p), cbind(0, a))
+
+# extract predictor values for points
+e <- extract(logo, xy[,2:3])
+
+# combine with response (excluding the ID column)
+v <- data.frame(cbind(pa=xy[,1], e))
+
+rm(p, a, xy, e)
+
+### with two output variables (probabilities for each class)
+v$pa <- as.factor(v$pa)
+rfm2 <- randomForest(formula=pa~., data=v)
+rfp <- predict(logo, rfm2, cores=2, type="prob", cpkgs="randomForest")
+
+f <- \(...) predict(...)[,2]
+rfp <- predict(logo, rfm2, cores=2, type="prob", cpkgs="randomForest", fun=f)
+
+```

### 2024-01-06 scripts/geomorpho90.R
adding intro skele

@@ -1,0 +1,38 @@
+setwd('/home/sagesteppe/Documents/SDM_restorations/scripts')
+
+library(sf)
+library(tidyverse)
+library(terra)
+
+boundaries <- read_sf('../data/raw/geomorph90_Boundary.kmz') %>% 
+  st_transform(4326)
+
+domain <- rast()
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
+crs(domain) <- "EPSG:4326"
+domain <- as.polygons(domain) %>% 
+  st_as_sf()
+
+weird <- data.frame(
+  ID = c(rep(LETTERS[1:2], each = 5)),
+  X = c(-105, -100, -100, -105, -105, -105, -100, -100, -105, -105),
+  Y = c(40, 40, 35, 35, 40, 35, 35, 30, 30, 35)
+) %>% 
+  st_as_sf(coords = c('X', 'Y'), crs = 4326) %>% 
+  group_by(ID) %>%  
+  dplyr::summarise() %>%
+  st_cast("POLYGON")
+
+poly <- bind_rows(
+  boundaries[ lengths( st_overlaps(boundaries, domain)) > 0, ],
+  boundaries[ unlist(st_contains(domain, boundaries)), ],
+  weird
+) %>% 
+  select(geometry)  %>% 
+  mutate(ID = 1:n())
+
+ggplot() +
+  geom_sf(data = poly) +
+  geom_sf(data = domain, fill = NA, color = 'red')
+

### 2024-01-06 scripts/Boruta_fun.R
citations

@@ -1,35 +1,0 @@
-cores <- parallel::detectCores() # run this outside fn to avoid repeat. 
-
-Boruta_var_selector <- function(x){
-  
-  BorutaRes <- Boruta::Boruta(Occurrence ~ ., data = x, num.threads = cores, doTrace = 1)
-  importance <- Boruta::attStats(BorutaRes)
-  rn <- rownames(importance)
-  important_vars <- Boruta::getSelectedAttributes(BorutaRes, withTentative = F)
-  
-  
-  ### Remove highly autocorrelated variables ###
-  co_pairs <- c('cfvo', 'phh2o', 'gdd', 'ngd', 'gdgfgd')
-  pair_removals <- unlist(lapply(co_pairs, function(x){ rn[min(grep(x, rn))]}))
-  
-  triplets <- c('sand', 'silt', 'clay')
-  triplet_removals <- unlist(lapply(triplets, function(x){ rn[grep(x, rn)[1:2]]}))
-
-  # reduce two the maximum of three soil textures
-  sand_imp <- mean(importance [ grep('sand', rn), 'meanImp'])
-  silt_imp <- mean(importance [ grep('silt', rn), 'meanImp'])
-  clay_imp <- mean(importance [ grep('clay', rn), 'meanImp'])
-  
-  texture2remove <- c('sand', 'silt', 'clay')[which.min(c(sand_imp, silt_imp, clay_imp))]
-  textures2remove <- rn[ grep(texture2remove, rn)]
-  
-  removals <- unique(c(pair_removals, triplet_removals, textures2remove))
-  
-  x_sub <- x[ , !names(x) %in% removals]# remove
-  x_sub <- x_sub[, names(x_sub) %in% c('Occurrence', important_vars)] # keep informative
-  
-  return(x_sub)
-}
-
-
-

### 2024-01-07 scripts/Introduction.Rmd
deriving patches
adding pdf
writing methods

@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output: html_document
+output: pdf_document
 ---
+
+# Introduction
+
+(identify difficulties)
+The lack of available plant germplasm is a primary challenge in effectively restoring terrestrial ecosystems.
+This challenge has two major components 1) the number of species available for projects, and 2) whether the seed is appropriately adapted to sites. 
+Enormous efforts are now underway, to wild harvest seed and increase it in agricultural settings, to increase the number of species available for restoration and the number of populations within these species.
+However, difficulties in both the wild harvest and increase of seed exist.
+
+(paragraph about why many species, why locally adapted, why it needs to be farmed, why decrease the possibility of bottlenecks while farming)
+
+(paragraph about difficulty to find/collect)
+While most species which are desired in restorations generally have relatively large geographic ranges, numbers of populations, and number of individuals per populations, collecting from an adequate amount of species and populations is lagging BEHIND GOALS...
+We posit that most of these challenges relate to the difficulty in finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed for agricultural increase.
+Tools which are capable of predicting a species geographic range, and presences across seed collection target units (such as empirical or ...
+seed transfer zones), such as Species Distribution Models offer promise to decrease the amount of time spent scouting populations, whilst increasing the number of populations which are suitable for collection in a single year.
+While SDM's generate hypothesis of whether areas have environmental conditions conducive to supporting populations of a species, they do not consider the probability of colonization of these patches. 
+However, connectivity analysis based on the predicted unsuitability of habitat, now offer a quick approach to proxy probabilities of occurrence to patches which are previously not known to support populations.
+We believe, based on recent correlations between habitat suitability predictions and abundance, properties of these patches (e.g.
+mean weighted suitability) offer a way to estimate census population size to help identify populations which can support seed collections (@weber2017there, @waldock2022quantitative ).
+
+(paragraph about difficulty to farm seed)
+The production of wild harvested seed in agricultural settings whilst preventing natural selection faces multiple challenges... 
+Not all farms suitable for all species ... increasing costs of human resources, water, cloth to warm seedlings, and perhaps decreasing yield. 
+Within species seed collected closer to harvest sites should do better ... 
+(just those two points... )
+
+## old intro
 
@@
 Finally, cell statistics are used to create aggregate scores for patches of suitable habitat which have been shown to correlate with abundance.
-
-Distance = aggregate
-
-% Species Cover on plot = sum(aggregate suitable cells)
-% Species Cover on plot = sqrt(sum(aggregate suitable cells))
-% Species Cover on plot = log(sum(aggregate suitable cells))

### 2024-01-07 scripts/methods.Rmd
writing methods

# 2 MATERIALS AND METHODS  
## 2.1 Study system

The 355 modelled species were selected by land managers in arid & semi-arid areas Western North America and are species they already or immediately intend to use in restoration. 
The spatial domain of modelling broadly encompasses the Western United States being bounded by the Pacific Ocean, the 50th parallel at north, -100 degrees at East and Mexico to the south. 
This domain has tremendous variation in amounts of elevation, temperature, and precipitation.

## Species Occurrence Records  
Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R (@chamberlain2024rgbif, @maitner2023bien, @michonneau2023idigbio). 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (bio1, 4, 10, 12, & 19; TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

Species absences were generated using three processes, and we sought to have a 1:1 ratio of presences to absences. 
True absences were acquired from a massive ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g. if 20% of land ownership across a species range was BLM, than the number of presences * 0.2 defined the number of these absences). 
Likely absences, representing 15% of records, were generated outside the known range of the species, but bounded to be within 50 miles of the extent extent of occurrences. 
Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for ((1 - (% BLM Land + 15 % LA records)) * 1.25).
Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used. The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables (@venables2002mass), identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables (@naimi2014usdm). 
As many records, classified by the LDA as originating from the Presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.

## Environmental Variables  
XX variables were used in generating the species distribution models (table X, @karger2017climatologies, @hengl2017soilgrids250m, @ivushkin2019global,  @tuanmu2014global, @yamazaki2017merit, @amatulli2020geomorpho90m, @sanderson2002human).
These variables were selected based on authors previous work, and represent variables commonly associated with the empirical distributions of taxa in arid and semi-arid regions. 
All variables were downloaded from source and re-sampled to 90m resolution using terra (@hijman2023terra).

## Species Distribution Modelling  
Random Forests models were generated for each species following several steps to reduce the number of features.
While superfluous features generally do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
The Boruta algorithm was used on the full stack of 44?? independent variables to remove un-informative variables @kursa2010boruta), and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.  
Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys (@kuhn2008caret, @liaw2002randomForest). 
Models were then predicted onto gridded surfaces which exceeded the species range by 50 miles using terra predict.

## Patch identification 
In order to identify patches which may support populations raster values less than 0.6 were masked as NA's and a rooks case search was conducted (@hijman2023terra).

To determine whether habitat suitability could predict species abundance mean weighted patch suitability $PS =  \sum_{\text{i = 1}}^{n} {x_{i}}$ was regressed against abundance using GLM... 
All AIM plots were queried for all occasions which a target species was observed via either the Species Richness, or Line-Point Intercept method (150 points along three equiangular transects over a 25m radius outside a 5m radius buffer zone). 
Occurrence of a species only along SR was considered 1% cover, while the canopy cover from LPI was used.

To determine whether connectivity could predict the presence of a population in a patch, connectivity was calculated between every patch, and jacknife-resampled for logistic regression where connectivity scaled from 0 to 100 served as a the independent variable.

### 2024-01-08 scripts/Introduction.Rmd
writing

--- 2023-06-24 scripts/Introduction.Rmd
@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output: html_document
+output: pdf_document
 ---
+
+# Introduction
+
+(identify difficulties)
+A primary challenge in effectively restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment).
+Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings.
+Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of these populations available for restoration @national2023assessment.
+However, difficulties in both the wild harvest and increase of seed exist are limiting our ability to develop adequate amounts of germplams.  
+Herein, we identify prospective methods to increase the 1) number of species, 2) populations, and 3) genetic diversity of collections and 4) increase the success of agriculturally produced seed lots using Species Distribution Models and subsequent analysis of them intended to pair them more aptly with species biology and ecology.
+
+(paragraph about why many species, why locally adapted, reluctant correlation between pop size genetic diversity, why seed needs to be farmed, why decrease the possibility of bottlenecks while farming)
+
+#### no species (@wilsey2021restoration)
+
+(paragraph about difficulty to find/collect)
+While most species which are desired in restorations historically have generally had relatively large geographic ranges, numbers of populations, and number of individuals per populations, the development of native germplasm remains behind targets (@national2023assessment).
+We posit that most of these challenges relate to the difficulty in finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed to being agricultural increase, complications borne of widespread habitat degradation and unnatural wildfires (cite on degradation and wildfires).
+Tools which are capable of predicting a species geographic range, the presence and size of populations across seed collection target units (such as empirical or provisional seed transfer zones or ecoregions), such as Species Distribution Models offer promise to decrease the amount of time spent by field crews scouting populations, whilst increasing their knowledge of the number of populations which are suitable for collection in a single year.
+
+(paragraph about resolving problems)
+However, while SDM's generate hypothesis of whether areas have environmental conditions similar to the observed environmental niche of the species they do not consider the probabilities of colonization, nor the populations census sizes.
+A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis.
+Utilizing the predicted unsuitability of habitat, with extreme barriers to dispersal, to create cost-resistance surfaces between patches with known populations and predicted populations allows for simulating the probability of occurrence.
+While previous correlations between habitat suitability and population size have often been low, we posit that patch level parameters offer a more useful prediction of population size (@weber2017there, @waldock2022quantitative)
+Patch metrics will more appropriately reflect parameters of a potential population, e.g.
+geographic extent, which we posit is a more informative proxy of population size than environmental niche correlation.
+
+Utilizing this approach will allow field crews to associate probabilities with previously unground truthed patches.
+
+(paragraph about difficulty to farm seed)
+The production of wild harvested seed in agricultural settings whilst preventing natural selection faces multiple challenges... 
+Not all farms suitable for all species ... increasing costs of human resources, water, cloth to warm seedlings, and perhaps decreasing yield. 
+Within species seed collected closer to harvest sites should do better ... 
+(just those two points... )
+
+Goals/Questions/hypotheses.
+
+Here we showcase the utility of SDM's to (1) predict suitable habitat for common species in natural settings and use those data to test two specific hypotheses.
+Firstly 2) that SDM's can identify more populations than exist in the occurrence based records they were derived from, and secondly 3) patch metrics derived from SDM's correlate with observed population census size.
+Further we showcase the possibility of SDM's as tools to predict areas which are suitable for agricultural grow-outs to increase seed lot sizes (4).
+
+## old intro
 

### 2024-01-09 scripts/generate_absences.Rmd
updated lda dropper to deal with 0 variance variables

@@
 
-out1 <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
+out1 <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15, rand_dist = 1000)
 
@@
   rename(taxon = Taxon) %>% 
-  mutate(date = as.Date('2023-12-18'), .before = geometry)
+  mutate(date = as.Date('2023-01-09'), .before = geometry)
 
-st_write(out, '../data/raw/occurrence/absences/absences.shp', append = FALSE)
+st_write(out, '../data/raw/occurrence/absences/absences1km.shp', append = FALSE)
 
@@
 
-```{r import raster stack}
+```{r import raster stack, eval = F}
 
-p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'
+p <- '~/Documents/WesternPlantPredictors2'
 paths <- file.path(p, list.files(p, recursive = F))
@@
 bio <- rast(bio)
-bio <- subset(bio, c(1:11, 14:17, 19)) # these 'empty' need investigate
+rm(p, paths, files)
 
-rm(p, paths, files)
+```
+
+```{r read in predictors2 }
+p <- '~/Documents/WesternPlantPredictors2'
+paths <- list.dirs(p, recursive = T)
+paths <- paths[2:length(paths)]
+paths <- paths[! grepl('chelsa$|veg_cover$|soilgrids$|sand$|silt$|clay$|phh2o$|cfvo$|soc$', paths) ]
+
+files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
+names(files) <- basename(paths)
+
+layers <- lapply(files, vrt)
+layers <- rast(layers)
+names(layers) <- basename(paths)
+
+rm(paths, files)
 ```
@@
 
-write_rds(collinear_probs, '../results/bioclim_VIF_PA_drop.rda')
+write_rds(collinear_probs, '../results/VIF_PA_drop.rda')
+
+rm(p, paths, files)
+```
+
+```{r subset rasterstack 2, eval = F}
+vif_vals <- spatSample(layers, size = 5000, values = TRUE) # this steps takes forever... 
+vif_vals <- vif_vals[ complete.cases(vif_vals), ] # 4148 records used
+collinear_probs <- usdm::vifstep(vif_vals, th = 10)
+layers <- usdm::exclude(layers, collinear_probs)
+
+write_rds(collinear_probs, '../results/VIF_PA_drop.rda')
 
@@
 
-collinear_probs <- read_rds('../results/bioclim_VIF_PA_drop.rda')
+collinear_probs <- read_rds('../results/VIF_PA_drop.rda')
 bio <- usdm::exclude(bio, collinear_probs)
@@
 
+set.seed(27)
+cn <- names(layers)
+splicies <- split(recs, f = recs$taxon)
+
+lapply(splicies[41:335], lda_PA_dropper, path = '../results/lda2/', col_names = cn)
+
+which(names(splicies) == 'Asclepias speciosa')
+
 rm(absences, presences, collinear_probs)

### 2024-01-09 scripts/functions.R

@@ -68,8 +68,11 @@ multiplesPERcell <- function(x){
   
 }
 
-absence_drawer <- function(x, bg_abs, terrestrial){
-  
+#' @param x sf/tibble/dataframe containing cleaned presences
+#' @param bg_abs the proportion of records which should be background absences, i..e come from outside the species known range
+#' @param terrestrial an object of terrestrial surfaces
+#' @param rand-dist the distance offset between the presences and random absences. 
+absence_drawer <- function(x, bg_abs, terrestrial, rand_dist){
   
   # create the observed hull, the buffered hull for random sample selection, 
   # and a hull to form the basis for a bounding box to generate regular absences in
@@ -127,7 +130,7 @@ absence_drawer <- function(x, bg_abs, terrestrial){
   
   # generate Random absences
   rand_recs <- nrow(x) * ((1 - bg_abs) * (1 - prcnt_blm))
-  buffered_presences <- sf::st_buffer(x, dist = 10000) |>
+  buffered_presences <- sf::st_buffer(x, dist = rand_dist) |>
     sf::st_union() |>
     sf::st_transform(sf::st_crs(blm_surf))
   
@@ -240,7 +243,7 @@ colon_balloon <- function(x){
 
 lda_PA_dropper <- function(x, path, col_names){
   
-  recs_w_vars <- terra::extract(bio, x)
+  recs_w_vars <- terra::extract(layers, x)
   
   recs_w_vars_scaled <- data.frame(apply(recs_w_vars[,2:ncol(recs_w_vars)], MARGIN = 2, FUN = scales::rescale, to = c(0,1)))
   recs_w_vars_scaled$ID <- x$ID
@@ -505,3 +508,22 @@ rfe_var_selector <- function(x){
   
 }
 
+
+
+
+# this bash script was made to write files to the sdd, and copy to an hdd periodically 
+
+#!/bin/bash
+# cd /home/steppe/Documents/suitability_maps
+#maps=(*) # save all file names into object
+#noFiles="$((${#maps[@]}-1))" # the number of completed files. 
+#if [ "$noFiles" -gt 1 ]; then
+#maps2move=${maps[@]::${#maps[@]}-1} # subset the completed files
+#  for f in $maps2move; do 
+#  mv "$f" "/media/steppe/hdd/SDM_restorations/results/suitability_maps/"
+#  done
+#  else
+#    :
+#    fi
+# we set up a cron job to run every24 hours to move the files
+# 0 */3 * * * sdd2hdd.sh

### 2024-01-10 scripts/Introduction.Rmd
playing with burning raster

@@
 
-(identify difficulties)
-A primary challenge in effectively restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment).
-Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings.
+A primary challenge in effectively restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment, @merritt2011restoration).
+Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings (@pedrini2020collection, @broadhurst2015seeding,  @national2023assessment).
 Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of these populations available for restoration @national2023assessment.
-However, difficulties in both the wild harvest and increase of seed exist are limiting our ability to develop adequate amounts of germplams.  
+However, difficulties exist in both the wild harvest and increase of seed which are limiting our ability to develop adequate amounts of germplasm material.  
 Herein, we identify prospective methods to increase the 1) number of species, 2) populations, and 3) genetic diversity of collections and 4) increase the success of agriculturally produced seed lots using Species Distribution Models and subsequent analysis of them intended to pair them more aptly with species biology and ecology.
@@
 (paragraph about why many species, why locally adapted, reluctant correlation between pop size genetic diversity, why seed needs to be farmed, why decrease the possibility of bottlenecks while farming)
+
+More species per restroation gooood. @wilsey2021restoration
+Each collection is generally only cultivated for 1-5 generations before various effects of selection start to reduce the efficacy of the lineage in restorations (@espeland2017evolution, @pizza2021eight, @conrady2023plants)
+
+@  "Sustainable seed harvesting in wild plant populations" different species can support different amounts of harvest. 
+@ theory of how less diversity can harm restorations
 

### 2024-01-10 scripts/burn_surface_raster.R

@@ -1,0 +1,54 @@
+setwd('~/Documents/SDM_restorations/scripts')
+getwd()
+
+library(tidyverse)
+library(sf)
+library(terra)
+
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+domain <- as.polygons(domain) |>  # convert to vector data
+  st_as_sf() 
+
+rivers <- st_read('../data/processed/rivers_noram/rivers_noram_37341.shp', quiet = TRUE) %>% 
+  st_crop(., domain) %>% 
+  select(Strahler, SUB_NAME, SUBBAS_ID) %>% 
+  filter(Strahler > 2) %>% 
+  group_by(SUBBAS_ID) %>% 
+  reframe(geometry = st_union(geometry)) %>% 
+  st_as_sf()
+
+rivers <- rivers[ as.numeric(st_length(rivers)) > 30000, ]
+rivers <- st_simplify(rivers) %>% 
+  st_transform(5070) %>% 
+  st_buffer(1000) %>% 
+  st_transform(4326)
+
+lakes10 <- rnaturalearth::ne_download(scale = 10, type = "lakes_north_america",
+                       category = "physical") %>% 
+  st_make_valid()
+
+lakes10 <- lakes10[ which(st_is_valid(lakes10) == T), ]
+lakes10 <- st_crop(lakes10, domain)
+
+
+ocean <- rnaturalearth::ne_download(scale = 10, type = "ocean",
+                                    category = "physical") 
+
+ggplot() + 
+  geom_sf(data = lakes10)  + 
+  geom_sf(data = rivers) +
+  theme_bw()
+
+domain <- rast(nrows = 25000, ncols = 25000) 
+# create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
+crs(domain) <- "EPSG:4326"
+values(domain) <- 1
+
+domain <- terra::mask(domain, vect(ocean), inverse = T)
+domain <- terra::mask(domain, vect(lakes10), inverse = T)
+domain <- terra::mask(domain, vect(rivers), inverse = T, touches = TRUE)
+
+plot(domain)

### 2024-01-10 scripts/predict_models.Rmd

@@
 
-lapply(models_occ[2], predict_wrapper)
+lapply(models_occ[30:335], predict_wrapper)
 
 ```
-
-```{r}
-achy <- rast('../results/suitability_maps/Achnatherum_hymenoides.tif')
-plot(achy[[2]])
-```
-
-```{r}
-logo <- rast(system.file("ex/logo.tif", package="terra"))   
-names(logo) <- c("red", "green", "blue")
-p <- matrix(c(48, 48, 48, 53, 50, 46, 54, 70, 84, 85, 74, 84, 95, 85, 
-              66, 42, 26, 4, 19, 17, 7, 14, 26, 29, 39, 45, 51, 56, 46, 38, 31, 
-              22, 34, 60, 70, 73, 63, 46, 43, 28), ncol=2)
-
-a <- matrix(c(22, 33, 64, 85, 92, 94, 59, 27, 30, 64, 60, 33, 31, 9,
-              99, 67, 15, 5, 4, 30, 8, 37, 42, 27, 19, 69, 60, 73, 3, 5, 21,
-              37, 52, 70, 74, 9, 13, 4, 17, 47), ncol=2)
-
-xy <- rbind(cbind(1, p), cbind(0, a))
-
-# extract predictor values for points
-e <- extract(logo, xy[,2:3])
-
-# combine with response (excluding the ID column)
-v <- data.frame(cbind(pa=xy[,1], e))
-
-rm(p, a, xy, e)
-
-### with two output variables (probabilities for each class)
-v$pa <- as.factor(v$pa)
-rfm2 <- randomForest(formula=pa~., data=v)
-rfp <- predict(logo, rfm2, cores=2, type="prob", cpkgs="randomForest")
-
-f <- \(...) predict(...)[,2]
-rfp <- predict(logo, rfm2, cores=2, type="prob", cpkgs="randomForest", fun=f)
-
-```

### 2024-01-16 scripts/Introduction.Rmd
intro and methods

--- 2024-01-08 scripts/Introduction.Rmd
@@
 ---
-title: "Introduction"
-date: '`r format(Sys.Date(), "%B %d, %Y")`'
+title: "Introduction - sdms for restoration"
 output: pdf_document
@@
 
-# Introduction
+# Abstract
+1) THE PROBLEM 
+2) WHAT WE DID
+3) BASIC RESULTS
+4) INSIGHTFUL RESULTS
+5) **Synthesis and applications**
 
-(identify difficulties)
-A primary challenge in effectively restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment).
-Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings.
-Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of these populations available for restoration @national2023assessment.
-However, difficulties in both the wild harvest and increase of seed exist are limiting our ability to develop adequate amounts of germplams.  
-Herein, we identify prospective methods to increase the 1) number of species, 2) populations, and 3) genetic diversity of collections and 4) increase the success of agriculturally produced seed lots using Species Distribution Models and subsequent analysis of them intended to pair them more aptly with species biology and ecology.
+# Introduction *(at most 1 1/4 pages)*
 
-(paragraph about why many species, why locally adapted, reluctant correlation between pop size genetic diversity, why seed needs to be farmed, why decrease the possibility of bottlenecks while farming)
+A primary challenge to restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment, @merritt2011restoration).
+Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings (@pedrini2020collection, @broadhurst2015seeding,  @national2023assessment).
+Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of these populations, available for restoration @national2023assessment.
+However, difficulties exist in both the wild harvest and increase of seed which are limiting our ability to develop adequate amounts of germplasm.
 
-#### no species (@wilsey2021restoration)
+While most species desired in restorations have historically had relatively large geographic ranges, numbers of populations, and number of individuals per populations, the development of native germplasm remains behind targets (@national2023assessment).
+We posit that in part this is due to the difficulty of finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed to being agricultural increase, a complications borne of widespread habitat degradation and unnatural wildfires *(cite on degradation and wildfires)*.
+Tools which are capable of predicting a species geographic range, the presence and size of populations across the range and in seed collection target units (such as empirical or provisional seed transfer zones or ecoregions), such as Species Distribution Models offer promise to increase the rate at which native germplasm can be developed.
 
-(paragraph about difficulty to find/collect)
-While most species which are desired in restorations historically have generally had relatively large geographic ranges, numbers of populations, and number of individuals per populations, the development of native germplasm remains behind targets (@national2023assessment).
-We posit that most of these challenges relate to the difficulty in finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed to being agricultural increase, complications borne of widespread habitat degradation and unnatural wildfires (cite on degradation and wildfires).
-Tools which are capable of predicting a species geographic range, the presence and size of populations across seed collection target units (such as empirical or provisional seed transfer zones or ecoregions), such as Species Distribution Models offer promise to decrease the amount of time spent by field crews scouting populations, whilst increasing their knowledge of the number of populations which are suitable for collection in a single year.
-
-(paragraph about resolving problems)
 However, while SDM's generate hypothesis of whether areas have environmental conditions similar to the observed environmental niche of the species they do not consider the probabilities of colonization, nor the populations census sizes.
@@
 geographic extent, which we posit is a more informative proxy of population size than environmental niche correlation.
-
 Utilizing this approach will allow field crews to associate probabilities with previously unground truthed patches.
 
-(paragraph about difficulty to farm seed)
-The production of wild harvested seed in agricultural settings whilst preventing natural selection faces multiple challenges... 
-Not all farms suitable for all species ... increasing costs of human resources, water, cloth to warm seedlings, and perhaps decreasing yield. 
-Within species seed collected closer to harvest sites should do better ... 
-(just those two points... )
-
-Goals/Questions/hypotheses.
-
-Here we showcase the utility of SDM's to (1) predict suitable habitat for common species in natural settings and use those data to test two specific hypotheses.
-Firstly 2) that SDM's can identify more populations than exist in the occurrence based records they were derived from, and secondly 3) patch metrics derived from SDM's correlate with observed population census size.
-Further we showcase the possibility of SDM's as tools to predict areas which are suitable for agricultural grow-outs to increase seed lot sizes (4).
-
-## old intro
-
-Drylands constitute roughly ~45% of earths surface (@pruavualie2016drylands). 
-These water-limited areas are commonly used as range or pasture lands, hence much of the land is maintained in natural or semi-natural conditions (@reynolds2007globa, @asner2004grazingl). 
-Accordingly drylands lands represent an enormous proportion of Earth's remaining natural areas, while fostering roughly 2.5 billion persons, however many of these areas are in ecological states which do not maximize their ecosystem services to either human or wildlife (@reynolds2007global, @bestelmeyer2015desertification).
-Historic actions and phenomena such as fire suppression, shrub encroachment, and overgrazing, in combination with climate change and drought are resulting in severe wildfires and their subsequent burns scars which may lead to a regime shifts wherein invasive species dominate burned sites for the foreseeable future (@bestelmeyer2015desertification).
-The prevalence of these invasive species decrease the ability of the land to serve multiple ecosystem services, such as to foster both wildlife and pastoral goals, and in extreme cases may preclude uses of the land by both wildlife and humans, and may have require high fiscal costs to return to a more desirable state.
-
-In order to maintain the ecosystem services of burned areas restoration plantings have long been used. 
-More recently a focus on utilizing native species, and now locally adapted genotypes, has been adopted as these species and populations have been shown to provide ecosystem services to a variety of wildlife and stakeholders, and they are often more capable of establishing at sites.
-However, the acquisition of native plant materials in the volumes required for restoration, especially in fire prone years, is not yet feasible. 
-In an attempt to provide for the demand of locally adapted native plant materials partnerships with farmers who are planting wild harvested seeds, and harvesting their seed for sale to restoration performing agencies has increased.
-However, the implementation of these goals still face significant challenges.
-
-We anticipate that three particular challenges which are currently being faced in developing and delivering a native seed supply in the Great Basin Ecoregion are ubiquitous to all supply chains. 
-1) Locating populations of adequate size that they can have enough viable seed collected from that that it is economically viable for farmers to grow out from the seed lots while not adversely affecting the demographic vital rates of the population.
-2) Minimizing the adverse affects of artificial selection of these seed lines while being increased in agricultural settings. 
-3) Ensuring that farmed seeds are applied to the areas in burn scars which they have the highest probability of success. 
-We posit that Species Distribution Modelling (SDMs) (or Environmental Niche Modelling (ENM)) may be used as an heuristic to increase the probability of success at each of these stages, and the development of these models are markedly cheaper than purely empirical approaches.
-
-Locating populations of native species which are large enough to support a seed collection which is large enough to establish an agricultural lot is difficult. 
-The areas over which native seed may feasibly be collected in semi-arid areas, which often times have higher proportion of natural areas relative to other settings, are enormous and have generally had their floras coarsely mapped.
-Restoration approaches for sourcing locally adapted seed such as Seed Transfer Zones (STZ's), predictive provincing, or Omernik Level 4 ecoregions, increase the relevancy of identifying many populations suitable for collections across a species range.
-Locating such populations is difficult given the lengths of growing seasons, human resources, and the length of time required to collect enough viable seed.
-
-Agricultural cultivation presents multiple steps which may alter the frequency of gene forms (alleles) within the seed lots via unintentional artificial selection.
-It is theorized that reducing these effects will result in seed lots with a proportion of more individuals with gene forms more typical of the planted wild population, than the offspring derived from it, and hence have more individuals likely to establish in a restoration.
-A method to reduce the demographic bottlenecks associated with recruitment of an individual, and to reduce over-representation of seed from species which are more fit in cultivated settings than the wild may be to reduce the environmental dissimilarity between the wild and cultivated settings.
-These models may also help farmers prioritize which species to grow to have the most successful crops, and to help recruit new farmers to dedicate portions of there fields to developing native plant materials.
-
-Planning a restoration following a wildfire is a difficult process, due to limitations of both fiscal budget, time, human resources, and knowledge of the heterogeneity of the affected areas. 
-Species Distribution Models offer a line of evidence which practitioners may use to identify species which are most suitable for the burned areas in a simple readily query friendly format. 
-The utilization of them ...
-
-Herein we develop simple Species Distribution Models, using a variety of predictors which are global in coverage - or for which global analogs exist, and species occurrences from publicly available and rangeland monitoring datasets, to assist in an ongoing effort to develop native plant materials in the Great Basin Ecoregion (U.S.A.).
-Two sets of models are generated for each species, one set which reflects the natural, in particular biophysical the settings of the species, and another set which more closely reflects the climatic envelope and soil properties of the species.
-The former are to assist in challenges of finding populations and restoration efforts while the latter is to support seed increase endeavors.
-In attempts to reduce the discrepancies between the fundamental and realized niche, to assist field workers in finding populations, known occurrences are combined with cost-surfaces and predictions of suitability models to provide a consensus surface and emulate the restrictions of dispersal to species distributions.
-Finally, cell statistics are used to create aggregate scores for patches of suitable habitat which have been shown to correlate with abundance.
+Here we showcase the utility of SDM's to predict suitable habitat for common species in natural settings and use those data to test two specific hypotheses. 
+Firstly that SDM's can identify more populations than exist in the occurrence based records they were derived from. 
+And secondly patch metrics derived from SDM's correlate with observed population census size. 
+In order to determine whether SDM's are useful in detecting new populations, over ~10~ field crews were given 50 putative populations to survey for the presence and absence of the modelled species. 
+To determine if patch metrics can be used to predict population census sizes, these crews noted whether the populations was large enough to support

### 2024-01-16 scripts/methods.Rmd
writing
intro and methods

@@
+---
+title: "Methods - sdms for restoration"
+output:
+  pdf_document: default
+  html_document: default
+---
 # 2 MATERIALS AND METHODS  
@@
 Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys (@kuhn2008caret, @liaw2002randomForest).
-Models were then predicted onto gridded surfaces which exceeded the species range by 50 miles using terra predict.
+Models were then predicted onto gridded surfaces which exceeded the species range by 50 miles using terra's predict.
 

### 2024-01-22 scripts/generate_RF_models.Rmd
perform rfe to subset predictors to reasonable size for prediction onto a gridded surface incroporated into modeller
ran all models

@@
 
-p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2'
+p <- '~/Documents/WesternPlantPredictors2'
 paths <- list.dirs(p, recursive = T)
@@
 cores <- parallel::detectCores()
-lapply(splicies[3:335], modeller)
+lapply(splicies, modeller, suffix = '1k')
 

### 2024-01-22 scripts/predict_models.Rmd
ran all models

@@
 
-# unix::rlimit_as(1e32)
-# setGDALconfig(c("GDAL_MAX_DATASET_POOL_SIZE"), c(""))
+setGDALconfig('GDAL_MAX_DATASET_POOL_SIZE=1024')
 ```
@@
          time = gsub('_', ' ', time), 
+         time = trimws(time),
+         time = gsub('^.1-|^1-', '', time), 
          time = as.POSIXct(time)) |>
-  group_by(species) |>
-  slice_max(order_by = time, n = 1) |>
+ # group_by(species) |>
+  filter(str_detect(species, '1k$')) |>
+ # slice_max(order_by = time, n = 1) #|>
   pull(path)
@@
 
+   terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
+
 }
 
-lapply(models_occ[30:335], predict_wrapper)
+lapply(models_occ[68:335], predict_wrapper)
 

### 2024-01-22 scripts/functions.R

@@ -355,7 +355,7 @@ Boruta_var_selector <- function(x){
 
 #' perform random forest modelling and save outputs to locations
 #' test_data output from Boruta_var_selector
-randomForests <- function(train_data, test_data, species){
+randomForests <- function(train_data, test_data, species, suffix){
   
   # identify the appropriate mtry
   opt_mtry <- data.frame(
@@ -370,18 +370,18 @@ randomForests <- function(train_data, test_data, species){
   # plot and save results
   species <- gsub(' ', '_', species)
   saveRDS(rf_model,
-          file = paste0('../results/rf_models/', species, '-', gsub(' ', '_', Sys.time()), '.rds'))
+          file = paste0('../results/rf_models/', species, suffix, '-', gsub(' ', '_', Sys.time()), '.rds'))
   
   vip::vip(rf_model) +
     theme_bw() +
     labs(title = species) + 
     theme(plot.title = element_text(hjust = 0.5))
-  ggsave(paste0('../results/vip_plots/', species, '.png'), device = 'png')
+  ggsave(paste0('../results/vip_plots/', species, suffix, '.png'), device = 'png')
   
   prediction_prob <- predict(rf_model, test_data, type = 'prob') # use this for auc/roc
   result.roc <- pROC::roc(test_data$Occurrence, prediction_prob[,1]) # Draw ROC curve.
   
-  png(file = paste0('../results/roc_plots/', species, '.png'))
+  png(file = paste0('../results/roc_plots/', species, suffix, '.png'))
   plot(
     result.roc,
     print.thres="best",
@@ -405,11 +405,12 @@ randomForests <- function(train_data, test_data, species){
     setNames(data.frame(as.numeric(result.roc$auc), 'AUC'), c('Value', 'Metric')), 
     setNames(data.frame((ncol(train_data)-1), 'Independent_Variables'), c('Value', 'Metric')), 
     summary, samples)
-  write.csv(summary, paste0('../results/summary/', species, '.csv'), row.names = F)
+  write.csv(summary, paste0('../results/summary/', species, suffix, '.csv'), row.names = F)
   
 }
 
-modeller <- function(x){
+
+modeller <- function(x, suffix){
   
   species <- sf::st_drop_geometry(x) %>% 
     dplyr::pull(taxon)
@@ -429,10 +430,11 @@ modeller <- function(x){
   
   TRAIN_DATA <- Boruta_var_selector(TRAIN)
[…]

### 2024-01-22 scripts/make_Impermeable_raster.R
working on burnt river raster

@@ -1,0 +1,35 @@
+setwd('/media/steppe/hdd/SDM_restorations/scripts')
+
+library(tidyverse)
+library(sf)
+library(terra)
+
+riv <- st_read('../data/processed/rivers_noram/rivers_noram_37341.shp', quiet = TRUE)|>
+  select(MAJ_NAME, Strahler) |>
+  filter(Strahler >= 3)
+
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
+crs(domain) <- "EPSG:4326"
+riv <- st_crop(riv, project(domain, crs(riv))) %>% 
+  st_simplify() %>% 
+  st_transform(5070) %>% 
+  st_simplify(dTolerance = 500) |>
+  st_buffer(2000)
+
+# import slope and identify large features... 
+
+p <- '~/Documents/WesternPlantPredictors2/slope'
+paths <- file.path(p, list.files(p))
+slope <- mosaic(
+  sprc(
+    lapply(paths, rast)
+    )
+  )
+
+# burn rivers to their own raster
+
+rivr <- terra::rasterize(vect(riv), rast(slope), touches = TRUE, update = TRUE)
+plot(rivr)
+
+writeRaster()

### 2024-01-22 scripts/make_Impermeable_raster.R
borders to split pops

@@ -14,8 +14,8 @@ crs(domain) <- "EPSG:4326"
 riv <- st_crop(riv, project(domain, crs(riv))) %>% 
   st_simplify() %>% 
   st_transform(5070) %>% 
-  st_simplify(dTolerance = 500) |>
-  st_buffer(2000)
+  st_simplify(dTolerance = 250) |>
+  st_buffer(140)
 
 # import slope and identify large features... 
 
@@ -30,6 +30,38 @@ slope <- mosaic(
 # burn rivers to their own raster
 
 rivr <- terra::rasterize(vect(riv), rast(slope), touches = TRUE, update = TRUE)
-plot(rivr)
+writeRaster(rivr, '../../Geospatial_data/river_resistance/rivR.tif', overwrite = TRUE)
 
-writeRaster()
+
+
+
+##############################################################
+## read and watershed boundaries and convert some to raster ##
+## to delineate groups (spatial structured populations, population, deme) ##
+p2gdb <- '../../Geospatial_data/WBD/WBD_National_GDB.gdb'
+st_layers(p2gdb)
+
+hu10 <- st_read(p2gdb, layer = 'WBDHU10', quiet = TRUE)  %>% 
+  st_make_valid() %>% 
+  st_crop(., project(domain, crs(.))) %>% 
+  st_simplify() %>% 
+  st_transform(5070) %>% 
+  st_simplify(dTolerance = 80)
+
+hu12 <- st_read(p2gdb, layer = 'WBDHU12', quiet = TRUE)  %>% 
+  st_make_valid() %>% 
+  st_crop(., project(domain, crs(.))) %>% 
+  st_simplify() %>% 
+  st_transform(5070) %>% 
+  st_simplify(dTolerance = 80)
+
+hu10 <- select(hu10, huc10)
+hu12 <- select(hu12, huc12)
+
+hu10R <- terra::rasterize(vect(hu10), rast(slope), update = TRUE, field = 'huc10')
+hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12')
+
+writeRaster(hu10R, '../../Geospatial_data/WBD_rast/hu10R.tif', overwrite = TRUE)
+writeRaster(hu12R, '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = TRUE)
+
+plot(hu12R)

### 2024-01-22 scripts/classify_patches.Rmd
aggregate cells before finding patches

---
title: "detect patches"
author: "steppe"
date: "2024-01-22"
output: html_document
---

```{r}
library(terra)
```

```{r}
p <- '../results/suitability_maps'
r <- rast(file.path(p, 'Achnatherum_aridum1k-2024-01-10_09:16:55.tif'))
msk <- ifel(r < 0.80, NA, r)
m <- mask(r, msk)
m1 <- aggregate(m, 3, fun = 'mean')

# aggregate the raster to speed up the patch detection process. 
```

```{r}
system.time({ob <- terra::patches(m1,  directions = 4)})
system.time({z =  zonal(cellSize(ob, unit="ha"), ob, sum, as.raster=TRUE) })
z = ifel(z < 2.0235, NA, z)

plot(z)
```

### 2024-01-22 scripts/make_Impermeable_raster.R

@@ -34,7 +34,6 @@ writeRaster(rivr, '../../Geospatial_data/river_resistance/rivR.tif', overwrite =
 
 
 
-
 ##############################################################
 ## read and watershed boundaries and convert some to raster ##
 ## to delineate groups (spatial structured populations, population, deme) ##
@@ -64,4 +63,3 @@ hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12
 writeRaster(hu10R, '../../Geospatial_data/WBD_rast/hu10R.tif', overwrite = TRUE)
 writeRaster(hu12R, '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = TRUE)
 
-plot(hu12R)

### 2024-01-23 scripts/classify_patches.Rmd
a few patch metrics before meeting

@@
 ```
+
+```{r Patch metics to predict presence}
+
+# euclidean nearest neighbor distance (enn)
+## (modification) log(x), sqrt(x)
+### Expectation: the number of colonized predicted patches decrease with increasing distance.
+#### why: long-distance dispersal is limited from colonized to suitable patches. 
+##### test: logistic regression, independent: distance, dependent: colonization (0|1)
+
+sqrt(c( 1000, 150))
+```
+
+```{r Patch metrics to predict abundance}
+library(landscapemetrics)
+
+# Patch level metrics. 
+print(
+  list_lsm(level = "patch"), 
+  n = 12)
+
+# Core Area Index (cai)  
+## (modification) arithmetic mean of core area cells
+### Expectation: higher cai higher abundance
+#### why: better within patch distribution, and a persistent 'source' to 'sink' peripheral cells. 
+##### test: glmm?(% cover ~ cai, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)
+
+# perimeter-area ratio (para)
+## (modification) none
+### Expectation: lower para higher abundance
+#### why: areas with 'large edges' will suffer dispersal limitations throughout them, as many seeds will go to unsuitable habitat.
+##### test: glmm?(% cover ~ para, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)
+
+# fractal dimension index (frac)
+## (modification) none
+### Expectation: lower fractal dimensions (less complex patch shapes) have higher abundances. 
+#### why: components of patches with complex shapes may suffer dispersal limitation. 
+##### test: glmm?(% cover ~ frac, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)
+
+## to all models a term for the previous 4 years SPEI may be incorporated.
+
+```

### 2024-01-23 scripts/burn_surface_raster.R
consolidated two scripts - deja vu - raster burning in one place now

@@ -1,54 +1,0 @@
-setwd('~/Documents/SDM_restorations/scripts')
-getwd()
-
-library(tidyverse)
-library(sf)
-library(terra)
-
-domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
-ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
-crs(domain) <- "EPSG:4326" # define the projection
-domain <- as.polygons(domain) |>  # convert to vector data
-  st_as_sf() 
-
-rivers <- st_read('../data/processed/rivers_noram/rivers_noram_37341.shp', quiet = TRUE) %>% 
-  st_crop(., domain) %>% 
-  select(Strahler, SUB_NAME, SUBBAS_ID) %>% 
-  filter(Strahler > 2) %>% 
-  group_by(SUBBAS_ID) %>% 
-  reframe(geometry = st_union(geometry)) %>% 
-  st_as_sf()
-
-rivers <- rivers[ as.numeric(st_length(rivers)) > 30000, ]
-rivers <- st_simplify(rivers) %>% 
-  st_transform(5070) %>% 
-  st_buffer(1000) %>% 
-  st_transform(4326)
-
-lakes10 <- rnaturalearth::ne_download(scale = 10, type = "lakes_north_america",
-                       category = "physical") %>% 
-  st_make_valid()
-
-lakes10 <- lakes10[ which(st_is_valid(lakes10) == T), ]
-lakes10 <- st_crop(lakes10, domain)
-
-
-ocean <- rnaturalearth::ne_download(scale = 10, type = "ocean",
-                                    category = "physical") 
-
-ggplot() + 
-  geom_sf(data = lakes10)  + 
-  geom_sf(data = rivers) +
-  theme_bw()
-
-domain <- rast(nrows = 25000, ncols = 25000) 
-# create a big empty raster, you can go in through sf too. 
-ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
-crs(domain) <- "EPSG:4326"
-values(domain) <- 1
-
-domain <- terra::mask(domain, vect(ocean), inverse = T)
-domain <- terra::mask(domain, vect(lakes10), inverse = T)
-domain <- terra::mask(domain, vect(rivers), inverse = T, touches = TRUE)
-
-plot(domain)

### 2024-01-23 scripts/make_Impermeable_raster.R

@@ -32,8 +32,27 @@ slope <- mosaic(
 rivr <- terra::rasterize(vect(riv), rast(slope), touches = TRUE, update = TRUE)
 writeRaster(rivr, '../../Geospatial_data/river_resistance/rivR.tif', overwrite = TRUE)
 
+##################################################
+### now burn lakes and oceans into the raster ###
+##################################################
 
+lakes10 <- rnaturalearth::ne_download(scale = 10, type = "lakes_north_america",
+                                      category = "physical") %>% 
+  st_as_sf() %>% 
+  st_make_valid() %>%
+  filter(featurecla == 'Lake') %>% 
+  select(featurecla)
 
+lakes10 <- lakes10[ which(st_is_valid(lakes10) == T), ]
+lakes10 <- st_crop(lakes10, domain)
+
+ocean <- rnaturalearth::ne_download(scale = 10, type = "ocean",
+                                    category = "physical") 
+
+rivr <- terra::mask(rivr, vect(ocean), touches = TRUE, update = TRUE)
+rivr <- terra::mask(rivr, vect(lakes10), touches = TRUE, update = TRUE)
+
+plot(lakes10[,1])
 ##############################################################
 ## read and watershed boundaries and convert some to raster ##
 ## to delineate groups (spatial structured populations, population, deme) ##

### 2024-01-24 scripts/classify_patches.Rmd
class metrics to test

@@
 m <- mask(r, msk)
-m1 <- aggregate(m, 3, fun = 'mean')
+m1 <- aggregate(m, 2, fun = 'mean')
 
@@
 ```
+
+aggregation by a factor of 2 allows for quick patch delineation and detection.
 
@@
 
-# euclidean nearest neighbor distance (enn)
-## (modification) log(x), sqrt(x)
+# euclidean nearest neighbor distance (enn) - (Patch)
+## (modification) log(enn), sqrt(enn)
 ### Expectation: the number of colonized predicted patches decrease with increasing distance.
 #### why: long-distance dispersal is limited from colonized to suitable patches. 
-##### test: logistic regression, independent: distance, dependent: colonization (0|1)
+##### test: glm(colonization(0|1) ~ enn,  family = binomial)
 
-sqrt(c( 1000, 150))
+# nearest neighbor distance (enn_mn) - (Class) # note other method of aggregation can be used too!
+## (modification) log(enn_mn), sqrt(enn_mn)
+### Expectation: the number of colonized predicted patches increases with increasing distance.
+#### why: this species is adept at long distance dispersal
+##### test: glm(colonization(0|1) ~ enn_mn,  family = binomial)
 ```
@@
 ```{r Patch metrics to predict abundance}
-library(landscapemetrics)
+# library(landscapemetrics)
 
@@
 print(
-  list_lsm(level = "patch"), 
-  n = 12)
+  landscapemetrics::list_lsm(level = "class"))
 

### 2024-01-24 scripts/patches.R

@@ -1,17 +1,0 @@
-library(terra)
-library(tidyverse)
-r <- rast(nrows=18, ncols=36, xmin=0)
-r[1:2, 5:8] <- 1.1
-r[5:8, 2:6] <- 1.2
-r[7:12, 22:36] <- 1.3
-r[15:16, 18:29] <- 1.4
-p <- patches(r)
-
-plot(p)
-
-tibble(
-  Patch = values(p)[!is.na(values(p))],
-  Suitability = values(r)[!is.na(values(p))]
-) %>% 
-  sample_n(15)
-

### 2024-01-25 scripts/classify_patches.Rmd
watershed bounds

@@
 p <- '../results/suitability_maps'
-r <- rast(file.path(p, 'Achnatherum_aridum1k-2024-01-10_09:16:55.tif'))
+r <- rast(file.path(p, 'Acamptopappus_sphaerocephalus1k-2024-01-10_08:57:46.tif'))
 msk <- ifel(r < 0.80, NA, r)
@@
 
+We need to identify and names patches.
+
 ```{r}
-system.time({ob <- terra::patches(m1,  directions = 4)})
-system.time({z =  zonal(cellSize(ob, unit="ha"), ob, sum, as.raster=TRUE) })
-z = ifel(z < 2.0235, NA, z)
+system.time({ob <- terra::patches(m1,  directions = 4, allowgaps = FALSE)})
+system.time({sizes =  zonal(cellSize(ob, unit="ha"), ob, sum, as.raster=TRUE) }) # use this has a mask
+size_mask = ifel(sizes < 2.0235, NA, sizes) # remove areas < 5 acres.
+ob <- mask(ob, z)
+plot(ob)
+```
 
-plot(z)
+```{r}
+
+p <- '../results/suitability_maps'
+f <- file.path(p, list.files(p, pattern = '1k-'))
+
+patcheR <- functions(x){
+
+  # read in file
+  r <- rast(f)
+  taxon <- gsub('1k.*$', '', basename(f))
+  pout <- '../results/patches'
+
+  # mask raster values < 0.8
+  r <- mask(r, ifel(r < 0.80, NA, r))
+  r1 <- aggregate(r, 2, fun = 'mean')
+
+  # burn away rivers, and burn patches to hydrologic units ~ populations
+  r1 <- mask(r1, hydro, inverse = TRUE)
+
+  # detect patches
+  pat <- terra::patches(r1,  directions = 4, allowgaps = FALSE)
+
+  # remove patches < 5 acres
+  sizes <- zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
+  size_mask <- ifel(sizes < 2.0235, NA, sizes)
+  r1 <- mask(r1, size_mask)
+
+  # resample the patches to the original raster resolution
+  resample(r1, r, threads = 16, method = 'near',
+           filename = file.path(pout, paste0(taxon, '.tif')))
+
+}
+
 ```
@@
 ##### test: glm(colonization(0|1) ~ enn_mn,  family = binomial)
+
 ```

### 2024-01-25 scripts/make_Impermeable_raster.R

@@ -53,6 +53,7 @@ rivr <- terra::mask(rivr, vect(ocean), touches = TRUE, update = TRUE)
 rivr <- terra::mask(rivr, vect(lakes10), touches = TRUE, update = TRUE)
 
 plot(lakes10[,1])
+
 ##############################################################
 ## read and watershed boundaries and convert some to raster ##
 ## to delineate groups (spatial structured populations, population, deme) ##
@@ -61,10 +62,13 @@ st_layers(p2gdb)
 
 hu10 <- st_read(p2gdb, layer = 'WBDHU10', quiet = TRUE)  %>% 
   st_make_valid() %>% 
+  select(huc10) %>% 
   st_crop(., project(domain, crs(.))) %>% 
   st_simplify() %>% 
   st_transform(5070) %>% 
   st_simplify(dTolerance = 80)
+
+hu10_line <- st_cast(hu10 , 'LINESTRING')
 
 hu12 <- st_read(p2gdb, layer = 'WBDHU12', quiet = TRUE)  %>% 
   st_make_valid() %>% 
@@ -73,12 +77,12 @@ hu12 <- st_read(p2gdb, layer = 'WBDHU12', quiet = TRUE)  %>%
   st_transform(5070) %>% 
   st_simplify(dTolerance = 80)
 
-hu10 <- select(hu10, huc10)
-hu12 <- select(hu12, huc12)
+hu10R <- terra::rasterize(vect(hu10), rast(slope), update = TRUE, field = 'huc10', 
+                          filename = '../../Geospatial_data/WBD_rast/hu10R.tif', overwrite = T)
 
-hu10R <- terra::rasterize(vect(hu10), rast(slope), update = TRUE, field = 'huc10')
-hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12')
+terra::rasterize(vect(hu10_line), rast(slope), update = TRUE, field = 'huc10', 
+                 filename = '../../Geospatial_data/WBD_rast/hu10R-line.tif', overwrite = T)
 
-writeRaster(hu10R, '../../Geospatial_data/WBD_rast/hu10R.tif', overwrite = TRUE)
-writeRaster(hu12R, '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = TRUE)
+hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12', 
+                          filename = '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = T)
 

### 2024-01-25 scripts/classify_patches.Rmd
fn to save patches rasters

@@
 m <- mask(r, msk)
+
+m <- mask(m, crop(hydr_bound, m), inverse = TRUE)
+plot(m)
 m1 <- aggregate(m, 2, fun = 'mean')
@@
 
-patcheR <- functions(x){
+gd <- '/media/steppe/hdd/Geospatial_data'
+hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu10R-line.tif') )
+hydr_ftr <-  rast( file.path(gd, 'river_resistance', 'water_ftR.tif') )
+
+patcheR <- function(x){
 
   # read in file
-  r <- rast(f)
-  taxon <- gsub('1k.*$', '', basename(f))
+  r <- rast(x)
+  taxon <- gsub('1k.*$', '', basename(x))
   pout <- '../results/patches'
@@
   # mask raster values < 0.8
-  r <- mask(r, ifel(r < 0.80, NA, r))
-  r1 <- aggregate(r, 2, fun = 'mean')
+  r <- terra::mask(r, ifel(r < 0.80, NA, r))
 
     # burn away rivers, and burn patches to hydrologic units ~ populations
-  r1 <- mask(r1, hydro, inverse = TRUE)
+  r <- terra::mask(r, terra::crop(hydr_bound, r), inverse = TRUE)
+  r <- terra::mask(r, terra::crop(hydr_ftr, r), inverse = TRUE)
 
   # detect patches
-  pat <- terra::patches(r1,  directions = 4, allowgaps = FALSE)
+  r1 <- terra::aggregate(r, 2, fun = 'mean')
+  pat <- terra::patches(r1,  directions = 4, allowGaps = FALSE)
 
   # remove patches < 5 acres
-  sizes <- zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
-  size_mask <- ifel(sizes < 2.0235, NA, sizes)
+  sizes <- terra::zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
+  size_mask <- terra::ifel(sizes < 4.0470, NA, sizes)
   r1 <- mask(r1, size_mask)
@@
   # resample the patches to the original raster resolution
-  resample(r1, r, threads = 16, method = 'near',
+  terra::resample(r1, r, threads = 16, method = 'near',
            filename = file.path(pout, paste0(taxon, '.tif')))
@@
 }
+
+lapply(f[4:335], patcheR)
 

### 2024-01-26 scripts/classify_patches.Rmd
rescaling patch detection to data size

@@
 ```{r}
-p <- '../results/suitability_maps'
-r <- rast(file.path(p, 'Acamptopappus_sphaerocephalus1k-2024-01-10_08:57:46.tif'))
-msk <- ifel(r < 0.80, NA, r)
-m <- mask(r, msk)
-
-m <- mask(m, crop(hydr_bound, m), inverse = TRUE)
-plot(m)
-m1 <- aggregate(m, 2, fun = 'mean')
-
-# aggregate the raster to speed up the patch detection process. 
-```
-
-aggregation by a factor of 2 allows for quick patch delineation and detection.
-
-We need to identify and names patches.
-
-```{r}
-system.time({ob <- terra::patches(m1,  directions = 4, allowgaps = FALSE)})
-system.time({sizes =  zonal(cellSize(ob, unit="ha"), ob, sum, as.raster=TRUE) }) # use this has a mask
-size_mask = ifel(sizes < 2.0235, NA, sizes) # remove areas < 5 acres.
-ob <- mask(ob, z)
-plot(ob)
-```
-
-```{r}
 
@@
 f <- file.path(p, list.files(p, pattern = '1k-'))
+info <- lapply(f, file.info)
+
+info1 <- info[[1]]$size/1e9
 
@@
   pout <- '../results/patches'
+  fsize <- file.info(f[1])$size/1e9
 
@@
 
-  # detect patches
-  r1 <- terra::aggregate(r, 2, fun = 'mean')
+  # detect patches - this scales to the extent of analysis. #
+  # Areas need to be aggregateed for computatations to be run in meaningful amount of
+  # time
+
+  if(fsize < 0.2){r1 <- r} else
+    if(fsize < 0.4){r1 <- terra::aggregate(r, 2, fun = 'mean')} else 
+      if(fsize < 0.7){r1 <- terra::aggregate(r, 3, fun = 'mean')} else
+        {r1 <- terra::aggregate(r, 4, fun = 'mean')}
+
   pat <- terra::patches(r1,  directions = 4, allowGaps = FALSE)
@@
 
+  terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
+
 }
 
-lapply(f[4:335], patcheR)
+lapply(f, patcheR)
 

### 2024-01-26 scripts/classify_patches.Rmd
fn to subset patches - tweaking aggregation factors for 10-20 mins per spp

@@
 f <- file.path(p, list.files(p, pattern = '1k-'))
-info <- lapply(f, file.info)
-
-info1 <- info[[1]]$size/1e9
 
@@
 hydr_ftr <-  rast( file.path(gd, 'river_resistance', 'water_ftR.tif') )
+rm(gd)
 
@@
   pout <- '../results/patches'
-  fsize <- file.info(f[1])$size/1e9
+  fsize <- file.info(x)$size/1e9
 
@@
 
-  if(fsize < 0.2){r1 <- r} else
-    if(fsize < 0.4){r1 <- terra::aggregate(r, 2, fun = 'mean')} else 
-      if(fsize < 0.7){r1 <- terra::aggregate(r, 3, fun = 'mean')} else
-        {r1 <- terra::aggregate(r, 4, fun = 'mean')}
+  if(fsize < 0.3){r1 <- terra::aggregate(r, 2, fun = 'mean')} else
+    if(fsize < 0.5){r1 <- terra::aggregate(r, 3, fun = 'mean')} else 
+      if(fsize < 0.7){r1 <- terra::aggregate(r, 4, fun = 'mean')} else
+        if(fsize > 0.9){r1 <- terra::aggregate(r, 5, fun = 'mean')}
 
@@
 
-lapply(f, patcheR)
+lapply(f[4:355], patcheR) # start 9:05
 

### 2024-01-30 scripts/classify_patches.Rmd
writing fn to calc landscape metrics

@@
 library(terra)
+library(landscapemetrics)
 ```
+
+Create coarse patches for 'meta-populations'
 
@@
 
-patcheR <- function(x){
+lapply(f, patcheR) # roughly 72 hours.
 
-  # read in file
-  r <- rast(x)
-  taxon <- gsub('1k.*$', '', basename(x))
-  pout <- '../results/patches'
-  fsize <- file.info(x)$size/1e9
+```
 
-  # mask raster values < 0.8
-  r <- terra::mask(r, ifel(r < 0.80, NA, r))
+```{r}
 
-    # burn away rivers, and burn patches to hydrologic units ~ populations
-  r <- terra::mask(r, terra::crop(hydr_bound, r), inverse = TRUE)
-  r <- terra::mask(r, terra::crop(hydr_ftr, r), inverse = TRUE)
+abvi <- rast('../results/patches/Abronia_villosa.tif')
 
-  # detect patches - this scales to the extent of analysis. #
-  # Areas need to be aggregateed for computatations to be run in meaningful amount of
-  # time
+gd <- '/media/steppe/hdd/Geospatial_data'
+hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu10R-line.tif') )
 
-  if(fsize < 0.3){r1 <- terra::aggregate(r, 2, fun = 'mean')} else
-    if(fsize < 0.5){r1 <- terra::aggregate(r, 3, fun = 'mean')} else 
-      if(fsize < 0.7){r1 <- terra::aggregate(r, 4, fun = 'mean')} else
-        if(fsize > 0.9){r1 <- terra::aggregate(r, 5, fun = 'mean')}
+landscapR <- function(x, pout, pRout){
 
-  pat <- terra::patches(r1,  directions = 4, allowGaps = FALSE)
+  x <- terra::ifel(x > 0.75, 1, NA)
+  x <- terra::mask(x, f_wbd, inverse = TRUE)
+  landscape <- landscapemetrics::get_patches(x, directions = 4)
+  calcs <- landscapemetrics::calculate_lsm(x, what = c('lsm_c_enn_mn', 'lsm_c_enn_cv',
+                "lsm_p_enn", "lsm_p_enn", "lsm_p_cai", "lsm_p_para", "lsm_p_frac"), 
+              neighbourhood = 4, directions = 4)
 
-  # remove patches < 5 acres
-  sizes <- terra::zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
-  size_mask <- terra::ifel(sizes < 4.0470, NA, sizes)
-  r1 <- mask(r1, size_mask)
+  p <- '../results/'
+  write.csv(
+    calcs[calcs$level == 'patch', c('id', 'metric', 'value')],
+    filename = paste0(p, pout, '/', taxon, '-', 'patch_vals.csv'),  row.names = F
+  )
 
-  # resample the patches to the original raster resolution
-  terra::resample(r1, r, threads = 16, method = 'near',
-           filename = file.path(pout, paste0(taxon, '.tif')))
+  write.csv(
+    calcs[calcs$level == 'class', c('id', 'metric', 'value')],
+    filename = paste0(p, pout, '/', taxon, '-', 'class_vals.csv'), row.names = F
+  )
 
-  terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
+  writeRaster(landscape[["layer_1"]][["class_1"]], 
+    filename = paste0(p, pRout, '/', taxon, '-', 'patchIDS.tif')
+  )
 
@@
 
-lapply(f[4:355], patcheR) # start 9:05
+landscapR(f, pout = 'patch_metrics', pRout = 'fine_patches')
 
@@
 
-# euclidean nearest neighbor distance (enn) - (Patch)
+# euclidean nearest neighbor distance (enn) - (Patch) {lsm_p_enn}
 ## (modification) log(enn), sqrt(enn)
@@
 
-# nearest neighbor distance (enn_mn) - (Class) # note other method of aggregation can be used too!
+# nearest neighbor distance (enn_mn) - (Class) {lsm_c_enn_mn} # note other method of aggregation can be used too!
 ## (modification) log(enn_mn), sqrt(enn_mn)
@@
 
-# Core Area Index (cai)  
+# Core Area Index (cai)  {lsm_p_cai}
 ## (modification) arithmetic mean of core area cells
@@
 
-# perimeter-area ratio (para)
+# perimeter-area ratio (para) {lsm_p_para}
 ## (modification) none
@@
 
-# fractal dimension index (frac)
+# fractal dimension index (frac) {lsm_p_frac}
 ## (modification) none

### 2024-01-30 scripts/functions.R

@@ -529,3 +529,45 @@ rfe_var_selector <- function(x){
 #    fi
 # we set up a cron job to run every24 hours to move the files
 # 0 */3 * * * sdd2hdd.sh
+
+
+
+
+patcheR <- function(x){
+  
+  # read in file
+  r <- rast(x)
+  taxon <- gsub('1k.*$', '', basename(x))
+  pout <- '../results/patches'
+  fsize <- file.info(x)$size/1e9
+  
+  # mask raster values < 0.8
+  r <- terra::mask(r, ifel(r < 0.80, NA, r))
+  
+  # burn away rivers, and burn patches to hydrologic units ~ populations
+  r <- terra::mask(r, terra::crop(hydr_bound, r), inverse = TRUE)
+  r <- terra::mask(r, terra::crop(hydr_ftr, r), inverse = TRUE)
+  
+  # detect patches - this scales to the extent of analysis. #
+  # Areas need to be aggregateed for computatations to be run in meaningful amount of
+  # time
+  
+  if(fsize < 0.3){r1 <- terra::aggregate(r, 2, fun = 'mean')} else
+    if(fsize < 0.5){r1 <- terra::aggregate(r, 3, fun = 'mean')} else 
+      if(fsize < 0.7){r1 <- terra::aggregate(r, 4, fun = 'mean')} else
+        if(fsize > 0.7){r1 <- terra::aggregate(r, 5, fun = 'mean')}
+  
+  pat <- terra::patches(r1,  directions = 4, allowGaps = FALSE)
+  
+  # remove patches < 5 acres
+  sizes <- terra::zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
+  size_mask <- terra::ifel(sizes < 4.0470, NA, sizes)
+  r1 <- mask(r1, size_mask)
+  
+  # resample the patches to the original raster resolution
+  terra::resample(r1, r, threads = 16, method = 'near',
+                  filename = file.path(pout, paste0(taxon, '.tif')))
+  
+  terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
+  
+}

### 2024-01-31 scripts/classify_patches.Rmd
push framing to parkland

@@
 
-```{r}
+```{r, eval = F}
 
@@
 
-abvi <- rast('../results/patches/Abronia_villosa.tif')
+f <- list.files('../results/patches/')
 
 gd <- '/media/steppe/hdd/Geospatial_data'
-hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu10R-line.tif') )
+hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu12R-line.tif') )
 
@@
 
-  x <- terra::ifel(x > 0.75, 1, NA)
-  x <- terra::mask(x, f_wbd, inverse = TRUE)
+  taxon <- gsub('[.]tif', '', basename(x))
+  x <- terra::rast(file.path('../results/patches', x))
+  x <- terra::ifel(x > 0, 1, NA)
+  x <- terra::mask(x, terra::crop(hydr_bound, x), inverse = TRUE)
   landscape <- landscapemetrics::get_patches(x, directions = 4)
   calcs <- landscapemetrics::calculate_lsm(x, what = c('lsm_c_enn_mn', 'lsm_c_enn_cv',
-                "lsm_p_enn", "lsm_p_enn", "lsm_p_cai", "lsm_p_para", "lsm_p_frac"), 
+                "lsm_p_enn", "lsm_p_cai", "lsm_p_para", "lsm_p_frac"), 
               neighbourhood = 4, directions = 4)
@@
     calcs[calcs$level == 'patch', c('id', 'metric', 'value')],
-    filename = paste0(p, pout, '/', taxon, '-', 'patch_vals.csv'),  row.names = F
+    file = paste0(p, pout, '/', taxon, '-', 'patch_vals.csv'),  row.names = F
   )
@@
   write.csv(
-    calcs[calcs$level == 'class', c('id', 'metric', 'value')],
-    filename = paste0(p, pout, '/', taxon, '-', 'class_vals.csv'), row.names = F
+    calcs[calcs$level == 'class', c('metric', 'value')],
+    file = paste0(p, pout, '/', taxon, '-', 'class_vals.csv'), row.names = F
   )
@@
 
-landscapR(f, pout = 'patch_metrics', pRout = 'fine_patches')
+lapply(f[4:20], FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
 

### 2024-01-31 scripts/make_Impermeable_raster.R

@@ -77,6 +77,8 @@ hu12 <- st_read(p2gdb, layer = 'WBDHU12', quiet = TRUE)  %>%
   st_transform(5070) %>% 
   st_simplify(dTolerance = 80)
 
+hu12_line <- st_cast(hu12 , 'LINESTRING')
+
 hu10R <- terra::rasterize(vect(hu10), rast(slope), update = TRUE, field = 'huc10', 
                           filename = '../../Geospatial_data/WBD_rast/hu10R.tif', overwrite = T)
 
@@ -86,3 +88,5 @@ terra::rasterize(vect(hu10_line), rast(slope), update = TRUE, field = 'huc10',
 hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12', 
                           filename = '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = T)
 
+terra::rasterize(vect(hu12_line), rast(slope), update = TRUE, field = 'huc12', 
+                 filename = '../../Geospatial_data/WBD_rast/hu12R-line.tif', overwrite = T)

### 2024-01-31 scripts/classify_patches.Rmd
adding skeletons for grabbing occurrence and abundance data

@@
 
-lapply(f[4:20], FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
+lapply(f[21:335], FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
 
@@
 ## (modification) log(enn_mn), sqrt(enn_mn)
-### Expectation: the number of colonized predicted patches increases with increasing distance.
-#### why: this species is adept at long distance dispersal
-##### test: glm(colonization(0|1) ~ enn_mn,  family = binomial)
+### Expectation: the proportion of predicted patches which are colonized decreases with increaing mean distance between patches
+#### why: further distances between suitable habitats prevent colonization and recolonization of habitat.
+##### test: glm(prcnt_patches_colonized ~ enn_mn,  family = (quasi)binomial)
 
@@
 ```
+
+To determine whether a patch is occupied the original training data are matched as an attribute to the fine patches.
+
+```{r}
+
+f <- file.path('../results/fine_patches', list.files('../results/fine_patches'))
+
+recs <- st_read('../data/raw/occurrence/combined_records') %>% 
+  filter(PtType == 1)
+
+patchTaggR <- function(x){
+
+  r <- terra::rast(f)
+  r_taxon <- gsub('[.]tif', '', basename(f))
+  occs <- dplyr::filter(recs, taxon == r_taxon) |>
+    dplyr::mutate(ID = 1:nrow(.)) |>
+    terra::vect()
+
+  occ_pts <- terra::extract(r, occs, bind = TRUE) |>
+    sf::st_as_sf()
+
+  # if pts were NA, buffer to 90m, many occupied areas were 'cut' away by the 
+  # hydrologic mapping. 
+  occ_poly <- dplyr::filter(occ_pts, is.na(RASTER)) |>
+    sf::st_buffer(occ_pts, 120) # we cut out 90, but let's go a bit further for corner cases.
+  occ_poly <- terra::extract(r, occ_poly, bind = TRUE, weights = TRUE) |>
+    sf::st_as_sf() |>
+    sf::st_centroid()  |>
+    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
+    dplyr::slice_max(WEIGHT) |>
+    dplyr::drop_na(RASTER) # these values actually NA.
+
+  occ_pops <- dplyr::bind_rows( # a dataframe of all raster patch ID's with populations known from within
+    # the last few decades in them. 
+    dplyr::drop_na(occ_pts), occ_poly) |>
+    dplyr::distinct(RASTER, .keep_all = TRUE) # this is multiple pts per patch - not neccessary for us.
+
+  write.csv(occ_pops, row.names = F,
+            file.path('../results/occupiedPatches', paste0(r_taxon, '.csv')))
+
+}
+
+```
+
+```{r}
+
+p1 <- '../data/raw/2021_AIM_Terrestrial/AIMTerrestrial9-1-22.gdb'
+st_layers(dsn = p1)
+
+AIM_points <- st_read(dsn = p1, 
+        layer = 'TerrADat', quiet = T) %>%
+  select(PrimaryKey, DateVisited, geometry = Shape) %>% 
+  st_transform(5070)
+
+lpi <- spp_richness <- st_read(dsn = p1, 
+        layer = 'tbl LINE POINT INTERCEPT', quiet = T) %>% 
+  select(SpeciesList, PrimaryKey) %>% 
+  right_join(AIM_points, by = 'PrimaryKey') %>% 
+  st_as_sf()  %>%
+
+  # NOW CALCULATE THE PERCENT COVER OF EACH SPECIES FROM LPI #
+  # ...
+
+  mutate(DateVisited = as.Date(DateVisited), .before = geometry) %>% 
+  arrange(SpeciesList) %>% 
+  right_join(., select(targets, taxon, symbol) %>% 
+               distinct(), by = c('SpeciesList' = 'symbol' ))
+
+# add the species names from USDA PLANTS CODES
+
+# ...
+
+AIM_abundance_retrievR <- function(x, aim_tables){
+
+}
+
+```

### 2024-02-06 scripts/classify_patches.Rmd
fn to note occupied patches

@@
 library(terra)
+library(tidyverse)
+library(sf)
 library(landscapemetrics)
@@
 
-```{r}
+Create finer scale patches which more or less are meant to represent populations
+
+```{r eval = F}
 
@@
 
-lapply(f[21:335], FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
+lapply(f, FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
 
@@
 f <- file.path('../results/fine_patches', list.files('../results/fine_patches'))
+p <- '../data/raw/occurrence/combined_records'
 
-recs <- st_read('../data/raw/occurrence/combined_records') %>% 
-  filter(PtType == 1)
+recs <- lapply(
+  file.path(p, list.files(p, pattern = '.shp$')), sf::st_read, quiet = TRUE) %>% 
+  bind_rows() %>% 
+  filter(Occurrence == 1)
+
+r1 <- rast(f[1])
 
@@
 
-  r <- terra::rast(f)
-  r_taxon <- gsub('[.]tif', '', basename(f))
+  r <- terra::rast(x)
+  r_taxon <- gsub('_', ' ', gsub('-patchIDS[.]tif', '', basename(x)))
   occs <- dplyr::filter(recs, taxon == r_taxon) |>
-    dplyr::mutate(ID = 1:nrow(.)) |>
+    dplyr::mutate(ID = dplyr::row_number()) |>
     terra::vect()
@@
   # hydrologic mapping. 
-  occ_poly <- dplyr::filter(occ_pts, is.na(RASTER)) |>
-    sf::st_buffer(occ_pts, 120) # we cut out 90, but let's go a bit further for corner cases.
-  occ_poly <- terra::extract(r, occ_poly, bind = TRUE, weights = TRUE) |>
-    sf::st_as_sf() |>
-    sf::st_centroid()  |>
+  occ_poly <- dplyr::filter(occ_pts, is.na(lyr.1)) |>
+    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
+
+  occ_poly <- terra::extract(r, occ_poly, weights = TRUE) %>% 
     dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
-    dplyr::slice_max(WEIGHT) |>
-    dplyr::drop_na(RASTER) # these values actually NA.
+    tidyr::drop_na(lyr.1) |> # these values actually NA. 
+    dplyr::slice_max(weight, with_ties = F) |>
+    dplyr::left_join(filter(occ_pts) |>
+                       select(-lyr.1), by = join_by('ID')) |>
+    dplyr::select(-weight)
 
-  occ_pops <- dplyr::bind_rows( # a dataframe of all raster patch ID's with populations known from within
+  occ_pops <- dplyr::bind_rows( # a data frame of all raster patch ID's with populations known from within
     # the last few decades in them. 
-    dplyr::drop_na(occ_pts), occ_poly) |>
-    dplyr::distinct(RASTER, .keep_all = TRUE) # this is multiple pts per patch - not neccessary for us.
+    tidyr::drop_na(occ_pts, lyr.1), occ_poly) |>
+    dplyr::rename(PatchID = lyr.1)
 
-  write.csv(occ_pops, row.names = F,
-            file.path('../results/occupiedPatches', paste0(r_taxon, '.csv')))
+  # write out summary values of the patch matching
+  data.frame(
+    noPres = , 
+    PresTopSuitability =, # the number of presences located in top suitability patches
+    UniquePatchID = , # the number of uniquelly identified patches
+    NoPatchesTopSuitability =  # the total number of most suitable patches from the RASTER
+  )
+
+  occ_pops <- dplyr::distinct(occ_pops, PatchID, .keep_all = TRUE) |> # this is multiple pts per patch - not necessary for us. 
+    sf::st_drop_geometry() |>
+    dplyr::select(date, PatchID) |>
+    dplyr::arrange(PatchID) |>
+    write.csv(row.names = F,
+            file.path('../results/occupiedPatches-topSuitability', paste0(r_taxon, '.csv')))
+
+  # for points which were not patched to a most suitable patch, put them into a hydrologic basin
 
@@
 
+patchTaggR(f[1])
 ```
@@
 ```{r}
+PLANTS <- read.csv('../data/raw/2024sos_species.csv') |>
+  select(genus:symbol) |>
+  unite(sctfcnm, genus:infraspecies, na.rm = TRUE)
 

### 2024-02-06 scripts/classify_patches.Rmd
working on fn to detect marginal populations

@@
 
+hu12 <- rast('../../Geospatial_data/WBD_rast/hu12R-line.tif')
+plot(hu12)
+
 patchTaggR <- function(x){
@@
   # write out summary values of the patch matching
+
+  fp <- '../results/occupiedPatches'
   data.frame(
-    noPres = , 
-    PresTopSuitability =, # the number of presences located in top suitability patches
-    UniquePatchID = , # the number of uniquelly identified patches
-    NoPatchesTopSuitability =  # the total number of most suitable patches from the RASTER
-  )
+    noPres = nrow(occs), 
+    PresTopSuitability = nrow(occ_pops), # the number of presences located in top suitability patches
+    UniquePatchID = nrow(dplyr::distinct(occ_pops, PatchID)), # the number of uniquelly identified patches
+    NoPatchesTopSuitability = terra::minmax(r)[2] # the total number of most suitable patches from the RASTER
+  ) |>
+    write.csv(row.names = F, 
+              file.path(fp, paste0(r_taxon, '-summary', '.csv')))
 
-  occ_pops <- dplyr::distinct(occ_pops, PatchID, .keep_all = TRUE) |> # this is multiple pts per patch - not necessary for us. 
+  dplyr::distinct(occ_pops, PatchID, .keep_all = TRUE) |> # this is multiple pts per patch - not necessary for us. 
     sf::st_drop_geometry() |>
@@
     write.csv(row.names = F,
-            file.path('../results/occupiedPatches-topSuitability', paste0(r_taxon, '.csv')))
+            file.path(fp, paste0(r_taxon, '-topSuitability', '.csv')))
 
   # for points which were not patched to a most suitable patch, put them into a hydrologic basin
+  marginal_pops <- sf::st_as_sf(occs)|>
+    dplyr::filter(!ID %in% occ_pops$ID)
 
+  # identify the drainages within the HU12 database
+  drain_nos <- terra::extract(r_drain, marginal_pops, bind = TRUE) |>
+    sf::st_as_sf()
+  r_drain_sub <- terra::mask()
+
+  # mask original suitability map to these areas, and filter >60% suitable habitat
+  f1 <- list.files('suitability_maps')
+
+  suitable <- terra::mask(suitable, r_drain_sub)
+  msk <- ifel(suitable > 55 & suitable < 80, NA, 1) # let's make sure these areas do not include existing clusters. 
+  suitable <- mask(suitable, msk) # this is now a second raster for occupied marginal populations
+
+  writeRaster(paste0('../results/marginal_habitat', r_taxon, '.tif'))
+  return(marginal_pops)
 }
 
-patchTaggR(f[1])
+ob <- patchTaggR(f[1])
 ```

### 2024-02-06 scripts/make_Impermeable_raster.R

@@ -85,7 +85,7 @@ hu10R <- terra::rasterize(vect(hu10), rast(slope), update = TRUE, field = 'huc10
 terra::rasterize(vect(hu10_line), rast(slope), update = TRUE, field = 'huc10', 
                  filename = '../../Geospatial_data/WBD_rast/hu10R-line.tif', overwrite = T)
 
-hu12R <- terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12', 
+terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12', 
                           filename = '../../Geospatial_data/WBD_rast/hu12R.tif', overwrite = T)
 
 terra::rasterize(vect(hu12_line), rast(slope), update = TRUE, field = 'huc12', 

### 2024-02-13 scripts/classify_patches.Rmd
text

--- 2024-02-06 scripts/classify_patches.Rmd
@@
 library(landscapemetrics)
+source('functions.R')
 ```
@@
 
-r1 <- rast(f[1])
+gd <- '/media/steppe/hdd/Geospatial_data'
+hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu12R-line.tif') )
+hu12 <- rast('../../Geospatial_data/WBD_rast/hu12R.tif')
 
-patchTaggR <- function(x){
+lapply(f, patchTaggR)
+```
 
-  r <- terra::rast(x)
-  r_taxon <- gsub('_', ' ', gsub('-patchIDS[.]tif', '', basename(x)))
-  occs <- dplyr::filter(recs, taxon == r_taxon) |>
-    dplyr::mutate(ID = dplyr::row_number()) |>
-    terra::vect()
+Combine all populated patches per species, something was amiss with identifying patches in marginal habitat, we will thin them down here.
 
-  occ_pts <- terra::extract(r, occs, bind = TRUE) |>
+```{r}
+
+f <- gsub('[.]tif$', '', list.files('../results/patches'))
+
+combine_pop_patches <- function(x){
+
+  pIDs <- read.csv(paste0('../results/occupiedPatches/', x, '-topSuitability.csv'))
+  top_patches <- terra::rast(paste0('../results/fine_patches/', x, '-patchIDS.tif'))
+  highest_no <- minmax(top_patches)[2]
+  top_patches <- terra::subst(top_patches, pIDs$PatchID,  pIDs$PatchID, others = NA)
+  names(top_patches)<- 'patches'
+  pts <- st_read(paste0('../data/raw/occurrence/combined_records/', x, '.shp'), quiet = TRUE) |>
+    dplyr::filter(Occurrence == 1)
+  marg_patch <- terra::rast(paste0('../results/marginal_habitat/', x, '.tif'))
+
+  fsize <- file.info(paste0('../results/marginal_habitat/', x, '.tif'))$size/1e9
+  if(fsize < 0.025){marg_patch <- terra::patches(marg_patch, directions = 4)} else
+    if(fsize > 0.025){marge1 <- terra::patches(terra::aggregate(marg_patch, 2, fun = 'mean'), directions = 4)} else
+      if(fsize > 0.04){marge1 <- terra::patches(terra::aggregate(marg_patch, 3, fun = 'mean'), directions = 4)} else
+        if(fsize > 0.05){marge1 <- terra::patches(terra::aggregate(marg_patch, 4, fun = 'mean'), directions = 4)} else
+          if(fsize > 0.06){marge1 <- terra::patches(terra::aggregate(marg_patch, 5, fun = 'mean'), directions = 4)} else
+            if(fsize > 0.07){marge1 <- terra::patches(terra::aggregate(marg_patch, 6, fun = 'mean'), directions = 4)}
+  if(exists('marge1')){marg_patch <- terra::resample(marge1, marg_patch, threads = 16, method = 'near')}
+
+  occ_pts <- terra::extract(marg_patch, terra::vect(pts), bind = TRUE) |>
     sf::st_as_sf()
+  occ_poly <- dplyr::filter(occ_pts, is.na(patches)) |>
+    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
+  occ_poly <- terra::extract(marg_patch, terra::vect(occ_poly), weights = TRUE) %>% 
+    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
+    tidyr::drop_na(patches) |> # these values actually NA. 
+    dplyr::slice_max(weight, with_ties = F)
 
-  # if pts were NA, buffer to 90m, many occupied areas were 'cut' away by the 
-  # hydrologic mapping. 
-  occ_poly <- dplyr::filter(occ_pts, is.na(lyr.1)) |>
-    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
-
-  occ_poly <- terra::extract(r, occ_poly, weights = TRUE) %>% 
-    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
-    tidyr::drop_na(lyr.1) |> # these values actually NA. 
-    dplyr::slice_max(weight, with_ties = F) |>
-    dplyr::left_join(filter(occ_pts) |>
-                       select(-lyr.1), by = join_by('ID')) |>
-    dplyr::select(-weight)
-
-  occ_pops <- dplyr::bind_rows( # a data frame of all raster patch ID's with populations known from within
-    # the last few decades in them. 
-    tidyr::drop_na(occ_pts, lyr.1), occ_poly) |>
-    dplyr::rename(PatchID = lyr.1)
-
-  # write out summary values of the patch matching
-  data.frame(
-    noPres = , 
-    PresTopSuitability =, # the number of presences located in top suitability patches
-    UniquePatchID = , # the number of uniquelly identified patches
-    NoPatchesTopSuitability =  # the total number of most suitable patches from the RASTER
+  occ_patches <- c(
+    dplyr::filter(occ_pts, ! is.na(patches)) |>
+      dplyr::pull(patches), 
+    occ_poly$patches
   )
 
-  occ_pops <- dplyr::distinct(occ_pops, PatchID, .keep_all = TRUE) |> # this is multiple pts per patch - not necessary for us. 
-    sf::st_drop_geometry() |>
-    dplyr::select(date, PatchID) |>
-    dplyr::arrange(PatchID) |>
-    write.csv(row.names = F,
-            file.path('../results/occupiedPatches-topSuitability', paste0(r_taxon, '.csv')))
+  if(length(occ_patches) > 0){
 
-  # for points which were not patched to a most suitable patch, put them into a hydrologic basin
+    marg_patch <- terra::subst(marg_patch, occ_patches,  occ_patches, others = NA)
+    old_ID <- unique(marg_patch)$patches # gather and relabel patches 
+    lkp <- data.frame(
+      old_ID = old_ID,
+      new_ID = highest_no + 1:length(old_ID)
+    )
+    marg_patch <- terra::subst(marg_patch, lkp$old_ID, lkp$new_ID)
+    final_patches <- terra::cover(top_patches, marg_patch)
+
+    terra::writeRaster(final_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
+  } else {
+    terra::writeRaster(top_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
+  }
+}
+
+# lentiformis  NTIF
+
+lapply(f[c(288:335)], combine_pop_patches) # 2:46
+```
+
+```{r create a polygon dataset, eval = F}
+
+  f <- paste0('../results/allOccupiedPatches/', 
+            list.files('../results/allOccupiedPatches/'))
+
+tif2shp <- function(x){
+
+  taxon <- gsub('.tif$', '',  basename(x))
+  pg <- terra::as.polygons(rast(x)) %>% 
+    sf::st_as_sf() %>% 
+    sf::st_make_valid()
+
+  sf::st_write(pg, paste0('../results/allOccupiedPatchesVector/', taxon, '.shp'), quiet = TRUE)
 
@@
 
-patchTaggR(f[1])
+lapply(f, tif2shp)
+
 ```

### 2024-02-13 scripts/functions.R

@@ -571,3 +571,81 @@ patcheR <- function(x){
   terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
   
 }
+
+
+
+#' identify patches which are known to have populations of a target species in them
+patchTaggR <- function(x){
+  
+  r <- terra::rast(x)
+  r_taxon <- gsub('_', ' ', gsub('-patchIDS[.]tif', '', basename(x)))
+  occs <- dplyr::filter(recs, taxon == r_taxon) |>
+    dplyr::mutate(ID = dplyr::row_number()) |>
+    terra::vect()
+  
+  occ_pts <- terra::extract(r, occs, bind = TRUE) |>
+    sf::st_as_sf()
+  
+  # if pts were NA, buffer to 90m, many occupied areas were 'cut' away by the 
+  # hydrologic mapping. 
+  occ_poly <- dplyr::filter(occ_pts, is.na(lyr.1)) |>
+    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
+  
+  occ_poly <- terra::extract(r, occ_poly, weights = TRUE) %>% 
+    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
+    tidyr::drop_na(lyr.1) |> # these values actually NA. 
+    dplyr::slice_max(weight, with_ties = F) |>
+    dplyr::left_join(filter(occ_pts) |>
+                       select(-lyr.1), by = join_by('ID')) |>
+    dplyr::select(-weight)
+  
+  occ_pops <- dplyr::bind_rows( # a data frame of all raster patch ID's with populations known from within
+    # the last few decades in them. 
+    tidyr::drop_na(occ_pts, lyr.1), occ_poly) |>
+    dplyr::rename(PatchID = lyr.1)
+  
+  # write out summary values of the patch matching
+  
+  fp <- '../results/occupiedPatches'
+  data.frame(
+    noPres = nrow(occs), 
+    PresTopSuitability = nrow(occ_pops), # the number of presences located in top suitability patches
+    UniquePatchID = nrow(dplyr::distinct(occ_pops, PatchID)), # the number of uniquelly identified patches
+    NoPatchesTopSuitability = terra::minmax(r)[2] # the total number of most suitable patches from the RASTER
+  ) |>
+    write.csv(row.names = F, 
[…]

### 2024-02-13 scripts/classify_patches.Rmd
r to vector

@@
 hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu12R-line.tif') )
-
-landscapR <- function(x, pout, pRout){
-
-  taxon <- gsub('[.]tif', '', basename(x))
-  x <- terra::rast(file.path('../results/patches', x))
-  x <- terra::ifel(x > 0, 1, NA)
-  x <- terra::mask(x, terra::crop(hydr_bound, x), inverse = TRUE)
-  landscape <- landscapemetrics::get_patches(x, directions = 4)
-  calcs <- landscapemetrics::calculate_lsm(x, what = c('lsm_c_enn_mn', 'lsm_c_enn_cv',
-                "lsm_p_enn", "lsm_p_cai", "lsm_p_para", "lsm_p_frac"), 
-              neighbourhood = 4, directions = 4)
-
-  p <- '../results/'
-  write.csv(
-    calcs[calcs$level == 'patch', c('id', 'metric', 'value')],
-    file = paste0(p, pout, '/', taxon, '-', 'patch_vals.csv'),  row.names = F
-  )
-
-  write.csv(
-    calcs[calcs$level == 'class', c('metric', 'value')],
-    file = paste0(p, pout, '/', taxon, '-', 'class_vals.csv'), row.names = F
-  )
-
-  writeRaster(landscape[["layer_1"]][["class_1"]], 
-    filename = paste0(p, pRout, '/', taxon, '-', 'patchIDS.tif')
-  )
-
-}
 
@@
 
-combine_pop_patches <- function(x){
-
-  pIDs <- read.csv(paste0('../results/occupiedPatches/', x, '-topSuitability.csv'))
-  top_patches <- terra::rast(paste0('../results/fine_patches/', x, '-patchIDS.tif'))
-  highest_no <- minmax(top_patches)[2]
-  top_patches <- terra::subst(top_patches, pIDs$PatchID,  pIDs$PatchID, others = NA)
-  names(top_patches)<- 'patches'
-  pts <- st_read(paste0('../data/raw/occurrence/combined_records/', x, '.shp'), quiet = TRUE) |>
-    dplyr::filter(Occurrence == 1)
-  marg_patch <- terra::rast(paste0('../results/marginal_habitat/', x, '.tif'))
-
-  fsize <- file.info(paste0('../results/marginal_habitat/', x, '.tif'))$size/1e9
-  if(fsize < 0.025){marg_patch <- terra::patches(marg_patch, directions = 4)} else
-    if(fsize > 0.025){marge1 <- terra::patches(terra::aggregate(marg_patch, 2, fun = 'mean'), directions = 4)} else
-      if(fsize > 0.04){marge1 <- terra::patches(terra::aggregate(marg_patch, 3, fun = 'mean'), directions = 4)} else
-        if(fsize > 0.05){marge1 <- terra::patches(terra::aggregate(marg_patch, 4, fun = 'mean'), directions = 4)} else
-          if(fsize > 0.06){marge1 <- terra::patches(terra::aggregate(marg_patch, 5, fun = 'mean'), directions = 4)} else
-            if(fsize > 0.07){marge1 <- terra::patches(terra::aggregate(marg_patch, 6, fun = 'mean'), directions = 4)}
-  if(exists('marge1')){marg_patch <- terra::resample(marge1, marg_patch, threads = 16, method = 'near')}
-
-  occ_pts <- terra::extract(marg_patch, terra::vect(pts), bind = TRUE) |>
-    sf::st_as_sf()
-  occ_poly <- dplyr::filter(occ_pts, is.na(patches)) |>
-    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
-  occ_poly <- terra::extract(marg_patch, terra::vect(occ_poly), weights = TRUE) %>% 
-    dplyr::group_by(ID) |> # now make sure we get the patch with the 'most' overlap with the polygon
-    tidyr::drop_na(patches) |> # these values actually NA. 
-    dplyr::slice_max(weight, with_ties = F)
-
-  occ_patches <- c(
-    dplyr::filter(occ_pts, ! is.na(patches)) |>
-      dplyr::pull(patches), 
-    occ_poly$patches
-  )
-
-  if(length(occ_patches) > 0){
-
-    marg_patch <- terra::subst(marg_patch, occ_patches,  occ_patches, others = NA)
-    old_ID <- unique(marg_patch)$patches # gather and relabel patches 
-    lkp <- data.frame(
-      old_ID = old_ID,
-      new_ID = highest_no + 1:length(old_ID)
-    )
-    marg_patch <- terra::subst(marg_patch, lkp$old_ID, lkp$new_ID)
-    final_patches <- terra::cover(top_patches, marg_patch)
-
-    terra::writeRaster(final_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
-  } else {
-    terra::writeRaster(top_patches, paste0('../results/allOccupiedPatches/', x, '.tif'))
-  }
-}
-
 # lentiformis  NTIF
@@
 
-```{r create a polygon dataset, eval = F}
+```{r create polygon datasets of patches, eval = F}
 
@@
 
+lapply(f, tif2shp)
+
+f <- paste0('../results/fine_patches/', 
+        list.files('../results/fine_patches/'))
+
+tif2shp <- function(x){
+
+  taxon <- gsub('.tif$', '',  basename(x))
+  pg <- terra::as.polygons(rast(x)) %>% 
+    sf::st_as_sf() %>% 
+    sf::st_make_valid()
+
+  sf::st_write(pg, paste0('../results/fine_patchesVector/', taxon, '.shp'), quiet = TRUE)
+
+}
 lapply(f, tif2shp)

### 2024-02-13 scripts/functions.R

@@ -649,3 +649,85 @@ patchTaggR <- function(x){
   
   terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
 }
+
+
+combine_pop_patches <- function(x){
+  
+  pIDs <- read.csv(paste0('../results/occupiedPatches/', x, '-topSuitability.csv'))
+  top_patches <- terra::rast(paste0('../results/fine_patches/', x, '-patchIDS.tif'))
+  highest_no <- minmax(top_patches)[2]
+  top_patches <- terra::subst(top_patches, pIDs$PatchID,  pIDs$PatchID, others = NA)
+  names(top_patches)<- 'patches'
+  pts <- st_read(paste0('../data/raw/occurrence/combined_records/', x, '.shp'), quiet = TRUE) |>
+    dplyr::filter(Occurrence == 1)
+  marg_patch <- terra::rast(paste0('../results/marginal_habitat/', x, '.tif'))
+  
+  fsize <- file.info(paste0('../results/marginal_habitat/', x, '.tif'))$size/1e9
+  if(fsize < 0.025){marg_patch <- terra::patches(marg_patch, directions = 4)} else
+    if(fsize > 0.025){marge1 <- terra::patches(terra::aggregate(marg_patch, 2, fun = 'mean'), directions = 4)} else
+      if(fsize > 0.04){marge1 <- terra::patches(terra::aggregate(marg_patch, 3, fun = 'mean'), directions = 4)} else
+        if(fsize > 0.05){marge1 <- terra::patches(terra::aggregate(marg_patch, 4, fun = 'mean'), directions = 4)} else
+          if(fsize > 0.06){marge1 <- terra::patches(terra::aggregate(marg_patch, 5, fun = 'mean'), directions = 4)} else
+            if(fsize > 0.07){marge1 <- terra::patches(terra::aggregate(marg_patch, 6, fun = 'mean'), directions = 4)}
+  if(exists('marge1')){marg_patch <- terra::resample(marge1, marg_patch, threads = 16, method = 'near')} 
+  
+  occ_pts <- terra::extract(marg_patch, terra::vect(pts), bind = TRUE) |>
+    sf::st_as_sf()
+  occ_poly <- dplyr::filter(occ_pts, is.na(patches)) |>
+    sf::st_buffer(250) # we cut out 90, but let's go a bit further for corner cases.
[…]

### 2024-02-13 scripts/classify_patches.Rmd
fn to rankpatches

@@
 
-lapply(f[c(288:335)], combine_pop_patches) # 2:46
+lapply(f, combine_pop_patches) # 2:46
 ```
@@
 }
+
 lapply(f, tif2shp)
+
+```
+
+detect neighboring patches
+
+```{r detect neighboring patches up to 5 orders}
+
+f <- list.files('../results/allOccupiedPatchesVector', pattern = 'shp$')
+
+#' identify contiguous neighbors and count them. 
+#' @param x 
+neigh_type <- function(x){
+
+  # assemble the data sets. 
+  taxon <- gsub('.shp$', '', basename(x))
+  pm <- read.csv(paste0('../results/patch_metrics/', taxon, '-patch_vals.csv')) %>% 
+    dplyr::filter(metric == 'cai' & value > 0) %>% 
+    dplyr::select(id)
+
+  occ <- sf::st_read(paste0('../results/allOccupiedPatchesVector/', taxon, '.shp'), quiet = T) |> 
+    dplyr::mutate(Presence = 'Known')
+  pred <- sf::st_read(paste0('../results/fine_patchesVector/', taxon, '-patchIDS.shp'), quiet = T) |> 
+    dplyr::mutate(Presence = 'Predicted') |> 
+    dplyr::filter(lyr_1 %in% pm$id) # filter for core area indexes
+
+  pred1 <- pred[ lengths(sf::st_covers(pred, occ)) == 0, ] # make this dataset distinct. 
+  pred2 <- sf::st_buffer(pred1, dist = 80) # for up north. 
+  pred2 <- sf::st_difference(pred2)
+
+  occ <- sf::st_buffer(occ, dist = 80) # for up north. 
+  occ <- sf::st_difference(occ)
+
+  patches <- dplyr::bind_rows(occ, pred2) %>% 
+    sf::st_difference()
+
+  # identify contiguous neighbors 
+  nblags <- spdep::nblag(
+    neighbours = spdep::poly2nb(patches, queen = TRUE), maxlag = 5)
+
+  # rows 1: here are presences. 
+  known_r <- nrow(patches[patches$Presence=='Known',])
+
+  # count the contiguous neighbors
+  nblags <- rapply(nblags, function(x){x <- length(as.numeric(x[x <= known_r & x != 0]))}, how = 'replace')
+  nblags <- data.table::rbindlist(nblags) %>% 
+    t() %>% 
+    as.data.frame() %>% 
+    setNames(., c('FirstOrder', 'SecondOrder', 'ThirdOrder', 'FourthOrder', 'FifthOrder')) 
+  rownames(nblags) <- 1:nrow(nblags)
+
+  # count the proximal neighbors
+  nbdist <- spdep::dnearneigh(x = st_centroid(patches), d1 = 0, d2 = 5000) # neighbors within 5k
+  nbdist <- lapply(nbdist, function(x){x <- length(as.numeric(x[x <= known_r & x != 0]))})
+
+  nbdist <- setNames(
+    data.frame(as.matrix(nbdist)), 'OccupiedNeighborsIn5k')
+  nbdist$OccupiedNeighborsIn5k <- as.numeric(nbdist$OccupiedNeighborsIn5k)
+
+  dat <- cbind(nbdist, nblags) %>% 
+    tibble::rownames_to_column('ROW') %>% 
+    tidyr::pivot_longer(FirstOrder:FifthOrder, names_to = 'OrderConnection', values_to = 'NoConnections') %>% 
+    dplyr::filter(ROW %in% 1:known_r | OccupiedNeighborsIn5k > 0 | NoConnections > 0)
+
+  ranks <- dplyr::mutate(dat, 
+         Rank = dplyr::case_when(
+           OrderConnection == 'FirstOrder' & NoConnections >= 2 ~ 2, 
+           OrderConnection == 'FirstOrder' & NoConnections == 1 ~ 3, 
+           OrderConnection == 'SecondOrder' & NoConnections >= 3 ~ 4, 
+           OrderConnection == 'SecondOrder' & NoConnections <= 2 ~ 5, 
+           OrderConnection == 'ThirdOrder' & NoConnections >= 4 ~ 6, 
+           OrderConnection == 'ThirdOrder' & NoConnections <= 3 ~ 7, 
+           OccupiedNeighborsIn5k >= 3 ~ 5, 
+           OccupiedNeighborsIn5k <= 2 ~ 6, 
+           .default = as.numeric(7)
+         )) |>
+    dplyr::slice_min(Rank, n = 1, by = 'ROW') |>
+    dplyr::select(ROW, Rank) |>
+    dplyr::mutate(Rank = dplyr::if_else(ROW %in% 1:known_r, 1, Rank))
+
+  x1 <- patches |>
+    tidyr::unite('PatchIDCombined', c('patches', 'lyr_1'), na.rm = T) |>
+    tibble::rownames_to_column('ROW') |>
+    dplyr::right_join( ranks, by = 'ROW') |>
+    dplyr::select(-ROW, -Presence, PchIDComb = PatchIDCombined)
+
+  write.csv(dat, paste0('../results/connections/', taxon, '.csv'), row.names = F)
+  sf::st_write(x1, paste0('../results/rankedPatches/', taxon, '.shp'), quiet = T, append = F)
+
+}
+
+lapply(f, neigh_type)
 

### 2024-02-14 scripts/functions.R
blm ownership simplified

@@ -731,3 +731,87 @@ landscapR <- function(x, pout, pRout){
   )
   
 }
+
+
+
+
+#' identify contiguous neighbors and count them. 
+#' @param x 
+neigh_type <- function(x){
+  
+  # assemble the data sets. 
+  taxon <- gsub('.shp$', '', basename(x))
+  pm <- read.csv(paste0('../results/patch_metrics/', taxon, '-patch_vals.csv')) %>% 
+    dplyr::filter(metric == 'cai' & value > 0) %>% 
+    dplyr::select(id)
+  
+  occ <- sf::st_read(paste0('../results/allOccupiedPatchesVector/', taxon, '.shp'), quiet = T) |> 
+    dplyr::mutate(Presence = 'Known')
+  pred <- sf::st_read(paste0('../results/fine_patchesVector/', taxon, '-patchIDS.shp'), quiet = T) |> 
+    dplyr::mutate(Presence = 'Predicted') |> 
+    dplyr::filter(lyr_1 %in% pm$id) # filter for core area indexes
+  
+  pred1 <- pred[ lengths(sf::st_covers(pred, occ)) == 0, ] # make this dataset distinct. 
+  pred2 <- sf::st_buffer(pred1, dist = 80) # for up north. 
+  pred2 <- sf::st_difference(pred2)
+  
+  occ <- sf::st_buffer(occ, dist = 80) # for up north. 
+  occ <- sf::st_difference(occ)
+  
+  patches <- dplyr::bind_rows(occ, pred2) %>% 
+    sf::st_difference()
+  
+  # identify contiguous neighbors 
+  nblags <- spdep::nblag(
+    neighbours = spdep::poly2nb(patches, queen = TRUE), maxlag = 5)
+  
+  # rows 1: here are presences. 
+  known_r <- nrow(patches[patches$Presence=='Known',])
+  
+  # count the contiguous neighbors
+  nblags <- rapply(nblags, function(x){x <- length(as.numeric(x[x <= known_r & x != 0]))}, how = 'replace')
+  nblags <- data.table::rbindlist(nblags) %>% 
+    t() %>% 
+    as.data.frame() %>% 
+    setNames(., c('FirstOrder', 'SecondOrder', 'ThirdOrder', 'FourthOrder', 'FifthOrder')) 
+  rownames(nblags) <- 1:nrow(nblags)
+  
+  # count the proximal neighbors
[…]

### 2024-02-14 scripts/make_Impermeable_raster.R

@@ -90,3 +90,50 @@ terra::rasterize(vect(hu12), rast(slope), update = TRUE, field = 'huc12',
 
 terra::rasterize(vect(hu12_line), rast(slope), update = TRUE, field = 'huc12', 
                  filename = '../../Geospatial_data/WBD_rast/hu12R-line.tif', overwrite = T)
+
+
+################################################################################
+### Raster of areas plants cannot grow
+r <- rast('../data/raw/nlcd/nlcd_2021_land_cover_l48_20230630.img')
+domain <- terra::project(domain, crs(r))
+r <- terra::crop(r, domain)
+ob <- terra::cats(r)
+
+will_mask <- c(11, 12, 21:24, 81, 82) 
+will_mask <- data.frame(ob[[1]]) %>% 
+  filter(value %in% will_mask) %>% 
+  pull(NLCD.Land.Cover.Class) 
+
+msk <- ifel(r %in% will_mask, 1, NA) 
+r_sub <- terra::aggregate(msk, factor = 3, fun = 'max', 
+                          filename = '../data/processed/nlcd.tif', overwrite = T) 
+
+################################################################################
+### Vector data of BLM surface administration
+
+p2gdb <- '/media/steppe/hdd/Geospatial_data/Public_land_ownership/PADUS3_0Geodatabase/PAD_US3_0.gdb'
+st_layers(p2gdb)
+
+ensure_multipolygons <- function(X) {
+  tmp1 <- tempfile(fileext = ".gpkg")
+  tmp2 <- tempfile(fileext = ".gpkg")
+  st_write(X, tmp1)
+  gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
+  Y <- st_read(tmp2)
+  st_sf(st_drop_geometry(X), geom = st_geometry(Y))
+}
+
+ownership <- st_read(p2gdb, 'PADUS3_0Fee', quiet = TRUE) %>% 
+  filter(Mang_Name == 'BLM' & State_Nm != 'AK') %>% 
+  select(Unit_Nm, Loc_Nm)
+ownership <- ensure_multipolygons(ownership) %>% 
+  st_make_valid()
+
+ownership <- st_crop(ownership, project(domain, crs(ownership)))
+own_simp <- ms_simplify(ownership, keep = 0.1, weighting = 0.6, keep_shapes = TRUE, sys = TRUE)
+
[…]

### 2024-02-15 scripts/classify_patches.Rmd
blm ownership simplified
working towards splitting data for crews

--- 2024-02-13 scripts/classify_patches.Rmd
@@
 }
+
 lapply(f, tif2shp)
 
+```
+
+detect neighboring patches
+
+```{r detect neighboring patches up to 5 orders, eval = F}
+
+f <- list.files('../results/allOccupiedPatchesVector', pattern = 'shp$')
+
+lapply(f, neigh_type)
+
+```
+
+These previous patches refer to pre-Colombian connectivity. We will reduce them to BLM surface areas.
+
+```{r crop patches to BLM, eval = F}
+blm_admin_surf <- st_read('/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp') %>% 
+  select(geometry)
+f <- paste0('../results/rankedPatches/', list.files('../results/rankedPatches/', pattern = '.shp$'))
+
+crop2blm <- function(x){
+
+  taxon <- gsub('.shp$', '', basename(x))
+  patches <- sf::st_read(x, quiet = TRUE) |> 
+    sf::st_make_valid()
+  blm_patches <- sf::st_intersection(patches, blm_admin_surf)
+  blm_patches <- sf::st_make_valid(blm_patches) |>
+    sf::st_collection_extract(type = c("POLYGON"))
+
+  blm_patches <- blm_patches[!sf::st_is_empty(blm_patches),]
+  sf::st_write(blm_patches, paste0('../results/PatchesClippedBLM/', taxon, '.shp'), quiet = TRUE)
+}
+
+lapply(f, crop2blm)
 ```

### 2024-02-15 scripts/subsetDataforCrews.Rmd
working towards splitting data for crews

---
title: "Download data"
author: "steppe"
date: "2023-10-14"
output: html_document
---

```{r load packages}
library(tidyverse)
library(terra)
library(sf)
source('functions.R')
```

Identify what areas crews require data for.

```{r Crew Info}

blm <- vect(
  '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp')

crew_locs <- read.csv('../data/raw/crew_locations.csv') %>% 
  mutate(across(.cols = everything(), ~ na_if(., ''))) %>% 
  pivot_longer(cols = FO1:FO5, names_to = 'FO', values_to = 'FieldOffice' ) %>% 
  drop_na() %>% 
  select(-FO)

teams <- filter(crew_locs, !str_detect(Contract, 'Senior')) %>% 
  mutate(FieldOffice = paste(FieldOffice, 'Field Office'))

seniors <- filter(crew_locs, str_detect(Contract, 'Senior')) %>% 
  rename(Senior = Contract, Contract = FieldOffice) %>% 
  left_join(., teams, by = 'Contract')

rm(crew_locs)
```

Identify which species crews require.

```{r}

spp_needed <- read.csv('../data/raw/2024sos_species.csv') %>% 
  filter(contract == 'SNDO') %>% 
  unite('binomial', genus:species) %>% 
  select(binomial, contract)

```

```{r import data}

p <- '/media/steppe/hdd/Geospatial_data/'

blm_surf <- st_read(paste0(p, 'Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp'), quiet = TRUE)
invasives <- rast(paste0(p, 'MRLC/', '2023_InvGrassCover.tif'))
change_intensity <- rast(paste0(p, 'MRLC/total_change_intensity_index/', 'TCII_3arc.tif'))
fire_perimeters <- st_read(
  paste0(p, 'NIFC_perimeters/', 'InterAgencyFirePerimeterHistory_All_Years_View-Simplified.shp'))

```

```{r}

#' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
#' @param blm_surf sf data set of surface management as multipolygon at most
#' @param fs_surf sf data set of surface USFS land management as multipolygon at most
#' @param fire sf data set of historic fires as multipopylgon at most
#' @param invasive sf dataset of invasive species relative cover estimates
#' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
#' @param roads sf dataset of roads with relevant attritbutes
#' @param seed_transfer sf dataset of seed transfer zones 
#' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
#' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month

```

### 2024-02-15 scripts/AssortedRasters.R

@@ -1,0 +1,61 @@
+setwd('/media/steppe/hdd/SDM_restorations/scripts')
+
+library(terra)
+library(sf)
+
+mrlc <- "/media/steppe/hdd/Geospatial_data/MRLC/"
+blm <- vect('/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp')
+list.files(mrlc, recursive = T)
+
+change <- rast(paste0(mrlc, 'total_change_intensity_index/rcmap_total_change_intensity_index.tif'))
+terra::aggregate(change, fact = 5, fun = 'mean', cores = parallel::detectCores(), overwrite = T, 
+                 filename = paste0(mrlc, 'total_change_intensity_index/TCII_3arc.tif'))
+change <- rast(paste0(mrlc, 'total_change_intensity_index/TCII_3arc.tif'))
+change <- mask(change, blm)
+writeRaster(change, filename = paste0(mrlc, 'total_change_intensity_index/TCII_3arc.tif'), overwrite = T)
+
+rm(change)
+
+may_inv <- rast(
+  paste0(mrlc, 'ExoticAnnualGrass_May152023_PercentCover/ExoticAnnualGrass_May152023_PercentCover.img'))
+may_inv <- ifel(may_inv > 100, NA, may_inv)
+
+june_inv <- rast(
+  paste0(mrlc, 'ExoticAnnualGrass_June192023_PercentCover/ExoticAnnualGrass_June192023_PercentCover.img'))
+june_inv <- ifel(june_inv > 100, NA, june_inv)
+
+plot(june_inv)
+
+inv_grass <- c(may_inv, june_inv)
+inv_grass$twentyThreeMean <- app(inv_grass, max, na.rm = F)
+twentyThreeMean <- inv_grass$twentyThreeMean
+
+terra::aggregate(twentyThreeMean, fact = 5, fun = 'mean', cores = parallel::detectCores(), overwrite = T, 
+                 filename = paste0(mrlc, '2023_InvGrassCover.tif'))
+
+inv_g <- rast(paste0(mrlc, '2023_InvGrassCover.tif'))
+writeRaster(inv_g, filename = paste0(mrlc, '2023_InvGrassCover.tif'), overwrite = T)
+
+inv_g <- mask(inv_g, blm)
+
+rm(june_inv, may_inv, inv_grass, twentyThreeMean, mrlc)
+
+###############################################
[…]

### 2024-02-15 scripts/functions.R

@@ -815,3 +815,144 @@ neigh_type <- function(x){
   sf::st_write(x1, paste0('../results/rankedPatches/', taxon, '.shp'), quiet = T, append = F)
   
 }
+
+
+############################################################################
+# write out portions of 
+
+#' Write out all spatial data for crews to a directory
+#'
+#' @param project_areas split sf data set, with each crews field office(s), and name
+#' @param target_species a dataframe containing possible target species
+#' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
+#' @param blm_surf sf data set of surface management as multipolygon at most
+#' @param fs_surf sf data set of surface USFS land management as multipolygon at most
+#' @param fire sf data set of historic fires as multipopylgon at most
+#' @param invasive sf dataset of invasive species relative cover estimates
+#' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
+#' @param roads sf dataset of roads with relevant attritbutes
+#' @param seed_transfer sf dataset of seed transfer zones 
+#' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
+#' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
+project_maker <- function(x, target_species,
+                          blm_surf, fs_surf, # ownership stuff
+                          fire,  invasive,
+                          sdm_stack,  occurrences, historic_SOS, 
+                          roads, seed_transfer, 
+                          drought6, drought12){
+  
+  defaultW <- getOption("warn")
+  options(warn = -1)
+  
+  #### Initiate Project Directories ####
+  
+  # create directory to hold all contents
+  x_df <- st_drop_geometry(x)
+  crew_dir <- paste0('../Crews/', x_df['Crew'][[1]][1])
+  
+  ifelse(!dir.exists(file.path('../Crews/')), dir.create(file.path('../Crews/')), FALSE)
+  ifelse(!dir.exists(file.path(crew_dir)), dir.create(file.path(crew_dir)), FALSE)
[…]

### 2024-02-15 scripts/make_Impermeable_raster.R

@@ -131,9 +131,43 @@ ownership <- ensure_multipolygons(ownership) %>%
 
 ownership <- st_crop(ownership, project(domain, crs(ownership)))
 own_simp <- ms_simplify(ownership, keep = 0.1, weighting = 0.6, keep_shapes = TRUE, sys = TRUE)
-
 own_simp <- own_simp[!st_is_empty(own_simp),]
+
+own_simp <- st_make_valid(own_simp) %>% 
+  st_collection_extract(type = c("POLYGON"))
+
 st_write(own_simp, 
          '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp', append = F)
 
-rm(p2gdb, ensure_multipolygons, domain)
+
+
+################################################################################
+### Vector data of USFS surface administration
+
+p2gdb <- '/media/steppe/hdd/Geospatial_data/Public_land_ownership/PADUS3_0Geodatabase/PAD_US3_0.gdb'
+st_layers(p2gdb)
+
+ensure_multipolygons <- function(X) {
+  tmp1 <- tempfile(fileext = ".gpkg")
+  tmp2 <- tempfile(fileext = ".gpkg")
+  st_write(X, tmp1)
+  gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
+  Y <- st_read(tmp2)
+  st_sf(st_drop_geometry(X), geom = st_geometry(Y))
+}
+
+ownership <- st_read(p2gdb, 'PADUS3_0Fee', quiet = TRUE) %>% 
+  filter(Mang_Name == 'USFS' & State_Nm != 'AK') %>% 
+  select(Unit_Nm, Loc_Nm)
+ownership <- ensure_multipolygons(ownership) %>% 
+  st_make_valid()
+
+ownership <- st_crop(ownership, project(domain, crs(ownership)))
+own_simp <- rmapshaper::ms_simplify(ownership, keep = 0.1, weighting = 0.6, keep_shapes = TRUE, sys = TRUE)
+own_simp <- own_simp[!st_is_empty(own_simp),]
+
+own_simp <- st_make_valid(own_simp) %>% 
+  st_collection_extract(type = c("POLYGON"))
+
+st_write(own_simp, 
+         '/media/steppe/hdd/Geospatial_data/Public_land_ownership/USFS_West_simplified/USFS_West_simplified.shp', append = F)

### 2024-02-16 scripts/subsetDataforCrews.Rmd
made roads

@@
 
-blm <- vect(
-  '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp')
+blm <- st_read( quiet = T,
+  '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp') %>% 
+  st_transform(5070)
 
@@
 
+blm <- filter(blm, Unit_Nm %in% seniors$FieldOffice)
+
 rm(crew_locs)
+```
+
+Gather together roads data
+```{r Gather Roads data}
+library(tigris)
+
+states <- states() %>% 
+  filter(STUSPS %in% c('NV', 'UT', 'CA', 'AZ', 'CO')) %>% 
+  select(STUSPS, ST_NAME = NAME, STATEFP) %>% 
+  st_drop_geometry()
+
+county <- counties(states$STUSPS) %>% 
+  st_transform(5070)
+county <- county[ lengths( st_intersects(county, blm)) > 0, ]
+
+county_rds <- roads(county$STATEFP[1], county$COUNTYFP[1]) %>% 
+  st_transform(st_crs(blm))
+
+p2 <- '/media/steppe/hdd/Geospatial_data/blmRoadsSOS24/raw/'
+
+road_setter <- function(x){
+
+  statefp <- x$STATEFP[1]
+  county_rds <- lapply(X = x$STATEFP, county = x$COUNTYFP, FUN = tigris::roads) %>% 
+    dplyr::bind_rows() %>% 
+    sf::st_transform(5070)
+  blm_roads <- sf::st_intersection(county_rds, blm)
+
+  st_write(blm_roads, paste0(p2, statefp, '.shp'))
+}
+
+rs <- split(county, f = county$STATEFP)
+lapply(rs, road_setter)
+
+roads <- lapply(paste0(p2, list.files(p2, pattern = 'shp$')), sf::st_read, quiet = TRUE) |>
+  dplyr::bind_rows() %>% 
+  select(LINEARID)
+
+roads_simp <- st_simplify(roads, dTolerance = 10)
+object.size(roads_simp)/10^9 # ~55% of original size. 
+st_write(roads_simp, '/media/steppe/hdd/Geospatial_data/blmRoadsSOS24/processed/BLM_roads.shp', append = F)
+
+rm(states, county, p2, roads, roads_simp, rs, road_setter)
 ```
@@
 blm_surf <- st_read(paste0(p, 'Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp'), quiet = TRUE)
+usfs_surf <- st_read(paste0(p, 'Public_land_ownership/USFS_West_simplified/USFS_West_simplified.shp'), quiet = TRUE)
 invasives <- rast(paste0(p, 'MRLC/', '2023_InvGrassCover.tif'))
 change_intensity <- rast(paste0(p, 'MRLC/total_change_intensity_index/', 'TCII_3arc.tif'))
-fire_perimeters <- st_read(
+fire_perimeters <- st_read( quiet = T,
   paste0(p, 'NIFC_perimeters/', 'InterAgencyFirePerimeterHistory_All_Years_View-Simplified.shp'))
+SWSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'desertSW/desert_southwest_provisional_seed_transfer_zones_USGS.shp')) %>% 
+  select(Zone)
+BOWERSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'WWETAC_STZ/GB_2013_revised_reduced.shp')) %>% 
+  select(seed_zone)
+
+```
+
+```{r}
 

### 2024-02-16 scripts/subsetDataforCrews.Rmd
importing crew data

--- 2024-02-15 scripts/subsetDataforCrews.Rmd
@@
 
+blm <- filter(blm, Unit_Nm %in% seniors$FieldOffice)
+
 rm(crew_locs)
+```
+
+```{r jitter the AIM data, eval = F}
+set.seed(125)
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records'
+occurrences <- lapply(
+  file.path(occ_p, list.files(occ_p, pattern = 'shp$')), st_read, quiet = T) |>
+  bind_rows() %>% 
+  filter(Occurrence == 1) %>% 
+  select(taxon, date, PtType)
+occ_aim <- filter(occurrences, PtType == 'AIM Presence') %>% 
+  st_jitter(amount = 250)
+occurrences <- bind_rows(
+  filter(occurrences, PtType != 'AIM Presence'), 
+  occ_aim) %>% 
+  mutate(
+    PtType = if_else(PtType == 'AIM Presence', 'jittered', 'exact'),
+    date = str_extract(date, '[0-9]{4}')
+    ) %>% 
+  rename(year = date, coord_status = PtType) %>% 
+  arrange(taxon, year, coord_status)
+
+table(occurrences$coord_status)
+st_write(occurrences, file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+
+rm(occ_p. occ_aim)
 ```
@@
 blm_surf <- st_read(paste0(p, 'Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp'), quiet = TRUE)
+usfs_surf <- st_read(paste0(p, 'Public_land_ownership/USFS_West_simplified/USFS_West_simplified.shp'), quiet = TRUE)
 invasives <- rast(paste0(p, 'MRLC/', '2023_InvGrassCover.tif'))
 change_intensity <- rast(paste0(p, 'MRLC/total_change_intensity_index/', 'TCII_3arc.tif'))
-fire_perimeters <- st_read(
+fire_perimeters <- st_read( quiet = T,
   paste0(p, 'NIFC_perimeters/', 'InterAgencyFirePerimeterHistory_All_Years_View-Simplified.shp'))
+SWSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'desertSW/desert_southwest_provisional_seed_transfer_zones_USGS.shp')) %>% 
+  select(seed_zone = Zone) %>% 
+  mutate(Dataset = 'SWSTZ', .before = geometry, seed_zone = as.character(seed_zone))
+BOWERSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'WWETAC_STZ/GB_2013_revised_reduced.shp')) %>% 
+  select(seed_zone) %>% 
+  mutate(Dataset = 'BowerRevised', .before = geometry)
+STZ <- bind_rows(BOWERSTZ, SWSTZ)
+roads <- st_read(quiet = T, paste0(p, 'blmRoadsSOS24/processed/BLM_roads.shp'))
+admin_boundaries <- st_read(paste0(p, 'Public_land_ownership/BLM_ADMU/BLM_ADMU_BOUNDARIES.shp'), quiet = TRUE)
+allotments <- st_read('Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp')
+
+patches_p <- '/media/steppe/hdd/SDM_restorations/results/PatchesClippedBLM'
+patches <- lapply(
+  file.path(patches_p, list.files(patches_p, pattern = 'shp$')), st_read, quiet = T) |>
+  bind_rows()
+
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records'
+occurrences <- st_read(file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+
+rm(SWSTZ, BOWERSTZ, occ_p)
+```
+
+```{r}
+
+ggplot() + 
+  geom_sf(data = BOWERSTZ, aes(fill = seed_zone)) +
+  geom_sf(data = SWSTZ, aes(color = Zone))
 
@@
 #' @param seed_transfer sf dataset of seed transfer zones
-#' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
-#' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
+
+lapply(CREWS_LIST, 
+       target_species = spp_needed,
+       blm_surf = blm_surf,
+       fs_surf = usfs_surf, 
+       fire = fire_perimeters,
+       invasive = invasives,
+       sdm_stack, 
+       occurrences,
+      # historic_SOS, 
+       roads = roads, seed_transfer = STZ)
 

### 2024-02-16 scripts/AssortedRasters.R

@@ -1,5 +1,6 @@
 setwd('/media/steppe/hdd/SDM_restorations/scripts')
 
+library(tidyverse)
 library(terra)
 library(sf)
 
@@ -43,19 +44,82 @@ rm(june_inv, may_inv, inv_grass, twentyThreeMean, mrlc)
 ###############################################
 # Create Roads Data set
 
+library(tigris)
 
-domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
-ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
-crs(domain) <- "EPSG:4326"
+states <- states() %>% 
+  filter(STUSPS %in% c('NV', 'UT', 'CA', 'AZ', 'CO')) %>% 
+  select(STUSPS, ST_NAME = NAME, STATEFP) %>% 
+  st_drop_geometry()
 
+county <- counties(states$STUSPS) %>% 
+  st_transform(5070)
+county <- county[ lengths( st_intersects(county, blm)) > 0, ]
 
-p <- '/media/steppe/hdd/Geospatial_data/Transportation_National_GDB/Transportation_National_GDB.gdb'
-roads <- st_read(p, layer = 'Trans_RoadSegment', quiet = TRUE) 
+county_rds <- roads(county$STATEFP[1], county$COUNTYFP[1]) %>% 
+  st_transform(st_crs(blm))
 
-st_crop(roads, domain) # restrict to the west
-# now restrict to BLM 
+p2 <- '/media/steppe/hdd/Geospatial_data/blmRoadsSOS24/raw/'
 
+road_setter <- function(x){
   
-# mask away urban areas. 
+  statefp <- x$STATEFP[1]
+  county_rds <- lapply(X = x$STATEFP, county = x$COUNTYFP, FUN = tigris::roads) %>% 
+    dplyr::bind_rows() %>% 
+    sf::st_transform(5070)
+  blm_roads <- sf::st_intersection(county_rds, blm)
+  
+  st_write(blm_roads, paste0(p2, statefp, '.shp'))
+}
+
+rs <- split(county, f = county$STATEFP)
+lapply(rs, road_setter)
+
+roads <- lapply(paste0(p2, list.files(p2, pattern = 'shp$')), sf::st_read, quiet = TRUE) |>
+  dplyr::bind_rows() %>% 
[…]

### 2024-02-16 scripts/functions.R

@@ -848,7 +848,8 @@ project_maker <- function(x, target_species,
   
   # create directory to hold all contents
   x_df <- st_drop_geometry(x)
-  crew_dir <- paste0('../Crews/', x_df['Crew'][[1]][1])
+  
+  crew_dir <- paste0('/media/steppe/hdd/2024SOS_CrewGeospatial', x_df['Crew'][[1]][1])
   
   ifelse(!dir.exists(file.path('../Crews/')), dir.create(file.path('../Crews/')), FALSE)
   ifelse(!dir.exists(file.path(crew_dir)), dir.create(file.path(crew_dir)), FALSE)
@@ -866,6 +867,7 @@ project_maker <- function(x, target_species,
   dir.create( file.path(crew_dir, 'Geodata/Disturb') ) 
   dir.create( file.path(crew_dir, 'Geodata/Disturb/Fire') ) 
   dir.create( file.path(crew_dir, 'Geodata/Disturb/Invasive') ) 
+  dir.create( file.path(crew_dir, 'Geodata/Disturb/Geodata/Disturb/TotalChangeIntensity') ) 
   
   # target species information
   dir.create( file.path(crew_dir, 'Geodata/Species') ) 
@@ -877,8 +879,8 @@ project_maker <- function(x, target_species,
   dir.create( file.path(crew_dir, 'Geodata/Roads')) 
   # Seed transfer zones
   dir.create( file.path(crew_dir, 'Geodata/STZ')) 
-  # drought 
-  dir.create( file.path(crew_dir, 'Geodata/Drought'))
+#  # drought 
+#  dir.create( file.path(crew_dir, 'Geodata/Drought'))
   
   
   #### Process geographic data to a mask of the field office ####
@@ -917,6 +919,10 @@ project_maker <- function(x, target_species,
   crop(invasives, focal_vect, mask = T, threads = T, filename =
          file.path(crew_dir, 'Geodata/Disturb/Invasive', 'Invasive.tif'))
   
+  # write out the change_intensity
+  crop(total_change, focal_vect, mask = T, threads = T, filename = 
+         file.path(crew_dir, 'Geodata/Disturb/TotalChangeIntensity', 'TCII.tif'))
+  
   # write out species occurrence data
   
   occurrences_sub <- filter(occurrences, species %in% t_spp$Species)
@@ -932,7 +938,7 @@ project_maker <- function(x, target_species,
   
   lapply(occurrences_list, occ_writer)
   
[…]

### 2024-02-16 scripts/subsetDataforCrews.Rmd
working towards writing out projects geodata

--- 2024-02-15 scripts/subsetDataforCrews.Rmd
@@
 
+blm <- filter(blm, Unit_Nm %in% seniors$FieldOffice)
+
 rm(crew_locs)
+```
+
+```{r jitter the AIM data, eval = F}
+set.seed(125)
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records'
+occurrences <- lapply(
+  file.path(occ_p, list.files(occ_p, pattern = 'shp$')), st_read, quiet = T) |>
+  bind_rows() %>% 
+  filter(Occurrence == 1) %>% 
+  select(taxon, date, PtType)
+occ_aim <- filter(occurrences, PtType == 'AIM Presence') %>% 
+  st_jitter(amount = 250)
+occurrences <- bind_rows(
+  filter(occurrences, PtType != 'AIM Presence'), 
+  occ_aim) %>% 
+  mutate(
+    PtType = if_else(PtType == 'AIM Presence', 'jittered', 'exact'),
+    date = str_extract(date, '[0-9]{4}')
+    ) %>% 
+  rename(year = date, coord_status = PtType) %>% 
+  arrange(taxon, year, coord_status)
+
+st_write(occurrences, file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+
+rm(occ_p. occ_aim)
 ```
@@
 spp_needed <- read.csv('../data/raw/2024sos_species.csv') %>% 
-  filter(contract == 'SNDO') %>% 
+  filter(contract %in% c('SNDO', 'EASTERN MOJAVE')) %>% 
   unite('binomial', genus:species) %>% 
@@
 blm_surf <- st_read(paste0(p, 'Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp'), quiet = TRUE)
+usfs_surf <- st_read(paste0(p, 'Public_land_ownership/USFS_West_simplified/USFS_West_simplified.shp'), quiet = TRUE)
 invasives <- rast(paste0(p, 'MRLC/', '2023_InvGrassCover.tif'))
 change_intensity <- rast(paste0(p, 'MRLC/total_change_intensity_index/', 'TCII_3arc.tif'))
-fire_perimeters <- st_read(
+fire_perimeters <- st_read( quiet = T,
   paste0(p, 'NIFC_perimeters/', 'InterAgencyFirePerimeterHistory_All_Years_View-Simplified.shp'))
+SWSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'desertSW/desert_southwest_provisional_seed_transfer_zones_USGS.shp')) %>% 
+  select(seed_zone = Zone) %>% 
+  mutate(Dataset = 'SWSTZ', .before = geometry, seed_zone = as.character(seed_zone))
+BOWERSTZ <- st_read( quiet = T,
+  paste0(p, 'provisional_SeedTransferZones/', 'WWETAC_STZ/GB_2013_revised_reduced.shp')) %>% 
+  select(seed_zone) %>% 
+  mutate(Dataset = 'BowerRevised', .before = geometry)
+STZ <- bind_rows(BOWERSTZ, SWSTZ)
+roads <- st_read(quiet = T, paste0(p, 'blmRoadsSOS24/processed/BLM_roads.shp'))
+admin_boundaries <- st_read(paste0(p, 'Public_land_ownership/BLM_ADMU/BLM_ADMU_BOUNDARIES.shp'), quiet = TRUE)
+allotments <- st_read(
+  file.path(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp'))
 
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records'
+occurrences <- st_read(file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+
+rm(SWSTZ, BOWERSTZ, occ_p)
 ```
@@
 
-#' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
-#' @param blm_surf sf data set of surface management as multipolygon at most
-#' @param fs_surf sf data set of surface USFS land management as multipolygon at most
-#' @param fire sf data set of historic fires as multipopylgon at most
-#' @param invasive sf dataset of invasive species relative cover estimates
-#' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
-#' @param roads sf dataset of roads with relevant attritbutes
-#' @param seed_transfer sf dataset of seed transfer zones 
-#' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
-#' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
+teams_list <- filter(teams, str_detect(Contract, 'SNDO|Eastern Mojave')) %>% 
+  split(., f = .$Contract)
+
+lapply(teams_list, project_maker,
+       target_species = spp_needed,
+       blm_surf = blm_surf,
+       fs_surf = usfs_surf, 
+       fire = fire_perimeters,
+       invasive = invasives,
+       total_change = change_intensity,
+      # sdm_stack, 
+       occurrences = occurrences,
+      # historic_SOS, 
+       roads = roads, seed_transfer = STZ)
 

### 2024-02-16 scripts/AssortedRasters.R

@@ -78,12 +78,12 @@ roads <- lapply(paste0(p2, list.files(p2, pattern = 'shp$')), sf::st_read, quiet
   dplyr::bind_rows() %>% 
   select(LINEARID)
 
-roads_simp <- st_simplify(roads, dTolerance = 10)
+roads_simp <- st_simplify(roads, dTolerance = 10) %>% 
+  st_transform(5070)
 object.size(roads_simp)/10^9 # ~55% of original size. 
 st_write(roads_simp, '/media/steppe/hdd/Geospatial_data/blmRoadsSOS24/processed/BLM_roads.shp', append = F)
 
 rm(states, county, p2, roads, roads_simp, rs, road_setter)
-
 
 
 ## simplify boundaries of allotments
@@ -93,8 +93,10 @@ allottments <- st_read(paste0(p,
   filter(ADMIN_ST != 'AK')
 
 allottments15.5 <- rmapshaper::ms_simplify(allottments, keep = 0.2, weighting = 0.6, keep_shapes = TRUE, sys = TRUE)
-allottments15.5 <- st_make_valid(allottments15.5)
+allottments15.5 <- st_make_valid(allottments15.5) %>% 
+  st_transform(5070)
 st_write(allottments15.5, paste0(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp'), append = F)
+
 #### CREATE boundaries of field offices. 
 
 p <- '/media/steppe/hdd/Geospatial_data/'
@@ -115,9 +117,10 @@ bounds <- bounds %>%
     ADMU_NAME = str_replace(ADMU_NAME, 'PALM SPRINGS/S. COAST FIELD OFFICE', 
                             'PALM SPRINGS/SOUTH COAST FIELD OFFICE'), 
     PARENT_NAME = str_replace(PARENT_NAME, 'DAKS', 'DAKOTAS')
-    )
+    ) %>% 
+  st_transform(5070)
 
-st_write(bounds, paste0(p, 'Public_land_ownership/BLM_ADMU/BLM_ADMU_BOUNDARIES.shp'))
+st_write(bounds, paste0(p, 'Public_land_ownership/BLM_ADMU/BLM_ADMU_BOUNDARIES.shp'), append = F)
 
 ggplot() +
   geom_sf(data = bounds)

### 2024-02-16 scripts/functions.R

@@ -832,14 +832,16 @@ neigh_type <- function(x){
 #' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
 #' @param roads sf dataset of roads with relevant attritbutes
 #' @param seed_transfer sf dataset of seed transfer zones 
+#' @param total_change Total Change Intensity Index from the MRLC
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 project_maker <- function(x, target_species,
                           blm_surf, fs_surf, # ownership stuff
-                          fire,  invasive,
-                          sdm_stack,  occurrences, historic_SOS, 
-                          roads, seed_transfer, 
-                          drought6, drought12){
+                          fire,  invasive,  occurrences, historic_SOS, 
+                          total_change, 
+                          roads, seed_transfer#, 
+                     #     drought6, drought12
+                     ){
   
   defaultW <- getOption("warn")
   options(warn = -1)
@@ -847,9 +849,8 @@ project_maker <- function(x, target_species,
   #### Initiate Project Directories ####
   
   # create directory to hold all contents
-  x_df <- st_drop_geometry(x)
   
-  crew_dir <- paste0('/media/steppe/hdd/2024SOS_CrewGeospatial', x_df['Crew'][[1]][1])
+  crew_dir <- paste0('/media/steppe/hdd/2024SOS_CrewGeospatial/', x['Contract'][[1]][1])
   
   ifelse(!dir.exists(file.path('../Crews/')), dir.create(file.path('../Crews/')), FALSE)
   ifelse(!dir.exists(file.path(crew_dir)), dir.create(file.path(crew_dir)), FALSE)
@@ -861,13 +862,14 @@ project_maker <- function(x, target_species,
   # write out BLM and Forest Service
   dir.create( file.path(crew_dir, 'Geodata/Admin') ) 
   dir.create( file.path(crew_dir, 'Geodata/Admin/Boundaries') ) 
+  dir.create( file.path(crew_dir, 'Geodata/Admin/Allotments'))
[…]

### 2024-02-16 scripts/make_Impermeable_raster.R

@@ -140,6 +140,10 @@ st_write(own_simp,
          '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp', append = F)
 
 
+own_simp <- st_read(
+  '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp') %>% 
+  st_transform(5070)
+
 
 ################################################################################
 ### Vector data of USFS surface administration
@@ -167,7 +171,8 @@ own_simp <- rmapshaper::ms_simplify(ownership, keep = 0.1, weighting = 0.6, keep
 own_simp <- own_simp[!st_is_empty(own_simp),]
 
 own_simp <- st_make_valid(own_simp) %>% 
-  st_collection_extract(type = c("POLYGON"))
+  st_collection_extract(type = c("POLYGON")) %>% 
+  st_transform(5070)
 
 st_write(own_simp, 
          '/media/steppe/hdd/Geospatial_data/Public_land_ownership/USFS_West_simplified/USFS_West_simplified.shp', append = F)

### 2024-02-16 scripts/subsetDataforCrews.Rmd
reworked project directory writing fn from last year SOS_gb

@@
 
+spp_needed <- read.csv('../data/raw/contract_lkp.csv') %>% 
+  left_join(spp_needed, ., by = c('contract' = 'Contract_colloquial')) %>% 
+  select(-contract)
 ```
@@
   mutate(Dataset = 'BowerRevised', .before = geometry)
-STZ <- bind_rows(BOWERSTZ, SWSTZ)
+STZ <- bind_rows(BOWERSTZ, SWSTZ) %>% 
+  st_transform(5070)
 roads <- st_read(quiet = T, paste0(p, 'blmRoadsSOS24/processed/BLM_roads.shp'))
@@
 allotments <- st_read(
-  file.path(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp'))
+  file.path(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp')) %>% 
+  select(ALLOT_NAME)
 
@@
 lapply(teams_list, project_maker,
+       admu = admin_boundaries, 
        target_species = spp_needed, 
@@
        total_change = change_intensity, 
-      # sdm_stack, 
        occurrences = occurrences, 
@@
 
+ob <- st_read('/media/steppe/hdd/2024SOS_CrewGeospatial/Eastern Mojave - P102073/Geodata/Species/SDM/Salazaria_mexicana.shp')
+
 ```

### 2024-02-16 scripts/functions.R

@@ -823,6 +823,7 @@ neigh_type <- function(x){
 #' Write out all spatial data for crews to a directory
 #'
 #' @param project_areas split sf data set, with each crews field office(s), and name
+#' @param admu administrative boundaries
 #' @param target_species a dataframe containing possible target species
 #' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
 #' @param blm_surf sf data set of surface management as multipolygon at most
@@ -836,6 +837,7 @@ neigh_type <- function(x){
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 project_maker <- function(x, target_species,
+                          admu, 
                           blm_surf, fs_surf, # ownership stuff
                           fire,  invasive,  occurrences, historic_SOS, 
                           total_change, 
@@ -860,22 +862,22 @@ project_maker <- function(x, target_species,
          dir.create(file.path(crew_dir, 'Data')), FALSE)
   
   # write out BLM and Forest Service
-  dir.create( file.path(crew_dir, 'Geodata/Admin') ) 
-  dir.create( file.path(crew_dir, 'Geodata/Admin/Boundaries') ) 
-  dir.create( file.path(crew_dir, 'Geodata/Admin/Allotments'))
-  dir.create( file.path(crew_dir, 'Geodata/Admin/Surface') ) # both BLM and Forest Service go in here
+  dir.create( file.path(crew_dir, 'Geodata/Admin'), showWarnings = F) 
+  dir.create( file.path(crew_dir, 'Geodata/Admin/Boundaries'), showWarnings = F ) 
+  dir.create( file.path(crew_dir, 'Geodata/Admin/Allotments'), showWarnings = F)
+  dir.create( file.path(crew_dir, 'Geodata/Admin/Surface'), showWarnings = F ) # both BLM and Forest Service go in here
   
   # fire and invasive species data
-  dir.create( file.path(crew_dir, 'Geodata/Disturb') ) 
[…]

### 2024-02-19 scripts/aggregateMaskMapsforPhen.R
creating surfaces for predicting phenology onto

@@ -1,0 +1,33 @@
+setwd('/media/steppe/hdd/SDM_restorations/scripts')
+
+library(terra)
+source('functions.R')
+
+p <- '../results/suitability_maps'
+f <- file.path(p, list.files(p, pattern = '1k-'))
+
+template <- rast('../data/processed/gddPCA.tif') |>
+  project(crs(rast(f[1])))
+noGrow <- rast('../data/processed/nlcd.tif') |>
+  resample(template) |>
+  project(crs(template))
+
+r1 <- rast(f[1])
+r1 <- terra::ifel(r1 < 0.5, NA, r1)
+ng1 <- terra::crop(noGrow, r1)
+plot(r1)
+
+terra::mask(r1, ng1, maskvalues = 1)
+
+phenAggregator <- function(x){
+  
+  taxon <- basename(x)
+  r <- terra::rast(x)
+  r <- terra::ifel(r < 0.5, NA, r)
+  r <- terra::resample(r, template, threads = parallel::detectCores())
+  terra::mask(r, terra::crop(noGrow, terra::ext(r)), inverse = T, 
+                   filename = file.path('../results/PhenPredSurfaces', taxon))
+  
+}
+
+lapply(f, phenAggregator)

### 2024-02-20 scripts/subsetDataforCrews.Rmd
mapping

@@
     PtType = if_else(PtType == 'AIM Presence', 'jittered', 'exact'),
-    date = str_extract(date, '[0-9]{4}')
+    date = as.numeric(str_extract(date, '[0-9]{4}'))
     ) %>% 
@@
 
-st_write(occurrences, file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+st_write(occurrences, file.path(occ_p, 'jittered_occurrences.shp'), quiet = T, append = F)
 
@@
 
-occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records'
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/jittered_occurrences'
 occurrences <- st_read(file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)

### 2024-02-20 scripts/aggregateMaskMapsforPhen.R

@@ -12,22 +12,16 @@ noGrow <- rast('../data/processed/nlcd.tif') |>
   resample(template) |>
   project(crs(template))
 
-r1 <- rast(f[1])
-r1 <- terra::ifel(r1 < 0.5, NA, r1)
-ng1 <- terra::crop(noGrow, r1)
-plot(r1)
-
-terra::mask(r1, ng1, maskvalues = 1)
-
 phenAggregator <- function(x){
   
-  taxon <- basename(x)
+  taxon <- basename(gsub('1k-.*$', '', x))
   r <- terra::rast(x)
-  r <- terra::ifel(r < 0.5, NA, r)
+  r <- terra::ifel(r < 0.5, NA, 1)
   r <- terra::resample(r, template, threads = parallel::detectCores())
-  terra::mask(r, terra::crop(noGrow, terra::ext(r)), inverse = T, 
-                   filename = file.path('../results/PhenPredSurfaces', taxon))
+  terra::mask(r, terra::crop(noGrow, terra::ext(r)), inverse = T, overwrite = T, 
+                   filename = file.path('../results/PhenPredSurfaces', paste0(taxon, '.tif')))
   
+  terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
 }
 
 lapply(f, phenAggregator)

### 2024-02-20 scripts/functions.R

@@ -960,7 +960,8 @@ project_maker <- function(x, target_species,
       sf::st_as_sf()
     f_crop <- st_intersection(f_crop, blm_surf_sub) %>% 
       select(-Unit_Nm,-Loc_Nm)
-    
+    f_crop <- f_crop[st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
+    f_crop <- rmapshaper::ms_simplify(f_crop)
     f_crop <- f_crop[st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
     
     binomial <- paste0(gsub(' ', '_', basename(x)))

### 2024-02-20 scripts/makeStaticMaps.R

@@ -1,0 +1,167 @@
+library(tidyverse)
+library(sf)
+library(lubridate)
+library(terra)
+library(tigris)
+library(ggthemes)
+
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.0, -100, 27, 50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+domain <- project(domain, "EPSG:5070") |>
+  as.polygons() |>
+  st_as_sf()
+#domain <- st_bbox(domain)
+
+setwd('/media/steppe/hdd/SDM_restorations/results')
+p <- '/media/steppe/hdd/SDM_restorations/results/suitability_maps'
+f <- file.path(p, list.files(p))
+f <- f[grep('1k-2024', f)]
+
+st <- tigris::states() %>% 
+  filter(REGION == '4' | STUSPS %in% c('SD', 'ND', 'OK', 'NE', 'TX')) %>% 
+  filter(! STUSPS %in% c('AK', 'HI')) %>% 
+  st_transform(5070) %>% 
+  st_intersection(., domain)
+
+ocean <- rnaturalearth::ne_download(scale = 10, type = "ocean", category = "physical") %>% 
+  st_as_sf() %>% 
+  st_transform(5070) %>% 
+  st_intersection(., domain)
+
+occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/jittered_occurrences'
+occurrences <- st_read(file.path(occ_p, 'jittered_occurrences.shp'), quiet = T)
+
+dem <- mosaic(
+  sprc(
+    lapply(file.path('/media/steppe/hdd/Geospatial_data/GMTED',
+                     list.files('/media/steppe/hdd/Geospatial_data/GMTED',
+                                recursive = T, pattern = 'mea300.tif'))
+           , rast))
+  ) |> 
+  project(crs(domain)) 
+
+dem <- crop(dem, domain)
+dem <- mask(dem, vect(ocean), inverse = T)
+slope <- terrain(dem, "slope", unit="radians")
+aspect <- terrain(dem, "aspect", unit="radians")
+hill <- shade(slope, aspect, 30, 315)
+
+rm(slope, aspect)
+
+
+#' plot many same sdm maps
+#' 
+#' @param x a raster stack of species to plot
+#' @param y an sf tibble of the same species occurrence data to plot
+#' @param path_dir directory to save the outfiles
+map_maker_fn <- function(x, y, path_dir){
[…]

### 2024-02-28 scripts/subsetDataforCrews.Rmd
fn to writee out results to tabular

@@
 allotments <- st_read(
-  file.path(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp')) %>% 
+  file.path(p, 'Public_land_ownership/BLM_Natl_Grazing_Allotment_Polygons/Allotment-simp.shp'), quiet = T) %>% 
   select(ALLOT_NAME)
@@
 
-```{r}
+```{r generate projects for crews}
 
@@
 ```
+
+Now we can generate estimates of flowering duration in certain field offices
+
+```{r generate phenology spreadsheet for crews}
+
+p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
+lapply(X = teams_list[1], 
+       FUN = phen_tabulator,
+       path = p2ts, 
+       admu = admin_boundaries, 
+       target_species = spp_needed
+       )
+
+focal_area <- dplyr::filter(admin_boundaries, ADMU_NA == str_to_upper('St. George Field Office' )) |>
+  sf::st_union() |>
+  sf::st_transform(5070) |>
+  vect()
+
+focal_r <- rast(file.path(path, f[[1]]))
+out_v <- terra::extract(focal_r, focal_area, method = 'simple', fun = table, na.rm = T, ID = FALSE)
+pos <- unlist(lapply(out_v, function(x){x1 <- colnames(x)[which.max(x)]})) # most common doy for event
+
+#names(out_v) <- c('Initation', 'Peak', 'Cessation')
+```

### 2024-02-28 scripts/functions.R

@@ -888,7 +888,7 @@ project_maker <- function(x, target_species,
   
   #### Process geographic data to a mask of the field office ####
   
-  focal_bbox <- filter(admin_boundaries, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) %>%  
+  focal_bbox <- filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) %>%  
     st_union() %>% 
     st_transform(5070)
   
@@ -987,4 +987,97 @@ project_maker <- function(x, target_species,
 #  writeRaster(sdm_fo, fnames)
   
   options(warn = defaultW)
+}
+
+############################################################################
+# write out whether a model says a species is flowering or not.
+#' @param x a list od data.frames of crews which need the estimates.
+#' @param path a path to the raster phenology estimate summaries from `spat_summarize` in sagesteppe::SeedPhenology
+#' @param admu 
+#' @param target_species 
+phen_tabulator <- function(x, path, project_areas, admu, target_species){
+  
+  # subset to field offices
+  focal_area <- dplyr::filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) |>
+    sf::st_union() |>
+    sf::st_transform(5070) |>
+    vect()
+  
+  # determine which species we want estimates for.
+  t_spp <- target_species |>
+    dplyr::filter(Requisition == x$Contract[1]) |>
+    dplyr::pull(binomial)
+  
+  # determine which species we have estimates for
+  f <- list.files(path, pattern = '.tif$', recursive = T)
+  f <- f[grep('summary_doys', f)]
+  f_name <- gsub('[.]tif', '', basename(f))
+  f <- file.path(path, na.omit(f[match(t_spp, f_name)]))
+  missing <- data.frame(
+    taxon = setdiff(t_spp, f_name))
+    
+  # basically lapply this through the taxa
+  
+  summarizer <- function(x){
+    focal_r <- terra::rast(x)
+   # focal_r <- terra::crop(focal_r, focal_area, mask = TRUE)
+    out_v <- terra::extract(focal_r, focal_area, method = 'simple', fun = table)
+  }
+  
+  test <- lapply(f[[1]], summarizer)
+  return(test)
+
+  # write out to this location
[…]

### 2024-02-28 scripts/makeStaticMaps.R

@@ -18,15 +18,18 @@ p <- '/media/steppe/hdd/SDM_restorations/results/suitability_maps'
 f <- file.path(p, list.files(p))
 f <- f[grep('1k-2024', f)]
 
+f1 <- rast(f[1])
+st_bbox(f1)
+
 st <- tigris::states() %>% 
   filter(REGION == '4' | STUSPS %in% c('SD', 'ND', 'OK', 'NE', 'TX')) %>% 
   filter(! STUSPS %in% c('AK', 'HI')) %>% 
-  st_transform(5070) %>% 
+  st_transform(st_crs(f1)) %>% 
   st_intersection(., domain)
 
 ocean <- rnaturalearth::ne_download(scale = 10, type = "ocean", category = "physical") %>% 
   st_as_sf() %>% 
-  st_transform(5070) %>% 
+  st_transform(st_crs(f1)) %>% 
   st_intersection(., domain)
 
 occ_p <- '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/jittered_occurrences'
@@ -49,7 +52,7 @@ hill <- shade(slope, aspect, 30, 315)
 
 rm(slope, aspect)
 
-
+st_bbox(st)
 #' plot many same sdm maps
 #' 
 #' @param x a raster stack of species to plot
@@ -67,13 +70,17 @@ map_maker_fn <- function(x, y, path_dir){
   colnames(dem) <- c('x', 'y', 'elevation')
   hillshade <- as.data.frame(hill_sub, xy = T)
   
-  focal_taxon <- gsub('1k.*$', '', gsub('_', ' ', basename(f[1])))
+  focal_taxon <- gsub('1k.*$', '', gsub('_', ' ', basename(x)))
 
-  occurrence <- filter(y, taxon == focal_taxon)
+  occurrence <- filter(y, taxon == focal_taxon) %>% 
+    st_crop(st_bbox(spp))
   pred <- as.data.frame(spp, xy= T)
   
-  ggplot() +
-    geom_sf(data = ocean, fill = 'blue') +
+  st_sub <- st_crop(st, st_bbox(spp))
+  ocean_sub <- st_crop(ocean, st_bbox(spp))
+  
+  ggplot() + #3C6997
+    geom_sf(data = ocean_sub, fill = '#3C6997') + # #97DFFC
     geom_tile(data = hillshade, aes(x = x, y = y, fill = hillshade))  +
     scale_fill_gradient(low = "grey50", high = "grey100") +
     guides(fill = 'none') +
[…]

### 2024-02-28 scripts/subsetDataforCrews.Rmd
fn to write out summaries

@@
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-lapply(X = teams_list[1], 
+oppy <- lapply(X = teams_list[1], 
        FUN = phen_tabulator,
@@
 
-focal_area <- dplyr::filter(admin_boundaries, ADMU_NA == str_to_upper('St. George Field Office' )) |>
-  sf::st_union() |>
-  sf::st_transform(5070) |>
-  vect()
+tdf <- oppy[[1]] |>
+  dplyr::arrange(taxon)
 
-focal_r <- rast(file.path(path, f[[1]]))
-out_v <- terra::extract(focal_r, focal_area, method = 'simple', fun = table, na.rm = T, ID = FALSE)
-pos <- unlist(lapply(out_v, function(x){x1 <- colnames(x)[which.max(x)]})) # most common doy for event
+tdf <- dplyr::left_join(tdf, dates[[1]], by = 'doy')
+
+dates <- date_maker()
 

### 2024-02-28 scripts/functions.R

@@ -1017,14 +1017,19 @@ phen_tabulator <- function(x, path, project_areas, admu, target_species){
     taxon = setdiff(t_spp, f_name))
     
   # basically lapply this through the taxa
-  
   summarizer <- function(x){
     focal_r <- terra::rast(x)
-   # focal_r <- terra::crop(focal_r, focal_area, mask = TRUE)
-    out_v <- terra::extract(focal_r, focal_area, method = 'simple', fun = table)
+    taxon <- gsub('[.]tif', '', basename(x))
+    out_v <- terra::extract(focal_r, focal_area, method = 'simple', fun = table, na.rm = T, ID = FALSE)
+    pos <- unlist(lapply(out_v, function(x){x1 <- colnames(x)[which.max(x)]})) # most common doy for event
+    summy <- data.frame(
+      taxon = taxon,
+      doy = as.numeric(pos), 
+      event = c('Initiation', 'Peak', 'Cessation'))
+    return(summy)
   }
   
-  test <- lapply(f[[1]], summarizer)
+  test <- dplyr::bind_rows(lapply(f, summarizer))
   return(test)
 
   # write out to this location

### 2024-02-28 scripts/subsetDataforCrews.Rmd
fn to write out when things peak flower

@@
 
-```{r generate projects for crews}
-
+```{r}
 teams_list <- filter(teams, str_detect(Contract, 'SNDO|Eastern Mojave')) %>% 
   split(., f = .$Contract)
+```
+
+```{r generate projects for crews, eval= F}
 
@@
 tdf <- oppy[[1]] |>
-  dplyr::arrange(taxon)
-
-tdf <- dplyr::left_join(tdf, dates[[1]], by = 'doy')
+  dplyr::arrange(taxon) |>
+  dplyr::distinct()
 
 dates <- date_maker()
+tdf1 <- dplyr::left_join(tdf, dates[[1]], by = 'doy') |>
+  dplyr::left_join(
+    dplyr::select(dates[[2]], -doy), by = c('reports_to' = 'reporting_event')) |>
+  dplyr::group_by(taxon) #|>
+  tidyr::pivot_wider(names_from = reports_to, values_from = event, id_cols = 'taxon',
+                     values_fill = NA)
+
+report_pts_rep <- sort(as.numeric(colnames(tdf1)[grep('[0-9]', colnames(tdf1))]))
+miss_wk <- setdiff(min(report_pts_rep):max(report_pts_rep), report_pts_rep) # see if some 
+# weeks are missing by default. 
+if(length(miss_wk)>0){
+
+  empty_cols <- setNames(
+    data.frame(
+      matrix(
+        NA, nrow = nrow(tdf1), ncol = length(miss_wk))), 
+    miss_wk)
+
+  tdf2 <- cbind(tdf1, empty_cols)
+  rm(empty_cols)
+}
+
+tdf2 <- tdf2[, c(
+  colnames(tdf2)[grep('[A-z]', colnames(tdf2))], 
+  sort(as.numeric(colnames(tdf2)[ grep('[0-9]', colnames(tdf2)) ]))) ]
+
+tdf2 <- tdf2[2:nrow(tdf2), ]
+
+tdf_peak <- dplyr::filter(tdf, event == 'Peak') %>% 
+  dplyr::mutate(
+    Early_Peak = reports_to - 1, 
+    Late_Peak = reports_to + 1 )
 

### 2024-02-29 scripts/subsetDataforCrews.Rmd
prototype fn to write out phenology metrics

--- 2024-02-28 scripts/subsetDataforCrews.Rmd
@@
 
-rm(occ_p. occ_aim)
+rm(occ_p, occ_aim)
 ```
@@
 
-spp_needed <- read.csv('../data/raw/2024sos_species.csv') %>% 
-  filter(contract %in% c('SNDO', 'EASTERN MOJAVE')) %>% 
+spp_needed <- read.csv('../data/raw/2024sos_species-FIXED.csv') %>% 
   unite('binomial', genus:species) %>% 
@@
   left_join(spp_needed, ., by = c('contract' = 'Contract_colloquial')) %>% 
-  select(-contract)
+  select(-contract) %>% 
+  distinct()
 ```
@@
 
-```{r generate projects for crews}
+```{r}
+teams_list <- split(teams, f = teams$Contract)
+```
 
-teams_list <- filter(teams, str_detect(Contract, 'SNDO|Eastern Mojave')) %>% 
-  split(., f = .$Contract)
+```{r generate projects for crews, eval= F}
 
@@
 
+spp_needed <- rename(spp_needed, taxon = binomial)
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-oppy <- lapply(X = teams_list[1], 
+lapply(X = teams_list, 
        FUN = phen_tabulator,
@@
 
-tdf <- oppy[[1]] |>
-  dplyr::arrange(taxon)
-
-tdf <- dplyr::left_join(tdf, dates[[1]], by = 'doy')
-
-dates <- date_maker()
-
-#names(out_v) <- c('Initation', 'Peak', 'Cessation')
 ```

### 2024-02-29 scripts/functions.R

@@ -960,14 +960,16 @@ project_maker <- function(x, target_species,
       sf::st_as_sf()
     f_crop <- st_intersection(f_crop, blm_surf_sub) %>% 
       select(-Unit_Nm,-Loc_Nm)
-    f_crop <- f_crop[st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
-    f_crop <- rmapshaper::ms_simplify(f_crop)
-    f_crop <- f_crop[st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
+    f_crop <- f_crop[sf::st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
+  #  f_crop <- f_crop[!sf::st_is_empty(f_crop), ]
+  #  f_crop <- rmapshaper::ms_simplify(f_crop, keep = 0.9)
+  #  f_crop <- f_crop[!sf::st_is_empty(f_crop), ]
+  #  f_crop <- f_crop[st_is(f_crop, c('POLYGON', 'MULTIPOLYGON')),]
     
     binomial <- paste0(gsub(' ', '_', basename(x)))
     if(nrow(f_crop) > 1){
     sf::st_write(f_crop, 
-                 file.path(crew_dir, 'Geodata/Species/SDM', binomial), quiet = TRUE)
+                 file.path(crew_dir, 'Geodata/Species/SDM', binomial), append = F, quiet = TRUE)
     }
   }
   
@@ -991,7 +993,7 @@ project_maker <- function(x, target_species,
 
 ############################################################################
 # write out whether a model says a species is flowering or not.
-#' @param x a list od data.frames of crews which need the estimates.
+#' @param x a list of data.frames of crews which need the estimates.
 #' @param path a path to the raster phenology estimate summaries from `spat_summarize` in sagesteppe::SeedPhenology
 #' @param admu 
 #' @param target_species 
@@ -1006,8 +1008,8 @@ phen_tabulator <- function(x, path, project_areas, admu, target_species){
   # determine which species we want estimates for.
   t_spp <- target_species |>
     dplyr::filter(Requisition == x$Contract[1]) |>
-    dplyr::pull(binomial)
-  
+    dplyr::pull(taxon)
+
   # determine which species we have estimates for
[…]

### 2024-03-01 scripts/subsetDataforCrews.Rmd
fn to wriite out rought flowering phenology

@@
   unite('binomial', genus:species) %>% 
-  select(binomial, contract)
+  select(binomial, contract) %>% 
+  mutate(binomial = str_trim(binomial))
 
@@
 spp_needed <- rename(spp_needed, taxon = binomial)
+
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-lapply(X = teams_list, 
+lapply(
+  X = teams_list, 
   FUN = phen_tabulator,

### 2024-03-01 scripts/functions.R

@@ -1032,19 +1032,22 @@ phen_tabulator <- function(x, path, project_areas, admu, target_species){
   }
   
   test <- dplyr::bind_rows(lapply(f, summarizer))
-  return(test)
+  
   dates <- date_maker()
   # now prepare the date in a format for reporting.
   tdf <- test |>
     dplyr::arrange(taxon) |>
     dplyr::distinct() |>
     dplyr::group_by(taxon) |>
-    dplyr::mutate( # peak cannot be the same as cessation or initation
+    dplyr::mutate( # peak cannot be the same as cessation or initiation
       doy = dplyr::if_else( 
-        dplyr::lead(doy, n = 1) == doy, doy-28, doy, missing = doy ), 
+        dplyr::lead(doy, n = 1) <= (doy - 27), doy-28, doy, missing = doy ), 
       doy = dplyr::if_else( 
-        dplyr::lag(doy, n = 1) == doy, doy+28, doy, missing = doy ))
-  
+        dplyr::lag(doy, n = 1) >= (doy - 27), doy+28, doy, missing = doy ), 
+      doy = dplyr::if_else(
+        (le = dplyr::lead(doy, n = 1)) - (la =dplyr::lag(doy, n = 1)) <= 57, ((le+la)/2), doy, missing = doy)
+      )
+
   dates <- date_maker()
   tdf <- dplyr::left_join(tdf, dates[[1]], by = 'doy') |>
     dplyr::left_join(
@@ -1071,13 +1074,15 @@ phen_tabulator <- function(x, path, project_areas, admu, target_species){
   ) |>
     dplyr::arrange(taxon, doy, event) |>
     dplyr::group_by(taxon, reports_to) |>
-    dplyr::mutate(
-      multiples = dplyr::n()) |>
-    dplyr::filter(! ( multiples==2 & str_detect(event, 'Late[.]|Early[.]'))) |>
+    dplyr::mutate(multiples = dplyr::n()) |>
+    dplyr::filter(! ( multiples>=2 & str_detect(event, 'Late[.]|Early[.]'))) |>
+    dplyr::mutate(multiples = dplyr::n()) |>
+    dplyr::filter(! ( multiples>=2 & str_detect(event, 'Peak'))) |>
     dplyr::select(-multiples) |>
     dplyr::arrange(doy) |>
     tidyr::pivot_wider(names_from = reports_to, values_from = event, id_cols = 'taxon',
[…]

### 2024-03-08 scripts/subsetDataforCrews.Rmd
fn for simple phen estimates based on kernel density

@@
 
-blm <- st_read( quiet = T,
+blm <- st_read(quiet = T,
   '/media/steppe/hdd/Geospatial_data/Public_land_ownership/BLM_West_simplified/BLM_West_simplified.shp') %>% 
@@
 
+```{r seed zones, eval = F}
+seed_zones <- read.csv('../data/raw/2024sos_species-FIXED.csv') %>% 
+    unite('binomial', genus:species)
+
+seed_zones %>% 
+  filter(stz == 'BowerRevised') %>% 
+  distinct(ind_stz) %>% 
+  mutate(
+    ind_stz = str_replace(ind_stz, '  ', ' ')
+  ) %>% 
+  separate_wider_delim(ind_stz, delim = ',', names= LETTERS[1:7], too_few = "align_start")
+
+  #  ind_stz = str_replace(ind_stz, ' -[0-9]|[0-9]- |[0-9]-[0-9]', ' - ') )
+```
+
 ```{r import data}
@@
 
-ob <- st_read('/media/steppe/hdd/2024SOS_CrewGeospatial/Eastern Mojave - P102073/Geodata/Species/SDM/Salazaria_mexicana.shp')
-
+table(STZ$Dataset) 
+head(STZ)
 ```
@@
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-lapply(
-  X = teams_list, 
+ott1 <- lapply(
+  X = teams_list[5], # 5 throws an error
   FUN = phen_tabulator,
@@
   target_species = spp_needed
-  )
+  ) # row 67
 
+ott1[[1]]
 ```

### 2024-03-08 scripts/functions.R

@@ -999,11 +999,12 @@ project_maker <- function(x, target_species,
 #' @param target_species 
 phen_tabulator <- function(x, path, project_areas, admu, target_species){
   
+  dates <- date_maker()
   # subset to field offices
-  focal_area <- dplyr::filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) |>
-    sf::st_union() |>
-    sf::st_transform(5070) |>
-    vect()
+  focal_area <- dplyr::filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) |> 
+    sf::st_union() |> 
+    sf::st_transform(5070) |> 
+    vect() 
   
   # determine which species we want estimates for.
   t_spp <- target_species |>
@@ -1018,6 +1019,58 @@ phen_tabulator <- function(x, path, project_areas, admu, target_species){
   missing <- data.frame(
     taxon = setdiff(t_spp, f_name))
   
+  # gather the simple density estimates. 
+
+  if(length(missing) >0){
+    
+  path <- '/media/steppe/hdd/SeedPhenology/data/processed/lowest_priority_sheets'
+  dat <- sf::st_read(file.path(path, 'lowest_priority_sheets.shp'), quiet = TRUE) |>
+    dplyr::filter(scntfcnm %in% gsub('_', ' ', missing$taxon)) |>
+    dplyr::group_by(scntfcnm) |>
+    tidyr::drop_na(doy) |>
+    dplyr::mutate(
+      q95 = quantile(doy, 0.95), # trim the last %5 
+      q25 = quantile(doy, 0.025)) |> # trim only the first 2.5%
+    dplyr::filter(doy > q25 & doy < q95) |>
+    dplyr::select(-q25, -q95) |>
+    sf::st_transform(sf::st_crs(focal_area))
+  
+  dat_sub <- sf::st_buffer(sf::st_as_sf(focal_area), 120000) |> # ~75 miles
+    sf::st_intersection(dat)
+  dat_sub <- split(dat_sub, f = dat_sub$scntfcnm)
+  
+  simple_summarizer <- function(x){
+    if(nrow(x) > 5){d <- density(x$doy)} else {d <- median(x$doy)}
+    if(length(d)==1){mid <- d} else{mid <- d$x[which.max(d$y)]}
+    
+    summy <- data.frame(
+      taxon = x$scntfcnm[1], 
[…]

### 2024-03-08 scripts/makeStaticMaps.R

@@ -117,44 +117,7 @@ map_maker_fn <- function(x, y, path_dir){
   
 }
 
-lapply(f[7], FUN = map_maker_fn, 
+lapply(f, FUN = map_maker_fn, 
        y = occurrences,
        path_dir = '/media/steppe/hdd/SDM_maps_CBG_guidebooks2024'
        )
-
-
-
-breaks <- c(121, 152, 182, 213, 244)
-labels <- c('May', 'June', 'July', 'Aug.', 'Sept.')
-
-spliecies <- split(records, records$species)
-
-phen_maker_fn <- function(x){
-  
-  species <- gsub('_', ' ', x$species[1])
-  
-  ggplot(x) +
-    geom_density(aes(x = DOY), fill = '#FF7B9C', color = '#FFC759', alpha = 0.5) +
-    theme_bw() + 
-    labs(title = 'Estimated Flowering') + 
-    theme(aspect.ratio = 6/16, 
-          plot.title = element_text(
-            hjust = 0.5, colour = "black", size = 7, face = "bold"),
-          axis.text.x= element_text(
-            family = "Tahoma", face = "bold", colour = "black", size=5),
-          panel.background = element_rect(fill='#607196'),
-          plot.background = element_rect(fill='#607196'),
-          panel.border = element_blank(),
-          axis.text.y = element_blank(),
-          axis.ticks.y = element_blank(),
-          axis.title.y = element_blank(),
-          axis.title.x = element_blank(),
-          panel.grid.major.x = element_blank(),
-          panel.grid.major.y = element_blank(),
-          panel.grid.minor.x = element_blank(), 
-          panel.grid.minor.y = element_blank()) +
-    scale_x_continuous(breaks = breaks, labels=labels, limits = c(120, 245) )
-  
-  ggsave(filename = paste0('../results/phen/', species, '.png'), plot = last_plot(), 
-         dpi = 150, width = 260, height = 150,  units = "px",  bg = 'transparent')
-}

### 2024-03-13 scripts/download_dependent_variables.Rmd
playing with burning raster
rerunninng modelling for missed az strip species

@@
 
+## these species missed, fixed on March 13th
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa)) %>% 
+  pull(taxon)
+
+targets <- target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')))
+
 ```
@@
 There is a paper by Maitner with Enquist, Wilson, Merow etc.
-about the need for open source devs to be rewarded academically, I take that has a sign of Brian stepping away from it.
+about the need for open source devs to be rewarded academically, I take that as a sign of Brian stepping away from it.
 We will use a for loop to download from BIEN.
+
 ```{r Perform bien queries, eval = F}
@@
 
+```{r stitch together BIEN queries 3, eval = F}
+
+p <- '../data/raw/occurrence/bien'
+files <- file.path(p, list.files(p))
+files <- files[grep(paste(paste0(gsub(' ', '_', target_taxa), '.csv'), collapse = '|'), basename(files))]
+
+bien_data <-  do.call(rbind, lapply(files, read.csv)) %>% 
+  filter(datasource != 'GBIF' & date_collected > '1950-01-01', ! scrubbed_species_binomial %in% c('Acer negundo', 'Prosopis glandulosa')) %>%
+  drop_na(date_collected, latitude) %>% 
+  select(scrubbed_species_binomial:datasource, observation_type)
+
+write.csv(bien_data, '../data/raw/occurrence/BIEN_records3.csv', row.names = FALSE)
+```
+
 ```{r Perform idigbio queries, eval = F}
@@
 idig <- bind_rows(ob) %>% 
-  filter(datecollected > '1950-01-01')
+  filter(year > 1950)
 
-write.csv(idig, '../data/raw/occurrence/IDIG2.csv')
+write.csv(idig, '../data/raw/occurrence/IDIG3.csv')
 ```
@@
 
-target_symbols <- c('ADFA', 'ADFAF', 'ADFAO', 'BREL', 'BRELE', 'KRGR', 'LOHU2')
-targets <- filter(targets, symbol %in% target_symbols) %>% 
-  distinct()
-
 spp_richness <- st_read(dsn = p1, 
@@
 
-st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records2.shp', append = F)
+spp_richness <- spp_richness[!st_is_empty(spp_richness),,drop=FALSE]
+
+st_write(spp_richness, '../data/raw/occurrence/AIM/AIM_records3.shp', append = F)
 
@@
 
+```{r Create consensus presence records 3, eval = F}
+
+base_p <- '../data/raw/occurrence'
+
+bien <- read.csv(file.path(base_p, 'BIEN_records3.csv')) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) %>% 
+  mutate(date = as.Date(date_collected)) %>% 
+  select(taxon = scrubbed_species_binomial, date)
+
+idig <- read.csv(file.path(base_p, 'IDIG3.csv')) %>% 
+  st_as_sf(coords = c('geopoint.lon', 'geopoint.lat'), crs = 4326) %>% 
+  mutate(date = as.Date(datecollected)) %>% 
+  select(taxon = scientificname, date)
+
+aim <- st_read(file.path(base_p, 'AIM', 'AIM_records3.shp'), quiet = TRUE) %>% 
+  mutate(date = as.Date(DatVstd)) %>% 
+  select(taxon, date, PrmryKy) %>% 
+  st_transform(4326)
+
+gbif3 <- read.delim(file.path(base_p, 'gbif4', 'occurrence.txt'), na.strings = "") %>% 
+  drop_na(decimalLongitude, decimalLatitude) %>% 
+  filter(str_detect(decimalLatitude, '[A-Z]|[a-z]', negate = T)) %>% 
+  st_as_sf(coords = c('decimalLongitude', 'decimalLatitude'), crs = 4326) %>% 
+  mutate(date = as.Date(eventDate), .before = geometry, 
+         coordinateUncertaintyInMeters = as.numeric(coordinateUncertaintyInMeters)) %>% 
+  filter(coordinateUncertaintyInMeters <= 250) %>% 
+  select(taxon = species, date)
+
+records <- bind_rows(aim, bien, gbif3,  idig) %>% 
+  drop_na(taxon) %>% 
+  mutate(taxon = str_to_sentence(taxon), 
+         taxon = if_else(taxon == 'Wyethia sagittata', 'Balsamorhiza sagittata', taxon)) %>%
+  arrange(taxon)
+
+rm(aim, bien, gbif3,  idig)
+
+splicies <- split(records, records$taxon)
+thinned <- lapply(splicies, ONEperCELL)
+thinned <- bind_rows(thinned)
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences3.shp', append = F)
+repeated_occurrences <- lapply(splicies, multiplesPERcell)
+repeated_occurrences <- purrr::discard(repeated_occurrences, ~nrow(.) < 2)
+repeated_occurrences <- bind_rows(repeated_occurrences)
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences3.shp', append = F)
+
+rm(ONEperCELL, multiplesPERcell, records, splicies, repeated_occurrences, base_p)
+```
+
 ```{r remove points outside analytical area, eval = F}
@@
 ```
+
+```{r remove points outside analytical area3, eval = F}
+library(terra)
+domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
+ext(domain) <- c( -125.5, -100, 27, 50) # set the extent
+crs(domain) <- "EPSG:4326" # define the projection
+
+domain <- as.polygons(domain) |>  # convert to vector data
+  st_as_sf() |> # convert to simple feature
+  st_transform(5070)
+
+thinned <- st_read('../data/raw/occurrence/thinned/thinned_occurrences3.shp', quiet = TRUE)
+thinned <- thinned[lengths(st_disjoint(thinned, domain)) ==  0, ]
+st_write(thinned, '../data/raw/occurrence/thinned/thinned_occurrences3.shp', append = FALSE)
+
+repeated_occurrences <- st_read('../data/raw/occurrence/repeated/repeated_occurrences3.shp', quiet = TRUE)
+repeated_occurrences <- thinned[lengths(st_disjoint(repeated_occurrences, domain)) ==  0, ]
+st_write(repeated_occurrences, '../data/raw/occurrence/repeated/repeated_occurrences3.shp', append = FALSE)
+```

### 2024-03-13 scripts/generate_RF_models_copy.Rmd
more rfe subset logic
writing process
rerunninng modelling for missed az strip species

--- 2023-12-31 scripts/generate_RF_models.Rmd
@@
 
-```{r read in response data and model}
+```{r read in response data and model, eval = F}
 p <- '../data/raw/occurrence/combined_records'
@@
 cores <- parallel::detectCores()
-lapply(splicies[3:335], modeller)
+lapply(splicies, modeller)
 
+```
+
+Model the missing AZ strip species
+```{r re run on missing az strip species, eval = F}
+
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
+
+target_taxa1 <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
+
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa1)) %>% 
+  pull(taxon)
+
+p <- '../data/raw/occurrence/combined_records'
+p2f <- file.path(p, list.files(p, pattern = '.shp$'))
+p2f <- p2f[grep(paste(paste0(gsub(' ', '_', target_taxa), '.shp'), collapse = '|'), basename(p2f))]
+
+train_dat <- lapply(p2f, sf::st_read, quiet = TRUE)
+train_dat <- bind_rows(train_dat) %>% 
+  st_transform(crs(layers))
+
+splicies <- split(train_dat, f = train_dat$taxon)
+
+cores <- parallel::detectCores()
+lapply(splicies, modeller) 
 ```

### 2024-03-13 scripts/generate_absences.Rmd
ran all models
rerunninng modelling for missed az strip species

@@
 
-records <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE)
+records <- st_read('../data/raw/occurrence/thinned/thinned_occurrences3.shp', quiet = TRUE)
 records <- records %>% 
@@
 
-st_write(out, '../data/raw/occurrence/absences/absences1km.shp', append = FALSE)
+st_write(out, '../data/raw/occurrence/absences/absences1km3.shp', append = FALSE)
 
@@
 collinear_probs <- read_rds('../results/VIF_PA_drop.rda')
-bio <- usdm::exclude(bio, collinear_probs)
+layers <- usdm::exclude(layers, collinear_probs)
 
-absences <- st_read('../data/raw/occurrence/absences/absences.shp', quiet = TRUE)
-presences <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE) %>% 
+absences <- st_read('../data/raw/occurrence/absences/absences1km3.shp', quiet = TRUE)
+presences <- st_read('../data/raw/occurrence/thinned/thinned_occurrences3.shp', quiet = TRUE) %>% 
   mutate(Occurrence = 1, 
@@
 
+recs <- recs %>% 
+  mutate(
+  taxon = str_replace(taxon, 'Aslcepias subulata', 'Asclepias subulata'),
+  taxon = str_replace(taxon, 'Elymus violaceus', 'Elymus trachycaulus'),
+  taxon = str_replace(taxon, 'Encelia actonii$', 'Encelia actoni'),
+  taxon = str_replace(taxon, 'Hillaria', 'Hilaria')
+  ) %>% 
+  filter(! taxon %in% c('Ephedra antisyphilitica', 'Ephedra fasciculata', 'Ephedra coryi', 'Ephedra cutleri',
+                        'Ephedra funerea', 'Ephedra nevadensis', 'Ephedra torreyana', 'Ephedra californica',
+                        'Ephedra trifurca', 'Ephedra trifurcata', 'Ephedra viridis', 'Ephedra aspera'))
+
 set.seed(27)
@@
 splicies <- split(recs, f = recs$taxon)
-
-lapply(splicies[41:335], lda_PA_dropper, path = '../results/lda2/', col_names = cn)
-
-which(names(splicies) == 'Asclepias speciosa')
+names(splicies)
+lapply(splicies[8:23], lda_PA_dropper, path = '../results/lda3/', col_names = cn)
 

### 2024-03-13 scripts/subsetDataforCrews.Rmd
rerunninng modelling for missed az strip species

@@
 ott1 <- lapply(
-  X = teams_list[5], # 5 throws an error
+  X = teams_list, # 5 throws an error
   FUN = phen_tabulator,
@@
 
-ott1[[1]]
 ```

### 2024-03-13 scripts/functions.R

@@ -1,10 +1,19 @@
 idig_records <- function(query){
   
-  result <- ridigbio::idig_search_records(rq = list(scientificname = query)) |> 
-    dplyr::select(scientificname, datecollected, collector, geopoint.lon, geopoint.lat) |> 
-    tidyr::drop_na(geopoint.lon, geopoint.lat) |> 
-    dplyr::mutate(datecollected = as.Date(datecollected)) |> 
-    dplyr::filter(datecollected > '1950-01-01')
+  
+  fields2getrecord <- c(
+    "data.dwc:scientificName",  "data.dwc:decimalLatitude", 
+    "data.dwc:decimalLongitude", "collector", "uuid",
+    "data.dwc:year", "data.dwc:month", "data.dwc:day")
+  names(fields2getrecord) <- c(
+    'scientificname', 'geopoint.lat', 'geopoint.lon', 'collector', 'uuid', 'year', 'month', 'day')
+  
+  result <- ridigbio::idig_search_records(rq = 
+                                            list(scientificname = query), fields = fields2getrecord) |> 
+    dplyr::rename(all_of(fields2getrecord)) |>
+    dplyr::select(scientificname, year, collector, geopoint.lon, geopoint.lat) |> 
+    tidyr::drop_na(geopoint.lon, geopoint.lat, year) |> 
+    dplyr::filter(year >= 1950)
   
   if(nrow(result) > 0) {result$status = 'found'} else {
     result <- data.frame(
@@ -156,8 +165,6 @@ absence_drawer <- function(x, bg_abs, terrestrial, rand_dist){
 
 
 
-
-
 quantile_flagger <- function(x, quant, flag_threshold){
   
   if(missing(quant)){quant <- 0.025}
@@ -295,7 +302,8 @@ lda_PA_dropper <- function(x, path, col_names){
     removals <- conflicted$ID
   }
   
-  if(!is.na(removals)){data_out <- dplyr::filter(x, ! ID %in% removals)}
+  
+  if(!is.na(removals[1])){data_out <- dplyr::filter(x, ! ID %in% removals)}
   
   metrics <- data.frame(
     Train_Presences = nrow(pres), 
@@ -309,6 +317,7 @@ lda_PA_dropper <- function(x, path, col_names){
     Flagged = nrow(conflicted), 
     Removed = length(removals)
   )
+  
   
   species <- x1$taxon[1]
   message(length(removals), ' records dropped from ', species)

### 2024-03-15 scripts/classify_patches.Rmd
creating surfaces for predicting phenology onto
arunning a second round, accidentally wrote over some original sections

@@
 lapply(f, patcheR) # roughly 72 hours. 
+```
 
+```{r coarse patches 2, eval = F}
+
+p <- '../results/suitability_maps'
+f <- file.path(p, list.files(p, pattern = '1k-'))
+
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
+
+target_taxa <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
+
+## these species missed, fixed on March 13th
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa)) %>% 
+  pull(taxon)
+
+f <- file.path(p, list.files(p, pattern = '1k-'))
+files <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
+
+rm(target_taxa)
+
+gd <- '/media/steppe/hdd/Geospatial_data'
+hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu10R-line.tif') )
+hydr_ftr <-  rast( file.path(gd, 'river_resistance', 'water_ftR.tif') )
+rm(gd)
+
+lapply(files, patcheR) # roughly 5 hours. 
 ```
@@
 
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
+
+target_taxa <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
+
+## these species missed, fixed on March 13th
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa)) %>% 
+  pull(taxon)
+
+files <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
+
 gd <- '/media/steppe/hdd/Geospatial_data'
@@
 
-lapply(f, FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
+lapply(files, FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
 

### 2024-03-15 scripts/predict_models.Rmd
aggregate cells before finding patches
arunning a second round, accidentally wrote over some original sections

@@
 
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
+
+target_taxa <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
+
+## these species missed, fixed on March 13th
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa)) %>% 
+  pull(taxon)
+
 pmodels <- '../results/rf_models'
 f <- file.path(pmodels, list.files(pmodels))
-f <- data.frame(
-  path = f,  
-  species = gsub('-.*$', '', basename(f)), 
-  time = gsub('[A-Za-z]', '', basename(f))
-  ) |>
-  mutate(time = gsub('_-|[.]', '', time),
-         time = gsub('_', ' ', time), 
-         time = trimws(time),
-         time = gsub('^.1-|^1-', '', time), 
-         time = as.POSIXct(time)) |>
- # group_by(species) |>
-  filter(str_detect(species, '1k$')) |>
- # slice_max(order_by = time, n = 1) #|>
-  pull(path)
+files <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
+
+rm(target_taxa)
 
 occpath <- '../data/raw/occurrence/combined_records'
-occ_files <- list.files(occpath, pattern = 'shp$')
+
+occ_files <- paste0( basename(gsub('1k.*', '', files)), '.shp')
 
 models_occ <- data.frame(
-  path2model = f, occurrence_data = occ_files
+  path2model = files, occurrence_data = occ_files
 ) %>% 
@@
 
-lapply(models_occ[68:335], predict_wrapper)
+lapply(models_occ, predict_wrapper)
+```
 
+```{r predict second set of models}
+
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>%
+
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon))
+
+target_taxa <- c('Brodiaea elegans', 'Krameria bicolor', 'Lotus humistratus', 'Adenostoma fasciculatum')
+
+## these species missed, fixed on March 13th
+target_taxa <- read.csv('../data/raw/2024sos_species-FIXED.csv', na.strings = "") %>% 
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) %>% 
+  filter(! taxon %in% c(targets$taxon, target_taxa)) %>% 
+  pull(taxon)
+
+pmodels <- '../results/rf_models'
+f <- file.path(pmodels, list.files(pmodels))
+files <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
+
+rm(target_taxa)
+
+occpath <- '../data/raw/occurrence/combined_records'
+
+occ_files <- paste0( basename(gsub('1k.*', '', files)), '.shp')
+
+models_occ <- data.frame(
+  path2model = files, occurrence_data = occ_files
+) %>% 
+  mutate(
+    species = gsub('-.*$', '', basename(path2model)),
+    match = if_else(species == gsub('[.]shp$', '', occurrence_data), T, F )
+  ) # all look good!
+
+models_occ <- split(models_occ, f = models_occ$species)
+
+predict_wrapper <- function(x){
+
+  sdm <- readRDS(x$path2model)
+  occ_path <- file.path(occpath, x$occurrence_data)
+
+  # read in occurrences and subset prediction to 100 miles from border
+  bound <- sf::st_read(occ_path, quiet = TRUE) |>
+    dplyr::filter(Occurrence == 1) |>
+    sf::st_union() |>
+    sf::st_convex_hull() |>
+    sf::st_buffer(80467) |>
+    sf::st_bbox()
+
+  pout <- '~/Documents/suitability_maps'
+  layer_subset <- terra::subset(layers, term_grab(sdm))
+  layer_subset <- terra::crop(layer_subset, bound)
+
+  OneClass <- \(...) predict(...)[,2]
+  terra::predict(
+    layer_subset, sdm, cpkgs="randomForest",
+    type = 'prob', cores = 1, f = OneClass,
+    filename = file.path(pout, paste0(x$species, '-', gsub(' ', '_', Sys.time()), '.tif')),
+    wopt = c(names = 'predicted_suitability'))
+
+   terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)
+
+}
+
+lapply(models_occ, predict_wrapper)
 ```

### 2024-03-15 scripts/functions.R

@@ -366,6 +366,7 @@ Boruta_var_selector <- function(x){
 #' test_data output from Boruta_var_selector
 randomForests <- function(train_data, test_data, species, suffix){
   
+  if(missing(suffix)){suffix <- '1k'}
   # identify the appropriate mtry
   opt_mtry <- data.frame(
     randomForest::tuneRF(train_data[,-1], train_data[,1], stepFactor=1.5))
@@ -421,6 +422,7 @@ randomForests <- function(train_data, test_data, species, suffix){
 
 modeller <- function(x, suffix){
   
+  if(missing(suffix)){suffix <- '1k'}
   species <- sf::st_drop_geometry(x) %>% 
     dplyr::pull(taxon)
   species <- species[1]

### 2024-03-31 scripts/classify_patches.Rmd
working on fn to write out projects for senior botanists

@@
 
-lapply(files, FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
-
+lapply(files[9:18], FUN = landscapR, pout = 'patch_metrics', pRout = 'fine_patches')
 ```
@@
 
+# recs <- filter(recs, taxon %in% target_taxa)
+# f <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
+
 gd <- '/media/steppe/hdd/Geospatial_data'
@@
 f <- gsub('[.]tif$', '', list.files('../results/patches'))
+# f <- f[grep(paste(paste0(gsub(' ', '_', target_taxa)), collapse = '|'), basename(f))]
 
-# lentiformis  NTIF
-
-lapply(f, combine_pop_patches) # 2:46
+lapply(f, combine_pop_patches) 
 ```
@@
 lapply(f, tif2shp)
-
 ```
@@
 lapply(f, neigh_type)
-
 ```

### 2024-03-31 scripts/generate_absences.Rmd
working on fn to write out projects for senior botanists

--- 2024-01-09 scripts/generate_absences.Rmd
@@
 
-lapply(splicies[41:335], lda_PA_dropper, path = '../results/lda2/', col_names = cn)
-
-which(names(splicies) == 'Asclepias speciosa')
+lapply(splicies, lda_PA_dropper, path = '../results/lda2/', col_names = cn)
 

### 2024-03-31 scripts/subsetDataforCrews.Rmd
working on fn to write out projects for senior botanists

@@
 teams_list <- split(teams, f = teams$Contract)
+seniors_list <- split(seniors, f = seniors$Senior)
 ```
@@
 
-table(STZ$Dataset) 
-head(STZ)
+lapply(seniors_list,  project_maker,
+       admu = admin_boundaries, intype = 'senior',
+       target_species = spp_needed, 
+       blm_surf = blm_surf, 
+       fs_surf = usfs_surf, 
+       fire = fire_perimeters, 
+       invasive = invasives, 
+       total_change = change_intensity, 
+       occurrences = occurrences, 
+      # historic_SOS, 
+       roads = roads, seed_transfer = STZ)
+
 ```
@@
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-ott1 <- lapply(
+lapply(
   X = teams_list, # 5 throws an error

### 2024-03-31 scripts/functions.R

@@ -621,7 +621,7 @@ patchTaggR <- function(x){
   data.frame(
     noPres = nrow(occs), 
     PresTopSuitability = nrow(occ_pops), # the number of presences located in top suitability patches
-    UniquePatchID = nrow(dplyr::distinct(occ_pops, PatchID)), # the number of uniquelly identified patches
+    UniquePatchID = nrow(dplyr::distinct(occ_pops, PatchID)), # the number of uniquely identified patches
     NoPatchesTopSuitability = terra::minmax(r)[2] # the total number of most suitable patches from the RASTER
   ) |>
     write.csv(row.names = F, 
@@ -835,6 +835,7 @@ neigh_type <- function(x){
 #'
 #' @param project_areas split sf data set, with each crews field office(s), and name
 #' @param admu administrative boundaries
+#' @param intype one of 'crew' or 'senior', defaults to crew
 #' @param target_species a dataframe containing possible target species
 #' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
 #' @param blm_surf sf data set of surface management as multipolygon at most
@@ -847,7 +848,7 @@ neigh_type <- function(x){
 #' @param total_change Total Change Intensity Index from the MRLC
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
-project_maker <- function(x, target_species,
+project_maker <- function(x, target_species, intype,
                           admu, 
                           blm_surf, fs_surf, # ownership stuff
                           fire,  invasive,  occurrences, historic_SOS, 
@@ -858,12 +859,14 @@ project_maker <- function(x, target_species,
   
   defaultW <- getOption("warn")
   options(warn = -1)
+  if(missing(intype)){intype <- 'crew'}
   
   #### Initiate Project Directories ####
-  
   # create directory to hold all contents
   
[…]

### 2024-04-03 scripts/GroundTruthSampling.Rmd
working oon 3 layer decisions

---
title: "Ground Truth Sampling Cells Colorado Plateau"
author: "steppe"
date: "2024-03-31"
output: html_document
---

To determine whether SDM's and hydrological basins can help find native plant populations field tests were conducted on the Northern Colorado Plateau.

```{r}
library(sf)
library(terra)
```

```{r import species data}
p <- '/media/steppe/hdd/2024SOS_CrewGeospatial/Plateau – Senior Botanist/Geodata'
f <- paste0(p, '/Species/SDM/', 
       list.files(paste0(p, '/Species/SDM/'), pattern = '.shp$'))
dat <- lapply(f, st_read, quiet = T) 
dat <- lapply(dat, '[', 2) 
dat <- lapply(dat, terra::vect)
names(dat) <- gsub('.shp$', '', basename(f))
# dat <- bind_rows(dat, .id = 'Taxon')

rm(f)
```

```{r import study extent and convert polygons to rasters}

blm_surf <- st_read(paste0(p, '/Admin/Surface/BLM_Surface.shp'), quiet = T) |>
  dplyr::filter(Unit_Nm %in% c('Grand Junction Field Office', 'Uncompahgre Field Office', 'Vernal Field Office', 'Price Field Office'))

r <- rast(blm_surf, resolution = 90, nlyrs = 3) # create 30 meter resolution raster
datr <- lapply(dat, rasterize, r, field = 'Rank')
datr <- rast(datr)
r[[1]] <- app(ifel(datr == 1, 1, NA), sum, na.rm = TRUE)
r[[2]] <- app(ifel(datr %in% c(2, 3, 4), 1, NA, na.rm = TRUE))
datr2 <-

rm(dat)
```

### 2024-04-03 scripts/GroundTruthSampling.Rmd
summary lyr 3

@@
 r <- rast(blm_surf, resolution = 90, nlyrs = 3) # create 30 meter resolution raster
+names(r) <- c('SumOccupiedPatches', 'Sum1stOrderNeighboringPatches', 'CommonessInversion')
 datr <- lapply(dat, rasterize, r, field = 'Rank')
 datr <- rast(datr)
+
+# count the occupied patches by overlap  
 r[[1]] <- app(ifel(datr == 1, 1, NA), sum, na.rm = TRUE)
-r[[2]] <- app(ifel(datr %in% c(2, 3, 4), 1, NA, na.rm = TRUE))
-datr2 <-
 
-rm(dat)
+# count the patches which are first order neighbors from a known-populated locality  
+r[[2]] <- app(ifel(datr %in% c(2, 3, 4), 1, NA), sum, na.rm = TRUE)
+
+# How common are occupied + 1st order neighbors for each species? i.e. which are less common
+# and should be considered more importantly during scouting ??
+
+datr <- ifel(datr <= 4, 1, NA)
+tab <- freq(datr)
+
+tab$InverseWt <- 1 - ( tab$count / max(tab$count))
+tab$NegativeInverseWt <- round( 1 / (max(tab$InverseWt+1) - tab$InverseWt), 2) * 100
+# extreme skew rescale with 'negative transform', keep digits few for memory.
+
+par(mfrow=c(1,2)) 
+hist(tab$InverseWt, main = 'Raw', xlab = 'Weight') 
+hist(tab$NegativeInverseWt, main = 'Negative Transform', xlab = 'Weight')
+
+for (i in seq_along(1:dim(datr)[3])){ 
+  datr[[i]] <- ifel(datr[[i]] == 1, tab$NegativeInverseWt[i], NA) 
+}
+
+r[[3]] <- app(datr, sum, na.rm = TRUE)
+names(r) <- c('SumOccupiedPatches', 'Sum1stOrderNeighboringPatches', 'SumInverseCommonness')
+plot(r)
+
+writeRaster(r, '../results/Summary_Rasters/COPL.tif')
+
+rm(datr, tab, i)
 ```

### 2024-04-03 scripts/GroundTruthSampling.Rmd
adding fn to select unique ish cells

@@
 library(terra)
+set.seed(23)
 ```
@@
 ```
+
+```{r mask areas with fewer predicted species}
+samples <- spatSample(r, 10000, na.rm = T)
+
+lyr1 <- as.numeric(quantile(samples$SumOccupiedPatches, probs = 0.75))
+lyr2 <- as.numeric(quantile(samples$Sum1stOrderNeighboringPatches, probs = 0.75))
+lyr3 <- as.numeric(quantile(samples$CommonessInversion, probs = 0.75))
+
+r[[1]] <- ifel(r[[1]] < lyr1, NA, r[[1]])
+r[[2]] <- ifel(r[[2]] < lyr2, NA, r[[2]])
+r[[3]] <- ifel(r[[3]] < lyr3, NA, r[[3]])
+
+# we will need to sample 2 '1's per species, 3 '2-4's per species.
+
+rm(lyr1, lyr2, lyr3)
+```
+
+```{r}
+datr <- lapply(dat, rasterize, r, field = 'Rank')
+datr <- rast(datr)
+```
+
+```{r}
+
+samplesFocal <- vector(mode = 'list', length = dim(datr)[3])
+samples1stOrder <- vector(mode = 'list', length = dim(datr)[3])
+
+for (i in seq_along(1:dim(datr)[3])){
+
+  samplesFocal[[i]] <- spatSample(ifel(datr[[i]] == 1, 1, NA), 50, na.rm = TRUE, exhaustive = TRUE, as.points = TRUE)
+  samplesFocal[[i]] <- extract(r, samplesFocal[[i]], method = 'bilinear', xy = TRUE)
+  samples1stOrder[[i]] <- spatSample(ifel(datr[[i]] %in% c(2, 3, 4), 1, NA), 50, na.rm = TRUE, exhaustive = TRUE, as.points = TRUE)
+  samples1stOrder[[i]] <- extract(r, samples1stOrder[[i]], method = 'bilinear', xy = TRUE)
+  names(samplesFocal) <- names(datr) ; names(samples1stOrder) <- names(datr)
+
+}
+
+# now we determine which are within X distance of each other... 
+ob <- data.frame(
+  rbind(
+  data.table::rbindlist(samplesFocal, idcol = 'taxon') |>
+    dplyr::mutate('Target' = 'Populated'),
+  data.table::rbindlist(samples1stOrder, idcol = 'taxon') |>
+    dplyr::mutate('Target' = '1stOrder')
+)) %>% 
+  dplyr::arrange(taxon) %>% 
+  sf::st_as_sf(coords = c('x', 'y'), crs = 5070)
+
+ob<- split(ob, f = ob$taxon)
+
+identify_unqiue_pts <- function(x){
+  # determine which pairs are unique. 
+  DAT <- nngeo::st_nn(x, x, k = 5, maxdist = 2500, sparse = TRUE, returnDist = TRUE)
+  names(DAT$nn) <- 1:length(DAT$nn)
+  dd <- DAT$nn[ lengths( DAT$nn ) >= 2 ]
+
+  # now we need to remove records which are within this distance from each other... 
+  pairs <- vector(mode = 'list', length = length(dd))
+  for (i in 1:length(dd)){
+    pairs[[i]] <- which ( lapply(dd, setequal, dd[[i]]) == TRUE)
+  }
+
+  chosen_pts <- c(
+    DAT$nn[ lengths( DAT$nn ) == 1 ],
+    DAT$nn[as.numeric(names( unlist(lapply(unique(pairs), sample, 1)) ) )]
+  ) |> 
+    names() |>
+    as.numeric() |>
+    sort()
+
+  out <- x[chosen_pts,]
+  return(out)
+
+}
+
+lapply(ob, identify_unqiue_pts)
+```

### 2024-05-03 scripts/GroundTruthSampling.Rmd
dispatch

@@
 
-```{r import study extent and convert polygons to rasters}
+```{r import study extent and convert polygons to rasters, eval = F}
 
@@
 ```{r mask areas with fewer predicted species}
+
+r <- rast( '../results/Summary_Rasters/COPL.tif')
 samples <- spatSample(r, 10000, na.rm = T)
 
-lyr1 <- as.numeric(quantile(samples$SumOccupiedPatches, probs = 0.75))
-lyr2 <- as.numeric(quantile(samples$Sum1stOrderNeighboringPatches, probs = 0.75))
-lyr3 <- as.numeric(quantile(samples$CommonessInversion, probs = 0.75))
+lyr1 <- as.numeric(quantile(samples$SumOccupiedPatches, probs = 0.5))
+lyr2 <- as.numeric(quantile(samples$Sum1stOrderNeighboringPatches, probs = 0.5))
+lyr3 <- as.numeric(quantile(samples$CommonessInversion, probs = 0.5))
 
@@
 
-```{r}
+```{r reload data to consider each species independently}
 datr <- lapply(dat, rasterize, r, field = 'Rank')
@@
 
-```{r}
+We will consider three attributes of each sampled population...
+
+```{r draw random points for each species}
 
@@
 
-lapply(ob, identify_unqiue_pts)
+ob1 <- lapply(ob, identify_unqiue_pts) |>
+  bind_rows()
 ```
+
+Let's determine which sites have useful info
+
+```{r}
+
+obbies <- ob1 %>% 
+  rowwise() %>% 
+  mutate(sum_na = sum(is.na(c_across(SumOccupiedPatches:CommonessInversion)))) %>% 
+  group_by(taxon, Target) %>% 
+  arrange(taxon, Target, sum_na, -SumOccupiedPatches, -Sum1stOrderNeighboringPatches, -CommonessInversion) %>% 
+  slice_head(n = 3)
+
+```
+
+Create site indexes for all sample points
+
+```{r}
+
+ggplot() + 
+  geom_sf(data = obbies)
+
+disties <- st_nearest_feature(obbies)  
+dd <- st_distance(obbies, obbies[disties,], by_element = TRUE)
+
+p <- '/media/steppe/hdd/2024SOS_CrewGeospatial/species_check_play/'
+st_write(obbies, dsn = paste0(p, 'test.shp'))
+
+```

### 2024-05-03 scripts/subsetDataforCrews.Rmd
dispatch

--- 2024-03-13 scripts/subsetDataforCrews.Rmd
@@
 ```{r jitter the AIM data, eval = F}
+
 set.seed(125)
@@
 
-```{r seed zones, eval = F}
+```{r seed zones}
 seed_zones <- read.csv('../data/raw/2024sos_species-FIXED.csv') %>% 
@@
 
-seed_zones %>% 
-  filter(stz == 'BowerRevised') %>% 
-  distinct(ind_stz) %>% 
+seed_zones <- seed_zones %>% 
   mutate(
@@
   ) %>% 
-  separate_wider_delim(ind_stz, delim = ',', names= LETTERS[1:7], too_few = "align_start")
+  mutate(commas = max(str_count(ind_stz, pattern = ','))+1) %>% 
+  separate_wider_delim(ind_stz, delim = ',', names= LETTERS[1:.$commas], too_few = "align_start") %>% 
+  mutate(across(everything(), ~ str_trim(.))) %>% 
+  relocate(contract, .before = 'A') %>% 
+  pivot_longer(cols = LETTERS[1:.$commas], values_to = 'st_zone') %>% 
+  select(-commas, -name) %>% 
+  drop_na(st_zone) %>% 
+  mutate(stz = case_when(
+    stz == 'Desert SW provisional' ~ 'DesertSW', 
+    stz == 'BowerRevised' ~ 'Bower2013', 
+    .default = as.character(stz)
+  ))
 
-  #  ind_stz = str_replace(ind_stz, ' -[0-9]|[0-9]- |[0-9]-[0-9]', ' - ') )
+seed_zones <- read.csv('../data/raw/contract_lkp.csv') %>% 
+  left_join(seed_zones, ., by = c('contract' = 'Contract_colloquial'))
+
 ```
@@
 
+prov_stz <- rast('/media/steppe/hdd/EmpiricalSeedZones/data/geodata/final_provisional/provisional_stack.tif')
+
 rm(SWSTZ, BOWERSTZ, occ_p)
@@
 teams_list <- split(teams, f = teams$Contract)
+seniors_list <- split(seniors, f = seniors$Senior)
+
+t <- filter(occurrences, taxon == 'Abronia villosa') |>
+  vect() |>
+  project(crs(prov_stz))
+
+class(t)[1] == 'sf' 
+unique(st_geometry_type(t))[1] == 'POINT'
+
+# need to be able to dispatch to both raster and vector data, and both point and
+# polygon data. 
+if(class(x) == 'SpatRaster'){
+
+  # dispatch method for raster
+
+} else if(class(x) == 'sf'){ # determine type of vector data and dispatch.
+
+  if(unique(sf::st_geometry_type(x))[1] == 'POINT'){
+
+  } else {
+
+  }
+} else {
+  message('This class not expected by this small function. Small overhaul needed?')
+  }
+
+# this identified the occurrence records which we want to keep. 
+which(
+  !is.na(
+    terra::extract(prov_stz[[2]], t, method = 'simple', ID = F)))
+
 ```
@@
 
-lapply(teams_list, project_maker,
+unique(seed_zones$stz)
+
+lapply(teams_list[13], project_maker,
        admu = admin_boundaries, 
@@
 
-table(STZ$Dataset) 
-head(STZ)
+lapply(seniors_list[3],  project_maker,
+       admu = admin_boundaries, intype = 'senior',
+       target_species = spp_needed, 
+       blm_surf = blm_surf, 
+       fs_surf = usfs_surf, 
+       fire = fire_perimeters, 
+       invasive = invasives, 
+       total_change = change_intensity, 
+       occurrences = occurrences, 
+      # historic_SOS, 
+       roads = roads, seed_transfer = STZ)
 ```
@@
 p2ts <- '/media/steppe/hdd/SeedPhenology/data/processed/timestamps'
-ott1 <- lapply(
-  X = teams_list, # 5 throws an error
+lapply(
+  X = teams_list[13], # 5 throws an error
   FUN = phen_tabulator,

### 2024-05-03 scripts/functions.R

@@ -890,6 +890,7 @@ project_maker <- function(x, target_species, intype,
   # target species information
   dir.create( file.path(crew_dir, 'Geodata/Species'), showWarnings = F ) 
   dir.create( file.path(crew_dir, 'Geodata/Species/SDM'), showWarnings = F ) 
+  dir.create( file.path(crew_dir, 'Geodata/Species/SDM-raw'), showWarnings = F ) 
   dir.create( file.path(crew_dir, 'Geodata/Species/Occurrences'), showWarnings = F ) 
   dir.create( file.path(crew_dir, 'Geodata/Species/Historic_SoS'), showWarnings = F)
   
@@ -901,7 +902,6 @@ project_maker <- function(x, target_species, intype,
 #  dir.create( file.path(crew_dir, 'Geodata/Drought'))
   
   #### Process geographic data to a mask of the field office ####
-  
   focal_bbox <- dplyr::filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) %>%  
     sf::st_union() %>% 
     sf::st_transform(5070)
@@ -946,6 +946,7 @@ project_maker <- function(x, target_species, intype,
   #  crop(drought12, focal_vect, mask = T, threads = T, filename =
   #         file.path(crew_dir, 'Geodata/Drought', 'drought-12.tif'))
   
+  
   # write out species occurrence data
   if(intype == 'crew'){ContractInfo <- x$Contract[1]} else{ContractInfo <- unique(x$Contract)}
 
@@ -958,6 +959,7 @@ project_maker <- function(x, target_species, intype,
   occurrences_sub <- dplyr::filter(occurrences, taxon %in% gsub('_', ' ', t_spp))
   occurrences_sub <- sf::st_intersection(occurrences_sub, blm_surf_sub)
   occurrences_list <- split(occurrences_sub, f = occurrences_sub$taxon)
+  
   occ_writer <- function(x){
     
     binomial <- paste0(gsub(' ', '_', sf::st_drop_geometry(x$taxon[1])), '.shp')
@@ -981,19 +983,38 @@ project_maker <- function(x, target_species, intype,
     f_crop <- st_intersection(f_crop, blm_surf_sub) %>% 
       select(-Unit_Nm,-Loc_Nm)
[…]

### 2024-05-03 scripts/subsetDataforCrews.Rmd
patchedthe dispatch into fn for subsetting to target sstzs

@@
 
-class(t)[1] == 'sf' 
-unique(st_geometry_type(t))[1] == 'POINT'
-
-# need to be able to dispatch to both raster and vector data, and both point and
-# polygon data. 
-if(class(x) == 'SpatRaster'){
-
-  # dispatch method for raster
-
-} else if(class(x) == 'sf'){ # determine type of vector data and dispatch.
-
-  if(unique(sf::st_geometry_type(x))[1] == 'POINT'){
-
-  } else {
-
-  }
-} else {
-  message('This class not expected by this small function. Small overhaul needed?')
-  }
-
-# this identified the occurrence records which we want to keep. 
-which(
-  !is.na(
-    terra::extract(prov_stz[[2]], t, method = 'simple', ID = F)))
-
 ```

### 2024-05-03 scripts/functions.R

@@ -1070,27 +1070,47 @@ function(x, target_stz, prov){
                             unique(prov_taxa[prov_taxa$stz_zone != 'any', 'stz_zone']), 
                           prov_crop, NA)) 
 
-  # first step, see if stz are irrelevant for the taxon 
-  # determine whether 'target zones' are absent, e.g. if any STZ is fine, we do 
-  # not to do anything with these records. 
-  if(! x$binomial %in% prov_taxa$binomial){return(y)} else{
+  # need to be able to dispatch to both raster and vector data, and both point and
+  # polygon data. 
+  if(class(x) == 'SpatRaster'){
     
-    # determine which provisional seed transfer zone to use so we are sub-setting the 
-    # raster stack correctly 
+    # we can simply mask non target seed zones in our stz rasters, and then mask
+    # our species raster to that raster.
+    msk <- terra::ifel(prov_crop %in% c( target_stz[binomial == unique(x$taxon), 'st_zone']), 1, NA)
+    targeted_st_zones <- terra::mask(prov_crop, msk, inverse = TRUE)
+    raw_sdm_masked_to_target_stzs <- terra::mask(x, targeted_st_zones)
+    return(raw_sdm_masked_to_target_stzs)
     
-    i = which(names(prov) == target_stz$stz)
+  } else if(class(x) == 'sf'){ # determine type of vector data and dispatch. 
     
-    # subset raster to the area of interest. 
-    m <- terra::ifel(prov_crop %in% c( target_stz[binomial == unique(x$taxon), 'st_zone']), 1, NA)
-    targeted <- terra::mask(prov_crop, m, inverse = TRUE)
-    
-    taxon_in_target_stz <- x[which( # identify point records overlapping target stz. 
-      !is.na(
-        terra::extract(prov_crop, terra::project(terra::vect(x), crs(prov_crop)), 
-                       method = 'simple', ID = F))), ]
-    
-    # species by species subset to the target zones
-    return(taxon_in_target_stz)
+    if(unique(sf::st_geometry_type(x[1,])) == 'POINT'){
+      
+      msk <- terra::ifel(prov_crop %in% c( target_stz[binomial == unique(x$taxon), 'st_zone']), 1, NA)
+      targeted <- terra::mask(prov_crop, msk, inverse = TRUE)
+      taxon_in_target_stz <- x[which( # identify point records overlapping target stz. 
+        !is.na(
[…]

### 2024-05-03 scripts/subsetDataforCrews.Rmd
working towards fn to split by stz

@@
 
-unique(seed_zones$stz)
-
 lapply(teams_list[13], project_maker,
@@
        occurrences = occurrences, 
-      # historic_SOS, 
        roads = roads, seed_transfer = STZ)

### 2024-05-03 scripts/functions.R

@@ -848,7 +848,7 @@ neigh_type <- function(x){
 #' @param total_change Total Change Intensity Index from the MRLC
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
-project_maker <- function(x, target_species, intype,
+project_maker_original <- function(x, target_species, intype,
                           admu, 
                           blm_surf, fs_surf, # ownership stuff
                           fire,  invasive,  occurrences, historic_SOS, 
@@ -1038,16 +1038,321 @@ project_maker <- function(x, target_species, intype,
   options(warn = defaultW)
 }
 
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+############################################################################
+# write out portions of 
+
+#' Write out all spatial data for crews to a directory
+#' @param x a project species list
+#' @param project_areas split sf data set, with each crews field office(s), and name
+#' @param admu administrative boundaries
+#' @param intype one of 'crew' or 'senior', defaults to crew
+#' @param target_species a dataframe containing possible target species
+#' @param crew_id column in project_areas holding the crews identifier info 'e.g. UFO'
+#' @param blm_surf sf data set of surface management as multipolygon at most
+#' @param fs_surf sf data set of surface USFS land management as multipolygon at most
+#' @param fire sf data set of historic fires as multipopylgon at most
+#' @param invasive sf dataset of invasive species relative cover estimates
+#' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
+#' @param roads sf dataset of roads with relevant attritbutes
+#' @param seed_transfer sf dataset of seed transfer zones 
[…]

### 2024-05-06 scripts/subsetDataforCrews.Rmd
make fn respect species

@@
        roads = roads, seed_transfer = STZ)
+
 ```

### 2024-05-06 scripts/functions.R

@@ -1242,11 +1242,6 @@ project_maker <- function(x, target_species, intype,
 #    sf::st_write( dsn = file.path(crew_dir, 'Geodata/STZ', 'STZ.shp'), quiet = T, append = F)
 
   
-  
-  
-  
-  
-  
   # write out species occurrence data
   if(intype == 'crew'){ContractInfo <- x$Contract[1]} else{ContractInfo <- unique(x$Contract)}
   
@@ -1257,11 +1252,14 @@ project_maker <- function(x, target_species, intype,
   if(intype == 'senior'){t_spp <- unique(unlist(t_spp))}
   
   occurrences_sub <- dplyr::filter(occurrences, taxon %in% gsub('_', ' ', t_spp))
-  print(x$Contract)
-  subsetdata <- subset2stz(occurrences_sub, target_stz = seed_zones, prov = prov_stz, 
-                           requisition = x$Contract, focal_bbox = focal_vect, 
-                           blm_surf_sub = blm_surf_sub)
-  occurrences_list <- split(subsetdata, f = subsetdata$taxon)
+  occurrences_list <- split(occurrences_sub, f = occurrences_sub$taxon)
+  
+  occurrences_list <- lapply(
+    X = occurrences_list, FUN = subset2stz, target_stz = seed_zones,
+    prov = prov_stz, requisition = x$Contract, focal_bbox = focal_vect, 
+    blm_surf_sub = blm_surf_sub)
+ 
+  occurrences_list <- Filter(function(x) nrow(x) > 1, occurrences_list)
   return(occurrences_list)
   
   occ_writer <- function(x){
@@ -1349,7 +1347,7 @@ project_maker <- function(x, target_species, intype,
 #' @param x the input species list
 #' @param target_stz a dataframe with the Taxons name, Contract, desired STZs, and STZ product related to them. 
 #' @param prov A STACK OF PROVISIONAL RASTERS, THIS SHOULD BE CROPPED TO THE AREA OF ANALYSIS IN THE FUNCTION AROUD THIS ONE. 
-#' @param requistion essentially x of the parent function, x$Contract
+
 #' @param focal_bbox focal bounding box for study area, supply as a vect() so can be projected if necessary. 
 #' @param blm_surf_sub subsetted to blm owernship. 
 subset2stz <- function(x, target_stz, prov, requisition, focal_bbox, blm_surf_sub){
@@ -1361,36 +1359,44 @@ subset2stz <- function(x, target_stz, prov, requisition, focal_bbox, blm_surf_su
[…]

### 2024-05-07 scripts/subsetDataforCrews.Rmd
fn

@@
 
-lapply(teams_list[13], project_maker,
+ob <- lapply(teams_list[13], project_maker,
        admu = admin_boundaries, 
@@
 
+plot( ab[["Uncompahgre - P102085"]][["Astragalus lonchocarpus"]] )
+
 lapply(seniors_list[3],  project_maker,

### 2024-05-07 scripts/functions.R

@@ -1214,32 +1214,9 @@ project_maker <- function(x, target_species, intype,
   # write out ownership details
   filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice)) |>
     sf::st_cast("MULTILINESTRING") |> 
-    sf::st_write( dsn = file.path(crew_dir, 'Geodata/Admin/Boundaries', 'Field_Office_Boundaries.shp' ), quiet = T, append = F)
+    sf::st_write(dsn = file.path(crew_dir, 'Geodata/Admin/Boundaries', 'Field_Office_Boundaries.shp' ), 
+                  quiet = T, append = F)
   blm_surf_sub <- sf::st_intersection(blm_surf, focal_bbox) 
-  sf::st_write(blm_surf_sub, dsn = file.path(crew_dir, 'Geodata/Admin/Surface', 'BLM_Surface.shp'), quiet = T, append = F)
-  sf::st_intersection(allotments, focal_bbox) |> 
-    sf::st_write(dsn = file.path(crew_dir, 'Geodata/Admin/Allotments', 'Allotments.shp'), quiet = T, append = F)
-  sf::st_intersection(fs_surf, focal_bbox) |>
-    sf::st_write(dsn = file.path(crew_dir, 'Geodata/Admin/Surface', 'USFS_Surface.shp'), quiet = T, append = F)
-  
-  # write out invasive species and fire
-  fire <- sf::st_intersection(fire, blm_surf_sub)
-  fire |> 
-    dplyr::mutate(FIRE_YEAR = as.numeric(FIRE_YEAR)) |>  
-    sf::st_write(dsn =  file.path(crew_dir, 'Geodata/Disturb/Fire', 'Fire.shp'), quiet = T, append = F)
-  
-#  terra::crop(invasives, focal_vect, mask = T, threads = T, overwrite = T, filename =
-#                file.path(crew_dir, 'Geodata/Disturb/Invasive', 'Invasive.tif'))
-  
-  # write out the change_intensity
-#  terra::crop(total_change, focal_vect, mask = T, threads = T, overwrite = T, filename = #
-#                file.path(crew_dir, 'Geodata/Disturb/TotalChangeIntensity', 'TCII.tif'))
-  
-  # write out assorted data 
-#  roads[lengths(sf::st_intersects(roads, focal_bbox)) > 0,] |>
[…]

### 2024-05-07 scripts/subsetDataforCrews.Rmd
fn tear down and restart

--- 2024-05-06 scripts/subsetDataforCrews.Rmd
@@
   distinct()
+
+# add on the paths to their SDM's and post-processed SDMs
+
+patches_p <- '/media/steppe/hdd/SDM_restorations/results/PatchesClippedBLM'
+sdm_files <- file.path(patches_p, list.files(patches_p, pattern = 'shp$'))
+sdm_files_basin <- data.frame(
+  path_basins = sdm_files, 
+  taxon = gsub('.shp', '', basename(sdm_files))
+  )
+
+raw_p <- '/media/steppe/hdd/SDM_restorations/results/suitability_maps'
+sdm_files <- file.path(raw_p, list.files(raw_p, pattern = 'tif$'))
+sdm_files <- sdm_files[grepl('1k-', sdm_files)]
+
+sdm_files <- data.frame(
+  path_raw_sdm = sdm_files, 
+  taxon = gsub('1k.*$', '', basename(sdm_files))
+  )
+
+spp_needed <- left_join(spp_needed, sdm_files, by = c('binomial' = 'taxon')) %>% 
+  left_join(., sdm_files_basin, by = c('binomial' = 'taxon')) %>% 
+  mutate(
+    occurrence_path = 
+      '/media/steppe/hdd/SDM_restorations/data/raw/occurrence/combined_records/jittered_occurrences.shp')
+
+rm(patches_p, sdm_files, sdm_files_basin, raw_p)
 ```
@@
 
+```{r}
+
+emps <- filter(seed_zones, stz == 'DesertSW' & Requisition == 'Uncompahgre - P102085') %>% 
+  pull(binomial) %>% 
+  unique()
+
+spp_needed_test <- filter(spp_needed, Requisition == 'Uncompahgre - P102085' & binomial %in% emps)
+spp_needed_test <- split(spp_needed_test, f = spp_needed_test$binomial)
+
+example_rast <- rast(spp_needed_test[[1]][,'path_raw_sdm'])
+ptest <- project(prov_stz[[2]], 'epsg:5070')
+ptest <- resample(ptest, example_rast)
+
+#### Process geographic data to a mask of the field office ####
+focal_bbox <- dplyr::filter(admin_boundaries, ADMU_NA %in% stringr::str_to_upper(teams_list[[13]]$FieldOffice) ) %>%  
+  sf::st_union() %>% 
+  sf::st_transform(5070)
+
+focal_bbox <- focal_bbox 
+blm_surf_sub <- sf::st_intersection(blm_surf, focal_bbox)
+
+ptest <- mask(ptest, blm_surf_sub)
+plot(ptest)
+
+a <- lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, seed_zones = seed_zones, pSTZs = ptest)
+
+b <- vect(spp_needed_test[[1]][,'path_basins'])
+
+ob <- terra::mask(b, ptest)
+
+plot(a[["Aristida_purpurea"]])
+```
+
+```{r}
+f <- system.file("ex/elev.tif", package="terra")
+r <- rast(f)
+msk <- ifel(r < 400, NA, 1)
+
+m <- mask(r, msk)
+
+f <- system.file("ex/lux.shp", package="terra")
+v <- vect(f)[1,]
+
+mv1 <- mask(r, v)
+plot(mv1)
+```
+
 ```{r generate projects for crews, eval= F}

### 2024-05-07 scripts/functions.R

@@ -1230,8 +1230,7 @@ project_maker <- function(x, target_species, intype,
   if(intype == 'senior'){t_spp <- unique(unlist(t_spp))}
   
   
-  
-  
+
   occurrences_sub <- dplyr::filter(occurrences, taxon %in% gsub('_', ' ', t_spp))
   occurrences_list <- split(occurrences_sub, f = occurrences_sub$taxon)
   
@@ -1421,7 +1420,6 @@ subset2stz <- function(x, target_stz, prov, requisition, focal_bbox, blm_surf_su
       
       } else { # now essentially intersect the areas of polygons. 
       
-        targeted <- terra::mask(prov_crop, msk, inverse = TRUE)
         targeted_v <- sf::st_as_sf(
           terra::as.polygons(targeted, dissolve = FALSE, values = FALSE))
       
@@ -1451,17 +1449,48 @@ subset2stz <- function(x, target_stz, prov, requisition, focal_bbox, blm_surf_su
 
 
 
+# lapply through the species which are needed. 
 
+empiricalSTZ_writer <- function(spp_needed, seed_zones, pSTZs){
+  
+  sz <- seed_zones[seed_zones$binomial == spp_needed$binomial & 
+                     seed_zones$Requisition == spp_needed$Requisition,]
+  if(all(
+      grepl('[A-z]', unique(sz$st_zone)))){
+    msk_vals <- unique(sz$st_zone)} else {
+      msk_vals <- as.numeric(unique(sz$st_zone))
+    }
+  
+  # mask to field office # 
+  msk <- terra::ifel(pSTZs %in% msk_vals, pSTZs, NA) 
+  msk <- terra::crop(pSTZs, msk, mask = TRUE) 
+  
+  ############# first write out the occurrence points ################
+  occ <- sf::st_read(spp_needed$occurrence_path, quiet = T) |>
+    dplyr::mutate(taxon = str_replace(taxon, ' ', '_')) |>
+    dplyr::filter(taxon == spp_needed$binomial)
+  
+  taxon_in_target_stz <- occ[which( # identify point records overlapping target stz. 
+    !is.na(
+      terra::extract(msk, terra::project(terra::vect(occ), crs(pSTZs)), 
+                     method = 'simple', ID = F))), ]
+  
+  ############ second write out the raw SDM ################
[…]

### 2024-05-07 scripts/subsetDataforCrews.Rmd
prototype to write sdm polys and points together

@@
 ptest <- mask(ptest, blm_surf_sub)
+trim(ptest)
 plot(ptest)
 
-a <- lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, seed_zones = seed_zones, pSTZs = ptest)
+a <- lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, 
+            seed_zones = seed_zones, pSTZs = ptest) # 24 pts, 245 polys
 
-b <- vect(spp_needed_test[[1]][,'path_basins'])
-
-ob <- terra::mask(b, ptest)
-
-plot(a[["Aristida_purpurea"]])
 ```

### 2024-05-07 scripts/functions.R

@@ -1476,20 +1476,19 @@ empiricalSTZ_writer <- function(spp_needed, seed_zones, pSTZs){
                      method = 'simple', ID = F))), ]
   
   ############ second write out the raw SDM ################
-#  raw_sdm <- terra::rast(spp_needed$path_raw_sdm)
-#  raw_sdm <- terra::crop(raw_sdm, msk, mask = TRUE)
+  raw_sdm <- terra::rast(spp_needed$path_raw_sdm)
+  raw_sdm <- terra::crop(raw_sdm, msk, mask = TRUE)
   
-#  raw_sdm <- terra::mask(raw_sdm, ifel(raw_sdm < 0.70, NA, raw_sdm)) # if suitability < 70% remove
+  raw_sdm <- terra::mask(raw_sdm, ifel(raw_sdm < 0.70, NA, raw_sdm)) # if suitability < 70% remove
   # now aggregate rasters to half resolutions
-#  raw_sdm <- terra::aggregate(raw_sdm, fact = 2, fun = 'mean', na.rm = TRUE)
-#  raw_sdm <- terra::trim(raw_sdm)
+  raw_sdm <- terra::aggregate(raw_sdm, fact = 2, fun = 'mean', na.rm = TRUE)
+  raw_sdm <- terra::trim(raw_sdm)
 
   ############ third write out the basins SDM ################
   basins <- terra::vect(spp_needed$path_basins)
-  drainages_in_target_stz <- terra::mask(basins, msk)
-  return(drainages_in_target_stz)
-  
- # return(c(taxon_in_target_stz, raw_sdm, drainages_in_target_stz))
+  drainages_in_target_stz <- sf::st_as_sf(terra::crop(basins, raw_sdm))
+
+  return(c(taxon_in_target_stz, raw_sdm, drainages_in_target_stz))
 }
 
 

### 2024-05-08 scripts/subsetDataforCrews.Rmd
fn to write out 3 sdm data products by species

@@
 ptest <- mask(ptest, blm_surf_sub)
-trim(ptest)
+ptest <- trim(ptest)
 plot(ptest)
 
-a <- lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, 
+crew_dir <-  '/media/steppe/hdd/2024SOS_CrewGeospatial/Uncompahgre - P102085'
+lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, 
             seed_zones = seed_zones, pSTZs = ptest) # 24 pts, 245 polys
 
-```
-
-```{r}
-f <- system.file("ex/elev.tif", package="terra")
-r <- rast(f)
-msk <- ifel(r < 400, NA, 1)
-
-m <- mask(r, msk)
-
-f <- system.file("ex/lux.shp", package="terra")
-v <- vect(f)[1,]
-
-mv1 <- mask(r, v)
-plot(mv1)
 ```

### 2024-05-08 scripts/functions.R

@@ -1474,28 +1474,38 @@ empiricalSTZ_writer <- function(spp_needed, seed_zones, pSTZs){
     !is.na(
       terra::extract(msk, terra::project(terra::vect(occ), crs(pSTZs)), 
                      method = 'simple', ID = F))), ]
+    
+  binomial <- gsub(' ', '_', sf::st_drop_geometry(spp_needed$binomial[1]))
+  sf::st_write(taxon_in_target_stz, dsn = file.path(
+    crew_dir, 'Geodata/Species/Occurrences', paste0(binomial,'.shp')),
+             quiet = T, append = F)
   
   ############ second write out the raw SDM ################
   raw_sdm <- terra::rast(spp_needed$path_raw_sdm)
   raw_sdm <- terra::crop(raw_sdm, msk, mask = TRUE)
   
-  raw_sdm <- terra::mask(raw_sdm, ifel(raw_sdm < 0.70, NA, raw_sdm)) # if suitability < 70% remove
+  # if suitability < 70% remove
+  raw_sdm <- terra::mask(raw_sdm, ifel(raw_sdm < 0.70, NA, raw_sdm)) 
   # now aggregate rasters to half resolutions
   raw_sdm <- terra::aggregate(raw_sdm, fact = 2, fun = 'mean', na.rm = TRUE)
   raw_sdm <- terra::trim(raw_sdm)
-
+  
+  terra::writeRaster(raw_sdm, filename = file.path(
+    crew_dir, 'Geodata/Species/SDM-raw', paste0(binomial,'.tif')),
+    overwrite = T)
+  
   ############ third write out the basins SDM ################
   basins <- terra::vect(spp_needed$path_basins)
-  drainages_in_target_stz <- sf::st_as_sf(terra::crop(basins, raw_sdm))
+  drainages_in_target_stz <- terra::crop(basins, raw_sdm)
+  msk <- terra::as.polygons(msk > -Inf)
+  drainages_in_target_stz <- sf::st_as_sf(
+    terra::mask(drainages_in_target_stz, msk))
 
-  return(c(taxon_in_target_stz, raw_sdm, drainages_in_target_stz))
+  sf::st_write(drainages_in_target_stz, dsn = file.path(
+    crew_dir, 'Geodata/Species/SDM', paste0(binomial,'.shp')),
+    quiet = T, append = F)
+
 }
-
-
-
-
-
-
 
 
 

### 2024-05-08 scripts/subsetDataforCrews.Rmd
rebuild project builder from the ground up

@@
 #### Process geographic data to a mask of the field office ####
-focal_bbox <- dplyr::filter(admin_boundaries, ADMU_NA %in% stringr::str_to_upper(teams_list[[13]]$FieldOffice) ) %>%  
+focal_bbox <- dplyr::filter(admin_boundaries, 
+                            ADMU_NA %in% 
+                              stringr::str_to_upper(teams_list[[13]]$FieldOffice)) %>%  
   sf::st_union() %>% 
@@
 blm_surf_sub <- sf::st_intersection(blm_surf, focal_bbox) 
+ptest <- mask(ptest, blm_surf_sub)
 
-ptest <- mask(ptest, blm_surf_sub)
-ptest <- trim(ptest)
-plot(ptest)
+spp_needed_test <- filter(spp_needed, Requisition == 'Uncompahgre - P102085')
+
+emp_prov_info <- identify_spp_emp_prov_sz(
+  spp_needed_test, seed_zones = seed_zones, pSTZs = prov_stz, 
+  focal_domain = focal_bbox, blm_surf_sub = blm_surf_sub)
 
 crew_dir <-  '/media/steppe/hdd/2024SOS_CrewGeospatial/Uncompahgre - P102085'
-lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, 
-            seed_zones = seed_zones, pSTZs = ptest) # 24 pts, 245 polys
+
+# here we can put in:
+taxa_w_provisional_stz <- split(emp_prov_info[["taxa_w_provisional_stz"]], 
+                                f = 1:nrow(emp_prov_info[["taxa_w_provisional_stz"]]))
+
+taxa_w_empirical_stz <- split(emp_prov_info[["taxa_w_empirical_stz"]], 
+                                f = 1:nrow(emp_prov_info[["taxa_w_empirical_stz"]]))
+
+lapply(X = spp_needed_test[3:10], FUN = provisionalSTZ_writer, 
+            seed_zones = seed_zones, pSTZs = masked_pSTZ_lyr) 
+p <- '/media/steppe/hdd/SDM_restorations/results/suitability_maps'
+example_rast <- rast(file.path(p,'Achnatherum_hymenoides1k-2024-01-10_09:42:26.tif') )
+prov_stz <- project(prov_stz, 'epsg:5070') 
+prov_stz <- resample(prov_stz, example_rast)
+
+lapply(teams_list[13], project_maker, 
+       admu = admin_boundaries, 
+       target_species = spp_needed, 
+       blm_surf = blm_surf, 
+       fs_surf = usfs_surf, 
+       fire = fire_perimeters, 
+       invasive = invasives, 
+       total_change = change_intensity, 
+       occurrences = occurrences, 
+       roads = roads, 
+       pSTZs = prov_stz,
+       seed_transfer = STZ)
 

### 2024-05-08 scripts/functions.R

@@ -1146,6 +1146,7 @@ project_maker_original <- function(x, target_species, intype,
 #' @param historic_SOS sf dataset with historic SoS endeavors and relevant info
 #' @param roads sf dataset of roads with relevant attritbutes
 #' @param seed_transfer sf dataset of seed transfer zones 
+#' @param pSTZs rasterstack of provisional seed transfer zones
 #' @param total_change Total Change Intensity Index from the MRLC
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
@@ -1153,21 +1154,67 @@ project_maker <- function(x, target_species, intype,
                           admu, 
                           blm_surf, fs_surf, # ownership stuff
                           fire,  invasive,  occurrences, historic_SOS, 
-                          total_change, 
+                          total_change, pSTZs,
                           roads, seed_transfer#, 
                           #     drought6, drought12
 ){
   
-  defaultW <- getOption("warn")
-  options(warn = -1)
   if(missing(intype)){intype <- 'crew'}
-  
-  #### Initiate Project Directories ####
-  # create directory to hold all contents
-  
   if(intype == 'crew'){
     crew_dir <- paste0('/media/steppe/hdd/2024SOS_CrewGeospatial/', x['Contract'][[1]][1])
   } else {crew_dir <- paste0('/media/steppe/hdd/2024SOS_CrewGeospatial/', x['Senior'][[1]][1])}
+  
+  
+  # create all of the directories to hold data. 
+  init_proj_dir(x = x, crew_dir)
+
+  #### Process geographic data to a mask of the field office ####
+  focal_bbox <- dplyr::filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice) ) |>  
+    sf::st_union() |> 
+    sf::st_transform(5070)
+  
+  # create domains in two formats for usage by different functions
+  focal_vect <- terra::vect(focal_bbox); focal_bbox <- focal_bbox 
+  
+  # write out ownership details 
+  filter(admu, ADMU_NA %in% stringr::str_to_upper(x$FieldOffice)) |> 
[…]

### 2024-05-08 scripts/subsetDataforCrews.Rmd
workable prototype for provisional stzs

--- 2024-05-08 scripts/subsetDataforCrews.Rmd
@@
 
-emps <- filter(seed_zones, stz == 'DesertSW' & Requisition == 'Uncompahgre - P102085') %>% 
-  pull(binomial) %>% 
-  unique()
+p <- '/media/steppe/hdd/SDM_restorations/results/suitability_maps'
+example_rast <- rast(file.path(p,'Achnatherum_hymenoides1k-2024-01-10_09:42:26.tif') )
+prov_stz <- project(prov_stz, 'epsg:5070') 
+prov_stz <- resample(prov_stz, example_rast)
 
-spp_needed_test <- filter(spp_needed, Requisition == 'Uncompahgre - P102085' & binomial %in% emps)
-spp_needed_test <- split(spp_needed_test, f = spp_needed_test$binomial)
+lapply( 
+  teams_list[2], project_maker, 
+  admu = admin_boundaries, 
+  target_species = spp_needed, 
+  blm_surf = blm_surf, 
+  fs_surf = usfs_surf, 
+  fire = fire_perimeters, 
+  invasive = invasives, 
+  total_change = change_intensity,  
+  occurrences = occurrences, 
+  roads = roads,  
+  pSTZs = prov_stz, 
+  seed_transfer = STZ
+  )
 
-example_rast <- rast(spp_needed_test[[1]][,'path_raw_sdm'])
-ptest <- project(prov_stz[[2]], 'epsg:5070')
-ptest <- resample(ptest, example_rast)
-
-#### Process geographic data to a mask of the field office ####
-focal_bbox <- dplyr::filter(admin_boundaries, ADMU_NA %in% stringr::str_to_upper(teams_list[[13]]$FieldOffice) ) %>%  
-  sf::st_union() %>% 
-  sf::st_transform(5070)
-
-focal_bbox <- focal_bbox 
-blm_surf_sub <- sf::st_intersection(blm_surf, focal_bbox)
-
-ptest <- mask(ptest, blm_surf_sub)
-ptest <- trim(ptest)
-plot(ptest)
-
-crew_dir <-  '/media/steppe/hdd/2024SOS_CrewGeospatial/Uncompahgre - P102085'
-lapply(X = spp_needed_test[1], FUN = empiricalSTZ_writer, 
-            seed_zones = seed_zones, pSTZs = ptest) # 24 pts, 245 polys
+plot(ob[["Bishop - P102079"]][["10"]])
 

### 2024-05-08 scripts/functions.R

@@ -1183,8 +1183,7 @@ project_maker <- function(x, target_species, intype,
       dsn = file.path(crew_dir, 'Geodata/Admin/Boundaries', 'Field_Office_Boundaries.shp' ), 
                   quiet = T, append = F) 
   blm_surf_sub <- sf::st_intersection(blm_surf, focal_bbox)  
-
-
+  
   # determine which Seed Transfer Zones to use 
   emp_prov_info <- identify_spp_emp_prov_sz(spp_needed = target_species,
     seed_zones = seed_zones, pSTZs = pSTZs, 
@@ -1319,6 +1318,7 @@ identify_spp_emp_prov_sz <- function(spp_needed, seed_zones, pSTZs, focal_domain
 #'zones. 
 provisionalSTZ_writer <- function(spp_needed, seed_zones, pSTZs, crew_dir){
   
+  binomial <- gsub(' ', '_', sf::st_drop_geometry(spp_needed$binomial[1]))
   sz <- seed_zones[seed_zones$binomial == spp_needed$binomial & 
                      seed_zones$Requisition == spp_needed$Requisition,]
   if(all(
@@ -1327,10 +1327,12 @@ provisionalSTZ_writer <- function(spp_needed, seed_zones, pSTZs, crew_dir){
       msk_vals <- as.numeric(unique(sz$st_zone))
     }
   
-  # mask the seed transfer zones to the ones of interest # 
-  msk <- terra::ifel(pSTZs %in% msk_vals, pSTZs, NA) 
-  msk <- terra::crop(pSTZs, msk, mask = TRUE) 
-  
+  if(all(msk_vals != 'any')){
+    # mask the seed transfer zones to the ones of interest # 
+    msk <- terra::ifel(pSTZs %in% msk_vals, pSTZs, NA) 
+    msk <- terra::crop(pSTZs, msk, mask = TRUE) 
+  } else {msk <- pSTZs}
+
   ############# first write out the occurrence points ################
   if(!is.na(spp_needed$occurrence_path)){
     
@@ -1343,7 +1345,6 @@ provisionalSTZ_writer <- function(spp_needed, seed_zones, pSTZs, crew_dir){
         terra::extract(msk, terra::project(terra::vect(occ), crs(pSTZs)), 
                        method = 'simple', ID = F))), ]
     
-    binomial <- gsub(' ', '_', sf::st_drop_geometry(spp_needed$binomial[1]))
[…]

### 2024-05-09 scripts/subsetDataforCrews.Rmd
fn for empirical

@@
 
+p <- '/media/steppe/hdd/EmpiricalSeedZones/data/geodata/rasters'
+f <- file.path(p, list.files(p, pattern = 'tif$'))
+eSTZs <- rast(f)
+names(eSTZs) <- gsub('-.*tif$', '', basename(f))
+eSTZs <- project(eSTZs, 'epsg:5070') 
+eSTZs <- resample(eSTZs, example_rast)
+rm(f, p)
+
 lapply( 
-  teams_list[2], project_maker, 
+  teams_list[13], project_maker, 
   admu = admin_boundaries, 
@@
   pSTZs = prov_stz, 
+  eSTZs = eSTZs, 
   seed_transfer = STZ 
@@
 
-plot(ob[["Bishop - P102079"]][["10"]])
+plot(eSTZs)
 

### 2024-05-09 scripts/functions.R

@@ -1147,6 +1147,7 @@ project_maker_original <- function(x, target_species, intype,
 #' @param roads sf dataset of roads with relevant attritbutes
 #' @param seed_transfer sf dataset of seed transfer zones 
 #' @param pSTZs rasterstack of provisional seed transfer zones
+#' @param eSTZs rasterstack of empirical seed transfer zones
 #' @param total_change Total Change Intensity Index from the MRLC
 #' @param drought6 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
 #' @param drought12 12 netcdf of drought dataset from SPEI website, we recommend using 6 and 12 month
@@ -1154,7 +1155,7 @@ project_maker <- function(x, target_species, intype,
                           admu, 
                           blm_surf, fs_surf, # ownership stuff
                           fire,  invasive,  occurrences, historic_SOS, 
-                          total_change, pSTZs,
+                          total_change, pSTZs, eSTZs, 
                           roads, seed_transfer#, 
                           #     drought6, drought12
 ){
@@ -1198,11 +1199,15 @@ project_maker <- function(x, target_species, intype,
                                 f = 1:nrow(emp_prov_info[["taxa_w_empirical_stz"]]))
   
   # write out the data for Occurrence Points, Hydrological Basins, and Raw SDMs
-  # for species with provisional seed transfer zones
+  # for species with PROVISIONAL seed transfer zones
   lapply(X = taxa_w_provisional_stz, FUN = provisionalSTZ_writer, 
          seed_zones = seed_zones, pSTZs = emp_prov_info[['masked_pSTZ_lyr']], 
          crew_dir = crew_dir)
 
+  # write out the data for Occurrence Points, Hydrological Basins, and Raw SDMs
+  # for species with EMPIRICAL seed transfer zones
+  lapply(X = taxa_w_empirical_stz, FUN = empiricalSTZ_writer,
+         seed_zones = seed_zones, eSTZs = eSTZs, crew_dir)
 }
 
 
@@ -1249,7 +1254,6 @@ init_proj_dir <- function(x, crew_dir){
   dir.create( file.path(crew_dir, 'Geodata/Drought'))
   
 }
-
 
 
 
[…]

### 2024-05-15 scripts/subsetDataforCrews.Rmd
writing

@@
 names(eSTZs) <- gsub('-.*tif$', '', basename(f))
-eSTZs <- project(eSTZs, 'epsg:5070') 
-eSTZs <- resample(eSTZs, example_rast)
+eSTZs <- resample(eSTZs, example_rast, method = 'near')
 rm(f, p)
@@
   )
-
-plot(eSTZs)
 

### 2024-05-15 scripts/functions.R

@@ -762,7 +762,7 @@ neigh_type <- function(x){
     dplyr::mutate(Presence = 'Predicted') |> 
     dplyr::filter(lyr_1 %in% pm$id) # filter for core area indexes
   
-  pred1 <- pred[ lengths(sf::st_covers(pred, occ)) == 0, ] # make this dataset distinct. 
+  pred1 <- pred[ lengths(sf::st_covers(pred, occ)) == 0, ] # make this data set distinct. 
   pred2 <- sf::st_buffer(pred1, dist = 80) # for up north. 
   pred2 <- sf::st_difference(pred2)
   

### 2024-08-12 scripts/subsetDataforCrews.Rmd
writing

@@
 
+crs(example_rast) == crs(eSTZs)
+
 p <- '/media/steppe/hdd/EmpiricalSeedZones/data/geodata/rasters'
 f <- file.path(p, list.files(p, pattern = 'tif$'))
-eSTZs <- rast(f)
-names(eSTZs) <- gsub('-.*tif$', '', basename(f))
-eSTZs <- resample(eSTZs, example_rast, method = 'near')
+eSTZs <- rast(f[1])
+plot(eSTZs)
+plot(example_rast)
+names(eSTZs) <- gsub('-.*tif$', '', basename(f[1]))
+
+eSTZs1 <- resample(eSTZs, example_rast, method = 'near', threads = 16)
+plot(eSTZs1)
 rm(f, p)
@@
 
-lapply(teams_list[13], project_maker, 
+lapply(teams_list[c(7,13)], project_maker, 
        admu = admin_boundaries, 
@@
 
-lapply(seniors_list[3],  project_maker,
+lapply(seniors_list[3],  project_maker_original,
        admu = admin_boundaries, intype = 'senior',
@@
 lapply(
-  X = teams_list[13], # 5 throws an error
+  X = teams_list[c(7,13)], # 5 throws an error
   FUN = phen_tabulator,

### 2024-08-13 scripts/Results.Rmd
importing lsm data

---
title: "Results"
author: "steppe"
date: "2024-08-12"
output: pdf_document
---

```{r load libraries}
library(tidyverse)
library(sf)
library(terra)
```

## Import and prepare data for analysis.

```{r Subset to relevant species}
scouting <- read.csv('../data/SOS/Scouting.csv') |>
  mutate(taxa = gsub(' ', '_', taxa)) |>
  select(c(1:10, 22, 23)) |>
  filter(taxa != 'Achnatherum_hymenoides')

f <- list.files('../results/suitability_maps')
f <- f[ str_detect(f, '1k')]

f <- data.frame(
  f_suitability = paste0('../results/suitability_maps/', f), 
  species = gsub('1k-.*', '', f)
)

scouting <- filter(scouting, taxa %in% f$species) |>
  sf::st_as_sf(coords = c('LONGITUDE_DECIMAL', 'LATITUDE_DECIMAL'), crs = 4326) |>
  st_transform(5070)

# here are the ranked patches
f_rp <- list.files('../results/rankedPatches', pattern = '.shp')
f_rp <- data.frame(
  f_rankedp = paste0('../results/rankedPatches/', f_rp), 
  species = gsub('[.]shp$', '', f_rp)
)

f_pm <- list.files('../results/patch_metrics')
f_pm <- data.frame(
  f_patchMetrics = paste0('../results/patch_metrics/', f_pm[grepl('patch', f_pm)]), 
  f_classMetrics = paste0('../results/patch_metrics/', f_pm[grepl('class', f_pm)]) 
) |>
  mutate(species = gsub('-.*$', '', basename(f_patchMetrics)))

files <- left_join(f, f_rp, by = 'species') |>
  left_join(f_pm, by = 'species') |>
  relocate(species, .before = 1) |>
  filter(species %in% scouting$taxa)

rm(f, f_rp, f_pm)
```

```{r Extract Values to Results}

scouting <- left_join(scouting, files, by = c('taxa' = 'species')) |>
  mutate(Presence = if_else(POPULATION_SIZE > 0, 1, 0), .after = POPULATION_SIZE)

scout_split <- split(scouting, scouting$taxa)
# we are going to discard species with fewer than 3 observations. 
scout_split <- scout_split[lapply(scout_split, nrow) >= 3]

#' return predicted suitability scores and patch ID's for scouting points
#' 
#' Use this function to return some modeled values from scouting points observations
#' @param x an sf/tibble/data frame split by species, and including columns with 
#' paths to required files. 
extractR <- function(x){

  suitability <- terra::rast(sf::st_drop_geometry(x$f_suitability[1]))
  patch_ranks <- sf::st_read(sf::st_drop_geometry(x$f_rankedp[1]), quiet = T)

  x <- sf::st_join(x, patch_ranks)
  suitability <- terra::extract(suitability, x)$predicted_suitability

  wide <- dplyr::mutate(
    x, 
    suitability = suitability, 
    .before = geometry) |>
    select(-f_suitability, -f_rankedp)

  # now we can join the class and patch metrics to each population 
  # which they were calculated for. 
  IDs <- tidyr::drop_na(wide, PchIDComb) |>
    dplyr::pull(PchIDComb)
  patch <- read.csv(sf::st_drop_geometry(x$f_patchMetrics[1]) ) |>
    dplyr::filter(id %in% IDs) |>
   # dplyr::rename(patch_metric = metric, patch_value = value) |>
    tidyr::pivot_wider(values_from = value, names_from = metric)
  class <- read.csv(sf::st_drop_geometry(x$f_classMetrics[1])) |>
    tidyr::pivot_wider(names_from = metric, values_from = value)

  ob <- cbind(patch, class)

  return(ob)

}

scouting <- lapply(scout_split[1], extractR)
scouting <- dplyr::bind_rows(scouting)

scouting[[1]] |> 
    tidyr::pivot_wider(names_from = metric, values_from = value)

rm(files)
```

## Determine which points were used in training the models

These points will need to be removed because they are already known to be presences...

## Analysis

1) What is the relationship between habitat suitability and presence of species?

```{r}

fp <- scout1[scouting$futurePotential=='Yes',]

quants <- quantile(fp$suitability, na.rm = TRUE, probs = c(0.25, 0.5, 0.75))
hist(fp$suitability)
abline(v = quants[[1]], lwd = 2)
abline(v = quants[[2]], lwd = 2)
abline(v = quants[[3]], lwd = 2)

```
```{r}
quants
```

2) What is the relationship between Patch Rank and presence of species?

```{r}
hist(fp$Rank)
```

3) What is the relationship between Patch metrics and presence of species?

```{r}

```

The second main part relates to population size

4) What is the relationship between habitat suitability and population size? 
```{r}
scout1 <- mutate(
  scout1, 
  taxa = as.factor(taxa), 
  COLL_ID = as.factor(COLL_ID))

plot(scout1$suitability, scout1$POPULATION_SIZE)

m <- lm(POPULATION_SIZE ~ suitability + taxa, data = scout1)
summary(m)

pois.mod <- lme4::glmer(POPULATION_SIZE ~ suitability + (1 | taxa), data = scout1, family = poisson)
summary(pois.mod)
anova(pois.mod)

pois.mod_int <- lme4::glmer(POPULATION_SIZE ~ suitability + (1 + suitability|taxa), data = scout1, family = poisson)
summary(pois.mod_int)

plot(pois.mod_int)
plot(residuals(pois.mod_int))

qqnorm(resid(pois.mod_int))# QQ-plot
qqline(resid(pois.mod_int))

AIC(pois.mod, pois.mod_int)
anova(pois.mod, pois.mod_int)
```

4) What is the relationship between patch metrics and population size? 
```{r}

```

### 2024-08-13 scripts/Results.Rmd
explore predicitons

@@
 
+  # now we can join the class and patch metrics to each population 
+  # which they were calculated for. 
+  IDs <- tidyr::drop_na(x, PchIDComb) |>
+    dplyr::pull(PchIDComb)
+
+  patch <- read.csv(sf::st_drop_geometry(x$f_patchMetrics[1]) ) |>
+    dplyr::filter(id %in% IDs) |>
+    tidyr::pivot_wider(values_from = value, names_from = metric)
+  class <- read.csv(sf::st_drop_geometry(x$f_classMetrics[1])) |>
+    tidyr::pivot_wider(names_from = metric, values_from = value)
+
+  if(nrow(patch) == 0){
+    patch <- data.frame(id = NA, cai = NA, enn = NA, frac = NA, para = NA )
+  }
+
+  lsm <- cbind(patch, class) |>
+    dplyr::mutate(id = as.character(id))
+
   wide <- dplyr::mutate(
@@
     .before = geometry) |>
-    select(-f_suitability, -f_rankedp)
+    dplyr::select(-f_suitability, -f_rankedp, -f_patchMetrics, -f_classMetrics) |>
+    dplyr::left_join(lsm, by = c('PchIDComb' = 'id' ))
 
-  # now we can join the class and patch metrics to each population 
-  # which they were calculated for. 
-  IDs <- tidyr::drop_na(wide, PchIDComb) |>
-    dplyr::pull(PchIDComb)
-  patch <- read.csv(sf::st_drop_geometry(x$f_patchMetrics[1]) ) |>
-    dplyr::filter(id %in% IDs) |>
-   # dplyr::rename(patch_metric = metric, patch_value = value) |>
-    tidyr::pivot_wider(values_from = value, names_from = metric)
-  class <- read.csv(sf::st_drop_geometry(x$f_classMetrics[1])) |>
-    tidyr::pivot_wider(names_from = metric, values_from = value)
-
-  ob <- cbind(patch, class)
-
-  return(ob)
+  return(wide)
 
@@
 
-scouting <- lapply(scout_split[1], extractR)
+scouting <- lapply(scout_split, extractR)
 scouting <- dplyr::bind_rows(scouting)
 
-scouting[[1]] |> 
-    tidyr::pivot_wider(names_from = metric, values_from = value)
+```
+
+```{r determine a null model for each species, eval = F}
+#' return a large sample from a raster to plot the amount of suitable habitat 
+#' 
+#' sample many rasters to make a plot showing the cumulative amount of suitable habitat
+#' @param x a data frame split by species including columns with paths to files. 
+prcntSuitable <- function(x){
+
+  r <- terra::rast(sf::st_drop_geometry(x$f_suitability[1]))
+  Suitability = spatSample(r, size = 50000, na.rm = TRUE)
+  q <- quantile(Suitability$predicted_suitability, probs = seq(0.01, 1, 0.01))
+
+  sample <- 
+    data.frame(
+      Species = sf::st_drop_geometry(x$taxa[1]), 
+      Percent = 1:100, 
+      Records = as.numeric(q)
+    )
+
+  return(sample)
+
+}
+
+suitability_amounts <- lapply(scout_split, prcntSuitable)
+saveRDS(suitability_amounts, file = '../data/processed/prcntSDMsurfaceSuitable.rds')
+```
+
+```{r}
+
+suitability_amounts <- readRDS('../data/processed/prcntSDMsurfaceSuitable.rds')
+suitability_amounts <- bind_rows(suitability_amounts)
+
+suitability_means <- suitability_amounts |>
+  group_by(Percent) |>
+  summarise(Mean = mean(Records))
+
+par(pty="s")
+plot(suitability_means$Mean, suitability_means$Percent, type = "l", col = "red", 
+     ylab = "% of all Raster Cells", xlab = "Cumulative Sum of Records at each Predicted Suitability")
+
+ggplot(suitability_amounts, aes(Percent, Records, color = Species, group = Species)) +
+  geom_line() +
+  scale_color_grey(guide = 'none') + 
+  geom_line(data = suitability_means, 
+            aes(Percent, Mean), inherit.aes = F, color = '#093824', lwd = 1) + 
+  labs(
+    y = "% of all Raster Cells",
+    x = 'Cumulative sum of records at each quantile of Predicted Suitability') + 
+  theme_classic() +
+  theme(aspect.ratio = 1/1)
+
+'#292E1E'
+'#1F271B'
+'#003E1F'
+'#003E1F'
+'#093824'
+'#04471C'
+'#1F2F16'
+'#137547'
+'#2A9134'
+
+'#254441'
+'#395B50'
+'#73BA9B'
+'#519872'
 
@@
 ```
+
 ```{r}
@@
 
+scout_long <- pivot_longer(scouting, cai:enn_mn)
+
+ggplot(scout_long, aes(y = POPULATION_SIZE, x = value)) +
+  geom_point() +
+  facet_grid(. ~ name)
+
 ```

### 2024-08-14 scripts/Results.Rmd
content

@@
 scouting <- left_join(scouting, files, by = c('taxa' = 'species')) |>
-  mutate(Presence = if_else(POPULATION_SIZE > 0, 1, 0), .after = POPULATION_SIZE)
+  mutate(Presence = if_else(POPULATION_SIZE > 0, 1, 0), .after = POPULATION_SIZE) |>
+
+  # penstemon palmeri are all seeded outside their native range and 'helped' along
+  # during establishment we will remove them from the analysis
+
+  filter(taxa != 'Penstemon_palmeri')
 
@@
 suitability_amounts <- bind_rows(suitability_amounts)
+remotes::install_github("bcgov/elucidate")
 
-suitability_means <- suitability_amounts |>
-  group_by(Percent) |>
-  summarise(Mean = mean(Records))
+y1 <- rnorm(1:10, 100, 10)
+elucidate::mean_ci(y1)
 
-par(pty="s")
-plot(suitability_means$Mean, suitability_means$Percent, type = "l", col = "red", 
-     ylab = "% of all Raster Cells", xlab = "Cumulative Sum of Records at each Predicted Suitability")
+recs <- filter(suitability_amounts, Percent == 50) |>
+  pull(Records)
 
-ggplot(suitability_amounts, aes(Percent, Records, color = Species, group = Species)) +
-  geom_line() +
+as.numeric( elucidate::mean_ci(recs, replicates = 1000, ci_level = 0.99) )
+
+prctn <- split(suitability_amounts, f = suitability_amounts$Percent)
+suitability_means <- lapply(prctn, \(x)
+  data.frame(as.list(
+    elucidate::mean_ci(x$Records, replicates = 1000, ci_level = 0.99))
+  )
+) |>
+  bind_rows() |>
+  mutate(Percent = 1:100)
+
+fp <- scouting[scouting$futurePotential=='Yes',] 
+fp <- arrange(fp, suitability) |>
+  st_drop_geometry() |>
+  select(suitability) |>
+  mutate(obs = 1:n() / n())
+
+linecols <- c(
+  'Species' = 'grey50',
+  "Mean" = "#093824", 
+  "95% CI" = "#73BA9B",
+  "Populations" = "#FF2C55"
+  )
+
+ggplot(suitability_amounts, aes(x = Percent, y = Records, color = Species, group = Species)) +
+  geom_line(lty = 1) +
   scale_color_grey(guide = 'none') + 
+  geom_ribbon(data = suitability_means, aes(ymin = lower, ymax = upper, x = Percent),
+              fill = '#73BA9B', inherit.aes = F, alpha = 0.5) + 
   geom_line(data = suitability_means, 
-            aes(Percent, Mean), inherit.aes = F, color = '#093824', lwd = 1) + 
+            aes(x = Percent, y =  mean, color = 'a'), color = '#093824', inherit.aes = F, lwd = 1) + 
+  geom_line(data = fp, aes(x = suitability*100, y = obs), color = '#FF2C55', inherit.aes = F,  lwd = 1)
+
   labs(
-    y = "% of all Raster Cells",
-    x = 'Cumulative sum of records at each quantile of Predicted Suitability') + 
+    title = 'Populations are found in\nareas with higher suitability',
+    x = "Raster Cells (cumulative)",
+    y = 'Predicted Suitability (cumulative)') + 
   theme_classic() +
-  theme(aspect.ratio = 1/1)
+  scale_y_continuous(breaks = seq(0, 1, by = 0.2), sec.axis = sec_axis(~.*100,
+    name = 'Populations Found (cumulative)',
+    breaks = seq(0, 100, by = 20), 
+    labels = \(x) paste0(x, "%"))
+  )  +
 
-'#292E1E'
-'#1F271B'
-'#003E1F'
-'#003E1F'
-'#093824'
-'#04471C'
-'#1F2F16'
-'#137547'
-'#2A9134'
+  scale_x_continuous(breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%")) + 
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    aspect.ratio = 1/1)
 
-'#254441'
-'#395B50'
-'#73BA9B'
-'#519872'
+rm(linecols)
 
-rm(files)
+hypothetical_curves
 ```
@@
 
-fp <- scout1[scouting$futurePotential=='Yes',]
+fp <- scouting[scouting$futurePotential=='Yes',] 
+quants <- quantile(fp$suitability, na.rm = TRUE, probs = c(0.25, 0.5, 0.75))
 
-quants <- quantile(fp$suitability, na.rm = TRUE, probs = c(0.25, 0.5, 0.75))
 hist(fp$suitability)
@@
 abline(v = quants[[2]], lwd = 2)
-abline(v = quants[[3]], lwd = 2)
 
@@
 ```{r}
-quants
+
 ```
@@
 
+- can we test moving the distance patches into their own categories ??? 
 ```{r}

### 2024-08-24 scripts/summarPlots-NAA-talk.R
plot for NAA talk

@@ -1,0 +1,47 @@
+setwd('/media/steppe/hdd/SDM_restorations/scripts')
+
+library(hrbrthemes)
+library(GGally)
+library(viridis)
+library(tidyverse)
+
+p2dat <- '../results/summary'
+f <- file.path(p2dat, list.files(p2dat)[ grep('1k.csv', list.files(p2dat))])
+
+dat <- do.call("rbind", sapply(f, read.csv, simplify = FALSE)) |>
+  rownames_to_column('Taxon') |>
+  mutate(Taxon = gsub('1k.*', '', basename(Taxon))) |>
+  filter(Metric %in% c('Independent_Variables', 'AUC', 'Sensitivity', 'Specificity',
+                       'Kappa', 'Balanced Accuracy')) |>
+  pivot_wider(names_from = Metric, values_from = Value) |>
+  mutate(AUC_copy = AUC)
+
+rm(p2dat, f)
+
+
+# Plot
+ggparcoord(dat,
+           columns = c( 4:5, 2, 6:7), groupColumn = 8,
+           showPoints = TRUE, scale = "globalminmax",
+           title = "Common evaluation metrics for each model",
+           alphaLines = 0.3 
+) + 
+  labs(y = 'Value', x = 'Measurement') + 
+  scale_color_gradient(high = '#426b69', low = '#8BB174') +
+
+  ylim(0, 1) + 
+  theme(
+    aspect.ratio = 4/3,
+    text = element_text(colour = "white"),
+    axis.text.x=element_text(colour="white"),
+    axis.text.y=element_text(colour="white"),
+    plot.title = element_text(hjust = 0.5),
+    legend.position="none",
+    panel.border = element_blank(),
+    panel.background = element_rect(fill='transparent'),
+    plot.background = element_rect(fill='transparent', color=NA),
+    legend.background = element_rect(fill='transparent'),
+    legend.box.background = element_rect(fill='transparent')
+  )
+
+ggsave('../plots/Eval_metrics.png', bg = 'transparent')

### 2024-08-25 scripts/summarPlots-NAA-talk.R
getting counts for data sources NAA talk

@@ -28,8 +28,6 @@ ggparcoord(dat,
 ) + 
   labs(y = 'Value', x = 'Measurement') + 
   scale_color_gradient(high = '#426b69', low = '#8BB174') +
-
-  ylim(0, 1) + 
   theme(
     aspect.ratio = 4/3,
     text = element_text(colour = "white"),
@@ -45,3 +43,84 @@ ggparcoord(dat,
   )
 
 ggsave('../plots/Eval_metrics.png', bg = 'transparent')
+
+
+
+#################################################################################
+####   TABLE OF HOW MANY RECORDS WERE USED AND WHAT THEIR SOURCES WERE    ######
+
+source('functions.R')
+
+base_p <- '../data/raw/occurrence'
+
+targets <- read.csv('../data/raw/2024sos_species.csv', na.strings = "") %>% 
+  ## we will not model a few species which are widely abundant, hence do not require modelling, and would utilize enormous computational resources.
+  filter(!species %in% c('tridentata', 'millefolium', 'dioica', 'negundo')) %>%  # purshia,
+  unite('taxon', genus:infraspecies, na.rm = T, sep = ' ') %>% # artemisia, and larrea !
+  mutate(taxon = str_squish(taxon)) |>
+  pull(taxon) |>
+  unique()
+
+  
+bien <- read.csv(file.path(base_p, 'BIEN_records.csv')) %>% 
+  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326) %>% 
+  mutate(date = as.Date(date_collected),
+         datasource = case_when(
+           datasource == 'VegBank' ~ 'VegBank', 
+           datasource == 'FIA' ~ 'FIA', 
+           .default = as.character('Herbarium')
+         )) |>
+  select(taxon = scrubbed_species_binomial, date, datasource)
+
+
+idig <- read.csv(file.path(base_p, 'IDIG.csv')) %>% 
+  st_as_sf(coords = c('geopoint.lon', 'geopoint.lat'), crs = 4326) %>% 
+  mutate(date = as.Date(datecollected), datasource = 'Herbarium') %>% 
+  select(taxon = scientificname, date, datasource)
+
+aim <- st_read(file.path(base_p, 'AIM', 'AIM_records.shp'), quiet = TRUE) %>% 
+  mutate(date = as.Date(DatVstd)) %>% 
[…]

### 2024-09-10 scripts/Manuscript.Rmd
layout manuscript for RE applied note

--- 
title: 'Using Species Distribution Models to inform native seed collection efforts' 
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  words          
keywords: |
  species distribution models, native seed development  
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/restoration-ecology.csl"
bibliography: ../citations/bibliography/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating} 
---

##  IMPLICATIONS FOR PRACTICE

up to six of these. let's shoot 3 with 2 sentence max per.

```{r echo = F}
knitr::opts_chunk$set(echo=F, warning = F, message = F)
```

# Introduction

A primary challenge in restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment, @merritt2011restoration). 
Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings (@pedrini2020collection, @broadhurst2015seeding,  @national2023assessment).
Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available for restoration @national2023assessment. 
However, numerous difficulties exist in both the wild harvest seed and it's increase which are limiting our ability to develop adequate amounts of germplasm.

While most species desired in restorations have historically had relatively large geographic ranges, numbers of populations, and number of individuals per populations, the development of native germplasm remains behind targets (@national2023assessment). 
We posit that in part this is due to the difficulty of finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed to being agricultural increase, a complications borne of widespread habitat degradation and unnatural wildfires *(@abatzoglou2021projected)*. 
Tools which are capable of predicting a species geographic range, the presence and size of populations across the range and in seed collection target units (such as empirical or provisional seed transfer zones or ecoregions), such as Species Distribution Models offer promise to increase the rate at which native germplasm can be developed.

However, while SDM's generate hypothesis of whether areas have environmental conditions similar to the observed environmental niche of the species they do not consider the probabilities of colonization, nor the populations census sizes. 
A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis. 
Utilizing the predicted unsuitability of habitat, with extreme barriers to dispersal, to create cost-resistance surfaces between patches with known populations and predicted populations allows for simulating the probability of occurrence. 
While previous correlations between habitat suitability and population size have often been low, we posit that patch level parameters offer a more useful prediction of population size (@weber2017there, @waldock2022quantitative). 
Patch metrics will more appropriately reflect parameters of a potential population, e.g. geographic extent, which we posit is a more informative proxy of population size than environmental niche correlation.  
Utilizing this approach will allow field crews to associate probabilities with previously unground-truthed patches.

Here we showcase the utility of SDM's to predict suitable habitat for common species in natural settings and use those data to test two specific hypotheses. 
Firstly that SDM's can identify more populations than exist in the occurrence based records they were derived from. 
And secondly patch metrics derived from SDM's correlate with observed population census size. 
In order to determine whether SDM's are useful in detecting new populations, over ~10~ field crews were given 50 putative populations to survey for the presence and absence of the modeled species. 
To determine if patch metrics can be used to predict population census sizes, these crews noted whether the populations was large enough to support

# 2 MATERIALS AND METHODS  
## 2.1 Study system

The 353 modelled species were selected by land managers in arid & semi-arid areas Western North America and are species they already or immediately intend to use in restoration. 
The spatial domain of modelling broadly encompasses the Western United States being bounded by the Pacific Ocean, the 50th parallel at north, -100 degrees at East and Mexico to the south. 
This domain has tremendous variation in amounts of elevation, temperature, and precipitation.

## Species Occurrence Records  
Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R (@chamberlain2024rgbif, @maitner2023bien, @michonneau2023idigbio). 
These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (bio1, 4, 10, 12, & 19; see TABLE X) and distance were flagged.
During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.

Species absences were generated using three processes, and we sought to have a 1:1 ratio of presences to absences. 
True absences were acquired from a massive ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g. if 20% of land ownership across a species range was BLM, than the number of presences * 0.2 defined the number of these absences). 
Likely absences, representing 15% of records, were generated outside the known range of the species, but bounded to be within 50 miles of the extent extent of occurrences. 
Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for ((1 - (% BLM Land + 15 % LA records)) * 1.25).
Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used. The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables (@venables2002mass), identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables (@naimi2014usdm). 
As many records, classified by the LDA as originating from the Presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.

## Environmental Variables  
Up to 44 variables were used in generating the species distribution models (table X, @karger2017climatologies, @hengl2017soilgrids250m, @ivushkin2019global,  @tuanmu2014global, @yamazaki2017merit, @amatulli2020geomorpho90m, @sanderson2002human). 
These variables were selected based on authors previous work, and represent variables commonly associated with the empirical distributions of taxa in arid and semi-arid regions. 
All variables were downloaded from source and re-sampled to 90m resolution using terra (@hijman2024terra).

## Species Distribution Modelling  
Random Forests models were generated for each species following several steps to reduce the number of features.
While superfluous features generally do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables @kursa2010boruta), and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.  
Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys (@kuhn2008caret, @liaw2002randomForest). 
Models were then predicted onto gridded surfaces which exceeded the species range by 50 miles (@hijman2024terra).

## Patch identification 
To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more (@usgs2004nhd), and the divides of HU10 watersheds (@usgs2023wbd) were 'burnt' away from the raster. 
The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a practical time period. 
Resultant patches < 5 acres were then discarded, and the rasters was resampled to it's input resolution.

Putative populations were identified using the patches generated above. These patches followed the same processing, except that HU12 watersheds were used to delineate populations.

Because the > 0.8 threshold used above did not capture all colonized patches, all patches with predicted suitability scores of >0.55 which contained species observations were then identified using Terra's patches to create a data set of known populations.

## Predicting Species Occurrence in Patches

All patches identified above were used to identify patches within 5 contiguous neighbors, or 5 kilometers, of a patch known to be occupied.
Occupied patches were determined using both the training and test occurrences data set used to generate the SDMs. 
To identify each patch within 5 orders of contiguous neighbors `nblag`, and to determine all patches within 5k of a populated patch `dnearneigh`, were used (@bivand2018spdep). 
For each of these non-occupied patches the number of occupied patches at different lags were counted.

Each non-occupied patch was assigned an arbitrary rank based upon whether they were contiguous with an occupied patch, and if contiguous than their lag number to the nearest occupied patch, and the number of occupied patches connected (TABLE XX). 
The arbitrarily assigned numbers increase from '1', for an occupied patch, to '7' for a patch which has fewer than 3 second-order contiguous neighbors to an occupied patch(es).

|  Connection   |  No. Connections   |   Rank   | 
| :-----------: | :----------------: | :------: | 
|     1^st^     |        >= 2        |    2     | 
|     1^st^     |         1          |    3     | 
|     2^nd^     |        >= 3        |    4     | 
|     2^nd^     |        <= 2        |    5     | 
|     3^rd^     |        >= 4        |    6     | 
|     3^rd^     |        <= 3        |    7     |

Patches which have no contiguous neighbors, but which have neighbors within 5km were also assigned rank values based in this system (TABLE XX).

|  No. Occupied Patches   |   Rank   | 
| :---------------------: | :------: | 
|           >= 3          |    5     |  
|           <= 2          |    6     |

# Results

# Discussion

# Acknowledgments

# Literature Cited

# Supporting Information

### 2024-09-11 scripts/Manuscript.Rmd
introduction

@@
   \noindent
-  words          
+  less than 250 words - no subsections.           
 keywords: |
@@
 
-A primary challenge in restoring Earths terrestrial ecosystems is the lack of available plant germplasm (@national2023assessment, @merritt2011restoration). 
-Given the scale of our restoration needs, in most scenarios, the only sustainable source of seed is from grow-outs of wild harvested seed in agricultural settings (@pedrini2020collection, @broadhurst2015seeding,  @national2023assessment).
-Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available for restoration @national2023assessment.
-However, numerous difficulties exist in both the wild harvest seed and it's increase which are limiting our ability to develop adequate amounts of germplasm.
+With the United Nations 'Decade of Restoration' well underway limitations in our capacity to carry out active restoration, which normally requires large volumes of seed, are becoming apparent [@national2023assessment].
+Given the scale of our restoration needs, in most scenarios, the only sustainable source of this quantity of seed are from the amplification ('grow-out') of wild harvested seed in agricultural settings [@pedrini2020collection; @broadhurst2015seeding; @national2023assessment].
+Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available from agricultural amplification for restoration projects [@national2023assessment, @merritt2011restoration].
+However, numerous difficulties exist in both the wild harvest of seed and it's subsequent  amplification which are limiting our ability to develop adequate amounts of germplasm.
 
-While most species desired in restorations have historically had relatively large geographic ranges, numbers of populations, and number of individuals per populations, the development of native germplasm remains behind targets (@national2023assessment).
-We posit that in part this is due to the difficulty of finding populations with the appropriate number of individuals, which are experiencing climatic conditions conducive to producing enough viable seed to being agricultural increase, a complications borne of widespread habitat degradation and unnatural wildfires *(@abatzoglou2021projected)*.
-Tools which are capable of predicting a species geographic range, the presence and size of populations across the range and in seed collection target units (such as empirical or provisional seed transfer zones or ecoregions), such as Species Distribution Models offer promise to increase the rate at which native germplasm can be developed.
+While most species desired in restorations have historically had relatively expansive geographic ranges, multitudes of populations - many of which had enormous census sizes, the development of native germplasm remains behind targets [@national2023assessment].
+We posit that in part this is due to the difficulty of finding populations with the appropriate number of individual plants, which are experiencing climatic conditions conducive to producing enough viable seed which can be sustainably harvested to begin economically feasible agricultural increase.
+Essentially the same reasons we need to enact active restoration in the first place,  widespread habitat degradation and unnatural wildfires spurred by climate change and ill-informed historic land management practices, are the same hindrances to developing a locally-adapted and affordable  germplasm  *[@abatzoglou2021projected)]*.
+Heuristic tools which are capable of predicting a species geographic range, the presence of populations across the range, and which can predict them across seed collection target units (e.g.
+seed transfer zones), such as Species Distribution Models offer promise to increase the rate at which native germplasm can be developed.
 
-However, while SDM's generate hypothesis of whether areas have environmental conditions similar to the observed environmental niche of the species they do not consider the probabilities of colonization, nor the populations census sizes.
+A Species Distribution Model (alternatively referred to as 'Environmental niche models' etc.) uses presence, and optionally absence, records of a species and correlates these records to environmental variables (e.g.
+climate, land cover, and soil), using various families of statistical models (e.g.
+GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface.
+While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@yee).
+In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g.
+they proliferate on uncommon soils and are outcompeted on more typical soil types.
+Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored.
+Testing the applicability of SDMs to modeling the range of common species is further valuable as much biodiversity planning incorporates common species which are oftentimes viewed as the fabric holding together ecosystems, and as they may assist with planning restoration plantings under climate change.
+
+While the goal of an SDM is to develop a testable hypothesis of where a self sustaining population of a species *can* grow, they generally lack an explicit link to estimating *where* a population will actually grow.
+We argue that where populations will occur is driven largely by dispersal limitations, rather than simply whether habitat is suitable. 
 A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis. 
-Utilizing the predicted unsuitability of habitat, with extreme barriers to dispersal, to create cost-resistance surfaces between patches with known populations and predicted populations allows for simulating the probability of occurrence.
-While previous correlations between habitat suitability and population size have often been low, we posit that patch level parameters offer a more useful prediction of population size (@weber2017there, @waldock2022quantitative).
-Patch metrics will more appropriately reflect parameters of a potential population, e.g.
-geographic extent, which we posit is a more informative proxy of population size than environmental niche correlation.
-Utilizing this approach will allow field crews to associate probabilities with previously unground-truthed patches.
+Connectivity analysis, as generally implemented in land management, often requires a variety of computationally intensive algorithms stemming from a variety of disciplines (e.g. circuit theory).
+However we posit that in the case of widespread species useful results can be obtained by simply detailing the occupancy of drainage catchments, and the number of catchments between these occupied sites, and drainages with hypothesized potential habitat.
 
-Here we showcase the utility of SDM's to predict suitable habitat for common species in natural settings and use those data to test two specific hypotheses. 
-Firstly that SDM's can identify more populations than exist in the occurrence based records they were derived from. 
-And secondly patch metrics derived from SDM's correlate with observed population census size. 
-In order to determine whether SDM's are useful in detecting new populations, over ~10~ field crews were given 50 putative populations to survey for the presence and absence of the modeled species. 
-To determine if patch metrics can be used to predict population census sizes, these crews noted whether the populations was large enough to support
+In this article we use empirical field verified data to explore two main concepts developed from a large number of SDMs across an expansive geographic range. 
+The first is that SDM's are useful for finding populations of species, and that the number of populations found increases with habitat suitability.  
+The second is that SDM's are most useful when combined with knowledge of distance from the predicted to the observed cell.
 
@@
 
-The 353 modelled species were selected by land managers in arid & semi-arid areas Western North America and are species they already or immediately intend to use in restoration. 
-The spatial domain of modelling broadly encompasses the Western United States being bounded by the Pacific Ocean, the 50th parallel at north, -100 degrees at East and Mexico to the south. 
+The 353 taxa modeled were selected by land managers working for the Bureau of Land Management in arid & semi-arid areas Western North America and are species they already use, or which they intend to develop for use, in restoration.
+The spatial extent of modelling broadly encompasses the Western United States being bounded by the Pacific Ocean, the 50th parallel at north, -100 degrees at East and Mexico to the south. 
 This domain has tremendous variation in amounts of elevation, temperature, and precipitation.
@@
 ## Species Occurrence Records  
-Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R (@chamberlain2024rgbif, @maitner2023bien, @michonneau2023idigbio).
+Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio].
 These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
@@
 Species absences were generated using three processes, and we sought to have a 1:1 ratio of presences to absences. 
-True absences were acquired from a massive ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g.
+True absences were acquired from an ecological monitoring program Assess, Inventory, and Monitor (AIM), these absences accounted for a percent equivalent to land managed by BLM within the species range (e.g.
 if 20% of land ownership across a species range was BLM, than the number of presences * 0.2 defined the number of these absences).
@@
 Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used.
-The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables (@venables2002mass), identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables (@naimi2014usdm).
+The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables [@venables2002mass], identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables [@naimi2014usdm].
 As many records, classified by the LDA as originating from the Presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.
@@
 ## Environmental Variables  
-Up to 44 variables were used in generating the species distribution models (table X, @karger2017climatologies, @hengl2017soilgrids250m, @ivushkin2019global,  @tuanmu2014global, @yamazaki2017merit, @amatulli2020geomorpho90m, @sanderson2002human).
+Up to 44 variables were used in generating the species distribution models (Appendix 1). 
 These variables were selected based on authors previous work, and represent variables commonly associated with the empirical distributions of taxa in arid and semi-arid regions. 
-All variables were downloaded from source and re-sampled to 90m resolution using terra (@hijman2024terra).
+The thematic contents of these variables largely relate to climate [@karger2017climatologies], landcover [@tuanmu2014global], topography and landform [@amatulli2020geomorpho90m ; @yamazaki2017merit], soil [@hengl2017soilgrids250m; @ivushkin2019global], and representations of anthropogenic impacts [@sanderson2002human].
+All variables were downloaded from source and re-sampled to 90m resolution using terra [@hijman2024terra].
 
@@
 While superfluous features generally do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
-The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables @kursa2010boruta), and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.
-Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys (@kuhn2008caret, @liaw2002randomForest).
-Models were then predicted onto gridded surfaces which exceeded the species range by 50 miles (@hijman2024terra).
+The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables [@kursa2010boruta], and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.
+Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys [@kuhn2008caret; @liaw2002randomForest].
+Models were then predicted onto raster surfaces which exceeded the species known range by 50 miles [@hijman2024terra].
 
 ## Patch identification 
-To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more (@usgs2004nhd), and the divides of HU10 watersheds (@usgs2023wbd) were 'burnt' away from the raster.
+To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more [@usgs2004nhd], and the divides of HU10 watersheds [@usgs2023wbd] were 'burnt' away from the raster.
 The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a practical time period.
@@
 Occupied patches were determined using both the training and test occurrences data set used to generate the SDMs. 
-To identify each patch within 5 orders of contiguous neighbors `nblag`, and to determine all patches within 5k of a populated patch `dnearneigh`, were used (@bivand2018spdep). 
+To identify each patch within 5 orders of contiguous neighbors `nblag`, and to determine all patches within 5k of a populated patch `dnearneigh`, were used [@bivand2018spdep]. 
 For each of these non-occupied patches the number of occupied patches at different lags were counted.
@@
 
+how many records removed manually?
+how many records flagged by LDA?
+
+Which variables were the most important rank positions across all models? 
+How were model evaluation statistics?
+
+Where were the populations crews ended up finding? (weird elbow plot)
+How did presences and absences interact? (logistic regression)
+
 # Discussion
@@
 
+The development of the species distribution models, and the connectivity data products, were funded on a contract awarded to the Chicago Botanic Garden from the Bureau of Land Management.
+Ground verification was conducted in the routine operations of 13 contract crews funded by the BLM on the same set of agreements. 
+Kristy Snyder is thanked for sharing Seeds of Success field data which were required for the analyses in this manuscript.
+
 # Literature Cited

### 2024-09-11 scripts/Results.Rmd
plot for NAA talk
introduction

@@
 
+```{r write out some species for further data acquisition, eval = F}
+sort( gsub('_', ' ', names(scout_split))) # these will be the species we make notes on.
+
+data.frame(
+  Species = sort( gsub('_', ' ', names(scout_split))), 
+  'Pres-Abs' = '',
+  Census = ''
+)|>
+  write_csv('../data/SOS/Absence_sampling.csv')
+
+```
+
 ```{r determine a null model for each species, eval = F}
@@
 
-ggplot(suitability_amounts, aes(x = Percent, y = Records, color = Species, group = Species)) +
+ggplot(suitability_amounts, aes(y = Percent, x = Records, color = Species, group = Species)) +
   geom_line(lty = 1) +
   scale_color_grey(guide = 'none') + 
-  geom_ribbon(data = suitability_means, aes(ymin = lower, ymax = upper, x = Percent),
+  geom_ribbon(data = suitability_means, aes(xmin = lower, xmax = upper, y = Percent),
               fill = '#73BA9B', inherit.aes = F, alpha = 0.5) + 
   geom_line(data = suitability_means, 
-            aes(x = Percent, y =  mean, color = 'a'), color = '#093824', inherit.aes = F, lwd = 1) + 
-  geom_line(data = fp, aes(x = suitability*100, y = obs), color = '#FF2C55', inherit.aes = F,  lwd = 1)
+            aes(y = Percent, x =  mean, color = 'a'), color = '#093824', inherit.aes = F, lwd = 1) + 
+  geom_line(data = fp, aes(y = suitability*100, x = obs), color = '#FF2C55', inherit.aes = F,  lwd = 1) +
 
   labs(
-    title = 'Populations are found in\nareas with higher suitability',
-    x = "Raster Cells (cumulative)",
-    y = 'Predicted Suitability (cumulative)') + 
+    title = 'Populations are found in areas\nwith higher predicted suitability',
+    y = "Raster Cells (cumulative)",
+    x = 'Predicted Suitability (cumulative)') + 
   theme_classic() +
-  scale_y_continuous(breaks = seq(0, 1, by = 0.2), sec.axis = sec_axis(~.*100,
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2), sec.axis = sec_axis(~.*100,
     name = 'Populations Found (cumulative)',
@@
 
-  scale_x_continuous(breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%")) + 
+  scale_y_continuous(breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%")) + 
   theme(
@@
 rm(linecols)
+```
 
-hypothetical_curves
+```{r}
+suitability_means[ which.min(abs(suitability_means$Percent-50)), 'mean'] # on average 50% of cells have lower probabilities of being suitable than 0.23
+
+suitability_means[ which.min(abs(suitability_means$Percent-80)), 'mean'] # on average 80% of cells have lower probabilities of being suitable than 0.614
+
+fp[ which.min(abs(fp$obs-0.2)), ] # 80% of all populations are found in areas with suitability > 65.2%
+
+fp[ which.min(abs(fp$obs-0.5)), ] # 50% of all populations are found in areas with suitability > 83.6%
+
+fp[ which.min(abs(fp$obs-0.8)), ] # the final 20 % of populations are found in areas with suitability > 94.6%
 ```

### 2024-09-11 scripts/functions.R

@@ -808,8 +808,8 @@ neigh_type <- function(x){
                            OrderConnection == 'SecondOrder' & NoConnections <= 2 ~ 5, 
                            OrderConnection == 'ThirdOrder' & NoConnections >= 4 ~ 6, 
                            OrderConnection == 'ThirdOrder' & NoConnections <= 3 ~ 7, 
-                           OccupiedNeighborsIn5k >= 3 ~ 5, 
-                           OccupiedNeighborsIn5k <= 2 ~ 6, 
+                           OccupiedNeighborsIn5k >= 3 ~ 8, 
+                           OccupiedNeighborsIn5k <= 2 ~ 9, 
                            .default = as.numeric(7)
                          )) |>
     dplyr::slice_min(Rank, n = 1, by = 'ROW') |>
@@ -822,8 +822,8 @@ neigh_type <- function(x){
     dplyr::right_join( ranks, by = 'ROW') |>
     dplyr::select(-ROW, -Presence, PchIDComb = PatchIDCombined)
   
-  write.csv(dat, paste0('../results/connections/', taxon, '.csv'), row.names = F)
-  sf::st_write(x1, paste0('../results/rankedPatches/', taxon, '.shp'), quiet = T, append = F)
+  write.csv(dat, paste0('../results/connections-manuscript/', taxon, '.csv'), row.names = F)
+  sf::st_write(x1, paste0('../results/rankedPatches-manuscript/', taxon, '.shp'), quiet = T, append = F)
   
 }
 

### 2024-09-11 scripts/Manuscript.Rmd
abstract

@@
 --- 
-title: 'Using Species Distribution Models to inform native seed collection efforts' 
+title: 'Using species distribution models to direct native seed collection efforts' 
 author:  |
@@
   \noindent
-  less than 250 words - no subsections.           
+  Developing many locally adapted germplasms for habitat restoration has been a challenging process. 
+  In particular, finding a large enough number of populations which can serve as source material for agricultural increase across the range of species is proving difficult. 
+  The population scouting efforts of native seed collection crews can be informed using Species Distribution Models, allowing them to prioritize areas with higher probabilities of successfully finding the target species.
+  However, SDM's themselves do not have a mechanism for addressing the discontuity between suitable habitat and dispersal limitation, requiring that some of connectivity analysis be performed as a post-processing step to aid in their interpretation.
+  Herein we detail the high-throughput and semi-automated development of SDM's for over 350 common taxa native to Western North America. 
+  We further employ simple connectivity analysis, based solely on drainage basins and neighborhood analysis, on these hypothesis to try and ensure that areas with higher probabilities of propagule dispersal and population establishment are visited.
+  Using the hypothesis of suitable habitat generated by these hypothesis, with ground verification data from 13 crews in 5 Western states, we show that SDMs are useful for crews to find populations of native species, and that they are most useful when combined with connectivity analysis.
+  Our workflow is readily replicable by users on a variety of low-cost computing systems, and our implementation strategies for field crews are briefly discussed. 
 keywords: |
@@
 
-##  IMPLICATIONS FOR PRACTICE
+##  Implications for Practice
 
-up to six of these. let's shoot 3 with 2 sentence max per.
+- Finding enough large populations across seed collection zones (e.g. seed transfer zones) for economically viable agricultural increase has proven to be difficult. 
+- Heuristic tools which can be generated cheaply and computationally such as Species Distribution Models, and connectivity analysis, offer a possibility to increase the number of populations eligible for wild land seed collection.
+- Our code and protocol for generating and utilizing these tools are available and widely applicable to other users.
 
@@
 With the United Nations 'Decade of Restoration' well underway limitations in our capacity to carry out active restoration, which normally requires large volumes of seed, are becoming apparent [@national2023assessment].
-Given the scale of our restoration needs, in most scenarios, the only sustainable source of this quantity of seed are from the amplification ('grow-out') of wild harvested seed in agricultural settings [@pedrini2020collection; @broadhurst2015seeding; @national2023assessment].
+Given the scale of our seed-based restoration needs, in most scenarios, the only sustainable source of this quantity of seed are from the amplification ('grow-out') of wild harvested seed in agricultural settings [@pedrini2020collection; @broadhurst2015seeding; @national2023assessment].
 Enormous efforts are now underway to increase the number of species, the number of populations within these species, and the genetic diversity of the seed available from agricultural amplification for restoration projects [@national2023assessment, @merritt2011restoration].
@@
 GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface.
-While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@yee).
+While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@a2022species).
 In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g.
@@
 ## Species Occurrence Records  
-Species presence records were collected from Natural History Museums, citizen science initiatives, and standardized ecological monitoring programmes using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio].
+Species presence records were collected from herbaria, citizen science initiatives (e.g.
+iNaturalist), and standardized ecological monitoring programmes (e.g.
+VegBank, Forest Inventory and Analysis) using R [@chamberlain2024rgbif; @maitner2023bien; @michonneau2023idigbio].
 These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
-All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (bio1, 4, 10, 12, & 19; see TABLE X) and distance were flagged.
+All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
 During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.
@@
 Likely absences, representing 15% of records, were generated outside the known range of the species, but bounded to be within 50 miles of the extent extent of occurrences. 
-Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for ((1 - (% BLM Land + 15 % LA records)) * 1.25).
+Pseudo-absences (PA) were randomly selected from areas in, or within XX km, of the species range but greater than 10km from an occurrence, these records accounted for $((1 - (\text{\% BLM Land} + 15 \text{\% LA records})) * 1.25)$.
 Because most of these species are highly abundant in order to reduce the probability that a PA was drawn within a species the environmental niche linear discriminant analysis (LDA) was used.
 The LDA utilized the presences, true absences and likely absences as the classes and several environmental variables [@venables2002mass], identified by vifstep with theta = 10 as displaying at most moderate collinearity, as independent variables [@naimi2014usdm].
-As many records, classified by the LDA as originating from the Presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.
+As many records, classified by the LDA as originating from the presence data set, from the pseudo-absences could be removed as to achieve a class balance between presences and absences.
 
-## Environmental Variables  
+## Independent Variable Processing 
 Up to 44 variables were used in generating the species distribution models (Appendix 1). 
@@
 Random Forests models were generated for each species following several steps to reduce the number of features.
-While superfluous features generally do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
-The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables [@kursa2010boruta], and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33.
-Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling with all default values except for optimized mtrys [@kuhn2008caret; @liaw2002randomForest].
-Models were then predicted onto raster surfaces which exceeded the species known range by 50 miles [@hijman2024terra].
+While superfluous features do not notably decrease the performance of random forests, they increase the amount of time required to predict models onto gridded surfaces. 
+In our experience, predicting models onto raster surfaces is the most time consuming step of SDM generation. 
+The Boruta algorithm was used on the full stack of 44 independent variables to remove un-informative variables [@kursa2010boruta], and the variable importance factor (VIF) scores were then used to subset the most informative features in several auto-correlated pairs, reducing the independent variables to at most 33 [@naimi2014usdm].
+Recursive Feature Elimination (rfe) was then used to identify the fewest number of independent variables which could either increase model performance (measured using accuracy), or only decrease it by 1.5% relative to the full set of variables, these variables were subset and used as features for random forest modelling, using the 'randomForest' package with all default values except for optimized mtrys [@kuhn2008caret; @liaw2002randomForest].
+Models were then predicted onto raster surfaces which exceeded the species known range, based on the training and test data, in all directions by 50 miles [@hijman2024terra].
 
@@
 To identify putative metapopulations raster cells predicted as having less than 0.8 probability of suitable habitat were masked as NA's, and then areas which were crossed by streams with Strahler orders of three of more [@usgs2004nhd], and the divides of HU10 watersheds [@usgs2023wbd] were 'burnt' away from the raster.
-The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a practical time period.
+The resulting rasters were aggregated by a factor of 2 to 5, depending of their sizes (< 300 MiB 2, < 500 MiB 3, < 700 MiB 4, > 700 MiB 5) to accommodate, a rooks case search using terra in a scalable time period.
 Resultant patches < 5 acres were then discarded, and the rasters was resampled to it's input resolution.

### 2024-09-12 scripts/Manuscript.Rmd
plot of variable importance

@@
 GLMs, MaxEnt, decision trees), which allows for the fit model to be predicted onto a gridded (raster) surface.
-While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous (@a2022species).
-In the few published instances where SDMs have been tested, they have oftentimes been for rare species which have rather obvious and consequential drivers of their distributions, e.g.
-they proliferate on uncommon soils and are outcompeted on more typical soil types.
+While Species Distribution Models (SDMs) only generate hypothesis of whether areas have conditions similar to the observed environmental niche of the species based on the training records these hypothesis are rarely tested, and the link between their predictions and ecological reality have been credibly questioned as being tenuous [@a2022species].
+In the few published instances where SDMs have been tested, they have oftentimes been for rare species which oftentimes have rather obvious and consequential drivers of their distributions, e.g.
+they proliferate on uncommon soils and are outcompeted on more typical soil types [@johnson2023field].
 Whether the hypotheses generated by these models may be useful for detecting populations or common species, whereby their distributions often appear more governed by more subtle cues, appears unexplored.
@@
 
-While the goal of an SDM is to develop a testable hypothesis of where a self sustaining population of a species *can* grow, they generally lack an explicit link to estimating *where* a population will actually grow.
-We argue that where populations will occur is driven largely by dispersal limitations, rather than simply whether habitat is suitable. 
+While the goal of an SDM is to develop a testable hypothesis of where a self sustaining population of a species *can* grow, they generally lack an explicit link to estimating *where* a population will actually grow [@guisan2005predicting)].
+We argue that where populations will occur is driven largely by dispersal limitations, rather than simply whether habitat is suitable [@franklin2010moving]. 
 A possible tool to associate a probability of occurrence of a species in a suitable habitat patch, is connectivity analysis. 
@@
 These records were filtered to only those collected after 1950, with coordinate uncertainty less than 250 meters, and only the most recent record per 90m cell was retained. 
-All of these records were then manually reviewed by species, where records with more 2 or more variables in the 2.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
+All of these records were then manually reviewed by species, where records with more 2 or more variables in the 1.5% quantiles of several environmental variables (BIO1, BIO4, BIO10, BIO12, & BIO19; see TABLE X) and distance were flagged.
 During subjective manual review all digitized herbarium specimens which were suspect were reviewed, while other records were dropped based on the analysts review of other occurrence records.
@@
 | :---------------------: | :------: | 
-|           >= 3          |    5     |  
-|           <= 2          |    6     |
+|           >= 3          |    8     |  
+|           <= 2          |    9     |
 
@@
 
-how many records removed manually?
-how many records flagged by LDA?
+```{r}
+source('functions.R')
+library(tidyverse)
+```
+
+```{r How many records were removed in the cleaning process?}
+
+thinned <- bind_rows(
+  sf::st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE),
+  sf::st_read('../data/raw/occurrence/thinned/thinned_occurrences3.shp', quiet = TRUE)
+) |>
+  sf::st_drop_geometry() |>
+  group_by(taxon) |>
+  summarize(totalRecords = n())
+
+pts2remove <- read.delim('../data/raw/species_manual_cleaning.txt', header = FALSE)
+colnames(pts2remove) <- 'Species'
+
+pos <- str_extract_all(pts2remove$Species, "\\([^()]+\\)")
+pos <- lapply(pos, function(x){substring(x, 2, nchar(x)-1)})
+exact <- lapply(pos, function(x){suppressWarnings({as.numeric(sub(' ' , '', unlist(strsplit(x,","))))})})
+ranges <- lapply(pos, colon_balloon)
+
+to_remove <- Map(c, exact, ranges)  
+to_remove <- lapply(lapply(to_remove, na.omit), sort)
+names(to_remove) <- stringr::str_remove(pts2remove$Species, ' c[(].*$')
+recs_removed <- lapply(to_remove, function(x){out <- length(na.omit(gsub('^0$', NA, x)))}) |>
+  cbind() |>
+  data.frame() |>
+  rownames_to_column() |>
+  setNames(c('taxon', 'recordsRemoved')) |>
+  mutate(recordsRemoved = as.numeric(recordsRemoved)) |>
+  inner_join(thinned, by = 'taxon') |>
+  mutate(propRemoved = if_else(recordsRemoved > 0, recordsRemoved/totalRecords, 0))
+
+rm(exact, ranges, pos, pts2remove, to_remove, thinned)
+rm(list = ls()[ sapply( ls(), exists, mode="function")])
+```
+
+```{r How many records flagged by LDA?}
+
+f <- file.path('../results/lda', list.files('../results/lda'))
+df <- do.call(rbind, lapply(f, function(x)
+  cbind(read.csv(x), name= gsub('[.]csv$', '', basename(x))))) |>
+  mutate(
+    removedFlagged = if_else(Removed >= Max_Recs_Removal, 1, 0),
+    propFlagged = Flagged/Abs_in
+    )
+
+```
+
+```{r}
+returnGini <- function(x){
+
+  f <- paste0(x, list.files(x, pattern = '1k-'))
+  mods <- lapply(f, readRDS)
+  mods <- parallel::mclapply(mods, randomForest::importance)
+
+  names(mods) <- gsub('1k-.*$', '', basename(f))
+  mods <- purrr::map(mods, data.frame) |>
+    purrr::map(tibble::rownames_to_column, var = 'Variable') |>
+    dplyr::bind_rows(.id = 'Taxon')
+
+  return(mods)
+}
+
+giniDecrease <- returnGini(x = '../results/rf_models/')
+
+gd_summ <- giniDecrease |>
+  group_by(Taxon) |>
+  mutate(
+    sumGini = sum(MeanDecreaseGini),
+    propDecreaseGini = MeanDecreaseGini/sumGini
+  ) |>
+  ungroup() |>
+  group_by(Variable) |>
+  summarize(
+    n = n(), 
+    mean = mean(propDecreaseGini)
+  ) |>
+  arrange(-n) |>
+  mutate(
+    Theme = case_when(
+      Variable %in% c(
+        'bio1', 'bio10', 'bio11', 'bio12', 'bio18', 'bio19', 'bio4', 'bio5',
+        'bio6', 'gdd5', 'gdgfgd5', 'hurs', 'ngd5', 'scd', 'swe', 'vpd') ~ 'Climate',
+      Variable %in% c('cfvo_30_60', 'depth2bedrock', 'phh2o_30_60', 'salinity', 
+                      'sand_5_15', 'silt_5_15', 'soc_0_5') ~ 'Soil',
+      Variable %in% c('aspect', 'cti', 'dem', 'slope', 'tpi', 'tri') ~ 'Topography',
+      Variable %in% c('herb', 'lastwild', 'shrub', 'tree') ~ 'Land Cover/Landuse'
+      )
+    )
+
+ggplot(data = gd_summ, aes(x = n, y = mean, color = Theme, label = Variable)) + 
+  geom_smooth(method = 'glm', formula = 'y ~ x', aes(x = n, y = mean), inherit.aes = F, color = 'black') +
+  scale_color_manual(
+    breaks = c("Climate", "Soil", "Land Cover/Landuse", "Topography"),
+    values=c("#773344", "#20A4F3", "#63A375", "#EA7317")
+  ) + 
+  geom_jitter() + 
+  ggrepel::geom_label_repel(max.overlaps = 17, alpha = 0.9) + 
+  labs(
+    title = 'Relationship between the number of models featuring\n a variable and the variables mean contribution to all models', 
+    x = 'Number of models including variable', 
+    y = ' Proportional mean decrease in Gini', 
+    caption = expression(~italic('Variables become more important as they increase the mean decrease in Gini'))
+    ) + 
+  theme_classic() + 
+  theme(
+    axis.line.y = element_blank(), 
+    axis.line.x = element_blank(),
+    plot.title = element_text(hjust = 0.5), 
+    legend.position = 'bottom'
+    )
+
+rm(ob, giniDecrease)
+```
+
+Of the `r round(sum(recs_removed$totalRecords) * 0.015, 0)` records which were flagged, and manually investigated by an analyst a total of `r sum(recs_removed$recordsRemoved)` were removed. 
+Of these `r nrow(recs_removed)` manually reviewed taxa, `r nrow(recs_removed[recs_removed$recordsRemoved==0,])` had no records removed, the mean proportion of records which were removed (based on the entire number of records present) was `r round(mean(recs_removed$propRemoved), 3)`, median = `r round(median(recs_removed$propRemoved), 3)`, with a maximum of `r round(max(recs_removed$propRemoved), 3)`.
+These values are skewed upwards due to a few species which serve as common 'dumping' grounds for difficult to identify taxa in species rich clades.
+
+While the training accuracy scores for the linear-discriminant analysis were high all across the board (mean = `r round(mean(df$ACC_train), 4)`, min = `r round(min(df$ACC_train), 3)`), the realized range on testing values were lower, and reflected values more typical of complex applications (mean = `r round(mean(df$ACC_test), 3)`, median = `r round(median(df$ACC_test), 3)`, min = `r round(min(df$ACC_test), 3)`).
+Of the randomly generated psuedo-absences which underwent linear discriminant analysis for `r nrow(df)` taxa, in only `r nrow(df[df$removedFlagged==0,])` instances did the number of records flagged for removal not exceed the number of records flagged.
+On average `r round(mean(df$propFlagged), 3)` records were flagged per species (median = `r round(median (df$propFlagged), 3)`).
 
 Which variables were the most important rank positions across all models?
+
 How were model evaluation statistics?
@@
 How did presences and absences interact? (logistic regression)
+
+```{r}
+rm(recs_removed)
+```
 

### 2024-09-12 scripts/summarPlots-NAA-talk.R

@@ -124,3 +124,80 @@ table(thinned$datasource)
 
 
 
+################################################################################
+#########    WHICH VARIABLES HAD MOST OF THE INFLUENCE ON THE MODELS?   ########
+returnGini <- function(x){
+  
+  f <- paste0(x, list.files(x, pattern = '1k-'))
+  mods <- lapply(f, readRDS)
+  mods <- parallel::mclapply(mods, randomForest::importance)
+  
+  names(mods) <- gsub('1k-.*$', '', basename(f))
+  mods <- purrr::map(mods, data.frame) |>
+    purrr::map(tibble::rownames_to_column, var = 'Variable') |>
+    dplyr::bind_rows(.id = 'Taxon')
+  
+  return(mods)
+}
+
+giniDecrease <- returnGini(x = '../results/rf_models/')
+
+gd_summ <- giniDecrease |>
+  group_by(Taxon) |>
+  mutate(
+    sumGini = sum(MeanDecreaseGini),
+    propDecreaseGini = MeanDecreaseGini/sumGini
+  ) |>
+  ungroup() |>
+  group_by(Variable) |>
+  summarize(
+    n = n(), 
+    mean = mean(propDecreaseGini)
+  ) |>
+  arrange(-n) |>
+  mutate(
+    Theme = case_when(
+      Variable %in% c(
+        'bio1', 'bio10', 'bio11', 'bio12', 'bio18', 'bio19', 'bio4', 'bio5',
+        'bio6', 'gdd5', 'gdgfgd5', 'hurs', 'ngd5', 'scd', 'swe', 'vpd') ~ 'Climate',
+      Variable %in% c('cfvo_30_60', 'depth2bedrock', 'phh2o_30_60', 'salinity', 
+                      'sand_5_15', 'silt_5_15', 'soc_0_5') ~ 'Soil',
+      Variable %in% c('aspect', 'cti', 'dem', 'slope', 'tpi', 'tri') ~ 'Topography',
+      Variable %in% c('herb', 'lastwild', 'shrub', 'tree') ~ 'Land Cover/Landuse'
+    )
+  )
+
+
+model <- lm(gd_summ$mean ~ gd_summ$n)
+
+ggplot(data = gd_summ, aes(x = n, y = mean, color = Theme, label = Variable)) + 
+  geom_smooth(method = 'glm', formula = 'y ~ x', aes(x = n, y = mean),
+              inherit.aes = F, color = 'black') +
[…]

### 2024-09-12 scripts/Manuscript.Rmd
results model diagnostics

@@
 rm(exact, ranges, pos, pts2remove, to_remove, thinned)
-rm(list = ls()[ sapply( ls(), exists, mode="function")])
 ```
@@
 f <- file.path('../results/lda', list.files('../results/lda'))
-df <- do.call(rbind, lapply(f, function(x)
+lda <- do.call(rbind, lapply(f, function(x)
   cbind(read.csv(x), name= gsub('[.]csv$', '', basename(x))))) |>
@@
 
-```{r}
-returnGini <- function(x){
-
-  f <- paste0(x, list.files(x, pattern = '1k-'))
-  mods <- lapply(f, readRDS)
-  mods <- parallel::mclapply(mods, randomForest::importance)
-
-  names(mods) <- gsub('1k-.*$', '', basename(f))
-  mods <- purrr::map(mods, data.frame) |>
-    purrr::map(tibble::rownames_to_column, var = 'Variable') |>
-    dplyr::bind_rows(.id = 'Taxon')
-
-  return(mods)
-}
+```{r Variable Importance}
 
@@
   ) |>
-  arrange(-n) |>
-  mutate(
-    Theme = case_when(
-      Variable %in% c(
-        'bio1', 'bio10', 'bio11', 'bio12', 'bio18', 'bio19', 'bio4', 'bio5',
-        'bio6', 'gdd5', 'gdgfgd5', 'hurs', 'ngd5', 'scd', 'swe', 'vpd') ~ 'Climate',
-      Variable %in% c('cfvo_30_60', 'depth2bedrock', 'phh2o_30_60', 'salinity', 
-                      'sand_5_15', 'silt_5_15', 'soc_0_5') ~ 'Soil',
-      Variable %in% c('aspect', 'cti', 'dem', 'slope', 'tpi', 'tri') ~ 'Topography',
-      Variable %in% c('herb', 'lastwild', 'shrub', 'tree') ~ 'Land Cover/Landuse'
-      )
-    )
+  arrange(-n)
 
-ggplot(data = gd_summ, aes(x = n, y = mean, color = Theme, label = Variable)) + 
-  geom_smooth(method = 'glm', formula = 'y ~ x', aes(x = n, y = mean), inherit.aes = F, color = 'black') +
-  scale_color_manual(
-    breaks = c("Climate", "Soil", "Land Cover/Landuse", "Topography"),
-    values=c("#773344", "#20A4F3", "#63A375", "#EA7317")
-  ) + 
-  geom_jitter() + 
-  ggrepel::geom_label_repel(max.overlaps = 17, alpha = 0.9) + 
-  labs(
-    title = 'Relationship between the number of models featuring\n a variable and the variables mean contribution to all models', 
-    x = 'Number of models including variable', 
-    y = ' Proportional mean decrease in Gini', 
-    caption = expression(~italic('Variables become more important as they increase the mean decrease in Gini'))
-    ) + 
-  theme_classic() + 
-  theme(
-    axis.line.y = element_blank(), 
-    axis.line.x = element_blank(),
-    plot.title = element_text(hjust = 0.5), 
-    legend.position = 'bottom'
-    )
+```
 
-rm(ob, giniDecrease)
+```{r Model Performance held out data}
+
+p2dat <- '../results/summary'
+f <- file.path(p2dat, list.files(p2dat)[ grep('1k.csv', list.files(p2dat))])
+
+mod_perform <- do.call("rbind", sapply(f, read.csv, simplify = FALSE)) |>
+  rownames_to_column('Taxon') |>
+  mutate(Taxon = gsub('1k.*', '', basename(Taxon))) |>
+  filter(Metric %in% c('Independent_Variables', 'AUC', 'Sensitivity', 'Specificity',
+                       'Kappa', 'Balanced Accuracy')) |>
+  pivot_wider(names_from = Metric, values_from = Value) |>
+  mutate(AUC_copy = AUC)
+
+rm(p2dat, f)
 ```
@@
 
-While the training accuracy scores for the linear-discriminant analysis were high all across the board (mean = `r round(mean(df$ACC_train), 4)`, min = `r round(min(df$ACC_train), 3)`), the realized range on testing values were lower, and reflected values more typical of complex applications (mean = `r round(mean(df$ACC_test), 3)`, median = `r round(median(df$ACC_test), 3)`, min = `r round(min(df$ACC_test), 3)`).
-Of the randomly generated psuedo-absences which underwent linear discriminant analysis for `r nrow(df)` taxa, in only `r nrow(df[df$removedFlagged==0,])` instances did the number of records flagged for removal not exceed the number of records flagged.
-On average `r round(mean(df$propFlagged), 3)` records were flagged per species (median = `r round(median (df$propFlagged), 3)`).
+While the training accuracy scores for the linear-discriminant analysis were high all across the board (mean = `r round(mean(lda$ACC_train), 4)`, min = `r round(min(lda$ACC_train), 3)`), the realized range on testing values were lower, and reflected values more typical of complex applications (mean = `r round(mean(lda$ACC_test), 3)`, median = `r round(median(lda$ACC_test), 3)`, min = `r round(min(lda$ACC_test), 3)`).
+Of the randomly generated psuedo-absences which underwent linear discriminant analysis for `r nrow(lda)` taxa, in only `r nrow(lda[lda$removedFlagged==0,])` instances did the number of records flagged for removal not exceed the number of records flagged.
+On average `r round(mean(lda$propFlagged), 3)` records were flagged per species (median = `r round(median (lda$propFlagged), 3)`).
 
-Which variables were the most important rank positions across all models?
+The final models utilized `r nrow(gd_summ)` variables, the variables which showed up in the most models, also tended to be the most important in driving decisions (FIGURE X). 
+In general climate variables, especially the Bioclim variables, tended to be the most important in decreasing mean Gini, with variables of other types also contributing to models in lesser capacities (FIGURE X).
+On the partitioned data, most models performed very well across a variety of diagnostic metrics (FIGURE XX). 
+Area under the receiver operator curve (AUC) values were generally the highest with a mean = `r round(mean(mod_perform$AUC), 3)`, median = `r round(median(mod_perform$AUC), 3)`, kappa scores were generally lowest with a mean = `r round(mean(mod_perform$Kappa), 3)`, median = `r round(median(mod_perform$Kappa), 3)`, and balanced accuracy scores were intermediate with mean = `r round(mean(mod_perform[['Balanced Accuracy']]), 3)`, median = `r round(median(mod_perform[['Balanced Accuracy']]), 3)`.
 
-How were model evaluation statistics?
-
-Where were the populations crews ended up finding? (weird elbow plot)
+*Where were the populations crews ended up finding?* weird elbow plot)
 How did presences and absences interact? (logistic regression)
@@
 rm(recs_removed)
+rm(list = ls()[ sapply( ls(), exists, mode="function")])
+
 ```

### 2024-09-12 scripts/functions.R

@@ -1788,3 +1788,19 @@ date_maker <- function(){
   
   return(list(dates, reporting_dates))
 }
+
+
+
+returnGini <- function(x){
+  
+  f <- paste0(x, list.files(x, pattern = '1k-'))
+  mods <- lapply(f, readRDS)
+  mods <- parallel::mclapply(mods, randomForest::importance)
+  
+  names(mods) <- gsub('1k-.*$', '', basename(f))
+  mods <- purrr::map(mods, data.frame) |>
+    purrr::map(tibble::rownames_to_column, var = 'Variable') |>
+    dplyr::bind_rows(.id = 'Taxon')
+  
+  return(mods)
+}

### 2024-09-12 scripts/summarPlots-NAA-talk.R

@@ -23,16 +23,17 @@ rm(p2dat, f)
 ggparcoord(dat,
            columns = c( 4:5, 2, 6:7), groupColumn = 8,
            showPoints = TRUE, scale = "globalminmax",
-           title = "Common evaluation metrics for each model",
+           title = "Evaluation metrics for models",
            alphaLines = 0.3 
 ) + 
-  labs(y = 'Value', x = 'Measurement') + 
+  labs(y = 'Value', x = 'Metric') + 
   scale_color_gradient(high = '#426b69', low = '#8BB174') +
   theme(
     aspect.ratio = 4/3,
-    text = element_text(colour = "white"),
-    axis.text.x=element_text(colour="white"),
-    axis.text.y=element_text(colour="white"),
+  #  text = element_text(colour = "white"),
+   # axis.text.x=element_text(colour="white"),
+  #  axis.text.y=element_text(colour="white"),
+    axis.text.x = element_text(angle = 25, vjust = 0.65),
     plot.title = element_text(hjust = 0.5),
     legend.position="none",
     panel.border = element_blank(),

### 2024-09-13 scripts/Results.Rmd
conceptual figs

--- 2024-08-14 scripts/Results.Rmd
@@
 saveRDS(suitability_amounts, file = '../data/processed/prcntSDMsurfaceSuitable.rds')
+```
+
+```{r}
+
+habitat_specialist <- data.frame(
+  suitability = c(
+    seq(from= 0, to  = 0.1, length.out = 70), 
+    seq(from= 0.1, to  = 0.4, length.out = 15), 
+    seq(from= 0.4, to  = 0.8, length.out = 10),
+    seq(from= 0.8, to  = 1, length.out = 5)
+  ),
+  percent = 1:100
+)
+
+habitat_generalist <- data.frame(
+  suitability = c(
+    seq(from= 0, to  = 0.5, length.out = 40), 
+    seq(from= 0.6, to  = 0.7, length.out = 30), 
+    seq(from= 0.7, to  = 0.9, length.out = 20),
+    seq(from= 0.9, to  = 1, length.out = 10)
+  ),
+  percent = 1:100
+)
+
+ggplot() + 
+  geom_smooth(data = habitat_specialist, aes(x = percent, y = suitability)) + 
+  geom_smooth(data = habitat_generalist, aes(x = percent, y = suitability)) +
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    aspect.ratio = 1/1)
+
 ```
@@
 
-ggplot(suitability_amounts, aes(x = Percent, y = Records, color = Species, group = Species)) +
+suitability_means <- suitability_means |>
+  mutate(Null_mod = mean*Percent)
+
+ggplot(suitability_amounts,
+       aes(x = Percent, y = Records, color = Species, group = Species)) +
   geom_line(lty = 1) +
@@
   geom_line(data = suitability_means, 
-            aes(x = Percent, y =  mean, color = 'a'), color = '#093824', inherit.aes = F, lwd = 1) + 
-  geom_line(data = fp, aes(x = suitability*100, y = obs), color = '#FF2C55', inherit.aes = F,  lwd = 1)
+            aes(x = Percent, y =  mean, color = 'a'), 
+            color = '#093824', inherit.aes = F, lwd = 1) + 
+  geom_line(data = suitability_means, aes(x = Percent, y = Null_mod/100),
+            color = 'black', inherit.aes = F,  lwd = 1) +
+  geom_line(data = fp, aes(x = suitability*100, y = obs),
+            color = '#FF2C55', inherit.aes = F,  lwd = 1) +
 
@@
 
-  scale_x_continuous(breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%")) + 
+  scale_x_continuous(breaks = seq(0, 100, by = 20), 
+                     labels = \(x) paste0(x, "%")) + 
   theme(

### 2024-09-13 scripts/Using species distribution models to direct native seed collection efforts.Rmd
conceptual figs

--- 2024-09-12 scripts/Manuscript.Rmd
@@
 
+## Ground Verification 
+13 native seed collection crews had three products shared with them for planning their field work to scout populations for collections. 
+1) All input training and test occurrence points used in generating the models.
+2) All SDM models, in raster format, clipped to > 80% probability of suitable habitat.
+3) All drainage basins ranked via an algorithm similar to the above, as vector data, but where the dis-contiguous patches were assigned ranks of '' and '' respectively. 
+Some crews, notable those with less amounts of standardized monitoring projects in their areas, which adversely affect product 3, were also given raw SDM output rasters, i.e.
+the rasters not masked to <80% suitability.
+
+Four crews, all located on the northern Colorado Plateau, were given a checklist for taxon specific presence-absence monitoring to begin in late August until the end of their terms in October.  
+These 27 species were selected as they already had multiple presence records (>3), crews were asked to fill out these forms 8 times or more each, across their field offices as time permitted. 
+This subset of taxa, and number of sampling events, were chosen so that each taxon could become an effect in a logistic regression model, and so that crews could collect these data relatively quickly.
+
 # Results
@@
 
+```{r Populations found across all crews}
+
+```
+
+```{r Presences and Absences from Colorado Plateau Crews}
+
+```
+
+```{r Populations relative to drainages}
+
+```
+
+\begin{wrapfigure}{R}{0.5\textwidth}  
+ \begin{center}
+    \includegraphics[width=.5\textwidth]{../plots/VariableImportance.png}  
+  \caption{bird is the word} 
+\end{center}
+\end{wrapfigure}
+
 Of the `r round(sum(recs_removed$totalRecords) * 0.015, 0)` records which were flagged, and manually investigated by an analyst a total of `r sum(recs_removed$recordsRemoved)` were removed. 
@@
 
+\begin{wrapfigure}{R}{0.35\textwidth}  
+ \begin{center}
+    \includegraphics[width=.35\textwidth]{../plots/Eval_metrics.png}  
+  \caption{word to all my birds} 
+\end{center}
+\end{wrapfigure}
+
 *Where were the populations crews ended up finding?* weird elbow plot)
+
 How did presences and absences interact? (logistic regression)
+
+Where were newly found populations in terms of drainage connections?
 
@@
 rm(list = ls()[ sapply( ls(), exists, mode="function")])
-
 ```

### 2024-09-13 scripts/Results.Rmd
crazy null plot

--- 2024-08-14 scripts/Results.Rmd
@@
 suitability_amounts <- bind_rows(suitability_amounts)
-remotes::install_github("bcgov/elucidate")
-
-y1 <- rnorm(1:10, 100, 10)
-elucidate::mean_ci(y1)
-
-recs <- filter(suitability_amounts, Percent == 50) |>
-  pull(Records)
-
-as.numeric( elucidate::mean_ci(recs, replicates = 1000, ci_level = 0.99) )
 
@@
 
+#################################################################################
 fp <- scouting[scouting$futurePotential=='Yes',] 
@@
 
-linecols <- c(
+### Here we prep some display text which will help users interpret the plot ####
+################################################################################
+sm50 <- suitability_means[ which.min(abs(suitability_means$Percent-50)), ] 
+# on average 50% of cells have lower probabilities of being suitable than 0.23
+
+sm80 <- suitability_means[ which.min(abs(suitability_means$Percent-80)), ] 
+# on average 80% of cells have lower probabilities of being suitable than 0.614
+
+fp_2 <- fp[ which.min(abs(fp$obs-0.2)), ] # 80% of all populations are found in areas with suitability > 65.2%
+fp_5 <- fp[ which.min(abs(fp$obs-0.5)), ] # 50% of all populations are found in areas with suitability > 83.6% 
+fp_8 <- fp[ which.min(abs(fp$obs-0.8)), ] # the final 20 % of populations are found in areas with suitability > 94.6%
+
+labs <- data.frame(
+  x = c(
+    round(fp_2$obs, 2)*100 -10, 
+    round(fp_5$obs, 2)*100 - 10, 
+    sm50$Percent+15, 
+    sm80$Percent+15),
+  y = c(
+    round(fp_2$suitability, 2) + 0.1,
+    round(fp_5$suitability, 2) + 0.1, 
+    sm50$mean - 0.15, 
+    sm80$mean - 0.15),
+  label = c(
+    paste0(100- (round(fp_2$obs, 2)*100), '% of pops.\nwere found in areas\nwith suitability > ', round(fp_2$suitability, 2)),  
+    paste0(round(fp_5$obs, 2)*100, '% of pops.\nwere found in areas\nwith suitability > ', round(fp_5$suitability, 2)),
+    paste0(sm50$Percent, '% of cells have\nprobabilities of being\nsuitable lower than ', round(sm50$mean, 2)), 
+    paste0(sm80$Percent, '% of cells have\nprobabilities of being\nsuitable lower than ', round(sm80$mean, 2))
+  )
+)
+
+################################################################################
+
+# Create the mean suitability values across all taxa, this will illustrate the
+# distribution of modeled suitable habitat across 'common species' in the West based
+# on our input data and modelling approach. 
+suitability_means <- suitability_means |>
+  mutate(Null_mod = mean*Percent)
+
+linecols <- c( # define our palette
   'Species' = 'grey50',
@@
 
-ggplot(suitability_amounts, aes(x = Percent, y = Records, color = Species, group = Species)) +
+ggplot(suitability_amounts,
+       aes(x = Percent, y = Records, color = Species, group = Species)) +
   geom_line(lty = 1) +
@@
   geom_line(data = suitability_means, 
-            aes(x = Percent, y =  mean, color = 'a'), color = '#093824', inherit.aes = F, lwd = 1) + 
-  geom_line(data = fp, aes(x = suitability*100, y = obs), color = '#FF2C55', inherit.aes = F,  lwd = 1)
+            aes(x = Percent, y =  mean, color = 'a'), 
+            color = '#093824', inherit.aes = F, lwd = 1) + 
+  geom_line(data = suitability_means, aes(y = Percent/100, x = Null_mod),
+            color = '#d95f02', inherit.aes = F,  lwd = 1, lty = 53, linemitre = 2) +
+  geom_line(data = fp, aes(y = suitability, x = obs*100),
+            color = '#7570b3', inherit.aes = F,  lwd = 1, lty = 53) +
 
@@
     x = "Raster Cells (cumulative)",
-    y = 'Predicted Suitability (cumulative)') + 
+    y = 'Predicted Suitability') + 
   theme_classic() +
-  scale_y_continuous(breaks = seq(0, 1, by = 0.2), sec.axis = sec_axis(~.*100,
-    name = 'Populations Found (cumulative)',
-    breaks = seq(0, 100, by = 20), 
-    labels = \(x) paste0(x, "%"))
+  scale_y_continuous(
+    breaks = seq(0, 1, by = 0.2)
   )  +
 
-  scale_x_continuous(breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%")) + 
+  scale_x_continuous(
+    breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%"), 
+          sec.axis = sec_axis(~., name = 'Populations Found (cumulative)', 
+                        breaks = seq(0, 100, by = 20), labels = \(x) paste0(x, "%"))) + 
   theme(
     plot.title = element_text(hjust = 0.5),
-    aspect.ratio = 1/1)
+    aspect.ratio = 1/1) +
 
-rm(linecols)
+  geom_segment(
+    aes(
+      xend = fp_2$obs*100, 
+      yend = fp_2$suitability, 
+      x = round(fp_2$obs, 2)*100 - 10, 
+      y = fp_2$suitability + 0.1),
+                  arrow = arrow(length = unit(0.25, "cm")), color = 'black') + 
+  geom_segment(
+    aes(
+      xend = fp_5$obs*100, 
+      yend = fp_5$suitability, 
+      x = round(fp_5$obs, 2)*100 - 10, 
+      y = fp_5$suitability + 0.1),
+    arrow = arrow(length = unit(0.25, "cm")), color = 'black') +
 
-hypothetical_curves
+  geom_segment(aes(xend = sm50$Percent, yend = sm50$mean, 
+                   x = sm50$Percent+15, y = sm50$mean - 0.15),
+                  arrow = arrow(length = unit(0.25, "cm")), color = 'black') +
+
+  geom_segment(aes(xend = sm80$Percent, yend = sm80$mean, 
+                   x = sm80$Percent+15, y = sm80$mean - 0.15),
+               arrow = arrow(length = unit(0.25, "cm")), color = 'black') +
+
+  geom_label(data = labs, aes(x = x, y = y, label = label), inherit.aes = F, size = 2.5)
+
+# rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
 ```
+
+Raster cells (X axis, lower), represent units of the entire landscape, and the landscape predicts the amount of suitable habitat (Y axis) available for a species. 
+Across all species, the amount of suitable habitat appears low, on average (green line) for any one species, 80% of the landscape has a less than 0.61 chance of being suitable. 
+So on average only 20% of the landscape has high odds of being suitable habitat for an average species. 
+We can multiple the probability of each raster cell having suitable habitat to generate an expectation of how many populations crews would find based on our models alone (orange line).
+Using empirical data (purple line) we show that crews mostly found populations in areas with high predicted suitability, 80% of populations were in areas with suitability > 0.65, 50% in areas with modeled suitability > 0.84$.
+Where the purple line is above the orange line, it shows that crews found fewer populations than would be expected by our model; I wager most of this deviation is due to them simply not recording populations - a sampling artifact.
 

### 2024-09-13 scripts/Results.Rmd
leaf plot

@@
   'Species' = 'grey50',
-  "Mean" = "#093824", 
+  "Mean (95% CI)" = "#093824", 
   "95% CI" = "#73BA9B",
-  "Populations" = "#FF2C55"
+  "Observed Populations" = "#7570b3",
+  'Theoretical Probability' = '#d95f02'
   )
@@
   scale_color_grey(guide = 'none') + 
-  geom_ribbon(data = suitability_means, aes(ymin = lower, ymax = upper, x = Percent),
-              fill = '#73BA9B', inherit.aes = F, alpha = 0.5) + 
+  ggnewscale::new_scale_color() +
+  geom_ribbon(data = suitability_means, aes(ymin = lower, ymax = upper, x = Percent, fill = '95% CI'),
+              inherit.aes = F, alpha = 0.5) + 
   geom_line(data = suitability_means, 
-            aes(x = Percent, y =  mean, color = 'a'), 
-            color = '#093824', inherit.aes = F, lwd = 1) + 
-  geom_line(data = suitability_means, aes(y = Percent/100, x = Null_mod),
-            color = '#d95f02', inherit.aes = F,  lwd = 1, lty = 53, linemitre = 2) +
-  geom_line(data = fp, aes(y = suitability, x = obs*100),
-            color = '#7570b3', inherit.aes = F,  lwd = 1, lty = 53) +
+            aes(x = Percent, y =  mean, color = 'Mean (95% CI)'), 
+            inherit.aes = F, lwd = 1) + 
+  geom_line(data = suitability_means, aes(y = Percent/100, x = Null_mod, color = 'Theoretical Probability'),
+            inherit.aes = F, lwd = 1) +
+  geom_line(data = fp, aes(y = suitability, x = obs*100, color = 'Observed Populations'), inherit.aes = F, lwd = 1) +
 
   labs(
-    title = 'Populations are found in\nareas with higher suitability',
+    title = 'Populations are found in areas\nwith higher predicted suitability',
     x = "Raster Cells (cumulative)",
@@
   )  +
+  scale_fill_manual(values = linecols, guide = 'none') + 
+  scale_colour_manual(values = linecols, name = NULL) +
 
@@
   theme(
+    legend.position=c(.9,.10),
     plot.title = element_text(hjust = 0.5),
@@
 
-  geom_label(data = labs, aes(x = x, y = y, label = label), inherit.aes = F, size = 2.5)
+  geom_label(data = labs, aes(x = x, y = y, label = label), 
+             inherit.aes = F, size = 2.5,
+             fill='transparent', color = 'white') +
 
-# rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
+  ######        USE THIS TO MAKE THE PLOT TRANSPARENT FOR THE TALK        #####
+
+  theme(
+    panel.background = element_rect(fill='transparent'), #transparent panel bg
+    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
+    panel.grid.major = element_blank(), #remove major gridlines
+    panel.grid.minor = element_blank(), #remove minor gridlines
+    legend.background = element_rect(fill='transparent'), #transparent legend bg
+    legend.box.background = element_rect(fill='transparent'), #transparent legend panel
+    text = element_text(colour = "white")
+  )
+
+ggsave('../plots/leafplot-trans.png', width = 9, height = 9, units = 'in')
+
+rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
 ```

### 2024-09-13 scripts/Results.Rmd
species added to legend of leafplot

@@
 
+## create a dummy species var to put the grey50 color on the legend. 
+dum_species <- data.frame(Percent = 1, Records = 0.1, Species = 'dum')
+
 linecols <- c( # define our palette
@@
   geom_line(data = fp, aes(y = suitability, x = obs*100, color = 'Observed Populations'), inherit.aes = F, lwd = 1) +
+  geom_line(data = dum_species, aes(color = 'Species'), lwd = 1) +
 
@@
   scale_fill_manual(values = linecols, guide = 'none') + 
-  scale_colour_manual(values = linecols, name = NULL) +
+  scale_colour_manual(
+    values = linecols, name = NULL, 
+    breaks = c('Species', 'Mean (95% CI)', 'Theoretical Probability', 'Observed Populations')) +
 
@@
   geom_label(data = labs, aes(x = x, y = y, label = label), 
-             inherit.aes = F, size = 2.5,
-             fill='transparent', color = 'white') +
+             inherit.aes = F, size = 2.5 ,
 
  ######        USE THIS TO MAKE THE PLOT TRANSPARENT FOR THE TALK        #####
+             fill='transparent', color = 'white') +
 

### 2024-09-15 scripts/Results.Rmd
removed scouting points within 1km of known occurrences

@@
 # here are the ranked patches
-f_rp <- list.files('../results/rankedPatches', pattern = '.shp')
+f_rp <- list.files('../results/rankedPatches-manuscript', pattern = '.shp')
 f_rp <- data.frame(
-  f_rankedp = paste0('../results/rankedPatches/', f_rp), 
+  f_rankedp = paste0('../results/rankedPatches-manuscript/', f_rp), 
   species = gsub('[.]shp$', '', f_rp)
@@
     .before = geometry) |>
-    dplyr::select(-f_suitability, -f_rankedp, -f_patchMetrics, -f_classMetrics) |>
+    dplyr::select(-f_rankedp, -f_patchMetrics, -f_classMetrics) |>
     dplyr::left_join(lsm, by = c('PchIDComb' = 'id' ))
@@
 
+ # for manuscript 193 records are NA.
+```
+
+Some of these visits undoubtedly arose due to revisits of known occurrences. 
+We will discard any point within 500m of an historic point which we know was distributed to the crews via QGIS. 
+```{r}
+p <- '../data/raw/occurrence/combined_records'
+f <- file.path(p, list.files(p, pattern = '.shp$'))
+
+f_occ <- data.frame(
+  f_rankedp = f, 
+  taxa = gsub('[.]shp$', '', basename(f))
+)
+
+files <- left_join(scouting, f_occ, by = 'taxa') 
+f <- split(files, f = files$f_rankedp)
+
+strip_occurrences <- function(x){
+
+  occurrences <- sf::st_read(sf::st_drop_geometry(x$f_rankedp[1]), quiet = T) |>
+    sf::st_buffer(1000)
+
+  ints <- sf::st_intersects(occurrences, x) # identify intersecting points
+  removals <- unique(unlist(ints[lengths(ints) > 0])) # remove them.
+
+  if(length(removals > 0)){
+    x <- x[-removals,]
+  }
+
+  return(dplyr::select(x, -f_rankedp))
+}
+
+files <- lapply(f, strip_occurrences) |>
+  bind_rows()
+
+rm(f, p, strip_occurrences)
 ```
@@
 
+scout_split <- split(files, files$taxa)
+# we are going to discard species with fewer than 3 observations. 
+scout_split <- scout_split[lapply(scout_split, nrow) >= 3]
+
 suitability_amounts <- lapply(scout_split, prcntSuitable)
@@
 
-suitability_amounts <- readRDS('../data/processed/prcntSDMsurfaceSuitable.rds')
-suitability_amounts <- bind_rows(suitability_amounts)
+suitability_amounts <- readRDS('../data/processed/prcntSDMsurfaceSuitable.rds') |>
+  bind_rows(suitability_amounts)
 

### 2024-09-15 scripts/functions.R

@@ -795,8 +795,8 @@ neigh_type <- function(x){
     data.frame(as.matrix(nbdist)), 'OccupiedNeighborsIn5k')
   nbdist$OccupiedNeighborsIn5k <- as.numeric(nbdist$OccupiedNeighborsIn5k)
   
-  dat <- cbind(nbdist, nblags) %>% 
-    tibble::rownames_to_column('ROW') %>% 
+  dat <- cbind(nbdist, nblags) |> 
+    tibble::rownames_to_column('ROW') |>  
     tidyr::pivot_longer(FirstOrder:FifthOrder, names_to = 'OrderConnection', values_to = 'NoConnections') %>% 
     dplyr::filter(ROW %in% 1:known_r | OccupiedNeighborsIn5k > 0 | NoConnections > 0)
   
@@ -810,7 +810,7 @@ neigh_type <- function(x){
                            OrderConnection == 'ThirdOrder' & NoConnections <= 3 ~ 7, 
                            OccupiedNeighborsIn5k >= 3 ~ 8, 
                            OccupiedNeighborsIn5k <= 2 ~ 9, 
-                           .default = as.numeric(7)
+                           .default = as.numeric(10)
                          )) |>
     dplyr::slice_min(Rank, n = 1, by = 'ROW') |>
     dplyr::select(ROW, Rank) |>

### 2024-09-16 scripts/Results.Rmd
psummarizing number and sizes of patches

@@
     .before = geometry) |>
-    dplyr::select(-f_rankedp, -f_patchMetrics, -f_classMetrics) |>
+    dplyr::select(-f_patchMetrics, -f_classMetrics) |>
     dplyr::left_join(lsm, by = c('PchIDComb' = 'id' ))
@@
 Some of these visits undoubtedly arose due to revisits of known occurrences. 
-We will discard any point within 500m of an historic point which we know was distributed to the crews via QGIS. 
+We will discard any point within 1000m of an historic point which we know was distributed to the crews via QGIS. 
 ```{r}
@@
 f_occ <- data.frame(
-  f_rankedp = f, 
+  f_occ = f, 
   taxa = gsub('[.]shp$', '', basename(f))
@@
 
-  occurrences <- sf::st_read(sf::st_drop_geometry(x$f_rankedp[1]), quiet = T) |>
+  occurrences <- sf::st_read(sf::st_drop_geometry(x$f_occ[1]), quiet = T) |>
     sf::st_buffer(1000)
@@
 
-  return(dplyr::select(x, -f_rankedp))
+  return(dplyr::select(x, -f_occ))
 }
@@
 suitability_amounts <- readRDS('../data/processed/prcntSDMsurfaceSuitable.rds') |>
-  bind_rows(suitability_amounts)
+  bind_rows()
 
@@
 
-## Determine which points were used in training the models
+Connectivity analysis
 
-These points will need to be removed because they are already known to be presences...
+```{r}
+ggplot () + 
+  geom_bar(data = files, aes(Rank))
+
+# What distribution of these ranks exist within the data sets? - Let's show 
+# the proportion which is returned by all.
+
+PatchSummary <- function(x){
+
+  patches <- sf::st_read(x, quiet = TRUE)
+
+  area_rank <- patches |>
+    dplyr::group_by(Rank) |>
+    dplyr::summarise(geometry = sf::st_union(geometry))
+
+  area_rank <- data.frame(
+    Taxon = gsub('[.]shp$', '', basename(x)),
+    dplyr::select(sf::st_drop_geometry(area_rank), 'Rank'),
+    Area_ha = as.numeric(units::set_units(sf::st_area(area_rank), 'ha'))
+  )
+
+  cts_rank <- patches |>
+    sf::st_drop_geometry() |> 
+    dplyr::select(Rank) |>
+    dplyr::count(Rank, name = 'Patches') |>
+    dplyr::left_join(area_rank, by = 'Rank') |>
+
+    # now calculate as proportions
+    dplyr::ungroup() |>
+    dplyr::mutate(
+      Patches = Patches/sum(Patches),
+      Area_ha = Area_ha/sum(Area_ha)
+    ) |>
+    dplyr::relocate(Taxon, .before = Rank)
+
+  return(cts_rank)
+
+}
+
+need_patches <- unique(files$f_rankedp)
+patch_summs <- lapply(need_patches, PatchSummary)
+
+```
 

### 2024-09-16 scripts/Results.Rmd
props

@@
 
-```{r}
-ggplot () + 
-  geom_bar(data = files, aes(Rank))
+```{r, eval = F}
 
@@
 need_patches <- unique(files$f_rankedp) 
-patch_summs <- lapply(need_patches, PatchSummary)
+patch_summs <- lapply(need_patches, PatchSummary)|> 
+  bind_rows()
 
+ggplot() + 
+  geom_point(data = patch_summs, aes(x = Rank, y = Area_ha))
+
+ggplot() + 
+  geom_point(data = patch_summs, aes(x = Rank, y = Patches))
+
+ps <- patch_summs |>
+  ungroup() |>
+  group_by(Rank) |>
+  summarize(
+    meanP = mean(Patches),
+    meanA = mean(Area_ha)
+    )
+
+f1 <- files |>
+  sf::st_drop_geometry() |>
+  count(Rank) |>
+  drop_na() |>
+  left_join(ps, 'Rank') |>
+  mutate(
+    meanP_prop = (1 - meanP) * n,
+    meanA_prop = (1 - meanA) * n
+  )
+
+ggplot () + 
+  geom_col(data = f1, aes(meanP_prop, x = Rank))
+
+ggplot () + 
+  geom_col(data = f1, aes(meanA_prop, x = Rank))
+
+saveRDS(patch_summs, file = '../data/processed/PatchCtsAreas.rds')
 ```
@@
 
-- can we test moving the distance patches into their own categories ??? 
 ```{r}

### 2024-10-02 scripts/Results.Rmd
mod

@@
   theme(
+    axis.text = element_text(color = 'white'),
+    axis.ticks = element_line(color = 'white'),
     panel.background = element_rect(fill='transparent'), #transparent panel bg
@@
 ```
+
+5) Where were the populations found relative to the patch numbers?
+
+```{r}
+
+patches <- read.csv('../data/SOS/PatchesPctChange.csv')
+
+# showcase the area and proportion of patches
+melted_patches <- select(patches, Rank, meanP, meanA, n_prop) |>
+  pivot_longer(names_to = 'Metric', values_to = 'Value', cols = -Rank) |>
+  mutate(Metric = case_when(
+    Metric == 'meanA' ~ 'Mean Area', 
+    Metric == 'meanP' ~ 'Mean Number',
+    Metric == 'n_prop' ~ 'Occurrences'
+  ))
+
+cols <- c('Occurrences' = '#d95f02', 'Mean Area' = '#093824', 'Mean Number' =  '#7570b3')
+shapes <- c('Occurrences' = 21, 'Mean Area' = 22, 'Mean Number' =  23)
+
+patch_pts <- ggplot(data = melted_patches, 
+                 aes(y = Rank, x = Value, fill = Metric, shape = Metric)) + 
+  geom_jitter(height = 0.1, size = 4) +  
+  theme_classic() + 
+  scale_y_reverse(breaks = 1:9) + 
+  scale_fill_manual(values = cols) + 
+  scale_shape_manual(values = shapes) + 
+  labs(
+    x = 'Proportion', 
+    y = 'Patch Rank', 
+    title = 'Relationship between
+    patches across landscape
+    and new occurrences'
+    ) + 
+  theme(
+    axis.text = element_text(color = 'white'),
+    axis.ticks = element_line(color = 'white'),
+    text = element_text(color = 'white'),
+    aspect.ratio = 16/9,
+    plot.title = element_text(hjust = 0.5), 
+    legend.title = element_text(hjust = 0.5),
+    legend.position = c(0.825, 0.2), 
+    panel.background = element_rect(fill='transparent'), 
+    plot.background = element_rect(fill='transparent', color=NA), 
+    panel.grid.major = element_blank(), 
+    panel.grid.minor = element_blank(), 
+    legend.background = element_rect(fill='transparent'),
+    legend.box.background = element_rect(fill='transparent') 
+    )
+
+ggsave('../plots/Ranks.png', patch_pts)
+```
+
+```{r}
+melted_metrics <- select(patches, Rank, A_pct_change, P_pct_change) |>
+  pivot_longer(names_to = 'Metric', values_to = 'Value', cols = -Rank) |>
+  mutate(y = rep(1:2, times = 7))
+
+ggplot(data = melted_patches, aes(x = Rank, y = y, fill = Value)) + 
+  geom_tile() + 
+  scale_fill_gradient2( 
+    'Percent\nChange', 
+    low = 'red', high = 'darkgreen', 
+    breaks = seq(-100, 400, by = 100), limits = c(-100, 410),
+    labels = paste0(seq(-100, 400, by = 100), '%')) + 
+  scale_y_continuous(breaks=c(1, 2),
+        labels=c("Area", "No. Patches")) + 
+  theme_void()
+
+rm(cols, shapes, melted_patches)
+```

### 2024-10-02 scripts/Results.Rmd
reworked leaf plot

@@
   geom_label(data = labs, aes(x = x, y = y, label = label), 
-             inherit.aes = F, size = 2.5 ,
+             inherit.aes = F, size = 3 ,
 
@@
 
-rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
+# rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
 ```

### 2024-10-03 scripts/loadingAbsenceData.R
read in field absence data

@@ -1,0 +1,50 @@
+setwd('~/Documents/SDM_restorations/scripts')
+library(tidyverse)
+
+p <- '../data/raw/AbsenceSampling'
+f <- file.path(p, list.files(p))
+
+PresAbs <- do.call("rbind", sapply(f, read.csv, simplify = FALSE, skip = 4)) |>
+  rownames_to_column(var = 'SampleEvent') |>
+  mutate(
+    SampleEvent = gsub('[.].*$','', basename(SampleEvent)),
+    Census = if_else(is.na(Census), 0, Census),
+    Pres.Abs = case_when(
+      Pres.Abs == 'P' ~ 1, 
+      Pres.Abs == 'A' ~ 0, 
+      .default = as.numeric(Pres.Abs)
+    )) 
+
+ob <- do.call("rbind", sapply(f, read.csv, simplify = FALSE))[,1:3]
+
+readheadR <- function(x){
+  
+  dat <- read.csv(x, header = F)[1:3,1:2] 
+  sampleEvent <- gsub('[A-z]|[.]', '', basename(x))
+  
+  colnames(dat) <- c('Variable', 'Response')
+  Crew <- gsub('CO150', 'UFO',
+       gsub('CO130', 'GJFO',
+          gsub('UT070|UT70', 'PFO',
+              gsub('UT080|UT80', 'VFO', dat$Response))))[3]
+  sampleEvent <- paste0(Crew, sampleEvent)
+  
+  dat <- tidyr::pivot_wider(dat, names_from = 'Variable', values_from = 'Response') 
+  
+  ob <- data.frame(
+    cbind(
+      matrix(
+        sapply(dat[c('Latitude', 'Longitude')], as.numeric), nrow = 1)
+    ), sampleEvent) |>
+    setNames(c('Latitude', 'Longitude', 'SampleEvent'))
+  
+  return(ob)
+  
+}
+
+SiteData <- lapply(f, readheadR) |>
+  bind_rows() |>
+  drop_na(Longitude) |>
+  sf::st_as_sf(coords = c('Longitude', 'Latitude'))
+
+left_join(PresAbs, SiteData, by = 'SampleEvent') 

### 2024-10-03 scripts/loadingAbsenceData.R
in case users submit longitude without negative sign

@@ -9,13 +9,12 @@ PresAbs <- do.call("rbind", sapply(f, read.csv, simplify = FALSE, skip = 4)) |>
   mutate(
     SampleEvent = gsub('[.].*$','', basename(SampleEvent)),
     Census = if_else(is.na(Census), 0, Census),
-    Pres.Abs = case_when(
+    Occurrence = case_when(
       Pres.Abs == 'P' ~ 1, 
       Pres.Abs == 'A' ~ 0, 
       .default = as.numeric(Pres.Abs)
-    )) 
-
-ob <- do.call("rbind", sapply(f, read.csv, simplify = FALSE))[,1:3]
+    )) |>
+  select(-Pres.Abs)
 
 readheadR <- function(x){
   
@@ -38,13 +37,17 @@ readheadR <- function(x){
     ), sampleEvent) |>
     setNames(c('Latitude', 'Longitude', 'SampleEvent'))
   
+  ob$Longitude <- as.numeric(paste0('-', abs(ob$Longitude)))
+  
   return(ob)
   
 }
 
-SiteData <- lapply(f, readheadR) |>
-  bind_rows() |>
-  drop_na(Longitude) |>
+SiteData <- lapply(f, readheadR) |> 
+  bind_rows() |> 
+  drop_na(Longitude) |> 
   sf::st_as_sf(coords = c('Longitude', 'Latitude'))
 
-left_join(PresAbs, SiteData, by = 'SampleEvent') 
+SiteData <- left_join(PresAbs, SiteData, by = 'SampleEvent') 
+
+rm(PresAbs, f, p, readheadR)

### 2024-10-11 scripts/Results.Rmd
check landscape metrics again

@@
 
-# rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
+rm(fp, fp_2, fp_5, fp_8, labs, prctn, sm50, sm80, linecols)
 ```
@@
 ```{r}
+write.csv(scout_long, '../data/SOS/LandscapeMetrics.csv')
 
 scout_long <- pivot_longer(scouting, cai:enn_mn)
+scout_long <- mutate(
+  scout_long, 
+  POP_CENSUS_SQRT = sqrt(POPULATION_SIZE),
+  POP_CENSUS_LOG = log(POPULATION_SIZE)
+  )
 
-ggplot(scout_long, aes(y = POPULATION_SIZE, x = value)) +
+write.csv(scout_long, '../data/SOS/LandscapeMetrics.csv')
+
+ggplot(scout_long, aes(y = POP_CENSUS_LOG, x = value)) +
   geom_point() +
-  facet_grid(. ~ name)
+  facet_grid(. ~ name, scales = 'free') +
+  geom_smooth(aes(group=name))
+
+scout_long
 
